### 2014-09-29 curvature.tex
Stub out tex.

\documentclass{amsart}

\usepackage{amsmath}
% \usepackage[german]{todonotes}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{example}[theorem]{Example}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\RR}{\mathbb R}

\begin{document}
\title{Ricci curvature of the phylogenetic subtree-prune-regraft graph}
\author[Whidden]{Christopher Whidden}
\author[Matsen]{Frederick Matsen}
\address{}
\thanks{}

\date{\today}

\begin{abstract}
\end{abstract}

\maketitle

% % Make a todo table of contents.
% \makeatletter
% \providecommand\@dotsep{5}
% \makeatother
% \listoftodos\relax

\section{Introduction}

\section{$k$-tubes}

% \bibliographystyle{plain}
% \bibliography{curvature}
\end{document}

### 2014-10-02 curvature.tex
rough k-tube construction procedure

@@
 \usepackage{amsmath}
-% \usepackage[german]{todonotes}
+\usepackage{ifthen}
+\usepackage{todonotes}
 
@@
 \newcommand{\RR}{\mathbb R}
+\newcommand{\TT}{\mathcal T}
+
+%---[ Parenthesization ]--------------------------------------------------------
+
+\newcommand{\newparentheses}[3]{%
+  \expandafter\newcommand\csname #1\endcsname[1]{#2##1#3}%
+  \expandafter\newcommand\csname #1L\endcsname[1]{\bigl#2##1\bigr#3}%
+  \expandafter\newcommand\csname #1XL\endcsname[1]{\Bigl#2##1\Bigr#3}%
+  \expandafter\newcommand\csname #1V\endcsname[1]{\left#2##1\right#3}}
+
+\newparentheses{parens}{(}{)}
+\newparentheses{floor}{\lfloor}{\rfloor}
+\newparentheses{ceil}{\lceil}{\rceil}
+\newparentheses{abs}{|}{|}
+\newparentheses{set}{\{}{\}}
+\newparentheses{size}{|}{|}
+
+%---[ Attributes ]--------------------------------------------------------------
+
+\makeatletter
+\newcommand{\onenewattribute}[3]{%
+  \@ifundefined{#1}{\let\@@def\newcommand}{\let\@@def\renewcommand}%
+  \expandafter\@@def\csname #1\endcsname[2][]{%
+    \ifthenelse{\equal{##1}{}}%
+    {#2\csname #3\endcsname{##2}}%
+    {#2_{##1}\csname #3\endcsname{##2}}}}
+\newcommand{\newattribute}[2]{%
+  \onenewattribute{#1}{#2}{parens}%
+  \onenewattribute{#1L}{#2}{parensL}%
+  \onenewattribute{#1XL}{#2}{parensXL}%
+  \onenewattribute{#1V}{#2}{parensV}}
+\makeatother
+
+%---[ Asymptotic notation ]-----------------------------------------------------
+
+\newattribute{OhOf}{\mathrm{O}}
+\newattribute{ThetaOf}{\Theta}
+\newattribute{OmegaOf}{\Omega}
+\newattribute{ohOf}{\mathrm{o}}
+\newattribute{omegaOf}{\omega}
+%\newattribute{min}{\text{min}}
+%\newattribute{max}{\text{max}}
+
+%---[ Distance measures, invocations, etc ]-------------------------------------
+\newattribute{dspr}{d_{\mathrm{SPR}}}
 
@@
 
+The curvature with respect to two topologies, $T_1$ and $T_2$, is primarily determined by a subset of the SPR graph in the localized neighborhood of these topologies.
+We first require some notation to define these subsets.
+Let $\TT$ be the full set of topologies with a given number of leaves.
+Denote the set of trees with SPR distance at most $k$ from a given tree $T$ by $N_k(T)$.
+These trees are the $k$-neighbors of $T$.
+That is, $$ N_k(T) = \set{t \in \TT \mid \dspr{T, t} \le k} .$$
+
+We now define the $k$-tube of two trees $T_1$ and $T_2$ to be the union of their $k$-neighbors and \emph{every tree on a shortest path between a $k$-neighbor of $T_1$ and a $k$-neighbor of $T_2$}.
+More formally, the $k$-tube for $T_1$ and $T_2$ is:
+\[ \set{t \in \TT \mid \dspr{t, t_1} + \dspr{t, t_2} = \dspr{t_1, t_2},\ \exists_{t_1 \in N_k(T_1)},\ \exists_{t_2 \in N_k(T_2)} } .\]
+We call $T_1$ and $T_2$ the centers of the $k$-tube.
+The distance between $k$-neighbors of different central trees is bounded by the maximum distance from each $k$-neighbor to its respective center and the maximum distance between the centers:
+
+\begin{lemma}
+	\label{lem:neighbor_distance}
+	$\dspr{t_1, t_2} \le \dspr{T_1, T_2} + 2k,\ \forall_{t_1 \in N_k(T_1)},\ \forall_{t_2 \in N_k(T_2)} } .$
+\end{lemma}
+
+To efficiently compute a $k$-tube, we first bound the maximum distance from members of the $k$-tube to the closest central tree.
+\begin{lemma}
+	\label{lem:k_tube_distance}
+	Any tree $t$ of the $k$-tube of two trees $T_1$ and $T_2$ satisfies:
+	\[ \min\left({\dspr{t, T_1}, \dspr{t, T_2}}\right) \le \floor{\dspr{T_1, T_2}/2} + 2k .\]
+\end{lemma}
+\begin{proof}
+Without loss of generality, assume that $t$ is on a shortest path between two trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$, and is either closer to $t_1$ than $t_2$ or equidistant from both.
+By Lemma~\ref{lem:neighbor_distance} there is a path of length at most $\dspr{T_1, T_2} + 2k$ between $t_1$ and $t_2$.
+Then $\dspr{t, t_1} \le \floor{(\dspr{T_1, T_2} + 2k) / 2}$.
+
+Along with the fact that $\dspr{T_1, t_1} \le k$, we have that
+\begin{align*}
+\dspr{t, T_1} &\le \dspr{t, t_1} + \dspr{t_1, T_1} \\
+&\le \floor{(\dspr{T_1, T_2} + 2k) / 2} + k \\
+&\le \floor{\dspr{T_1, T_2} / 2} + 2k.
+\end{align*}
+
+\end{proof}
+We can thus restrict our search for members of the $k$-tube to topologies of distance at most $\floor{\dspr{T_1, T_2}/2} + 2k$ from the central trees.
+We call this set of trees $k$-tube candidates.
+Note that there are $\OhOf{n^{2(\floor{\dspr{T_1, T_2}/2} + 2k)}}$ $k$-tube candidates.
+We identify these candidates by iterating over increasingly large $k$-neighbourhoods of $T_1$ and $T_2$, and removing duplicate topologies through hashing.
+\todo{We might be able to do better by using ``edge protection'' to avoid duplicates.
+I need to check how accurately neighbourhood sizes follow the naive $\OhOf(n^{2k})$ bound.}
+
+In the second part of our procedure for computing a $k$-tube, we identify $k$-tube candidates that are not on the shortest path between two $k$-neighbors of the central trees.
+We first compute shortest path lengths from each member of $N_k(T_1)$ to each $k$-tube candidate using breadth-first search (BFS).
+For each BFS traversal tree, we backtrack from members of $N_k(T_2)$, identifying topologies in the $k$-tube.
+We can use unique tree indices here, rather than hashing, to avoid duplicates. 
+The size of $N_k(T_1)$ is $\OhOf{n^{2k}}$, so this takes
+$\OhOf( n^{2k} M)$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.
+\todo{We could probably do something smarter by starting from trees that are 0, 1, etc SPRs from $T_1$.}
+
+This procedure identifies exactly the trees in the $k$-tube, so we have the following lemma:
+\begin{lemma}
+	\label{lem:k_tube_correctness}
+	This procedure correctly identifies the $k$-tube.
+\end{lemma}
+\begin{proof}
+	First, suppose that this procedure identifies a tree $x$ that is not in the $k$-tube.
+	Assume, without loss of generality, that the BFS tree $B$ from tree $t_1 \in N_k(T_1)$ transitioned from $x$ to a $k$-tube tree $t$.
+	Moreover, there must be a path in $B$ from $t$ to a tree $t_2 \in N_k(T_2)$ such that $t$ is on a shortest path from $t_1$ to $t_2$.
+	Then, by the properties of BFS, $x$ is on this shortest path and belongs in the $k$-tube.
+
+	Now, suppose that this procedure does not identify a tree $x$ that belongs in the $k$-tube.
+	$x$ must be on a shortest path between some trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$.
+	Then, by the properties of BFS, $x$ would have been on the path from $t_1$ to $t_2$ in the $t_1$ BFS tree.
+\end{proof}
+
 % \bibliographystyle{plain}

### 2014-10-06 ricci-tangle.py
ricci-tangle!

@@ -1,0 +1,37 @@
+#!/usr/bin/env sage
+
+import time
+from sage.all import *
+load("tree-fun.py")
+load("gricci/code.py")
+load("gricci/nb-fun.py")
+load("../tangle-fun.py")
+
+for n in (int(a) for a in sys.argv[1:]):
+    old_time = time.time()
+    idx_fname = "tangles/tangle{}.idx".format(n)
+    matrix_fname = "matrices/matrix_{}".format(n)
+    print idx_fname
+    assert(os.path.exists(idx_fname))
+    assert(os.path.exists(matrix_fname))
+    adj_graph = Graph(matrix_of_csv(matrix_fname))
+
+    def process_line(line):
+        row = list(int(x) for x in line)
+        t1_idx = row[0]
+        t2_idx = row[1]
+        if t1_idx == t2_idx:
+            row.append("-")
+        else:
+            row.append(ric_unif_rw(adj_graph, source=row[0], target=row[1]).ric)
+        return row
+
+    with open(idx_fname, 'rb') as csvfile:
+        idx_reader = csv.reader(csvfile, delimiter='\t', quotechar="'")
+        results = list(process_line(line) for line in idx_reader)
+        results.sort(key=lambda r: r[3])
+        with open("ricci"+str(n)+".mat", "w") as f_out:
+            for r in results:
+                f_out.write("\t".join(str(e) for e in r)+"\n")
+
+    print "{}\t{}".format(n, time.time() - old_time)

### 2014-10-07 ricci-tangle.py
Not sorting results in ricci-tangle.

@@ -17,19 +17,18 @@ for n in (int(a) for a in sys.argv[1:]):
     adj_graph = Graph(matrix_of_csv(matrix_fname))
 
     def process_line(line):
-        row = list(int(x) for x in line)
-        t1_idx = row[0]
-        t2_idx = row[1]
+        row = list(line)
+        t1_idx = int(row[0])
+        t2_idx = int(row[1])
         if t1_idx == t2_idx:
             row.append("-")
         else:
-            row.append(ric_unif_rw(adj_graph, source=row[0], target=row[1]).ric)
+            row.append(ric_unif_rw(adj_graph, source=t1_idx, target=t2_idx).ric)
         return row
 
     with open(idx_fname, 'rb') as csvfile:
         idx_reader = csv.reader(csvfile, delimiter='\t', quotechar="'")
         results = list(process_line(line) for line in idx_reader)
-        results.sort(key=lambda r: r[3])
         with open("ricci"+str(n)+".mat", "w") as f_out:
             for r in results:
                 f_out.write("\t".join(str(e) for e in r)+"\n")

### 2014-10-07 ricci-tangle.py
Recording SPR dist between chosen points.

@@ -21,9 +21,10 @@ for n in (int(a) for a in sys.argv[1:]):
         t1_idx = int(row[0])
         t2_idx = int(row[1])
         if t1_idx == t2_idx:
-            row.append("-")
+            row.extend([0, "-"])
         else:
-            row.append(ric_unif_rw(adj_graph, source=t1_idx, target=t2_idx).ric)
+            calc = ric_unif_rw(adj_graph, source=t1_idx, target=t2_idx)
+            row.extend([calc.dist, calc.ric])
         return row
 
     with open(idx_fname, 'rb') as csvfile:

### 2014-10-07 prep_figures/distribution/distribution.Rmd
Adding a shiny visualization.

---
title: "Ricci"
runtime: shiny
output: html_document
---

```{r, echo=FALSE}

library(ggvis)

d <- read.table("../../ricci5.mat", stringsAsFactors=FALSE)
colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
d <- subset(d, kappa_str != "-")
d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
d$both_nwk = paste(d$t1_nwk, d$t2_nwk, sep="<br />")

d %>%
  ggvis(x = ~kappa, y = ~dist) %>%
  layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
  # A dummy layer so that we can use text in the tooltip.
  layer_text(text := ~both_nwk, fontSize := 0) %>%
  add_axis("x", grid = FALSE) %>%
  add_axis("y", grid = FALSE, values = 1:3) %>%
  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
  add_legend(c("stroke"), properties=legend_props(legend=list(y=70))) %>%
  add_legend(c("size"), properties=legend_props(legend=list(y=120))) %>%
  set_options(height = 150) %>% 
  add_tooltip(function(df) df$both_nwk)
```

### 2014-10-07 ricci-tangle.py
Adding multiprocessing to ricci-tangle.

@@ -1,6 +1,8 @@
 #!/usr/bin/env sage
 
 import time
+from multiprocessing import Pool
+from sys import stdout
 from sage.all import *
 load("tree-fun.py")
 load("gricci/code.py")
@@ -27,9 +29,19 @@ for n in (int(a) for a in sys.argv[1:]):
             row.extend([calc.dist, calc.ric])
         return row
 
+    def process_line_status(line):
+        stdout.write("*")
+        stdout.flush()
+        return process_line(line)
+
+    p = Pool(processes=8)
+
     with open(idx_fname, 'rb') as csvfile:
-        idx_reader = csv.reader(csvfile, delimiter='\t', quotechar="'")
-        results = list(process_line(line) for line in idx_reader)
+        lines = list(csv.reader(csvfile, delimiter='\t', quotechar="'"))
+        print "There are {} tangles to process.".format(len(lines))
+        print "|"+"-"*(len(lines)-2)+"|"
+        results = p.map(process_line_status, lines)
+        print ""
         with open("ricci"+str(n)+".mat", "w") as f_out:
             for r in results:
                 f_out.write("\t".join(str(e) for e in r)+"\n")

### 2014-10-07 prep_figures/distribution/distribution.Rmd
fancier tooltip.

@@
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-d$both_nwk = paste(d$t1_nwk, d$t2_nwk, sep="<br />")
+d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
 
@@
   # A dummy layer so that we can use text in the tooltip.
-  layer_text(text := ~both_nwk, fontSize := 0) %>%
+  layer_text(text := ~info, fontSize := 0) %>%
   add_axis("x", grid = FALSE) %>%
@@
   add_legend(c("size"), properties=legend_props(legend=list(y=120))) %>%
-  set_options(height = 150) %>% 
-  add_tooltip(function(df) df$both_nwk)
+  set_options(width=500, height = 150) %>%
+  add_tooltip(function(df) df$info)
 ```

### 2014-10-08 prep_figures/distribution/distribution.Rmd
Ricci plots 4-6.

--- 2014-10-07 prep_figures/distribution/distribution.Rmd
@@
 
-d <- read.table("../../ricci5.mat", stringsAsFactors=FALSE)
+d <- read.table("../../results/ricci6.mat", stringsAsFactors=FALSE)
 colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
@@
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-d$both_nwk = paste(d$t1_nwk, d$t2_nwk, sep="<br />")
+d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
+d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
 
 d %>%
-  ggvis(x = ~kappa, y = ~dist) %>%
+  ggvis(x = ~kappa, y = ~dist_jitter) %>%
   layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
   # A dummy layer so that we can use text in the tooltip.
-  layer_text(text := ~both_nwk, fontSize := 0) %>%
+  layer_text(text := ~info, fontSize := 0) %>%
   add_axis("x", grid = FALSE) %>%
@@
   add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
-  add_legend(c("stroke"), properties=legend_props(legend=list(y=70))) %>%
-  add_legend(c("size"), properties=legend_props(legend=list(y=120))) %>%
-  set_options(height = 150) %>% 
-  add_tooltip(function(df) df$both_nwk)
+  add_legend(c("stroke"), properties=legend_props(legend=list(y=100))) %>%
+  add_legend(c("size"), properties=legend_props(legend=list(y=200))) %>%
+  add_tooltip(function(df) df$info)
 ```

### 2014-10-28 ricci-tangle.py
Enabling batching of ricci.

@@ -1,49 +1,84 @@
 #!/usr/bin/env sage
 
+import argparse
+import math
 import time
+import sys
 from multiprocessing import Pool
-from sys import stdout
 from sage.all import *
 load("tree-fun.py")
 load("gricci/code.py")
 load("gricci/nb-fun.py")
 load("../tangle-fun.py")
 
-for n in (int(a) for a in sys.argv[1:]):
-    old_time = time.time()
-    idx_fname = "tangles/tangle{}.idx".format(n)
-    matrix_fname = "matrices/matrix_{}".format(n)
-    print idx_fname
-    assert(os.path.exists(idx_fname))
-    assert(os.path.exists(matrix_fname))
-    adj_graph = Graph(matrix_of_csv(matrix_fname))
 
-    def process_line(line):
-        row = list(line)
-        t1_idx = int(row[0])
-        t2_idx = int(row[1])
-        if t1_idx == t2_idx:
-            row.extend([0, "-"])
-        else:
-            calc = ric_unif_rw(adj_graph, source=t1_idx, target=t2_idx)
-            row.extend([calc.dist, calc.ric])
-        return row
+parser = argparse.ArgumentParser(description='ricci curvature of tangles',
+                                 prog='ricci-tangle.py')
 
-    def process_line_status(line):
-        stdout.write("*")
-        stdout.flush()
-        return process_line(line)
+parser.add_argument('--size', dest='n',
+                    type=int, help='Number of leaves.')
+parser.add_argument('--n-batches', dest='n_batches',
+                    type=int, help='Number of batches.')
+parser.add_argument('--batch', dest='batch', default=None,
+                    type=int, help='Which batch, zero indexed.')
 
-    p = Pool(processes=8)
+arguments = parser.parse_args()
+n = arguments.n
+n_batches = arguments.n_batches
+batch = arguments.batch
 
-    with open(idx_fname, 'rb') as csvfile:
-        lines = list(csv.reader(csvfile, delimiter='\t', quotechar="'"))
-        print "There are {} tangles to process.".format(len(lines))
-        print "|"+"-"*(len(lines)-2)+"|"
-        results = p.map(process_line_status, lines)
-        print ""
-        with open("ricci"+str(n)+".mat", "w") as f_out:
-            for r in results:
[…]

### 2014-10-29 ricci-tangle.py
New strategy for splitting ricci jobs.

@@ -15,29 +15,16 @@ load("../tangle-fun.py")
 parser = argparse.ArgumentParser(description='ricci curvature of tangles',
                                  prog='ricci-tangle.py')
 
-parser.add_argument('--size', dest='n',
-                    type=int, help='Number of leaves.')
-parser.add_argument('--n-batches', dest='n_batches',
-                    type=int, help='Number of batches.')
-parser.add_argument('--batch', dest='batch', default=None,
-                    type=int, help='Which batch, zero indexed.')
-
-arguments = parser.parse_args()
-n = arguments.n
-n_batches = arguments.n_batches
-batch = arguments.batch
-
-assert(n is not None)
-assert(n_batches is not None)
-assert(batch is not None)
-
-idx_fname = "tangles/tangle{}.idx".format(n)
-matrix_fname = "matrices/matrix_{}".format(n)
-assert(os.path.exists(idx_fname))
-assert(os.path.exists(matrix_fname))
-adj_graph = Graph(matrix_of_csv(matrix_fname))
-assert(batch >= 0)
-assert(batch < n_batches)
+parser.add_argument('idx_path',
+                    type=str, help='Path to .idx file.')
+parser.add_argument('--matrix', dest='matrix_path',
+                    type=str, help='Matrix path.')
+parser.add_argument('--out', dest='out_path',
+                    type=str, help='Output path.')
+args = parser.parse_args()
+assert(os.path.exists(args.idx_path))
+assert(os.path.exists(args.matrix_path))
+adj_graph = Graph(matrix_of_csv(args.matrix_path))
 
 
 def process_line(line):
@@ -58,27 +45,13 @@ def process_line_status(line):
     return process_line(line)
 
 
-def chunks(l, n_chunks):
-    """ Yield successive n-sized chunks from l.
-    http://stackoverflow.com/questions/312443/how-do-you-split-a-list-into-evenly-sized-chunks-in-python
-    """
-    for i in xrange(0, len(l), n_chunks):
-        yield l[i:i+n_chunks]
+p = Pool(processes=4)
 
-
-p = Pool(processes=8)
-
-with open(idx_fname, 'rb') as csvfile:
+with open(args.idx_path, 'rb') as csvfile:
     lines = list(csv.reader(csvfile, delimiter='\t', quotechar="'"))
-    batch_size = int(len(lines) / n_batches)+1
-    batch_start = range(0, len(lines), batch_size)[batch]
[…]

### 2014-10-29 ricci-tangle.py
sbatch script werkin.

@@ -45,7 +45,7 @@ def process_line_status(line):
     return process_line(line)
 
 
-p = Pool(processes=4)
+p = Pool(processes=8)
 
 with open(args.idx_path, 'rb') as csvfile:
     lines = list(csv.reader(csvfile, delimiter='\t', quotechar="'"))

### 2014-10-29 srun-ricci.sh

@@ -2,19 +2,34 @@
 set -e
 set -u
 
-n_leaves=5
-split_size=15
+test -z $1 && exit 1
+
+n_leaves=$1
+split_size=20
 slurm_config="-c 4 -n 1 -w gizmof[1-180]"
+slurm_config="-c 8 -n 1"
 
-out_path="results/ricci$n_leaves.mat"
-files=$(split -l $split_size tangles/tangle$n_leaves.idx --filter="echo \$FILE")
+# Prepare vars and split file.
+out_path="ricci$n_leaves.mat"
+splitargs="-l $split_size tangles/tangle$n_leaves.idx"
+files=$(split $splitargs --filter="echo \$FILE")
+# Apparently split doesn't actually split when given the --filter flag, so we run it again.
+split $splitargs
 
-
-for i in $files; do
+# Run!
+for i in $files
+do
     export SLURM_JOB_NAME=ricci-$i
-    cmd="srun $slurm_config ./ricci-tangle.py --matrix matrices/matrix_$n_leaves --out $out_path-$i $i"
+    cmd="./ricci-tangle.py --matrix matrices/matrix_$n_leaves --out $out_path-$i $i"
     echo $cmd
-    $cmd
+    echo "#!/bin/sh" > $i.sh
+    echo $cmd >> $i.sh
+    sbatch $slurm_config $i.sh
 done
 
-
+# Wait until done.
+while test 0 -ne $(squeue -u matsen -h | wc -l)
+do
+    squeue -u matsen
+    sleep 5
+done

### 2014-10-29 sbatch-ricci.sh
Collating in sbatch script.

@@ -33,3 +33,10 @@ do
     squeue -u matsen
     sleep 5
 done
+
+rm results/$out_path
+touch results/$out_path
+for i in $files
+do
+    cat $out_path-$i >> results/$out_path
+done

### 2014-10-29 simple-ricci.sh
Adding a simple ricci runner.

@@ -1,0 +1,12 @@
+#!/bin/sh
+set -e
+set -u
+
+test -z $1 && exit 1
+
+n_leaves=$1
+
+idx_path="tangles/tangle$n_leaves.idx"
+out_path="results/ricci$n_leaves.mat"
+
+./ricci-tangle.py --matrix matrices/matrix_$n_leaves --out $out_path $idx_path

### 2014-10-29 matrices/csv2graphsobj.py
Dropping matrices in favor of graph objects.

@@ -1,0 +1,13 @@
+#!/usr/bin/env sage
+
+from sage.all import *
+load("../gricci/nb-fun.py")
+
+for i in range(7,8):
+    print i
+    matrix_path = "matrix_{}".format(i)
+    adj_graph = Graph(matrix_of_csv(matrix_path))
+    sage.structure.sage_object.save(
+        adj_graph,
+        filename=(matrix_path+".sobj"))
+

### 2014-10-29 ricci-tangle.py

@@ -17,14 +17,14 @@ parser = argparse.ArgumentParser(description='ricci curvature of tangles',
 
 parser.add_argument('idx_path',
                     type=str, help='Path to .idx file.')
-parser.add_argument('--matrix', dest='matrix_path',
-                    type=str, help='Matrix path.')
+parser.add_argument('--adjacency', dest='adjacency_path',
+                    type=str, help='Path to graph in SAGE object format.')
 parser.add_argument('--out', dest='out_path',
                     type=str, help='Output path.')
 args = parser.parse_args()
 assert(os.path.exists(args.idx_path))
-assert(os.path.exists(args.matrix_path))
-adj_graph = Graph(matrix_of_csv(args.matrix_path), immutable=True)
+assert(os.path.exists(args.adjacency_path))
+adj_graph = load(args.adjacency_path)
 
 
 def process_line(line):
@@ -45,7 +45,7 @@ def process_line_status(line):
     return process_line(line)
 
 
-p = Pool(processes=8)
+p = Pool(processes=4)
 
 with open(args.idx_path, 'rb') as csvfile:
     lines = list(csv.reader(csvfile, delimiter='\t', quotechar="'"))

### 2014-10-29 sbatch-ricci.sh

@@ -5,9 +5,8 @@ set -u
 test -z $1 && exit 1
 
 n_leaves=$1
-split_size=20
-slurm_config="-c 4 -n 1 -w gizmof[1-180]"
-slurm_config="-c 8 -n 1"
+split_size=50
+slurm_config="-c 4 -n 1"
 
 # Prepare vars and split file.
 out_path="ricci$n_leaves.mat"
@@ -20,7 +19,7 @@ split $splitargs
 for i in $files
 do
     export SLURM_JOB_NAME=ricci-$i
-    cmd="./ricci-tangle.py --matrix matrices/matrix_$n_leaves --out $out_path-$i $i"
+    cmd="./ricci-tangle.py --adjacency matrices/matrix_$n_leaves.sobj --out $out_path-$i $i"
     echo $cmd
     echo "#!/bin/sh" > $i.sh
     echo $cmd >> $i.sh
@@ -30,7 +29,7 @@ done
 # Wait until done.
 while test 0 -ne $(squeue -u matsen -h | wc -l)
 do
-    squeue -u matsen
+    squeue -h -u matsen | wc -l
     sleep 5
 done
 

### 2014-10-29 simple-ricci.sh

@@ -9,4 +9,4 @@ n_leaves=$1
 idx_path="tangles/tangle$n_leaves.idx"
 out_path="results/ricci$n_leaves.mat"
 
-./ricci-tangle.py --matrix matrices/matrix_$n_leaves --out $out_path $idx_path
+./ricci-tangle.py --adjacency matrices/matrix_$n_leaves.sobj --out $out_path $idx_path

### 2014-10-30 matrices/csv2graphsobj.py
Organizing & committing .sobj files.

@@ -1,13 +1,0 @@
-#!/usr/bin/env sage
-
-from sage.all import *
-load("../gricci/nb-fun.py")
-
-for i in range(7,8):
-    print i
-    matrix_path = "matrix_{}".format(i)
-    adj_graph = Graph(matrix_of_csv(matrix_path))
-    sage.structure.sage_object.save(
-        adj_graph,
-        filename=(matrix_path+".sobj"))
-

### 2014-10-30 rspr/csv2graphsobj.py

@@ -1,0 +1,12 @@
+#!/usr/bin/env sage
+
+from sage.all import *
+load("../gricci/nb-fun.py")
+
+for i in range(3,8):
+    print i
+    adj_graph = Graph(matrix_of_csv("matrix_{}".format(i)))
+    sage.structure.sage_object.save(
+        adj_graph,
+        filename=("graph_{}.sobj".format(i)))
+

### 2014-10-30 ricci-tangle.py
s/adjacency/graph/

@@ -17,14 +17,14 @@ parser = argparse.ArgumentParser(description='ricci curvature of tangles',
 
 parser.add_argument('idx_path',
                     type=str, help='Path to .idx file.')
-parser.add_argument('--adjacency', dest='adjacency_path',
+parser.add_argument('--graph', dest='graph_path',
                     type=str, help='Path to graph in SAGE object format.')
 parser.add_argument('--out', dest='out_path',
                     type=str, help='Output path.')
 args = parser.parse_args()
 assert(os.path.exists(args.idx_path))
-assert(os.path.exists(args.adjacency_path))
-adj_graph = load(args.adjacency_path)
+assert(os.path.exists(args.graph_path))
+g = load(args.graph_path)
 
 
 def process_line(line):
@@ -34,7 +34,7 @@ def process_line(line):
     if t1_idx == t2_idx:
         row.extend([0, "-"])
     else:
-        calc = ric_unif_rw(adj_graph, source=t1_idx, target=t2_idx)
+        calc = ric_unif_rw(g, source=t1_idx, target=t2_idx)
         row.extend([calc.dist, calc.ric])
     return row
 

### 2014-10-30 sbatch-ricci.sh

@@ -19,7 +19,7 @@ split $splitargs
 for i in $files
 do
     export SLURM_JOB_NAME=ricci-$i
-    cmd="./ricci-tangle.py --adjacency matrices/matrix_$n_leaves.sobj --out $out_path-$i $i"
+    cmd="./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path-$i $i"
     echo $cmd
     echo "#!/bin/sh" > $i.sh
     echo $cmd >> $i.sh

### 2014-10-30 simple-ricci.sh

@@ -9,4 +9,4 @@ n_leaves=$1
 idx_path="tangles/tangle$n_leaves.idx"
 out_path="results/ricci$n_leaves.mat"
 
-./ricci-tangle.py --adjacency matrices/matrix_$n_leaves.sobj --out $out_path $idx_path
+./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path

### 2014-10-30 prep_figures/distribution/distribution.Rmd
ricci7.svg

--- 2014-10-07 prep_figures/distribution/distribution.Rmd
@@
 
-d <- read.table("../../ricci5.mat", stringsAsFactors=FALSE)
+d <- read.table("../../results/ricci7.mat", stringsAsFactors=FALSE)
 colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
 d <- subset(d, kappa_str != "-")
+d <- sample(d)  # Shuffle rows of d.
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-d$both_nwk = paste(d$t1_nwk, d$t2_nwk, sep="<br />")
+d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
+d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
 
 d %>%
-  ggvis(x = ~kappa, y = ~dist) %>%
-  layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
+  ggvis(x = ~kappa, y = ~dist_jitter) %>%
+  layer_points(size = 5, fill = ~t1_nwk, stroke = ~t2_idx, strokeOpacity := 0) %>%
+#  layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
   # A dummy layer so that we can use text in the tooltip.
-  layer_text(text := ~both_nwk, fontSize := 0) %>%
+  layer_text(text := ~info, fontSize := 0) %>%
   add_axis("x", grid = FALSE) %>%
@@
   add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
-  add_legend(c("stroke"), properties=legend_props(legend=list(y=70))) %>%
-  add_legend(c("size"), properties=legend_props(legend=list(y=120))) %>%
-  set_options(height = 150) %>% 
-  add_tooltip(function(df) df$both_nwk)
+  #add_legend(c("stroke"), properties=legend_props(legend=list(y=100))) %>%
+  #add_legend(c("size"), properties=legend_props(legend=list(y=200))) %>%
+  add_tooltip(function(df) df$info)
 ```

### 2014-10-30 prep_figures/distribution/ggvis.Rmd
Also looking via ggplot.

--- 2014-10-30 prep_figures/distribution/distribution.Rmd
@@
 
-d <- read.table("../../results/ricci7.mat", stringsAsFactors=FALSE)
-colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
-d <- subset(d, kappa_str != "-")
-d <- sample(d)  # Shuffle rows of d.
-d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
-d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
+source("process.R")
+
+d <- get_results("../../results/ricci7.mat")
 

### 2014-10-30 prep_figures/distribution/process.R

@@ -1,0 +1,11 @@
+
+get_results <- function(path) {
+    d <- read.table(path, stringsAsFactors=FALSE)
+    colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
+    d <- subset(d, kappa_str != "-")
+    d <- sample(d)  # Shuffle rows of d.
+    d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
+    d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
+    d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
+    d
+}

### 2014-10-31 prep_figures/distribution/process.R
Heat maps of curvature.

@@ -1,3 +1,7 @@
+library(ape, apTreeshape)
+
+shape_of_str <- function(s) as.treeshape(read.tree(text=s))
+
 
 get_results <- function(path) {
     d <- read.table(path, stringsAsFactors=FALSE)
@@ -7,5 +11,16 @@ get_results <- function(path) {
     d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
     d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
     d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
+
+    # Tree shape information.
+    d$t1_shape <- lapply(d$t1_nwk, shape_of_str)
+    d$t2_shape <- lapply(d$t2_nwk, shape_of_str)
+    d$t1_colless <- sapply(d$t1_shape, colless)
+    d$t2_colless <- sapply(d$t2_shape, colless)
+
     d
 }
+
+
+
+

### 2014-10-31 rspr/adjlist2graphsobj.py
adjlist2graphsobj converter.

@@ -1,0 +1,16 @@
+#!/usr/bin/env sage
+
+from sage.all import *
+load("../gricci/nb-fun.py")
+
+for i in range(8, 9):
+    print i
+    with open(("adj_list_{}".format(i)), 'r') as f:
+        def tuple_generator():
+            for line in f:
+                line.rstrip("\n")
+                yield tuple(line.split(","))
+        adj_graph = Graph(tuple_generator())
+        sage.structure.sage_object.save(
+            adj_graph,
+            filename=("graph_{}.sobj".format(i)))

### 2014-12-05 sbatch-ricci.sh
Flip-flopping tangle and curvature.

@@ -10,7 +10,7 @@ slurm_config="-c 4 -n 1"
 
 # Prepare vars and split file.
 out_path="ricci$n_leaves.mat"
-splitargs="-a 4 -l $split_size tangles/tangle$n_leaves.idx"
+splitargs="-a 4 -l $split_size tangle/rooted-symmetric/tangle$n_leaves.idx"
 files=$(split $splitargs --filter="echo \$FILE")
 # Apparently split doesn't actually split when given the --filter flag, so we run it again.
 split $splitargs

### 2014-12-05 simple-ricci.sh

@@ -6,7 +6,7 @@ test -z $1 && exit 1
 
 n_leaves=$1
 
-idx_path="tangles/tangle$n_leaves.idx"
+idx_path="tangle/rooted-symmetric/tangle$n_leaves.idx"
 out_path="results/ricci$n_leaves.mat"
 
 ./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path

### 2014-12-05 sbatch-ricci.sh
More organizational improvement.

@@ -33,9 +33,9 @@ do
     sleep 5
 done
 
-rm results/$out_path
-touch results/$out_path
+rm results-rspr/$out_path
+touch results-rspr/$out_path
 for i in $files
 do
-    cat $out_path-$i >> results/$out_path
+    cat $out_path-$i >> results-rspr/$out_path
 done

### 2014-12-05 simple-ricci.sh

@@ -7,6 +7,6 @@ test -z $1 && exit 1
 n_leaves=$1
 
 idx_path="tangle/rooted-symmetric/tangle$n_leaves.idx"
-out_path="results/ricci$n_leaves.mat"
+out_path="results-rspr/ricci$n_leaves.mat"
 
 ./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path

### 2014-12-09 prep_figures/commute/curvature-MCT.Rmd
new file:   curvature-MCT.Rmd

---
title: "Ricci and commute time"
runtime: shiny
output: html_document
---

```{r, echo=FALSE}

library(ggvis)
library(ape)
library(apTreeshape)

source("process.R")

d <- get_results("../../results-rspr/commute6.mat")
d$dist <- as.factor(d$dist)

d %>%
  ggvis(x = ~kappa, y = ~MCT) %>%
  layer_points(size = 5, fill = ~dist) %>%
  # A dummy layer so that we can use text in the tooltip.
  layer_text(text := ~info, fontSize := 0) %>%
  add_axis("x", grid = FALSE) %>%
  add_axis("y", grid = FALSE) %>%
  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
  add_tooltip(function(df) df$info)
```

### 2014-12-09 prep_figures/commute/process.R

@@ -1,0 +1,20 @@
+library(ape, apTreeshape)
+
+shape_of_str <- function(s) as.treeshape(read.tree(text=s))
+
+
+get_results <- function(path) {
+    d <- read.table(path, stringsAsFactors=FALSE)
+    colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MCT")
+    d <- subset(d, !is.na(kappa))
+    d <- sample(d)  # Shuffle rows of d.
+    d$info= paste(d$t1_nwk, d$t2_nwk, sep="<br />")
+
+    # Tree shape information.
+    d$t1_shape <- lapply(d$t1_nwk, shape_of_str)
+    d$t2_shape <- lapply(d$t2_nwk, shape_of_str)
+    d$t1_colless <- sapply(d$t1_shape, colless)
+    d$t2_colless <- sapply(d$t2_shape, colless)
+
+    d
+}

### 2014-12-13 prep_figures/distribution/process.R
Shuffling later.

@@ -6,8 +6,8 @@ shape_of_str <- function(s) as.treeshape(read.tree(text=s))
 get_results <- function(path) {
     d <- read.table(path, stringsAsFactors=FALSE)
     colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "mult", "dist", "kappa_str")
+    d$row = 1:nrow(d)
     d <- subset(d, kappa_str != "-")
-    d <- sample(d)  # Shuffle rows of d.
     d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
     d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
     d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)

### 2014-12-13 prep_figures/commute/curvature-MCT.Rmd
Checking out commute time results.

@@
 ---
-title: "Ricci and commute time"
+title: "Ricci and transit times"
 runtime: shiny
@@
 
-d <- get_results("../../results-rspr/commute6.mat")
-d$dist <- as.factor(d$dist)
+# d <- get_results("../../results-rspr/commute-nodata/commute6.mat")
+# d$dist <- as.factor(d$dist)
+
+d5 <- get_results("../../results-rspr/commute-onesplit/123_45-1v5c.tsv")
+d6 <- get_results("../../results-rspr/commute-onesplit/123_456-1v5c.tsv")
+d <- d6
 
 d %>%
-  ggvis(x = ~kappa, y = ~MCT) %>%
-  layer_points(size = 5, fill = ~dist) %>%
+  ggvis(x = ~kappa, y = ~MAT) %>%
+  #ggvis(x = ~kappa, y = ~MCT) %>%
+  layer_points() %>%
   # A dummy layer so that we can use text in the tooltip.
@@
   add_axis("y", grid = FALSE) %>%
-  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
   add_tooltip(function(df) df$info)

### 2014-12-13 prep_figures/commute/process.R

@@ -5,9 +5,8 @@ shape_of_str <- function(s) as.treeshape(read.tree(text=s))
 
 get_results <- function(path) {
     d <- read.table(path, stringsAsFactors=FALSE)
-    colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MCT")
-    d <- subset(d, !is.na(kappa))
-    d <- sample(d)  # Shuffle rows of d.
+    # colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MCT") # no stderr
+    colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MAT_stderr", "MCT", "MAT_stderr")
     d$info= paste(d$t1_nwk, d$t2_nwk, sep="<br />")
 
     # Tree shape information.

### 2014-12-18 prep_figures/commute/commute-time-detail.py
Detailed summaries of commute times.

@@ -1,0 +1,44 @@
+#!/usr/bin/env python
+
+import argparse
+import collections
+import string
+
+parser = argparse.ArgumentParser(description='Find commute times',
+                                 prog='gen-tangles.py')
+
+parser.add_argument('tracefiles', nargs='+')
+parser.add_argument('--s1', action='store', help='The first state.', required=True)
+parser.add_argument('--s2', action='store', help='The second state.', required=True)
+
+args = parser.parse_args()
+s1 = args.s1
+s2 = args.s2
+
+interesting = [s1, s2]
+
+
+def posixify_newick(s):
+    return s.translate(string.maketrans("(),;", "CD_ "))
+
+
+for fname in args.tracefiles:
+    last_seen = {s: 0 for s in interesting}
+    transitions = {}
+    for a in interesting:
+        transitions[a] = {}
+        for b in interesting:
+            transitions[a][b] = collections.Counter()
+    with open('-'.join([fname, posixify_newick(s1), posixify_newick(s2)])+'.tab', 'w') as fout:
+        with open(fname, 'r') as fin:
+            for l in fin:
+                (_, idx, curr) = l.split()
+                idx = int(idx)
+                if curr in interesting:
+                    for prev in interesting:
+                        transitions[curr][prev][idx - last_seen[prev]] += 1
+                    last_seen[curr] = idx
+        for a in interesting:
+            for b in interesting:
+                for time, freq in transitions[a][b].items():
+                    fout.write('\t'.join([a, b, str(time), str(freq)])+'\n')

### 2014-12-18 prep_figures/commute/commute-detail.Rmd

--- 2014-12-13 prep_figures/commute/curvature-MCT.Rmd
@@
 d6 <- get_results("../../results-rspr/commute-onesplit/123_456-1v5c.tsv")
-d <- d6
+d <- d5
 

### 2014-12-18 prep_figures/commute/commute-detail.Rmd
commute time plots

--- 2014-12-09 prep_figures/commute/curvature-MCT.Rmd
@@
 ---
-title: "Ricci and commute time"
-runtime: shiny
+title: "Transit time distributions."
+runtime: knitr
 output: html_document
@@
 
-library(ggvis)
-library(ape)
-library(apTreeshape)
+library(ggplot2)
 
-source("process.R")
+grab_data <- function(fname) {
+  df <- read.table(fname, stringsAsFactors = FALSE)
+  colnames(df) <- c("t1", "t2", "time", "count")
+  df$cmp <- paste(df$t1, df$t2)
+  df$dens <- df$count / sum(df$count)
+  df
+}
 
-d <- get_results("../../results-rspr/commute6.mat")
-d$dist <- as.factor(d$dist)
+d <- rbind(
+  grab_data("runs/run1-uniq_trees_T-col-CCC1_2D_3D_C4_5DD-CC1_C2_3DD_C4_5DD.tab"),
+  grab_data("runs/run1-uniq_trees_T-col-CCCC1_2D_3D_4D_5D-CCCC1_2D_3D_5D_4D.tab"),
+  grab_data("runs/run1-uniq_trees_T-col-CCCC1_2D_3D_4D_5D-CCC1_2D_3D_C4_5DD.tab"))
 
-d %>%
-  ggvis(x = ~kappa, y = ~MCT) %>%
-  layer_points(size = 5, fill = ~dist) %>%
-  # A dummy layer so that we can use text in the tooltip.
-  layer_text(text := ~info, fontSize := 0) %>%
-  add_axis("x", grid = FALSE) %>%
-  add_axis("y", grid = FALSE) %>%
-  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
-  add_tooltip(function(df) df$info)
+sub_d <- subset(d, t1 != t2)
+
+ggplot(sub_d, aes(x=time, y=dens, color=cmp)) + geom_line() + scale_x_continuous(limits=c(0,100))
 ```

### 2014-12-18 prep_figures/commute/commute-time-detail.py
Making logic work correctly at beginning.

@@ -2,28 +2,33 @@
 
 import argparse
 import collections
+import re
 import string
 
 parser = argparse.ArgumentParser(description='Find commute times',
                                  prog='gen-tangles.py')
 
 parser.add_argument('tracefiles', nargs='+')
-parser.add_argument('--s1', action='store', help='The first state.', required=True)
-parser.add_argument('--s2', action='store', help='The second state.', required=True)
+parser.add_argument('--tp', help='A pair of trees of interest. String is trimmed between first open paren and last semicolon.', required=True)
 
 args = parser.parse_args()
-s1 = args.s1
-s2 = args.s2
+tp_str = args.tp
+tp_str = re.sub('^[^(]*', '', tp_str)
+tp_str = re.sub(';[^;]*$', '', tp_str)
 
-interesting = [s1, s2]
+# The states of interest to us.
+interesting = [s.rstrip(';') for s in tp_str.split()]
+(s1, s2) = interesting
+
+print "Searching for: " + str(interesting)
 
 
 def posixify_newick(s):
-    return s.translate(string.maketrans("(),;", "CD_ "))
+    return s.translate(string.maketrans("(),", "CD_"))
 
 
 for fname in args.tracefiles:
-    last_seen = {s: 0 for s in interesting}
+    last_seen = {s: None for s in interesting}
     transitions = {}
     for a in interesting:
         transitions[a] = {}
@@ -36,9 +41,15 @@ for fname in args.tracefiles:
                 idx = int(idx)
                 if curr in interesting:
                     for prev in interesting:
-                        transitions[curr][prev][idx - last_seen[prev]] += 1
+                        if last_seen[prev] is not None:
+                            # idx - last_seen[prev] is the number of steps
+                            # between the last time we saw prev and the current
+                            # index.
+                            transitions[prev][curr][idx - last_seen[prev]] += 1
                     last_seen[curr] = idx
+        print 'count\tfrom\tto'
         for a in interesting:
             for b in interesting:
                 for time, freq in transitions[a][b].items():
                     fout.write('\t'.join([a, b, str(time), str(freq)])+'\n')
+                print "\t".join([str(sum(transitions[a][b].values())), a, b])

### 2014-12-18 prep_figures/commute/commute-time-detail.py
Better CLI for commute-time-detail.

@@ -1,15 +1,26 @@
 #!/usr/bin/env python
+'''
+Suggested use:
+./commute-time-detail.py --tp \
+    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" <mcmc-trace-file>
+'''
 
 import argparse
 import collections
+import os
 import re
 import string
 
 parser = argparse.ArgumentParser(description='Find commute times',
                                  prog='gen-tangles.py')
 
-parser.add_argument('tracefiles', nargs='+')
-parser.add_argument('--tp', help='A pair of trees of interest. String is trimmed between first open paren and last semicolon.', required=True)
+parser.add_argument('tracefile')
+parser.add_argument(
+    '--tp',
+    help='A pair of trees of interest. \
+    String is trimmed between first open paren and last semicolon.',
+    required=True)
+parser.add_argument('-o', help='Out path')
 
 args = parser.parse_args()
 tp_str = args.tp
@@ -19,37 +30,43 @@ tp_str = re.sub(';[^;]*$', '', tp_str)
 # The states of interest to us.
 interesting = [s.rstrip(';') for s in tp_str.split()]
 (s1, s2) = interesting
-
 print "Searching for: " + str(interesting)
 
 
 def posixify_newick(s):
     return s.translate(string.maketrans("(),", "CD_"))
 
+trace = args.tracefile
+if args.o is None:
+    outpath = '-'.join([os.path.basename(trace),
+                        posixify_newick(s1),
+                        posixify_newick(s2)])+'.tab'
+else:
+    outpath = args.o
 
-for fname in args.tracefiles:
-    last_seen = {s: None for s in interesting}
-    transitions = {}
+last_seen = {s: None for s in interesting}
+transitions = {}
+for a in interesting:
+    transitions[a] = {}
+    for b in interesting:
+        transitions[a][b] = collections.Counter()
+
+with open(outpath, 'w') as fout:
+    with open(trace, 'r') as fin:
+        for l in fin:
+            (_, idx, curr) = l.split()
+            idx = int(idx)
+            if curr in interesting:
+                for prev in interesting:
+                    if last_seen[prev] is not None:
+                        # idx - last_seen[prev] is the number of steps
+                        # between the last time we saw prev and the current
+                        # index.
+                        transitions[prev][curr][idx - last_seen[prev]] += 1
+                last_seen[curr] = idx
+    print 'count\tfrom\tto'
     for a in interesting:
-        transitions[a] = {}
[…]

### 2014-12-18 prep_figures/commute/curvature-MCT.Rmd
A nice pipeline, working for 5 taxon data.

@@
 
-# d <- get_results("../../results-rspr/commute-nodata/commute6.mat")
-# d$dist <- as.factor(d$dist)
-
 d5 <- get_results("../../results-rspr/commute-onesplit/123_45-1v5c.tsv")
 d6 <- get_results("../../results-rspr/commute-onesplit/123_456-1v5c.tsv")
-d <- d6
+d <- d5[d5$MAT < 1000,]
 
@@
   #ggvis(x = ~kappa, y = ~MCT) %>%
-  layer_points() %>%
+  layer_text(text := ~num) %>%
   # A dummy layer so that we can use text in the tooltip.

### 2014-12-18 prep_figures/commute/get-runs-detail.sh

@@ -1,0 +1,6 @@
+
+#for ntangle in 7 14 66 68; do
+for ntangle in 69; do
+    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
+    ./commute-time-detail.py --tp "$($sedcmd)" ~/curvature-paths/123_45-1v5c/run1-uniq_trees_T-col -o run1.${ntangle}.tab
+done

### 2014-12-18 prep_figures/commute/process.R

@@ -7,6 +7,7 @@ get_results <- function(path) {
     d <- read.table(path, stringsAsFactors=FALSE)
     # colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MCT") # no stderr
     colnames(d) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "dist", "kappa", "MAT", "MAT_stderr", "MCT", "MAT_stderr")
+    d$num <- c(1:nrow(d))
     d$info= paste(d$t1_nwk, d$t2_nwk, sep="<br />")
 
     # Tree shape information.

### 2014-12-19 prep_figures/commute/commute-time-detail.py
Ignoring the new column from @cwhidden.

@@ -54,7 +54,7 @@ for a in interesting:
 with open(outpath, 'w') as fout:
     with open(trace, 'r') as fin:
         for l in fin:
-            (_, idx, curr) = l.split()
+            (_, idx, _, curr) = l.split()
             idx = int(idx)
             if curr in interesting:
                 for prev in interesting:

### 2014-12-19 prep_figures/commute/get-runs-detail.sh

@@ -1,6 +1,9 @@
 
-#for ntangle in 7 14 66 68; do
-for ntangle in 69; do
+rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
+
+for ntangle in 7 14 66 68 69; do
     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-    ./commute-time-detail.py --tp "$($sedcmd)" ~/curvature-paths/123_45-1v5c/run1-uniq_trees_T-col -o run1.${ntangle}.tab
+    ./commute-time-detail.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run1/uniq_trees_T -o runs/run1.${ntangle}.tab
 done
+
+#123_456-1v5c.tre

### 2014-12-19 prep_figures/commute/access-distn.Rmd
Less bad names for access distribution stuff.

--- 2014-12-18 prep_figures/commute/commute-detail.Rmd
@@
 
-d <- rbind(
-  grab_data("runs/run1-uniq_trees_T-col-CCC1_2D_3D_C4_5DD-CC1_C2_3DD_C4_5DD.tab"),
-  grab_data("runs/run1-uniq_trees_T-col-CCCC1_2D_3D_4D_5D-CCCC1_2D_3D_5D_4D.tab"),
-  grab_data("runs/run1-uniq_trees_T-col-CCCC1_2D_3D_4D_5D-CCC1_2D_3D_C4_5DD.tab"))
+data_of_num <- function(n) {
+  df <- grab_data(paste0("runs5/run1.",as.character(n),".tab"))
+  df$num <- as.factor(n)
+  df
+}
+
+d <- do.call(rbind, lapply(c(7,14,66,68,69), data_of_num))
 
 sub_d <- subset(d, t1 != t2)
+sub_d <- subset(d, t1 < t2)
 
-ggplot(sub_d, aes(x=time, y=dens, color=cmp)) + geom_line() + scale_x_continuous(limits=c(0,100))
+ggplot(sub_d, aes(x=time, y=dens, color=num)) + geom_line() + scale_x_continuous(limits=c(0,25))
 ```

### 2014-12-19 prep_figures/commute/auto-acess-time-count.sh

@@ -3,7 +3,7 @@ rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
 
 for ntangle in 7 14 66 68 69; do
     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-    ./commute-time-detail.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run1/uniq_trees_T -o runs/run1.${ntangle}.tab
+    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run1/uniq_trees_T -o 5-taxon-access-distn/run1.${ntangle}.tab
 done
 
 #123_456-1v5c.tre

### 2014-12-19 prep_figures/commute/access-time-count.py
Allowing multiple tracefiles.

@@ -2,25 +2,23 @@
 '''
 Suggested use:
 ./commute-time-detail.py --tp \
-    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" <mcmc-trace-file>
+    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" <mcmc-trace-files>
 '''
 
 import argparse
 import collections
-import os
 import re
-import string
 
 parser = argparse.ArgumentParser(description='Find commute times',
                                  prog='gen-tangles.py')
 
-parser.add_argument('tracefile')
+parser.add_argument('tracefiles', nargs='+')
 parser.add_argument(
     '--tp',
     help='A pair of trees of interest. \
     String is trimmed between first open paren and last semicolon.',
     required=True)
-parser.add_argument('-o', help='Out path')
+parser.add_argument('-o', help='Out path', required=True)
 
 args = parser.parse_args()
 tp_str = args.tp
@@ -32,39 +30,30 @@ interesting = [s.rstrip(';') for s in tp_str.split()]
 (s1, s2) = interesting
 print "Searching for: " + str(interesting)
 
-
-def posixify_newick(s):
-    return s.translate(string.maketrans("(),", "CD_"))
-
-trace = args.tracefile
-if args.o is None:
-    outpath = '-'.join([os.path.basename(trace),
-                        posixify_newick(s1),
-                        posixify_newick(s2)])+'.tab'
-else:
-    outpath = args.o
-
-last_seen = {s: None for s in interesting}
 transitions = {}
 for a in interesting:
     transitions[a] = {}
     for b in interesting:
         transitions[a][b] = collections.Counter()
 
-with open(outpath, 'w') as fout:
-    with open(trace, 'r') as fin:
-        for l in fin:
-            (_, idx, _, curr) = l.split()
-            idx = int(idx)
-            if curr in interesting:
-                for prev in interesting:
-                    if last_seen[prev] is not None:
-                        # idx - last_seen[prev] is the number of steps
-                        # between the last time we saw prev and the current
-                        # index.
-                        transitions[prev][curr][idx - last_seen[prev]] += 1
-                last_seen[curr] = idx
-    print 'count\tfrom\tto'
+with open(args.o, 'w') as fout:
+    for trace in args.tracefiles:
+        print trace
+        # Forget if we saw it in another trace file.
+        last_seen = {s: None for s in interesting}
+        with open(trace, 'r') as fin:
[…]

### 2014-12-19 prep_figures/commute/auto-acess-time-count.sh

@@ -3,7 +3,7 @@ rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
 
 for ntangle in 7 14 66 68 69; do
     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run1/uniq_trees_T -o 5-taxon-access-distn/run1.${ntangle}.tab
+    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/run1.${ntangle}.tab
 done
 
 #123_456-1v5c.tre

### 2014-12-19 prep_figures/commute/access-time-count.py
Aggregating commute times across runs for 5 taxa.

@@ -53,7 +53,7 @@ with open(args.o, 'w') as fout:
                             # index.
                             transitions[prev][curr][idx - last_seen[prev]] += 1
                     last_seen[curr] = idx
-        print 'count\tfrom\tto'
+    print 'count\tfrom\tto'
     for a in interesting:
         for b in interesting:
             for time, freq in transitions[a][b].items():

### 2014-12-19 prep_figures/commute/auto-acess-time-count.sh
6-taxon-access-distn

@@ -1,9 +1,13 @@
 
 rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
 
-for ntangle in 7 14 66 68 69; do
-    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/run1.${ntangle}.tab
+# for ntangle in 7 14 66 68 69; do
+#     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
+#     ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/${ntangle}.tab
+# done
+
+for ntangle in 662 665 795; do
+    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle6.idx"
+    ./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-1v5c/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/${ntangle}.tab
 done
 
-#123_456-1v5c.tre

### 2014-12-19 prep_figures/commute/curvature-MCT.Rmd
access-distn WIP.

@@
 d <- d5[d5$MAT < 1000,]
+d <- d6[d6$MAT < 1000,]
 

### 2014-12-19 prep_figures/commute/access-distn.R

@@ -1,0 +1,40 @@
+library(ggplot2)
+
+grab_data <- function(fname) {
+  df <- read.table(fname, stringsAsFactors = FALSE)
+  colnames(df) <- c("t1", "t2", "time", "count")
+  df$cmp <- paste(df$t1, df$t2)
+  df$dens <- df$count / sum(df$count)
+  df
+}
+
+n_taxa <- 5; favorite_tangles <- c(7,14,66,68,69)
+n_taxa <- 6; favorite_tangles <- c(662,665,795)
+sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
+data_of_num <- function(n) {
+  df <- grab_data(paste0(as.character(n_taxa),"-taxon-access-distn/",as.character(n),".tab"))
+  df$num <- as.factor(n)
+  df
+}
+
+
+adjacency <- read.csv(sub_n_taxa("../../rspr/matrix_NTAXA"))
+degree = rowSums(adjacency)
+tangles <- read.table(
+    sub_n_taxa("../../results-rspr/ricciNTAXA.mat"),
+    stringsAsFactors = FALSE)
+colnames(tangles) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "coset", "dist", "kappa_str")
+tangles$t1_deg <- sapply(tangles$t1_idx, function(i) degree[i+1]) # tangleX.idx uses 0 indexing.
+tangles$t2_deg <- sapply(tangles$t2_idx, function(i) degree[i+1]) # tangleX.idx uses 0 indexing.
+tangles$coset <- NULL
+tangles$num <- c(1:nrow(tangles))
+head(tangles)
+
+tail(d)
+d <- do.call(rbind, lapply(favorite_tangles, data_of_num))
+d <- merge(d, tangles, by="num")
+sub_d <- subset(d, t1 != t2)
+sub_d <- subset(d, t1 < t2)
+ggplot(sub_d, aes(x=time, y=count, color=num)) + geom_line() + scale_x_continuous(limits=c(0,25)) + facet_wrap(~ t1_deg)
+
+subset(sub_d, num==66)

### 2014-12-19 prep_figures/commute/access-distn.R
Fixing header degree bug, thanks to @cwhidden.

@@ -18,7 +18,7 @@ data_of_num <- function(n) {
 }
 
 
-adjacency <- read.csv(sub_n_taxa("../../rspr/matrix_NTAXA"))
+adjacency <- read.csv(sub_n_taxa("../../rspr/matrix_NTAXA"), header=FALSE)
 degree = rowSums(adjacency)
 tangles <- read.table(
     sub_n_taxa("../../results-rspr/ricciNTAXA.mat"),
@@ -30,11 +30,14 @@ tangles$coset <- NULL
 tangles$num <- c(1:nrow(tangles))
 head(tangles)
 
-tail(d)
 d <- do.call(rbind, lapply(favorite_tangles, data_of_num))
 d <- merge(d, tangles, by="num")
+d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
 sub_d <- subset(d, t1 != t2)
 sub_d <- subset(d, t1 < t2)
-ggplot(sub_d, aes(x=time, y=count, color=num)) + geom_line() + scale_x_continuous(limits=c(0,25)) + facet_wrap(~ t1_deg)
+p <- ggplot(sub_d, aes(x=time, y=count, group=num, color=kappa)) + geom_line() + scale_x_continuous(limits=c(0,25)) 
+p + facet_wrap(~ t1_deg)
+
++ facet_grid(t1_deg ~ t2_deg)
 
 subset(sub_d, num==66)

### 2014-12-19 prep_figures/commute/auto-acess-time-count.sh
More access times.

@@ -1,13 +1,15 @@
 
 rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
 
-# for ntangle in 7 14 66 68 69; do
-#     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-#     ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/${ntangle}.tab
-# done
+#for ntangle in 7 14 66 68 69; do
+#    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
+#    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-equal/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/equal/${ntangle}.tab
+#    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/123_45-1v5c/${ntangle}.tab
+#done
 
 for ntangle in 662 665 795; do
     sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle6.idx"
-    ./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-1v5c/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/${ntangle}.tab
+    ./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-equal/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/equal/${ntangle}.tab
+    #./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-1v5c/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/123_456-1v5c/${ntangle}.tab
 done
 

### 2014-12-19 prep_figures/commute/access-distn.R
May the lord forgive me, for I have shelled.

@@ -12,7 +12,7 @@ n_taxa <- 5; favorite_tangles <- c(7,14,66,68,69)
 n_taxa <- 6; favorite_tangles <- c(662,665,795)
 sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
 data_of_num <- function(n) {
-  df <- grab_data(paste0(as.character(n_taxa),"-taxon-access-distn/",as.character(n),".tab"))
+  df <- grab_data(paste0(as.character(n_taxa),"-taxon-access-distn/equal/",as.character(n),".tab"))
   df$num <- as.factor(n)
   df
 }
@@ -35,9 +35,11 @@ d <- merge(d, tangles, by="num")
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
 sub_d <- subset(d, t1 != t2)
 sub_d <- subset(d, t1 < t2)
-p <- ggplot(sub_d, aes(x=time, y=count, group=num, color=kappa)) + geom_line() + scale_x_continuous(limits=c(0,25)) 
+p <- ggplot(sub_d, aes(x=time, y=count, group=num, color=kappa)) + geom_line() + scale_x_continuous(limits=c(0,15)) 
 p + facet_wrap(~ t1_deg)
 
-+ facet_grid(t1_deg ~ t2_deg)
+summary(d$dist)
+
+p + facet_grid(t1_deg ~ t2_deg)
 
 subset(sub_d, num==66)

### 2014-12-19 prep_figures/commute/annotate-curvature.sh

@@ -1,0 +1,8 @@
+fname=$1
+
+s1=$(mktemp)
+cut -f 3 $fname | nw_topology -IL - | nw_order -c n - > $s1
+s2=$(mktemp)
+cut -f 4 $fname | nw_topology -IL - | nw_order -c n - > $s2
+
+paste $fname $s1 $s2 > annotated-rspr/$(basename $1)

### 2014-12-19 prep_figures/commute/same-shape-access-time-count.sh

@@ -1,0 +1,7 @@
+rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
+size=5
+
+for ntangle in $(awk '{if($8 == $9) print NR}' annotated-rspr/ricci${size}.mat); do
+    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle${size}.idx"
+    ./access-time-count.py --tp "$($sedcmd)" $rw_path/${size}-taxa/*-equal/run[1-9]/uniq_trees_T -o ${size}-taxon-access-distn/equal/${ntangle}.tab
+done

### 2014-12-19 prep_figures/commute/access-distn.R
Code for access distn plots.

@@ -8,8 +8,12 @@ grab_data <- function(fname) {
   df
 }
 
-n_taxa <- 5; favorite_tangles <- c(7,14,66,68,69)
+all_equal_shape_5_taxon = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 51, 52, 53, 54, 61, 62, 63, 64, 65, 66, 67, 68, 69)
+original_5_taxon <- c(7,14,66,68,69)
+n_taxa <- 5; favorite_tangles <- all_equal_shape_5_taxon
+
 n_taxa <- 6; favorite_tangles <- c(662,665,795)
+
 sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
 data_of_num <- function(n) {
   df <- grab_data(paste0(as.character(n_taxa),"-taxon-access-distn/equal/",as.character(n),".tab"))
@@ -28,17 +32,24 @@ tangles$t1_deg <- sapply(tangles$t1_idx, function(i) degree[i+1]) # tangleX.idx 
 tangles$t2_deg <- sapply(tangles$t2_idx, function(i) degree[i+1]) # tangleX.idx uses 0 indexing.
 tangles$coset <- NULL
 tangles$num <- c(1:nrow(tangles))
-head(tangles)
+tail(tangles)
 
 d <- do.call(rbind, lapply(favorite_tangles, data_of_num))
+d <- subset(d, t1 != t2)
+head(d)
 d <- merge(d, tangles, by="num")
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-sub_d <- subset(d, t1 != t2)
 sub_d <- subset(d, t1 < t2)
-p <- ggplot(sub_d, aes(x=time, y=count, group=num, color=kappa)) + geom_line() + scale_x_continuous(limits=c(0,15)) 
-p + facet_wrap(~ t1_deg)
 
-summary(d$dist)
+# Plot in terms of kappa.
+p <- ggplot(sub_d, aes(x=time, y=dens, group=num, color=kappa)) + geom_line()
+p <- p + scale_x_continuous(limits=c(0,7))
+p + facet_grid(dist ~ t1_deg, scales="free")
+
+# Plot showing exponential part of drop.
[…]

### 2014-12-20 prep_figures/commute/access-distn.R
access-distn to make ribbon plots.

@@ -1,4 +1,6 @@
+library(dplyr)
 library(ggplot2)
+library(magrittr)
 
 grab_data <- function(fname) {
   df <- read.table(fname, stringsAsFactors = FALSE)
@@ -8,16 +10,21 @@ grab_data <- function(fname) {
   df
 }
 
-all_equal_shape_5_taxon = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 51, 52, 53, 54, 61, 62, 63, 64, 65, 66, 67, 68, 69)
+all_equal_shape_5_taxon <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 51, 52, 53, 54, 61, 62, 63, 64, 65, 66, 67, 68, 69)
 original_5_taxon <- c(7,14,66,68,69)
 n_taxa <- 5; favorite_tangles <- all_equal_shape_5_taxon
+data <- "123_45-1v5c/"
+data <- "equal/"
 
-n_taxa <- 6; favorite_tangles <- c(662,665,795)
[…]

### 2015-01-07 prep_figures/commute/access-distn.R
Better plots for access distribution.

@@ -2,6 +2,8 @@ library(dplyr)
 library(ggplot2)
 library(magrittr)
 
+theme_set(theme_bw(16))
+
 grab_data <- function(fname) {
   df <- read.table(fname, stringsAsFactors = FALSE)
   colnames(df) <- c("t1", "t2", "time", "count")
@@ -47,16 +49,47 @@ head(d)
 d <- merge(d, tangles, by="num")
 d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
 
-d %>% 
+
+# Calculate the expectation of the commute time,
+# (compare below where we are looking at the complete distribution).
+expectations <- d %>%
+  group_by(num) %>%
+  summarise(expectation = sum(time*dens))
+expectations <- merge(expectations, tangles, by="num")
+
+p <- ggplot(expectations, aes(x=t1_deg, y=expectation, color=factor(t2_deg)))
+p <- p + geom_point(size=3.5)
+p
+
+
+# There are two entries for each (num, time) pair for the two directions,
+# assuming as we do that the trees are distinct.
+# Here we average their densities.
+# Alternatively, we could look at the two times individually,
+# but they are the same in the limit of lots of samples.
+means <- d %>%
   group_by(num, time) %>%
-  summarise(min_count = min(count), max_count = max(count), mean_count = mean(count)) -> count_extremes
-d <- merge(d, count_extremes)
+  summarise(mean_count = mean(count))
+d <- merge(d, means)
+# Because we are looking at means, we only take one representative.
+d <- subset(d, t1 < t2)
+
 
 # Plot in terms of kappa.
-p <- ggplot(d, aes(x=time, y=mean_count, ymin=min_count, ymax=max_count, group=num, color=kappa, fill=kappa)) 
-p <- p + geom_ribbon() + geom_line()
-p <- p + scale_x_continuous(limits=c(0,7)) + scale_y_log10()
+p <- ggplot(d, aes(x=time, y=mean_count, group=num, color=kappa, fill=kappa))
+p <- p + geom_line()
+p <- p + scale_x_continuous(limits=c(0,4)) + scale_y_log10()
 p + facet_grid(dist ~ t1_deg, scales="free")
+
+# Plot showing exponential part of drop.
[…]

### 2015-01-07 prep_figures/commute/access-distn.R
access-distn.R now makes a degree bar chart.

@@ -12,19 +12,24 @@ grab_data <- function(fname) {
   df
 }
 
-all_equal_shape_5_taxon <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 51, 52, 53, 54, 61, 62, 63, 64, 65, 66, 67, 68, 69)
 original_5_taxon <- c(7,14,66,68,69)
-n_taxa <- 5; favorite_tangles <- all_equal_shape_5_taxon
+all_equal_shape_5_taxon <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 51, 52, 53, 54, 61, 62, 63, 64, 65, 66, 67, 68, 69)
+all_5_taxon <- 1:69
+n_taxa <- 5; favorite_tangles <- all_5_taxon
 data <- "123_45-1v5c/"
-data <- "equal/"
+data <- "123_45-equal/"
 
 all_equal_shape_6_taxon <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 71, 72, 73, 74, 75, 76, 77, 78, 79, 80, 81, 82, 83, 84, 85, 86, 87, 88, 89, 90, 91, 92, 93, 94, 95, 96, 97, 98, 99, 100, 101, 102, 103, 104, 105, 106, 107, 108, 109, 110, 111, 112, 113, 114, 115, 449, 450, 451, 452, 453, 454, 455, 456, 457, 458, 459, 460, 461, 462, 548, 549, 550, 551, 552, 553, 554, 555, 556, 557, 558, 559, 560, 561, 562, 563, 564, 565, 566, 567, 568, 569, 570, 571, 572, 573, 574, 575, 576, 577, 578, 579, 580, 581, 582, 583, 584, 585, 586, 662, 665, 691, 692, 693, 694, 695, 696, 697, 698, 
[…]

### 2015-01-07 prep_figures/commute/auto-access-time-count.sh
Improving access time scripts.

@@ -2,24 +2,27 @@ set -eu
 
 rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
 
-# type="*-equal"
-
 # size=5
-# type=123_45-1v5c
+# type=123_45-equal
 
 size=6
-type=123_456-1v5c
+type=123_456-equal
+# type=123_456-1v5c
 
 
-for ntangle in $(awk '{if($8 == $9) print NR}' annotated-rspr/ricci${size}.mat); do
-    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle${size}.idx"
+tangle_file=../../tangle/rooted-symmetric/tangle${size}.idx
+
+for ntangle in $(seq $(wc -l $tangle_file | cut -f 1 -d' ')); do # loop over all tangles (yes, self-defeating)
+    sedcmd="sed -n ${ntangle}p $tangle_file"
     outdir=${size}-taxon-access-distn/${type}
     mkdir -p $outdir
     ./access-time-count.py --tp "$($sedcmd)" \
         $rw_path/${size}-taxa/${type}/run[1-9]/uniq_trees_T \
         -o $outdir/${ntangle}.tab &
 
-    while test $(ps -u matsen | grep python | wc -l) -gt 13; do
+    while test $(ps -u matsen | grep python | wc -l) -gt 12; do
         sleep 5;
     done
 done
+
+

### 2015-01-07 prep_figures/commute/auto-acess-time-count.sh

@@ -1,15 +1,0 @@
-
-rw_path=/shared/silo_researcher/Matsen_F/MatsenGrp/working/cwhidden/random_walks
-
-#for ntangle in 7 14 66 68 69; do
-#    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle5.idx"
-#    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-equal/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/equal/${ntangle}.tab
-#    ./access-time-count.py --tp "$($sedcmd)" $rw_path/5-taxa/123_45-1v5c/run[1-9]/uniq_trees_T -o 5-taxon-access-distn/123_45-1v5c/${ntangle}.tab
-#done
-
-for ntangle in 662 665 795; do
-    sedcmd="sed -n ${ntangle}p ../../tangle/rooted-symmetric/tangle6.idx"
-    ./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-equal/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/equal/${ntangle}.tab
-    #./access-time-count.py --tp "$($sedcmd)" $rw_path/6-taxa/123_456-1v5c/run[1-9]/uniq_trees_T -o 6-taxon-access-distn/123_456-1v5c/${ntangle}.tab
-done
-

### 2015-01-08 prep_figures/commute/access-distn.R
Colored bar plot for degrees.

@@ -29,7 +29,7 @@ data <- "123_456-1v5c/"
 
 sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
 data_of_num <- function(n) {
-  cat(n)
+  cat(paste0(n, ", "))
   df <- grab_data(paste0(as.character(n_taxa),"-taxon-access-distn/",data,as.character(n),".tab"))
   df$num <- as.numeric(n)
   df
@@ -62,8 +62,8 @@ expectations <- d %>%
   summarise(expectation = sum(time*dens))
 expectations <- merge(expectations, tangles, by="num")
 
-p <- ggplot(expectations, aes(x=expectation))
-p <- p + geom_histogram(binwidth=1) + labs(title=paste(n_taxa, "taxon access times"))
+p <- ggplot(expectations, aes(x=expectation, fill=factor(dist)))
+p <- p + geom_bar(binwidth=0.5) + labs(title=paste(n_taxa, "taxon access times"))
 p <- p + facet_grid(t1_deg ~ t2_deg) + scale_x_continuous(labels = function (x) floor(x))
 p + theme(panel.grid.minor =   element_blank(),
           panel.grid.major =   element_line(size=1.))

### 2015-01-08 prep_figures/commute/access-distn.R
ggsaving.

@@ -24,8 +24,8 @@ original_6_taxon <-  c(662,665,795)
 all_6_taxon <- 1:807
 n_taxa <- 6; favorite_tangles <- all_6_taxon
 data <- "equal/"
-data <- "123_456-equal/"
 data <- "123_456-1v5c/"
+data <- "123_456-equal/"
 
 sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
 data_of_num <- function(n) {
@@ -34,7 +34,6 @@ data_of_num <- function(n) {
   df$num <- as.numeric(n)
   df
 }
-
 
 adjacency <- read.csv(sub_n_taxa("../../rspr/matrix_NTAXA"), header=FALSE)
 degree = rowSums(adjacency)
@@ -65,9 +64,9 @@ expectations <- merge(expectations, tangles, by="num")
 p <- ggplot(expectations, aes(x=expectation, fill=factor(dist)))
 p <- p + geom_bar(binwidth=0.5) + labs(title=paste(n_taxa, "taxon access times"))
 p <- p + facet_grid(t1_deg ~ t2_deg) + scale_x_continuous(labels = function (x) floor(x))
-p + theme(panel.grid.minor =   element_blank(),
-          panel.grid.major =   element_line(size=1.))
-ggsave(paste0(n_taxa,"-taxon-access-by-degree.svg"))
+p <- p + theme(panel.grid.minor =   element_blank(),
+               panel.grid.major =   element_line(size=1.))
+ggsave(paste0(n_taxa,"-taxon-access-by-degree.svg"), p)
 
 # There are two entries for each (num, time) pair for the two directions,
 # assuming as we do that the trees are distinct.

### 2015-01-08 prep_figures/distribution/scatter-ggvis.Rmd
Tidying up scatterplots.

---
title: "Ricci"
runtime: shiny
output: html_document
---

```{r, echo=FALSE}

library(ggvis)
library(ape)
library(apTreeshape)

source("process.R")
d <- get_results("../../results-rspr/ricci4.mat")
d <- sample(d)  # Shuffle rows of d.

d %>%
  ggvis(x = ~kappa, y = ~dist_jitter) %>%
#  layer_points(size = 5, fill = ~t1_nwk, stroke = ~t2_idx, strokeOpacity := 0) %>%
  layer_points(size = 12, fill = ~t1_nwk) %>%
#  layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
  # A dummy layer so that we can use text in the tooltip.
  # layer_text(text := ~info, fontSize := 0) %>%
  add_axis("x", grid = FALSE) %>%
  add_axis("y", grid = FALSE, values = 1:3) %>%
  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) 
  #add_legend(c("stroke"), properties=legend_props(legend=list(y=100))) %>%
  #add_legend(c("size"), properties=legend_props(legend=list(y=200))) %>%
  #add_tooltip(function(df) df$info)
```

### 2015-01-15 curvature.tex
Tex fix.
gh-4: first go at proving flatness for small moves.

@@
 
+\section{Curvature}
+
+Note that we have to write out simple random walk versus degree-informed.
+Can we just say that degree differences are small relative to $n$? Is that true?
+
+\begin{proposition}
+Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
+Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence but with $R$ cut off and attached at a different location for each $T_n$.
+Then $\lim_{n \rightarrow \infty} \kappa(T_n, S_n) = 0$ on the SPR graph.
+\end{proposition}
+\begin{proof}
+First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(O(n)))/q_n$, where $q_n$ is the number of neighbors of $T_n$ (OK, this gets messy when $T_n$ and $S_n$ have different numbers of neighbors).
+Let $T'_n$ be any SPR move away from $T_n$ where the SPR move does not cut $R$ into two and doesn't exclusively move $R$.
+In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ (need to show this makes sense), such that $\dspr(T'_n, S'_n) = 1$.
+This works for every tree that doesn't have $R$ cut in two.
+The number of trees that \emph{do} have $R$ cut in two or moved by its root is $(2k-1)(O(n))$, and their distance is at most 3.
+
+A lower bound is also available because that we can't do better than distance 1 for all $(T'_n, S'_n)$ pairs.
+If we completely ignore the $(2k-1)(O(n))$ trees that cut $R$ in two, then we get a lower bound of $(q_n-(2k-1)(O(n)))/q_n$.
+
+The proposition follows because $q_n$ is quadratic in $n$.
+\end{proof}
+
 \section{$k$-tubes}
@@
 	\label{lem:neighbor_distance}
-	$\dspr{t_1, t_2} \le \dspr{T_1, T_2} + 2k,\ \forall_{t_1 \in N_k(T_1)},\ \forall_{t_2 \in N_k(T_2)} } .$
+	$\dspr{t_1, t_2} \le \dspr{T_1, T_2} + 2k,\ \forall_{t_1 \in N_k(T_1)},\ \forall_{t_2 \in N_k(T_2)}.$
 \end{lemma}

### 2015-01-21 curvature.tex
statement of degree change lemma

@@
 \newattribute{omegaOf}{\omega}
+\newattribute{lca}{\text{LCA}}
 %\newattribute{min}{\text{min}}
@@
 \section{Introduction}
+
+\section{Rooted SPR Neighborhoods}
+
+\begin{lemma}
+Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
+The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+If multiple SPR moves transform $T$ into $S$ (i.e. an NNI\todo{lemma}), then assume that $V$ is the aunt of $R$ in $T$.
+Let $L$ be the $\lca{U,V}$ in $T$.
+Let $a$ be the distance from $U$ to $L$ in $T$, and $b$ the distance from $V$ to $L$ in $T$.
+Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
+Then the degree of $T$ and $S$ differs by:
+$$(2k-1)(a-b-1) + 2(i - j).$$
+
+\end{lemma}
+\begin{proof}
+
+Proof strategy:
+
+Gain $(2k-1)(a-2)$ moves from the $a$-path to $R$.
+
+Lose $(2k-1)(b-1)$ moves from the $b$-path to $R$.
+
+Gain $2(i-j)$ moves involving the parent of $R$ to $U$ and $V$.
+
+\end{proof}
 

### 2015-01-21 ricci-tangle.py
Proper paths for *-fun.py

@@ -1,14 +1,12 @@
 #!/usr/bin/env sage
 
 import argparse
-import math
-import time
 import sys
 from multiprocessing import Pool
 from sage.all import *
-load("tree-fun.py")
+load("gricci/all-hail-sage/tree-fun.py")
 load("gricci/code.py")
-load("gricci/nb-fun.py")
+load("gricci/all-hail-sage/nb-fun.py")
 
 
 parser = argparse.ArgumentParser(description='ricci curvature of tangles',

### 2015-01-21 ricci-tangle.py
Using more generalized gricci.

@@ -31,7 +31,7 @@ def process_line(line):
     if t1_idx == t2_idx:
         row.extend([0, "-"])
     else:
-        calc = ric_unif_rw(g, source=t1_idx, target=t2_idx)
+        calc = ricci('lazy_unif', g, source=t1_idx, target=t2_idx)
         row.extend([calc.dist, calc.ric])
     return row
 

### 2015-01-21 ricci-tangle.py
Generalizing to allow for other walks from shell.

@@ -18,6 +18,8 @@ parser.add_argument('--graph', dest='graph_path',
                     type=str, help='Path to graph in SAGE object format.')
 parser.add_argument('--out', dest='out_path',
                     type=str, help='Output path.')
+parser.add_argument('--walk', choices=['lazy_unif', 'unif_prior_mh'],
+                    default='lazy_unif', help='What random walk to use.')
 args = parser.parse_args()
 assert(os.path.exists(args.idx_path))
 assert(os.path.exists(args.graph_path))
@@ -31,7 +33,7 @@ def process_line(line):
     if t1_idx == t2_idx:
         row.extend([0, "-"])
     else:
-        calc = ricci('lazy_unif', g, source=t1_idx, target=t2_idx)
+        calc = ricci(args.walk, g, source=t1_idx, target=t2_idx)
         row.extend([calc.dist, calc.ric])
     return row
 

### 2015-01-21 sbatch-ricci.sh

@@ -4,12 +4,16 @@ set -u
 
 test -z $1 && exit 1
 
+# walk=lazy_unif
+walk=unif_prior_mh
+mkdir -p results-rspr/$walk
+
 n_leaves=$1
 split_size=50
 slurm_config="-c 4 -n 1"
 
 # Prepare vars and split file.
-out_path="ricci$n_leaves.mat"
+out_path="results-rspr/$walk/ricci$n_leaves.mat"
 splitargs="-a 4 -l $split_size tangle/rooted-symmetric/tangle$n_leaves.idx"
 files=$(split $splitargs --filter="echo \$FILE")
 # Apparently split doesn't actually split when given the --filter flag, so we run it again.
@@ -19,7 +23,7 @@ split $splitargs
 for i in $files
 do
     export SLURM_JOB_NAME=ricci-$i
-    cmd="./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path-$i $i"
+    cmd="./ricci-tangle.py --walk $walk --graph rspr/graph_$n_leaves.sobj --out $out_path-$i $i"
     echo $cmd
     echo "#!/bin/sh" > $i.sh
     echo $cmd >> $i.sh

### 2015-01-21 simple-ricci.sh

@@ -4,9 +4,13 @@ set -u
 
 test -z $1 && exit 1
 
+# walk=lazy_unif
+walk=unif_prior_mh
+mkdir -p results-rspr/$walk
+
 n_leaves=$1
 
 idx_path="tangle/rooted-symmetric/tangle$n_leaves.idx"
-out_path="results-rspr/ricci$n_leaves.mat"
+out_path="results-rspr/$walk/ricci$n_leaves.mat"
 
-./ricci-tangle.py --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path
+./ricci-tangle.py --walk $walk --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path

### 2015-01-21 curvature.tex
SPR degree change correction and proof

@@
 The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
-If multiple SPR moves transform $T$ into $S$ (i.e. an NNI\todo{lemma}), then assume that $V$ is the aunt of $R$ in $T$.
 Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the distance from $U$ to $L$ in $T$, and $b$ the distance from $V$ to $L$ in $T$.
+Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$ excluding endpoints.
+Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
 Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
 Then the degree of $T$ and $S$ differs by:
-$$(2k-1)(a-b-1) + 2(i - j).$$
+$$2(k(a-b) + i - j).$$
 
@@
 
-Proof strategy:
+Observe that the set of permissible moves changes after the movement of $R$ in four different ways: 1) Subtrees on the path from $U$ to $L$ may now be moved within $R$ and its newly introduced parent, 2) Subtrees on the path from $V$ to $L$ may no longer be moved within $R$ and its parent, 3) $R$'s parent subtree may now be moved within $U$, and 4) $R$'s parent subtree may no longer be moved within $V$.
+Note that no additional moves are introduced are blocked by the SPR operation.
 
-Gain $(2k-1)(a-2)$ moves from the $a$-path to $R$.
+In the first case, there are $a$ subtrees that can now be moved to the $2k$ positions within $R$, including its newly introduced parent node, for a total gain of $2ka$ moves.
+Similarly, we lose $2kb$ moves in the second case.
 
-Lose $(2k-1)(b-1)$ moves from the $b$-path to $R$.
+In the third case, $R's$ parent subtree may now make $2(i-1)$ moves within $U$.
+Similarly, we lose $2(j-1)$ moves in the fourth case.
 
-Gain $2(i-j)$ moves involving the parent of $R$ to $U$ and $V$.
-
+The difference in SPR degree is thus:
+\begin{align*}
+&2ka - 2kb + 2(i-1) - 2(j-1) \\
+=\  &2(k(a-b) + i - j)
+\end{align*}
 \end{proof}

### 2015-01-21 sbatch-ricci.sh
Fix sbatch-ricci.sh and first results.

@@ -37,9 +37,9 @@ do
     sleep 5
 done
 
-rm results-rspr/$out_path
-touch results-rspr/$out_path
+rm -f $out_path
+touch $out_path
 for i in $files
 do
-    cat $out_path-$i >> results-rspr/$out_path
+    cat $out_path-$i >> $out_path
 done

### 2015-01-22 curvature.tex
lemma on min size of paired trees, rewrote curvature limit proof

@@
 \begin{lemma}
+\label{lem:degree_change}
 Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
@@
 Observe that the set of permissible moves changes after the movement of $R$ in four different ways: 1) Subtrees on the path from $U$ to $L$ may now be moved within $R$ and its newly introduced parent, 2) Subtrees on the path from $V$ to $L$ may no longer be moved within $R$ and its parent, 3) $R$'s parent subtree may now be moved within $U$, and 4) $R$'s parent subtree may no longer be moved within $V$.
-Note that no additional moves are introduced are blocked by the SPR operation.
+Note that no additional moves are introduced or blocked by the SPR operation.
 
@@
 
+\begin{corollary}
+\label{cor:paired_neighbors}
+Let the degrees of $T$ and $S$ be $q$ and $p$ respectively.
+Then at least $o = q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that:
+\begin{enumerate}
+\item $\dspr{T',S'} = 1$ for each $(T',S')$ pair, and
+\item each neighbor of $T$ and each neighbor of $S$ is paired to exactly one tree.
+\end{enumerate}
+\end{corollary}
+\begin{proof}
+By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
+For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
+\end{proof}
+
 \section{Curvature}
@@
 \begin{proof}
-First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(O(n)))/q_n$, where $q_n$ is the number of neighbors of $T_n$ (OK, this gets messy when $T_n$ and $S_n$ have different numbers of neighbors).
-Let $T'_n$ be any SPR move away from $T_n$ where the SPR move does not cut $R$ into two and doesn't exclusively move $R$.
-In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ (need to show this makes sense), such that $\dspr(T'_n, S'_n) = 1$.
-This works for every tree that doesn't have $R$ cut in two.
-The number of trees that \emph{do} have $R$ cut in two or moved by its root is $(2k-1)(O(n))$, and their distance is at most 3.
+First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(\OhOf{n}))/q_n$, where $q_n$ is the number of neighbors of $T_n$ and $S_n$.
+Let $T'_n$ be any of the $o$ SPR moves from a pairing following Corollary~\ref{cor:paired_neighbors}.
+In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ such that $\dspr{T'_n, S'_n} = 1$.
+The number of trees unmatched by this pairing is $\OhOf{kn}$, and their distance is at most 3.
+Thus, $\kappa$ is bounded above by $(q_n + 2\OhOf{kn}) / q_n$.
 
-A lower bound is also available because that we can't do better than distance 1 for all $(T'_n, S'_n)$ pairs.
-If we completely ignore the $(2k-1)(O(n))$ trees that cut $R$ in two, then we get a lower bound of $(q_n-(2k-1)(O(n)))/q_n$.
+A lower bound is also available because we can't do better than distance 1 for all trees unless they move $R$.
+There are only $\OhOf{n}$ pairs with trees that move $R$.
+By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$.
 
 The proposition follows because $q_n$ is quadratic in $n$.
+
+Now, we prove the proposition when the number of neighbors differ.
+Let $q_n$ and $p_n$ be the number of neighbors of $T_n$ and $S_n$, respectively.
+Assume without loss of generality that $q_n < p_n$.
+By Lemma~\ref{lem:degree_change}, $p_n - q_n = 2(k(a-b) + i - j)$, where $a,b,i,j < n$.
+Thus, $p_n - q_n = \OhOf{kn}$.
+We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $p_n > q_n$ we can only account for at most $q_n / p_n$ of the mass directly and may have to move the $(p_n - q_n) / p_n$ remainder to trees at distance at most 3.
+Thus, $\kappa$ is bounded above by $(q_n + 3\OhOf{kn}) / p_n = (p_n + 2\OhOf{kn}) / p_n$.
+We again bound $\kappa$ from below with $(q_n - \OhOf{n}) / q_n$ by ignoring the mass in trees that may overlap.
+The proposition again follows because $q_n$ is quadratic in $n$.
+
 \end{proof}

### 2015-01-22 prep_figures/distribution/lazy-RW-vs-unif-MH.R
Scatterplots comparing lazy to MH.

@@ -1,0 +1,29 @@
+#! /usr/bin/Rscript
+
+library(ggplot2)
+library(ape)
+library(apTreeshape)
+
+source("process.R")
+
+get_df <- function(n_taxa, walk) {
+  d <- get_results(paste0("../../results-rspr/", walk, "/ricci", n_taxa,".mat"))
+  d <- sample(d)  # Shuffle rows of d.
+  d$n_taxa <- n_taxa
+  colnames(d)[colnames(d) == "kappa"] <- paste0("kappa_", walk)
+  colnames(d)[colnames(d) == "kappa_str"] <- paste0("kappa_str_", walk)
+  d$dist_jitter <- NULL
+  d
+}
+
+theme_set(theme_bw(16))
+
+n_taxa <- 7
+d <- merge(get_df(n_taxa, "lazy_unif"), get_df(n_taxa, "unif_prior_mh"))
+
+p <- ggplot(d, aes(kappa_lazy_unif, kappa_unif_prior_mh, color=t1_nwk))
+p <- p + geom_point()
+p <- p + scale_x_continuous(name="κ for lazy random walk")
+p <- p + scale_y_continuous(name="κ for MH sampling of uniform prior")
+
+ggsave(paste0("kappa", n_taxa, "-lazy-RW-vs-unif-MH.svg"), p, width=9)

### 2015-01-22 prep_figures/distribution/scatter-ggvis.Rmd
Tidying up plotting.

@@
 source("process.R")
-d <- get_results("../../results-rspr/ricci4.mat")
+
+get_df <- function(n_taxa, walk) {
+  d <- get_results(paste0("../../results-rspr/", walk, "/ricci", n_taxa,".mat"))
   d <- sample(d)  # Shuffle rows of d.
+  d$n_taxa <- n_taxa
+  d
+}
+
+n_taxa <- 6
+
+walk <- "unif_prior_mh"
+walk <- "lazy_unif"
+d <- get_df(n_taxa, walk)
+
+props <- axis_props(labels=list(fontSize=16), title=list(fontSize=24))
 
@@
   # A dummy layer so that we can use text in the tooltip.
-  # layer_text(text := ~info, fontSize := 0) %>%
-  add_axis("x", grid = FALSE) %>%
-  add_axis("y", grid = FALSE, values = 1:3) %>%
-  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) 
+  layer_text(text := ~info, fontSize := 0) %>%
+  add_axis("x", title = "κ", grid = FALSE, properties=props) %>%
+  add_axis("y", title = "distance", grid = FALSE, values=1:4, properties=props) %>%
+#  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
   #add_legend(c("stroke"), properties=legend_props(legend=list(y=100))) %>%
   #add_legend(c("size"), properties=legend_props(legend=list(y=200))) %>%
-  #add_tooltip(function(df) df$info)
+  export_svg(paste0("ricci", n_taxa, "-", walk,".svg"))
+#  add_tooltip(function(df) df$info)
+#  function (x) x # null
 ```

### 2015-01-22 prep_figures/distribution/process.R

@@ -9,7 +9,8 @@ get_results <- function(path) {
     d$row = 1:nrow(d)
     d <- subset(d, kappa_str != "-")
     d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
-    d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
+    d$info= paste(d$t1_nwk, d$t2_nwk, sep="<br />")
+    # d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
     d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
 
     # Tree shape information.

### 2015-01-23 prep_figures/commute/access-distn.R
Greatly improving access time plots.

@@ -2,7 +2,7 @@ library(dplyr)
 library(ggplot2)
 library(magrittr)
 
-theme_set(theme_bw(16))
+theme_set(theme_bw(22))
 
 grab_data <- function(fname) {
   df <- read.table(fname, stringsAsFactors = FALSE)
@@ -38,7 +38,7 @@ data_of_num <- function(n) {
 adjacency <- read.csv(sub_n_taxa("../../rspr/matrix_NTAXA"), header=FALSE)
 degree = rowSums(adjacency)
 tangles <- read.table(
-    sub_n_taxa("../../results-rspr/ricciNTAXA.mat"),
+    sub_n_taxa("../../results-rspr/unif_prior_mh/ricciNTAXA.mat"),
     stringsAsFactors = FALSE)
 colnames(tangles) <- c("t1_idx", "t2_idx", "t1_nwk", "t2_nwk", "coset", "dist", "kappa_str")
 tangles$t1_deg <- sapply(tangles$t1_idx, function(i) degree[i+1]) # tangleX.idx uses 0 indexing.
@@ -66,6 +66,7 @@ p <- p + geom_bar(binwidth=0.5) + labs(title=paste(n_taxa, "taxon access times")
 p <- p + facet_grid(t1_deg ~ t2_deg) + scale_x_continuous(labels = function (x) floor(x))
 p <- p + theme(panel.grid.minor =   element_blank(),
                panel.grid.major =   element_line(size=1.))
+p <- p + scale_fill_discrete("distance")
 ggsave(paste0(n_taxa,"-taxon-access-by-degree.svg"), p)
 
 # There are two entries for each (num, time) pair for the two directions,
@@ -76,34 +77,29 @@ ggsave(paste0(n_taxa,"-taxon-access-by-degree.svg"), p)
 means <- d %>%
   group_by(num, time) %>%
   summarise(mean_count = mean(count))
+means$mean_dens <- means$mean_count / sum(means$mean_count)
 d <- merge(d, means)
 # Because we are looking at means, we only take one representative.
 d <- subset(d, t1 < t2)
 
+make_lims <- function(max_index, column) {
+  relevants <- d[d$time %in% 0:max_index, colnames(d) == column]
+  lims <- c(min(relevants), max(relevants))
+}
 
 # Plot in terms of kappa.
[…]

### 2015-01-23 curvature.tex
Pass through flatness proposition.
Please check, @cwhidden.

@@
 \usepackage{todonotes}
+\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
 
@@
 \begin{document}
-\title{Ricci curvature of the phylogenetic subtree-prune-regraft graph}
+\title{Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph}
 \author[Whidden]{Christopher Whidden}
@@
 
+This isn't quite the right graph to be looking at (rooted).
+
+We are talking about rooted.
+Parent and parent subtree.
+Define adjacency of subtrees.
+
 \section{Rooted SPR Neighborhoods}
@@
 Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$ excluding endpoints.
+Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
 Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
 Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
-Then the degree of $T$ and $S$ differs by:
-$$2(k(a-b) + i - j).$$
+Then the degrees of $T$ and $S$ differ by:
+$$2\left(k(a-b) + i - j\right).$$
 
@@
 
-Observe that the set of permissible moves changes after the movement of $R$ in four different ways: 1) Subtrees on the path from $U$ to $L$ may now be moved within $R$ and its newly introduced parent, 2) Subtrees on the path from $V$ to $L$ may no longer be moved within $R$ and its parent, 3) $R$'s parent subtree may now be moved within $U$, and 4) $R$'s parent subtree may no longer be moved within $V$.
-Note that no additional moves are introduced or blocked by the SPR operation.
+Observe that the set of permissible SPR moves changes after the movement of $R$ in four different ways:
+1) Subtrees including vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
+2) Subtrees including vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
+3) $R$'s parent subtree may now be moved into $U$, and
+4) $R$'s parent subtree may no longer be moved into $V$.
+No additional moves are introduced or blocked by the original SPR operation on $R$.
 
-In the first case, there are $a$ subtrees that can now be moved to the $2k$ positions within $R$, including its newly introduced parent node, for a total gain of $2ka$ moves.
+Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges (note that we are excluding any ``root edge'' in this number, so this tree has a degree two root).
+In the first case there are $a$ subtrees that can now be moved to the $2k$ positions into $R$ (including its newly introduced parent node and one of the newly subdivided root edge of $V$) for a total gain of $2ka$ distinct moves.
 Similarly, we lose $2kb$ moves in the second case.
-
-In the third case, $R's$ parent subtree may now make $2(i-1)$ moves within $U$.
+In the third case, $R's$ parent subtree may now make $2(i-1)$ moves into $U$.
 Similarly, we lose $2(j-1)$ moves in the fourth case.
 
-The difference in SPR degree is thus:
-\begin{align*}
-&2ka - 2kb + 2(i-1) - 2(j-1) \\
-=\  &2(k(a-b) + i - j)
-\end{align*}
+The difference in SPR degree is thus $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
 \end{proof}
@@
 \label{cor:paired_neighbors}
-Let the degrees of $T$ and $S$ be $q$ and $p$ respectively.
-Then at least $o = q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that:
+Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ nodes in the SPR graph be $q$ and $p$ respectively.
+Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that:
 \begin{enumerate}
@@
 \section{Curvature}
-
 Note that we have to write out simple random walk versus degree-informed.
@@
 
+For the purposes of this paper, $\kappa_n$ is curvature on the $n$-taxon rSPR graph and $W_{1,n}$ is the corresponding mass transport term.
+
 \begin{proposition}
 Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
-Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence but with $R$ cut off and attached at a different location for each $T_n$.
-Then $\lim_{n \rightarrow \infty} \kappa(T_n, S_n) = 0$ on the SPR graph.
+Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
+Then $\lim_{n \rightarrow \infty} \kappa_n(T_n, S_n) = 0$ for \todo{TBA} random walk on the SPR graph.
 \end{proposition}
 \begin{proof}
-First we show that $\kappa$ is bounded above by $(q_n+3(2k-1)(\OhOf{n}))/q_n$, where $q_n$ is the number of neighbors of $T_n$ and $S_n$.
+Because $d(T_n, S_n) = 1$, we will prove the proposition by showing that the mass transport term $W_{1,n}$ sits between two bounds with limit 1 as $n$ goes to infinity.
+
+To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
+First we show that $W_{1,n}$ is bounded above by $(q_n+3(2k-1)(\OhOf{n}))/q_n$, where $q_n$ is the number of neighbors of $T_n$ and $S_n$.
 Let $T'_n$ be any of the $o$ SPR moves from a pairing following Corollary~\ref{cor:paired_neighbors}.
-In the mass transport, we pair $T'_n$ with $S'_n$, which is the same move applied to $S_n$ such that $\dspr{T'_n, S'_n} = 1$.
+In the mass transport, we pair $T'_n$ with $S'_n$, which results from the same move applied to $S_n$ such that $\dspr{T'_n, S'_n} = 1$.
 The number of trees unmatched by this pairing is $\OhOf{kn}$, and their distance is at most 3.
-Thus, $\kappa$ is bounded above by $(q_n + 2\OhOf{kn}) / q_n$.
+Thus, $W_{1,n}$ is bounded above by $(q_n + 2\OhOf{kn}) / q_n$.
 
@@
 There are only $\OhOf{n}$ pairs with trees that move $R$.
-By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$.
+By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$ for $W_{1,n}$.
 
-The proposition follows because $q_n$ is quadratic in $n$.
+The desired control of $W_{1,n}$ is thus obtained because $q_n$ is quadratic in $n$.
 
@@
 Thus, $p_n - q_n = \OhOf{kn}$.
-We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $p_n > q_n$ we can only account for at most $q_n / p_n$ of the mass directly and may have to move the $(p_n - q_n) / p_n$ remainder to trees at distance at most 3.
-Thus, $\kappa$ is bounded above by $(q_n + 3\OhOf{kn}) / p_n = (p_n + 2\OhOf{kn}) / p_n$.
-We again bound $\kappa$ from below with $(q_n - \OhOf{n}) / q_n$ by ignoring the mass in trees that may overlap.
+We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $p_n > q_n$ we can only account for at most $q_n / p_n$ of the mass directly and may have to move the $(p_n - q_n) / p_n$ remainder to trees a distance at most 3.
+Thus, $W_{1,n}$ is bounded above by $(q_n + 3\OhOf{kn}) / p_n = (p_n + 2\OhOf{kn}) / p_n$.
+We again bound $W_{1,n}$ from below with $(q_n - \OhOf{n}) / q_n$ by ignoring the mass in trees that may overlap.
 The proposition again follows because $q_n$ is quadratic in $n$.

### 2015-01-23 prep_figures/commute/access-distn.R
access-distn fixups

@@ -88,10 +88,10 @@ make_lims <- function(max_index, column) {
 }
 
 # Plot in terms of kappa.
-max_index <- 7
+max_index <- 5
 p <- ggplot(d, aes(x=time, y=mean_dens, group=num, color=kappa, fill=kappa))
 p <- p + geom_line()
-p <- p + scale_x_continuous(limits=c(0,max_index))
+p <- p + scale_x_continuous(limits=c(1,max_index))
 p <- p + scale_y_continuous("density", breaks=c())
 # p <- p + scale_y_continuous("density", limits=make_lims(max_index, "mean_dens"), breaks=c())
 p <- p + facet_grid(dist ~ t1_deg)
@@ -99,7 +99,8 @@ ggsave(paste0(as.character(n_taxa),"-taxon-access-time-kappa-short.svg"), p)
 
 # Plot showing long time info.
 p <- ggplot(d, aes(x=time, y=mean_dens, group=num, color=kappa)) + geom_line()
-p <- p + scale_x_continuous(limits=c(0,500)) + scale_y_continuous("density", breaks=c())
+p <- p + scale_x_continuous(limits=c(1,2500)) + scale_y_continuous("density", breaks=c())
 p <- p + facet_grid(dist ~ t1_deg, scales="free")
-ggsave(paste0(as.character(n_taxa),"-taxon-access-time-kappa-long.svg"), p, width=9.5)
+p <- p + theme(axis.text.x=element_text(angle=-90))
+ggsave(paste0(as.character(n_taxa),"-taxon-access-time-kappa-long.png"), p, width=9.5)
 

### 2015-01-24 prep_figures/commute/access-distn.R
access-distn: just taking one panel.

@@ -19,13 +19,14 @@ n_taxa <- 5; favorite_tangles <- all_5_taxon
 data <- "123_45-1v5c/"
 data <- "123_45-equal/"
 
+if(FALSE) {
 all_equal_shape_6_taxon <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 71, 72, 73, 74, 75, 76, 77, 78, 79, 80, 81, 82, 83, 84, 85, 86, 87, 88, 89, 90, 91, 92, 93, 94, 95, 96, 97, 98, 99, 100, 101, 102, 103, 104, 105, 106, 107, 108, 109, 110, 111, 112, 113, 114, 115, 449, 450, 451, 452, 453, 454, 455, 456, 457, 458, 459, 460, 461, 462, 548, 549, 550, 551, 552, 553, 554, 555, 556, 557, 558, 559, 560, 561, 562, 563, 564, 565, 566, 567, 568, 569, 570, 571, 572, 573, 574, 575, 576, 577, 578, 579, 580, 581, 582, 583, 584, 585, 586, 662, 665, 691, 692, 693, 694, 695, 696, 697, 698, 699, 700, 701, 702, 703, 704, 705, 706, 707, 708, 709, 710, 711, 712, 713, 714, 715, 716, 717, 718, 719, 720, 721, 722, 723, 724, 725, 726, 727, 728, 729, 776, 777, 778, 779, 780, 781, 782, 794, 795, 796, 797, 798, 799, 800, 801, 802, 803, 804, 805, 806, 807)
 original_6_taxon <-  c(662,665,795)
 all_6_taxon <- 1:807
 n_taxa <- 6; favorite_tangles <- all_6_taxon
-data <- "equal/"
 data <- "123_456-1v5c/"
 data <- "123_456-equal/"
+}
 
 sub_n_taxa <- function(s) sub("NTAXA", as.character(n_taxa), s)
 data_of_num <- function(n) {
[…]

### 2015-01-26 prep_figures/distribution/scatter-ggvis.Rmd
Better static plots.

--- 2015-01-08 prep_figures/distribution/scatter-ggvis.Rmd
@@
 source("process.R")
-d <- get_results("../../results-rspr/ricci4.mat")
-d <- sample(d)  # Shuffle rows of d.
+
+props <- axis_props(labels=list(fontSize=16), title=list(fontSize=24))
+
+n_taxa <- 7
+# walk <- "lazy_unif"
+walk <- "unif_prior_mh"
+
+d <- get_results(paste0("../../results-rspr/", walk, "/ricci", n_taxa,".mat"))
 
@@
 #  layer_points(size = 5, fill = ~t1_nwk, stroke = ~t2_idx, strokeOpacity := 0) %>%
-  layer_points(size = 12, fill = ~t1_nwk) %>%
+  layer_points(size = 12, fill = ~avg_colless) %>%
 #  layer_points(size = ~mult, fill = ~t1_nwk, stroke = ~t2_idx) %>%
   # A dummy layer so that we can use text in the tooltip.
-  # layer_text(text := ~info, fontSize := 0) %>%
-  add_axis("x", grid = FALSE) %>%
-  add_axis("y", grid = FALSE, values = 1:3) %>%
-  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) 
+  layer_text(text := ~info, fontSize := 0) %>%
+  add_axis("x", title = "κ", grid = FALSE, properties=props) %>%
+  add_axis("y", title = "distance", grid = FALSE, values=1:4, properties=props) %>%
+#  add_legend(c("fill"), properties=legend_props(legend=list(y=0))) %>%
   #add_legend(c("stroke"), properties=legend_props(legend=list(y=100))) %>%
   #add_legend(c("size"), properties=legend_props(legend=list(y=200))) %>%
-  #add_tooltip(function(df) df$info)
+#  add_tooltip(function(df) df$info)
+  function (x) x # null
 ```

### 2015-01-26 prep_figures/distribution/process.R

@@ -11,13 +11,20 @@ get_results <- function(path) {
     d$kappa <- sapply(d$kappa_str, function(s) eval(parse(text=s)))
     d$info= paste(d$t1_nwk, d$t2_nwk, sep="<br />")
     # d$info= paste(d$t1_nwk, d$t2_nwk, paste("kappa:", d$kappa_str), sep="<br />")
-    d$dist_jitter = d$dist + runif(nrow(d), min = -0.35, max = 0.35)
+    d$dist_jitter = d$dist + runif(nrow(d), min = -0.25, max = 0.25)
 
     # Tree shape information.
     d$t1_shape <- lapply(d$t1_nwk, shape_of_str)
     d$t2_shape <- lapply(d$t2_nwk, shape_of_str)
     d$t1_colless <- sapply(d$t1_shape, colless)
     d$t2_colless <- sapply(d$t2_shape, colless)
+    d$avg_colless <- (d$t1_colless + d$t2_colless)/2
+    d$t1_sackin <- sapply(d$t1_shape, sackin)
+    d$t2_sackin <- sapply(d$t2_shape, sackin)
+    d$avg_sackin <- (d$t1_sackin + d$t2_sackin)/2
+
+    # Shuffle rows of d.
+    d <- sample(d)
 
     d
 }

### 2015-01-26 prep_figures/distribution/scatter-ggplot.R

@@ -1,0 +1,24 @@
+library(ggplot2)
+library(ape)
+library(apTreeshape)
+
+source("process.R")
+
+theme_set(theme_bw(22))
+
+n_taxa <- 7
+# walk <- "lazy_unif"; sub_title <- " taxa, lazy uniform random walk"
+walk <- "unif_prior_mh"; sub_title <- " taxa, uniform prior MCMC"
+
+d <- get_results(paste0("../../results-rspr/", walk, "/ricci", n_taxa,".mat"))
+
+p <- ggplot(d, aes(kappa, dist_jitter, color=avg_sackin))
+p <- p + geom_point()
+p <- p + ggtitle(paste0(n_taxa, sub_title))
+p <- p + scale_x_continuous("κ") + scale_y_continuous("SPR distance")
+p <- p + scale_colour_gradient(name="Sackin\nmeasure\n",
+                         breaks=c(min(d$avg_sackin), max(d$avg_sackin)),
+                         labels=c("balanced", "imbalanced"))
+
+ggsave(paste0("scatter-ricci-", walk, "-",n_taxa,".svg"), p, width=9.5, height=7)
+ 

### 2015-02-06 curvature.tex
degree change lemmas and proofs

@@
 \newattribute{dspr}{d_{\mathrm{SPR}}}
+\newattribute{degree}{deg}
 
@@
 \section{Rooted SPR Neighborhoods}
+
+Song (2003) proved the following:
+
+\begin{lemma}
+	\label{lem:degree_extremes}
+	Let $T$ be a rooted phylogenetic tree with $n$ leaves. Then
+	\begin{enumerate}
+		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree\todo{notation? Song used $\lambda$-balanced},
+		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
+		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
+	\end{enumerate}
+\end{lemma}
+
+We now bound the relative and exact difference in SPR degree between two trees with $n$ leaves.
+
+\begin{lemma}
+	\label{lem:degree_max_delta}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
+	Then:
+	\begin{enumerate}
+		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
+		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
+	\end{enumerate}
+\end{lemma}
+\begin{proof}
+	We first prove (1).
+	Clearly $\degree{T} / \degree{S}$ is minimized when $T$ is of minimum degree and $S$ of maximum degree.
+	Then, by Lemma~\ref{lem:degree_extremes}, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$.
+	We thus have:
+	\begin{align}
+		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
+		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
+		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
+		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
+		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\
+		\ge\ &\frac{3}{4} &\forall n \ge 3
+	\end{align}
+
+	We now prove (2).
+	$\degree{S} - \degree{T}$  is maximized when $T$ and $S$ are of minimum and maximum degree.
+	We thus have:
+	\begin{align}
+		\degree{S} - \degree{T}
+		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
+		- (3n^2 - 13n + 14) \\
+		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
+		\le\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
+		\le\ &n^2 - 5n + 6
+	\end{align}
+\end{proof}
 
@@
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
+\end{proof}
+
+We can now use Lemma~\ref{lem:degree_change} to improve on our bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
+
+\begin{lemma}
+	\label{lem:degree_max_delta_adjacent}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
+	Then:
+	\begin{enumerate}
+		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
+		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
+		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
+	\end{enumerate}
+\end{lemma}
+\begin{proof}
+	We first prove (1).
+	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
+	This value is maximized by minimizing $b$, namely by setting $b=0$.
+	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ and balancing the sides of the square $ka$ with as many leaves as possible.
+	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not. If not, then we set $j=1$ and balance the square with the remaining $n-b-i-j = n-2$ leaves
+	Thus, $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
+	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the rootmost leaf of the tree.
+
+	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf to balance the square but increase $j$.
+	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
+	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
+	If we move the additional leaf, we have:
+$\degree{S} - \degree{T} \le 2\ceil{\frac{n}{2}}\floor{\frac{n-2}{2}} -  (\floor{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.
+Similarly, if we do not move the additional leaf, we have:
+$\degree{S} - \degree{T} \le 2\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} -  (\ceil{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.
+
+	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is maximized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.
+	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
+	We then have:
+	\begin{align}
+		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
+		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
+		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
+	\end{align}
+	Thus (2) and (3) follow.
+
 \end{proof}

### 2015-02-06 curvature.tex
a few fixes to the proof of lem:degree_max_delta_adjacent

@@
   \expandafter\newcommand\csname #1XL\endcsname[1]{\Bigl#2##1\Bigr#3}%
+  \expandafter\newcommand\csname #1XXL\endcsname[1]{\biggl#2##1\biggr#3}%
   \expandafter\newcommand\csname #1V\endcsname[1]{\left#2##1\right#3}}
@@
 	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not. If not, then we set $j=1$ and balance the square with the remaining $n-b-i-j = n-2$ leaves
-	Thus, $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
+	Thus, $$\degree{S} - \degree{T} \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2.$$
 	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the rootmost leaf of the tree.
@@
 	If we move the additional leaf, we have:
-$\degree{S} - \degree{T} \le 2\ceil{\frac{n}{2}}\floor{\frac{n-2}{2}} -  (\floor{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.
+$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
 Similarly, if we do not move the additional leaf, we have:
-$\degree{S} - \degree{T} \le 2\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} -  (\ceil{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.
+$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
 

### 2015-02-06 curvature.tex
adjacent tree overlap and adjacent tree max curvature lemmas

@@
 
+Next, we bound the number of neighbors shared by two adjacent trees.
+
+\begin{lemma}
+	\label{lem:shared_neighbors}
+Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
+Then $\size{N(T) \cap N(S)} \le 6n - 17$.
+\end{lemma}
+
+\begin{proof}
+	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
+	Let $U$ be a tree in $N(T)$ and $N(S)$.
+	Then either (1) $T$ and $U$ differ in the location of $R$, or (2) $T$ and $U$ differ in the location of another subtree $P$.
+	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P) - S|(X \setminus L(P)$.
+	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
+	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $P$, $Q$, and $R$ are themselves a ``triple'' of subtrees.
+	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
+	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
+	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
+	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
+\end{proof}
+
 \section{Curvature}
@@
 
-A lower bound is also available because we can't do better than distance 1 for all trees unless they move $R$.
-There are only $\OhOf{n}$ pairs with trees that move $R$.
+A lower bound is also available because we can't do better than distance 1 for all trees but the $\OhOf{n}$ pairs from Lemma~\ref{lem:shared_neighbors}.
 By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$ for $W_{1,n}$.
@@
 
+\end{proof}
+
+Next we bound the maximum curvature of two adjacent trees.
+
+\begin{lemma}
+	The maximum curvature of the uniform random walk between two adjacent trees is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
+\end{lemma}
+\begin{proof}
+Let $T$ and $S$ be two adjacent trees.
+The maximum curvature occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
+By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
+The maximum curvature is thus $1 - \frac{\degree{T} - 6n-17}{\degree{T}} = \frac{6n - 17}{\degree{T}}$.
+This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
 \end{proof}

### 2015-02-09 curvature.tex
close relationship between lazy and uniform curvature

@@
 	Then either (1) $T$ and $U$ differ in the location of $R$, or (2) $T$ and $U$ differ in the location of another subtree $P$.
-	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P) - S|(X \setminus L(P)$.
+	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P)) = S|(X \setminus L(P))$.
 	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
@@
 \begin{lemma}
-	The maximum curvature of the uniform random walk between two adjacent trees is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
+	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
 \end{lemma}
@@
 This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
+\end{proof}
+
+We next bound the effect of laziness on curvature. Let $\kappa_{l,p}$ be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$. 
+For example, $\kappa_{l,\frac{1}{4}}$ moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
+\todo{should be defined earlier, particularly the part where the curvature is divided by $p$.}
+
+\begin{lemma}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
+	Then:
+	\begin{enumerate}
+		\item	$\kappa_{l,p}(T,S) = \kappa_{l,0}(T,S)$, if $\dspr{T,S} > 1$, and
+		\item	$\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,S) \le \kappa_{l,0}(T,S) + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+	\end{enumerate}
+\end{lemma}
+
+\begin{proof}
+	We first prove the lower bound, that is $\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,s)$.
+	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
+	Recall that $\kappa_{l,0}(T,S) = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\kappa_{l,p}(T,S) = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
+	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
+	Then:
+	\begin{align}
+		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
+		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
+		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
+		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
+		&\ge \kappa_{l,0}(T,S)
+	\end{align}
+
+	For the upper bound, we observe that $$W_1'(T,S) \ge p(W_1(T,S) + (1-p)\left(\dspr{T,S}\right) - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $\frac{1}{\max(\degree{T},\degree{S})}$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
+	The upper bound then follows analogously to the lower bound.
+	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
 \end{proof}

### 2015-02-10 curvature.tex
curvature lower bound

@@
 
+Note that this bound is tight and has been verified computationally for $n \le 7$.
+
+Next, we bound the minimum curvature of two adjacent trees.
+
+\begin{lemma}
+	The curvature of the uniform random walk between two adjacent trees with $n$ leaves is at least
+	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
+\end{lemma}
+\begin{proof}
+	Let $T$ and $S$ be two adjacent trees.
+	We can bound the mass transport cost from above by $1 + 2m$, where $m$ is the maximum proportion of mass that can not be moved between adjacent trees. 
+	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
+	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that must be paired at a distance $>1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
+	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
+	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees.
+	Then, let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
+	$T$ has degree at most $3.5n^2 -15n + 16$.
+	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
+	Then the maximum $m$ is:
+	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
+	The claim follows from multiplying this value by $-2$.
+\end{proof}
+
+We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
+Exhaustive tests with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees\todo{cite one of the Ollivier papers}.
+This directly leads to the following corollary:
+
+\begin{corollary}
+	The curvature of the uniform random walk between two phylogenetic trees is at least $-\frac{2}{5}$.
+\end{corollary}
+
+Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
+
 We next bound the effect of laziness on curvature. Let $\kappa_{l,p}$ be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$. 

### 2015-02-10 curvature.tex
rough MH curvature bound

@@
 	We thus have:
-	\begin{align}
+	\begin{align*}
 		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
@@
 		\ge\ &\frac{3}{4} &\forall n \ge 3
-	\end{align}
+	\end{align*}
 
@@
 	We thus have:
-	\begin{align}
+	\begin{align*}
 		\degree{S} - \degree{T}
@@
 		\le\ &n^2 - 5n + 6
-	\end{align}
+	\end{align*}
 \end{proof}
@@
 	We then have:
-	\begin{align}
+	\begin{align*}
 		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
@@
 		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
-	\end{align}
+	\end{align*}
 	Thus (2) and (3) follow.
@@
 	Then:
-	\begin{align}
+	\begin{align*}
 		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
@@
 		&\ge \kappa_{l,0}(T,S)
-	\end{align}
+	\end{align*}
 
@@
 	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
+\end{proof}
+
+We continue to explore the effect of biased random walks by considering the curvature of Metropolis Hastings random walks.
+The Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+We now bound the difference between the curvature of uniform random walks $\kappa$, and that of MH random walks, $\kappa_{MH}$.
+
+\begin{lemma}
+	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
+	\begin{enumerate}
+	$$\kappa(T,S) - \frac{1}{3\dspr{T,S}}
+	\le \kappa_{MH}(T,S)
+	\le \kappa(T,S) + \frac{1}{3\dspr{T,S}}, \text{ and }$$ 
+	$$\kappa(T,S) - \frac{1}{6}
+	\le \kappa_{MH}(T,S)
+	\le \kappa(T,S) + \frac{1}{6}.$$
+\end{lemma}
+\begin{proof}
+	We first prove the lower bound.
+	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
+	Thus, the hastings ratio is always $\ge \frac{5}{6}$.
+	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
+	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
+	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T's$ neighbors to $S$.
+	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
+	Moreover, the affected mass must be moved through at most two additional trees.
+	Then, $W_1' &\le W_1 + \frac{2}{6}$.
+	We now have:
+	\begin{align*}
+		\kappa_{MH}(T,S) & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
+		& \ge \kappa(T,S) - \frac{1}{3\dspr{T,S}}
+	\end{align*}
+
+	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
+	We thus obtain the lower bound of $\kappa(T,S) - \frac{1}{6}$ in this case.
+
+	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e.
+directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
+	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
+
 \end{proof}

### 2015-02-12 curvature.tex
Making it compile.
A beginning of an intro.

@@
 \begin{document}
-\title{Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph}
+\title{Ricci-Ollivier curvature of the data-free random walks on rooted phylogenetic subtree-prune-regraft graph}
 \author[Whidden]{Christopher Whidden}
@@
 
-This isn't quite the right graph to be looking at (rooted).
+Statistical phylogenetics is.
+Likelihood of generating the observed sequences via a continuous time Markov chain.
+Likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
+Bayesian versus maximum likelihood (ML).
 
-We are talking about rooted.
+Phylogenetic algorithms do MCMC or heuristic local search.
+In both settings, one needs a means of describing "nearby" trees.
+There are a number of such
+SPR graph.
+For MCMC moves, one would like to ensure good mixing.
+For ML moves, we are concerned about local optima.
+We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
+vs. branch length effects?
+
+Researchers have noticed SPR graph effects in MCMC.
+\cite{Mossel2005-ly,Mossel2006-fo}
+\cite{Ronquist2006-fv}
+\cite{Stefankovic2011-hu}
+We recently showed that graph structure has an effect of MCMC mixing using real data and programs.
+\cite{Whidden2015-yi}
+Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves \cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object called the compact real tree which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
+
+ML also has graph effects, which manifest as local maxima.
+There have been investigation of local maxima, but concern branch lengths.
+\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}
+The work on SPR moves in the ML context has been focused on increasing efficiency of the SPR moves in finding a well-defined maximum (via simulated data) rather than investigating local maxima \cite{Hordijk2005-dl,Stamatakis2005-rh,Guindon2010-lo,Price2010-fi}.
+
+The SPR graph is very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
+\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rSPR graph.
+\cite{Ding2011-bj} showed that the diameter of the rSPR graph is $n - \Theta(\sqrt n)$.
+\todo{CW: what is known about the uSPR graph?}
+
+Recently, a new approach to proving theorems about random walks on discrete spaces has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
+
+\section{Preliminaries}
+\[
+\kappa(x, y) := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
+\]
+
+We are talking about rooted, and would like to make it clear that this is not the most realistic graph to consider.
+Our motivations for considering this graph are because it is understand-able, and some programs do rSPR-type moves.
 Parent and parent subtree.
 Define adjacency of subtrees.
+
+\section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
+
+After the chain has achieved stationarity, one sees a
 
@@
 	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $P$, $Q$, and $R$ are themselves a ``triple'' of subtrees.
-	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
+	By Lemma~ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
 	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
@@
 	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
-	\begin{enumerate}
 	$$\kappa(T,S) - \frac{1}{3\dspr{T,S}}
@@
 	Moreover, the affected mass must be moved through at most two additional trees.
-	Then, $W_1' &\le W_1 + \frac{2}{6}$.
+	Then, $W_1' \le W_1 + \frac{2}{6}$.
 	We now have:
@@
 
-% \bibliographystyle{plain}
-% \bibliography{curvature}
+\section{Conclusion and future work}
+This isn't quite the right graph to be looking at (rooted).
+
+Want to investigate curvature of MCMC steps with a nontrivial likelihood function.
+MCMC does both branch length and topology changes, thus this isn't the real thing.
+Nevertheless, graph effects still matter.
+\cite{Stefankovic2011-hu} take a chain where the likelihoods are the likelihoods maximized across branch length.
+
+Ranked trees \cite{Song2006-xe}
+Partitions \cite{Gusfield2002-il}
+
+\bibliographystyle{plain}
+\bibliography{curvature}
 \end{document}

### 2015-02-12 curvature.tex
A bit more intro material.

@@
 \usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
+\usepackage{cite}
 
@@
 \begin{document}
-\title{Ricci-Ollivier curvature of the data-free random walks on rooted phylogenetic subtree-prune-regraft graph}
+\title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
 \author[Whidden]{Christopher Whidden}
@@
 In both settings, one needs a means of describing "nearby" trees.
-There are a number of such
+There are a number of such means of doing so.
+SPR.
 SPR graph.
@@
 We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
-vs. branch length effects?
+We use "random walks" as being a general notion of movement around a graph, including MCMC.
 
@@
 \cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}
-The work on SPR moves in the ML context has been focused on increasing efficiency of the SPR moves in finding a well-defined maximum (via simulated data) rather than investigating local maxima \cite{Hordijk2005-dl,Stamatakis2005-rh,Guindon2010-lo,Price2010-fi}.
+The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another \cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
+We are not aware of any general conclusions concerning hill-climbing on the SPR or related graphs that has come from this work.
 
@@
 
-Recently, a new approach to proving theorems about random walks on discrete spaces has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
+Recently, a new approach to quantifying properties of random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
+They define a notion of curvature on a discrete or continuous case.
+Formalizes idea of the extent to which random walking brings points together, and is generalization of the notion of Ricci curvature on manifolds.
+
+In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and connections to access times for those random walks.
+For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic random walk and MCMC sampling from the uniform prior on trees.
+By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
+By exploiting symmetries in the rSPR graph, we are able to calculate all of the curvatures for pairs of trees with up to seven leaves.
+Using simulation of these processes, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
 

### 2015-02-12 curvature.tex
TBR and uSPR diameter bounds

@@
 \cite{Ding2011-bj} showed that the diameter of the rSPR graph is $n - \Theta(\sqrt n)$.
-\todo{CW: what is known about the uSPR graph?}
-
+\todo{CW: what is known about the uSPR graph? EM: the best bounds are in the same paper}
+\cite{Ding2011-bj} showed that
+$$ n - 2\ceil{\sqrt{n}} + 1
+\le \Delta_{\text{TBR}}(n)
+\le \Delta_{\text{SPR}}(n)
+\le \floorV{\frac{\sqrt{n - 2} - 1}{2}}.
+$$
 Recently, a new approach to quantifying properties of random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.

### 2015-02-13 curvature.tex
Edits. Please review, @cwhidden.

--- 2015-02-12 curvature.tex
@@
 \newcommand{\TT}{\mathcal T}
+
+\newcommand{\cuttable}[1]{#1} % In case we need to be succinct for limits.
 
@@
 \newattribute{dspr}{d_{\mathrm{SPR}}}
-\newattribute{degree}{deg}
+\newattribute{degree}{\operatorname{deg}}
 
@@
 \cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rSPR graph.
-\cite{Ding2011-bj} showed that the diameter of the rSPR graph is $n - \Theta(\sqrt n)$.
-\todo{CW: what is known about the uSPR graph?}
+\cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
+$$ n - 2\ceil{\sqrt{n}} + 1
+\le \Delta_{\text{uSPR}}(n)
+\le n - 3 - \floorV{\frac{\sqrt{n - 2} - 1}{2}}.
+$$
 
@@
 
-In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and connections to access times for those random walks.
+In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and make connections to access times for those random walks.
 For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic random walk and MCMC sampling from the uniform prior on trees.
@@
 
+For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
+A \emph{parent (sub)tree} of a subtree $U$ is the smallest subtree strictly containing $U$.
+A \emph{parent edge} of a subtree $U$ is the edge connecting $U$ to the rest of the tree.
+The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
+A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
+LCA of subtrees.
+
 \section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
@@
 
-Song (2003) proved the following:
+In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of nodes in the graph.
+Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.
 
-\begin{lemma}
+\begin{lemma}{Song, \cite{Song2003-gf}.}
 	\label{lem:degree_extremes}
@@
 	\begin{enumerate}
-		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree\todo{notation? Song used $\lambda$-balanced},
+		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree,
 		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
@@
 \begin{proof}
-	We first prove (1).
-	Clearly $\degree{T} / \degree{S}$ is minimized when $T$ is of minimum degree and $S$ of maximum degree.
-	Then, by Lemma~\ref{lem:degree_extremes}, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$.
-	We thus have:
+	To prove (1), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
 	\begin{align*}
@@
 		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
-		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
-		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
-		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\
-		\ge\ &\frac{3}{4} &\forall n \ge 3
+		\cuttable{= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\}
+		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
+		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\}
 	\end{align*}
-
-	We now prove (2).
-	$\degree{S} - \degree{T}$  is maximized when $T$ and $S$ are of minimum and maximum degree.
-	We thus have:
+	which is greater than 3/4 when $n \ge 3$.
+    Similarly for (2):
 	\begin{align*}
-		\degree{S} - \degree{T}
+		\cuttable{\degree{S} - \degree{T}
 		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
-		- (3n^2 - 13n + 14) \\
+		- (3n^2 - 13n + 14) \\}
 		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
-		\le\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
-		\le\ &n^2 - 5n + 6
+		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
+		=\ &n^2 - 5n + 6.
 	\end{align*}
@@
 \label{lem:degree_change}
-Let $T$ and $S$ be phylogenetic trees such that $\dspr{T,S}=1$, that is $T$ can be transformed into $S$ by moving a subtree $R$ with $k$ leaves.
-The corresponding SPR operation moves $R$ from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
 Let $L$ be the $\lca{U,V}$ in $T$.
@@
 Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
+\todo{My impression was that the leaves of $R$ and $U$ (or $V$) were disjoint.
+If that's the case, then this "excluding" clause is more confusing than helpful.}
 Then the degrees of $T$ and $S$ differ by:
@@
 \begin{proof}
-
-Observe that the set of permissible SPR moves changes after the movement of $R$ in four different ways:
-1) Subtrees including vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
-2) Subtrees including vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
+The set of permissible SPR moves changes in four different ways due to the movement of $R$:
+1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
+2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,\todo{let's choose "node" or "vertex"}
 3) $R$'s parent subtree may now be moved into $U$, and
@@
 
-Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges (note that we are excluding any ``root edge'' in this number, so this tree has a degree two root).
-In the first case there are $a$ subtrees that can now be moved to the $2k$ positions into $R$ (including its newly introduced parent node and one of the newly subdivided root edge of $V$) for a total gain of $2ka$ distinct moves.
+Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges (recall that we are excluding any ``root edge'' in these calculations).
+In the first case there are $a$ subtrees that can now be moved onto the $2k$ edges in $R$ (including its newly introduced parent edge and one of the newly subdivided root edges of $V$) for a total gain of $2ka$ distinct moves.
 Similarly, we lose $2kb$ moves in the second case.
@@
 Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ nodes in the SPR graph be $q$ and $p$ respectively.
-Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that:
-\begin{enumerate}
-\item $\dspr{T',S'} = 1$ for each $(T',S')$ pair, and
-\item each neighbor of $T$ and each neighbor of $S$ is paired to exactly one tree.
-\end{enumerate}
+Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
 \end{corollary}
@@
 
-We can now use Lemma~\ref{lem:degree_change} to improve on our bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
-
+We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
 \begin{lemma}
@@
 	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
-	This value is maximized by minimizing $b$, namely by setting $b=0$.
-	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ and balancing the sides of the square $ka$ with as many leaves as possible.
-	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not. If not, then we set $j=1$ and balance the square with the remaining $n-b-i-j = n-2$ leaves
-	Thus, $$\degree{S} - \degree{T} \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2.$$
+	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
+	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) and balancing the sides of the product $ka$ with as many leaves as possible.
+
+	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not.
+    If not, then we set $j=1$ and balance the root with the remaining $n-b-i-j = n-2$ leaves, giving (1).
 	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the rootmost leaf of the tree.
 
-	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf to balance the square but increase $j$.
+	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf\todo{Hm, I'm not sure where this additional leaf is!} to balance the product but increase $j$.
 	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
@@
 
-	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is maximized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.
+	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.\todo{This sort of thing does not appear true in general.
+Please explain why it works here.}
 	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.

### 2015-02-13 curvature.tex
more explanation in some proofs

@@
 Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
-Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
-\todo{My impression was that the leaves of $R$ and $U$ (or $V$) were disjoint.
-If that's the case, then this "excluding" clause is more confusing than helpful.}
+Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
+\todo{EM: if we move R to an ancestor (e.g.
+the root) then the subtree $V$ in $T$ contains both $R$ and $U$.
+$R$ and $U$ are disjoint in $T$, though so I dropped one excluding.}
 Then the degrees of $T$ and $S$ differ by:
@@
 
-	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf\todo{Hm, I'm not sure where this additional leaf is!} to balance the product but increase $j$.
+	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf \todo{Hm, I'm not sure where this additional leaf is!
+EM: does this make more sense?} to balance the product at the cost of increasing $j$.
 	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
@@
 
-	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.\todo{This sort of thing does not appear true in general.
-Please explain why it works here.}
-	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
+The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
+This is is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximum.
+\todo{This sort of thing does not appear true in general.
+Please explain why it works here.
+EM: Is this better?
+Think of $\frac{x}{1+x}$ which is an increasing function.}
+	That is, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = \degree{T} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
 	We then have:

### 2015-02-13 curvature.tex
flesh out second half of lem:degree_max_delta_adjacent, s/node/vertex; s/nodes/vertices

@@
 The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
-A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
+A \emph{ladder tree} is a tree such that every internal vertex has a leaf as a direct descendant.
 LCA of subtrees.
@@
 
-In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of nodes in the graph.
+In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
 Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.
@@
 Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
-Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
+Let $a$ be the number of intermediate vertices on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
+Similarly, let $b$ be the number of intermediate vertices on the path from $V$ to $L$ in $T$, excluding endpoints.
 Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
-\todo{EM: if we move R to an ancestor (e.g.
-the root) then the subtree $V$ in $T$ contains both $R$ and $U$.
-$R$ and $U$ are disjoint in $T$, though so I dropped one excluding.}
 Then the degrees of $T$ and $S$ differ by:
@@
 The set of permissible SPR moves changes in four different ways due to the movement of $R$:
-1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
-2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,\todo{let's choose "node" or "vertex"}
+1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent vertex,
+2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent vertex,
 3) $R$'s parent subtree may now be moved into $U$, and
@@
 \label{cor:paired_neighbors}
-Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ nodes in the SPR graph be $q$ and $p$ respectively.
+Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ vertices in the SPR graph be $q$ and $p$ respectively.
 Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
@@
 \begin{proof}
-By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
+By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target vertices.
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
@@
 
-	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf \todo{Hm, I'm not sure where this additional leaf is!
-EM: does this make more sense?} to balance the product at the cost of increasing $j$.
+	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product at the cost of increasing $j$.
 	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
@@
 $$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
+Therefore claim (1) holds.
 
 The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
-This is is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximum.
-\todo{This sort of thing does not appear true in general.
-Please explain why it works here.
-EM: Is this better?
-Think of $\frac{x}{1+x}$ which is an increasing function.}
-	That is, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = \degree{T} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
+By (1), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2 $.
+Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
+This bound is maximized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
 	We then have:
 	\begin{align*}
-		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
-		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
+		\frac{\degree{T}}{\degree{S}} &\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
 		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}

### 2015-02-15 curvature.tex
fixed a spurious maximized.
Abstract and first paragraph.

@@
 \author[Whidden]{Christopher Whidden}
-\author[Matsen]{Frederick Matsen}
+\author[Matsen]{Frederick A Matsen IV}
 \address{}
@@
 \begin{abstract}
+Statistical phylogenetics inference methods use tree rearrangement operations to perform either local hill-climbing search or Markov chain Monte Carlo across tree topologies.
+The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
+Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the graph.
+Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
+In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
+This notion of curvature is especially relevant to stochastic search on the graph.
+By proving theorems and direct calculation, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
+This indicates significant structure of the graph, and a greater understanding of this structure could ultimately lead to improved strategies for tree search.
+
+We find that indeed there is
 \end{abstract}
@@
 \section{Introduction}
+Phylogenetic methods, which reconstruct evolutionary trees from DNA or RNA data, are of fundamental importance to modern biology.
+Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
+The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution (cite{Felsenstein}).
+The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
+Thus likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
+Within this framework researchers typically choose either a Bayesian approach, in which a computer program approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which a computer program searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
 
-Statistical phylogenetics is.
-Likelihood of generating the observed sequences via a continuous time Markov chain.
-Likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
-Bayesian versus maximum likelihood (ML).
-
+Here we focus on the discrete search part.
 Phylogenetic algorithms do MCMC or heuristic local search.
@@
 Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
-This bound is maximized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
+This bound is minimized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
 	We then have:

### 2015-02-16 curvature.tex
A bit more introduction.

@@
 \section{Introduction}
-Phylogenetic methods, which reconstruct evolutionary trees from DNA or RNA data, are of fundamental importance to modern biology.
+Molecular phylogenetic methods, which reconstruct evolutionary trees from DNA or RNA data, are of fundamental importance to modern biology.
 Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
@@
 Thus likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
-Within this framework researchers typically choose either a Bayesian approach, in which a computer program approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which a computer program searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
+Within this framework researchers typically choose either a Bayesian approach, in which an algorithm (typically Markov chain Monte Carlo or MCMC) approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which an algorithm searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
 
-Here we focus on the discrete search part.
-Phylogenetic algorithms do MCMC or heuristic local search.
-In both settings, one needs a means of describing "nearby" trees.
+In order to get accurate such estimates, MCMC samplers must sufficiently explore the collection of trees, or heuristic local search algorithms must find a way to the ML tree, avoiding and escaping local peaks.
+In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
+% Here we focus on the discrete search part.
 There are a number of such means of doing so.

### 2015-02-17 curvature.tex
Bracket typo.
Pass through to the beginning of curvature.

@@
 \newtheorem{definition}[theorem]{Definition}
+
+\renewcommand{\theenumi}{\roman{enumi}}
 
@@
 \section{Preliminaries}
-\[
+\begin{equation}
 \kappa(x, y) := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
-\]
+\label{eq:curvatureDef}
+\end{equation}
 
+Rooted $X$-trees.
+Leaf restriction notation with $|$.
 We are talking about rooted, and would like to make it clear that this is not the most realistic graph to consider.
-Our motivations for considering this graph are because it is understand-able, and some programs do rSPR-type moves.
+Our motivations for considering this graph are because it is understandable, and some programs do rSPR-type moves.
+Note that in rSPR you can move subtrees to the root.
 Parent and parent subtree.
@@
 
-\begin{lemma}{Song, \cite{Song2003-gf}.}
+\begin{lemma}[{Song \cite{Song2003-gf}}]
+    \pushQED{\qed}
 	\label{lem:degree_extremes}
@@
 	\end{enumerate}
+    \popQED
 \end{lemma}
@@
 \begin{proof}
-	To prove (1), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
+	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
 	\begin{align*}
@@
 	which is greater than 3/4 when $n \ge 3$.
-    Similarly for (2):
+    Similarly for (ii):
 	\begin{align*}
@@
 The set of permissible SPR moves changes in four different ways due to the movement of $R$:
-1) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent vertex,
-2) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent vertex,
-3) $R$'s parent subtree may now be moved into $U$, and
-4) $R$'s parent subtree may no longer be moved into $V$.
+(i) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent vertex,
+(ii) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent vertex,
+(iii) $R$'s parent subtree may now be moved into $U$, and
+(iv) $R$'s parent subtree may no longer be moved into $V$.
 No additional moves are introduced or blocked by the original SPR operation on $R$.
@@
 \begin{proof}
-	We first prove (1).
+	We first prove (i).
 	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
 	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
-	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) and balancing the sides of the product $ka$ with as many leaves as possible.
+	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.
 
-	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not.
-    If not, then we set $j=1$ and balance the root with the remaining $n-b-i-j = n-2$ leaves, giving (1).
-	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the rootmost leaf of the tree.
+	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root or not.
+    If not, then we set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
+	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.
 
-	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product at the cost of increasing $j$.
+	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
 	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
@@
 	If we move the additional leaf, we have:
-$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
-Similarly, if we do not move the additional leaf, we have:
-$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
-Therefore claim (1) holds.
+$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}$$
+like before.
+Similarly, if we do not move the additional leaf, we also have:
+$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
+proving (i).
 
-The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
-By (1), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2 $.
-Thus, $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
-This bound is minimized when $\degree{T}$ is minimized and, again, $\degree{T} \ge 3n^2 - 13n + 14$ by Lemma~\ref{lem:degree_extremes}.
-	We then have:
+The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})}$.
+By (i), we have that $\degree{S} - \degree{T} \le \frac{1}{2} (n-2)^2$,
+so $\frac{\degree{T}}{\degree{S}} \ge \frac{\degree{T}}{\degree{T} + \frac{1}{2} (n-2)^2} $.
+This bound is minimized when $\degree{T}$ is minimized, and recall by Lemma~\ref{lem:degree_extremes} that $\degree{T}$ is bounded below by $3n^2 - 13n + 14$.
+	Thus
 	\begin{align*}
 		\frac{\degree{T}}{\degree{S}} &\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
-		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
+		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}.
 	\end{align*}
-	Thus (2) and (3) follow.
+	Statements (ii) and (iii) follow from this bound.
+    % 3*16 - 13*4 + 14 = 10
+    % 3.5*16 - 15*4 + 16 = 12
 
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
+Let $N(Q)$ be the set of neighbors of a tree $Q$ (this does not include $Q$).
 
@@
 	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
-	Let $U$ be a tree in $N(T)$ and $N(S)$.
-	Then either (1) $T$ and $U$ differ in the location of $R$, or (2) $T$ and $U$ differ in the location of another subtree $P$.
-	In the latter case, the facts that $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$, imply that $T|(X \setminus L(P)) = S|(X \setminus L(P))$.
+	Pick a neighbor $Q \in N(T) \cap N(S)$ of both $T$ and $S$ \todo{Um, why can't this intersection be empty?}.
+	Then either (i) $T$ and $Q$ differ in the location of $R$, or (ii) $T$ and $Q$ differ in the location of another subtree $P$.
+	In the latter case, $T|(X \setminus L(P)) = S|(X \setminus L(P))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$.
+    \todo{FAM stopped here}
 	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
@@
 \section{Curvature}
-Note that we have to write out simple random walk versus degree-informed.
-Can we just say that degree differences are small relative to $n$? Is that true?
-
-For the purposes of this paper, $\kappa_n$ is curvature on the $n$-taxon rSPR graph and $W_{1,n}$ is the corresponding mass transport term.
+For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term.
 
@@
 \begin{proof}
-Because $d(T_n, S_n) = 1$, we will prove the proposition by showing that the mass transport term $W_{1,n}$ sits between two bounds with limit 1 as $n$ goes to infinity.
+Because $d(T_n, S_n) = 1$, we will prove the proposition by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.
 
 To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
-First we show that $W_{1,n}$ is bounded above by $(q_n+3(2k-1)(\OhOf{n}))/q_n$, where $q_n$ is the number of neighbors of $T_n$ and $S_n$.
-Let $T'_n$ be any of the $o$ SPR moves from a pairing following Corollary~\ref{cor:paired_neighbors}.
-In the mass transport, we pair $T'_n$ with $S'_n$, which results from the same move applied to $S_n$ such that $\dspr{T'_n, S'_n} = 1$.
-The number of trees unmatched by this pairing is $\OhOf{kn}$, and their distance is at most 3.
-Thus, $W_{1,n}$ is bounded above by $(q_n + 2\OhOf{kn}) / q_n$.
+First we claim that $W_{1,n}$ is bounded above by $(|N(T_n)|+\OhOf{kn})/|N(T_n)|$ by exhibiting a mass transport program satisfying that bound.
+Let $(T'_n, S'_n)$ be any of the $o$ pairs of trees which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
+We pair the these trees in the mass transport.
+The number of trees unmatched by this pairing is $\OhOf{kn}$, and we can pair each of them arbitrarily with another tree of distance at most 3.
+Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + \OhOf{kn}) / |N(T_n)|$.
 
-A lower bound is also available because we can't do better than distance 1 for all trees but the $\OhOf{n}$ pairs from Lemma~\ref{lem:shared_neighbors}.
-By ignoring these trees we get a lower bound of $(q_n -(\OhOf{n}))/q_n$ for $W_{1,n}$.
+A lower bound is also available because we can't do better than distance 1 for all trees except for the $\OhOf{n}$ pairs from Lemma~\ref{lem:shared_neighbors}.
+By ignoring these trees we get a lower bound of $(|N(T_n)| - (\OhOf{n}))/|N(T_n)|$ for $W_{1,n}$.
 
-The desired control of $W_{1,n}$ is thus obtained because $q_n$ is quadratic in $n$.
+The desired control of $W_{1,n}$ is thus obtained because $|N(T_n)|$ is quadratic in $n$.
 
-Now, we prove the proposition when the number of neighbors differ.
-Let $q_n$ and $p_n$ be the number of neighbors of $T_n$ and $S_n$, respectively.
-Assume without loss of generality that $q_n < p_n$.
-By Lemma~\ref{lem:degree_change}, $p_n - q_n = 2(k(a-b) + i - j)$, where $a,b,i,j < n$.
-Thus, $p_n - q_n = \OhOf{kn}$.
-We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $p_n > q_n$ we can only account for at most $q_n / p_n$ of the mass directly and may have to move the $(p_n - q_n) / p_n$ remainder to trees a distance at most 3.
-Thus, $W_{1,n}$ is bounded above by $(q_n + 3\OhOf{kn}) / p_n = (p_n + 2\OhOf{kn}) / p_n$.
-We again bound $W_{1,n}$ from below with $(q_n - \OhOf{n}) / q_n$ by ignoring the mass in trees that may overlap.
-The proposition again follows because $q_n$ is quadratic in $n$.
+Now we prove the proposition when the number of neighbors differ.
+Assume without loss of generality that $|N(T_n)| < |N(S_n)|$.
+By Lemma~\ref{lem:degree_change}, $|N(S_n)| - |N(T_n)| = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
+Thus, $|N(S_n)| - |N(T_n)| = \OhOf{kn}$.
+We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $|N(T_n)| < |N(S_n)|$ we can only account for at most $|N(T_n)| / |N(S_n)|$ of the mass directly and may have to move the $(|N(S_n)| - |N(T_n)|) / |N(S_n)|$ remainder to trees a distance at most 3.
+Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + 3\OhOf{kn}) / |N(S_n)| = (|N(S_n)| + 2\OhOf{kn}) / |N(S_n)|$.
+We again bound $W_{1,n}$ from below with $(|N(T_n)| - \OhOf{n}) / |N(T_n)|$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
+The proposition again follows because $|N(T_n)|$ is quadratic in $n$.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{curvature}

### 2015-02-19 scripts/access-time-count.py
-

@@ -2,7 +2,8 @@
 '''
 Suggested use:
 ./access-time-count.py --tp \
-    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" <mcmc-trace-files>
+    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" \
+    <mcmc-trace-files>
 '''
 
 import argparse

### 2015-02-19 curvature.tex
Typo.
Little twiddles in intro.

@@
 The SPR graph is very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
-\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rSPR graph.
+\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rSPR graph, and corresponding bounds on degree.
 \cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
@@
 $$
+Despite the centrality of the SPR graph to phylogenetic inference, we are not aware of any other work investigating its properties.
 
-Recently, a new approach to quantifying properties of random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
+Recently, a new approach to quantifying properties of discrete and continuous spaces with respect to random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
 They define a notion of curvature on a discrete or continuous case.
@@
 By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
-The maximum curvature is thus $1 - \frac{\degree{T} - 6n-17}{\degree{T}} = \frac{6n - 17}{\degree{T}}$.
+The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}$.
 This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

### 2015-02-19 curvature.tex
Changing letters around so they form an order.
Q, R: subtrees
S, T, U: full-size trees.

cc @cwhidden

@@
 Next, we bound the number of neighbors shared by two adjacent trees.
-Let $N(Q)$ be the set of neighbors of a tree $Q$ (this does not include $Q$).
+Let $N(U)$ be the set of neighbors of a tree $U$ (this does not include $U$).
 
@@
 	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
-	Pick a neighbor $Q \in N(T) \cap N(S)$ of both $T$ and $S$ \todo{Um, why can't this intersection be empty?}.
-	Then either (i) $T$ and $Q$ differ in the location of $R$, or (ii) $T$ and $Q$ differ in the location of another subtree $P$.
-	In the latter case, $T|(X \setminus L(P)) = S|(X \setminus L(P))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,P} = \dspr{S,P} = 1$.
-    \todo{FAM stopped here}
-	Then leaves $r' \in R$, $p' \in P$, and $q' \in Q$, for some subtree $Q$, form a triple of $T$ and a different triple in $S$.
-	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $P$, $Q$, and $R$ are themselves a ``triple'' of subtrees.
+	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ \todo{Um, why can't this intersection be empty?}.
+	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
+	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
+	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
+	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
 	By Lemma~ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.

### 2015-02-19 curvature.tex
A bit more revision.

@@
 In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and make connections to access times for those random walks.
-For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic random walk and MCMC sampling from the uniform prior on trees.
+For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic\todo{Shall we use uniform or isotropic?
+The former is more intuitive, but the latter seems more specific to me.} random walk and MCMC sampling from the uniform prior on trees.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
@@
 To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
+\todo[inline]{Do you think that we should talk about $3 \OhOf{kn}$, or just talk about $\OhOf{kn}$?}
 First we claim that $W_{1,n}$ is bounded above by $(|N(T_n)|+\OhOf{kn})/|N(T_n)|$ by exhibiting a mass transport program satisfying that bound.
-Let $(T'_n, S'_n)$ be any of the $o$ pairs of trees which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
+Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
 We pair the these trees in the mass transport.
-The number of trees unmatched by this pairing is $\OhOf{kn}$, and we can pair each of them arbitrarily with another tree of distance at most 3.
+There are $\OhOf{kn}$ trees unmatched by this pairing, and we can pair each of them arbitrarily with another tree of distance at most 3.
 Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + \OhOf{kn}) / |N(T_n)|$.
 
-A lower bound is also available because we can't do better than distance 1 for all trees except for the $\OhOf{n}$ pairs from Lemma~\ref{lem:shared_neighbors}.
+A lower bound is also available because we can't do better than distance 1 for all trees except for shared neighbors, of which there are $\OhOf{n}$ by Lemma~\ref{lem:shared_neighbors}.
 By ignoring these trees we get a lower bound of $(|N(T_n)| - (\OhOf{n}))/|N(T_n)|$ for $W_{1,n}$.
@@
 Thus, $|N(S_n)| - |N(T_n)| = \OhOf{kn}$.
-We again pair $T'_n$ with $S'_n$ such that $\dspr{T'_n, S'_n}=1$ but, as $|N(T_n)| < |N(S_n)|$ we can only account for at most $|N(T_n)| / |N(S_n)|$ of the mass directly and may have to move the $(|N(S_n)| - |N(T_n)|) / |N(S_n)|$ remainder to trees a distance at most 3.
+We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $|N(T_n)| < |N(S_n)|$ we can only account for at most $|N(T_n)| / |N(S_n)|$ of the mass directly and may have to move the $(|N(S_n)| - |N(T_n)|) / |N(S_n)|$ remainder to trees a distance at most 3.
 Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + 3\OhOf{kn}) / |N(S_n)| = (|N(S_n)| + 2\OhOf{kn}) / |N(S_n)|$.
@@
 Next we bound the maximum curvature of two adjacent trees.
-
 \begin{lemma}
-	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$. \todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$}
+	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
+\todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$.
+CW: I'm a little confused here.
+Doesn't the curvature of the lazy random walk depend on how lazy it is?}
 \end{lemma}
 \begin{proof}
-Let $T$ and $S$ be two adjacent trees.
-The maximum curvature occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
+The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
 By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
@@
 \end{proof}
-
-Note that this bound is tight and has been verified computationally for $n \le 7$.
+This bound is tight and has been verified computationally for $n \le 7$.
 
@@
 	Let $T$ and $S$ be two adjacent trees.
-	We can bound the mass transport cost from above by $1 + 2m$, where $m$ is the maximum proportion of mass that can not be moved between adjacent trees.
+	Because pairs of neighbors of adjacent trees are at most distance 3 apart, we can bound the mass transport cost from above by $1 + 2m$, where $m$ is the maximum proportion of mass that can not be moved between adjacent trees.
 	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
-	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that must be paired at a distance $>1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
+	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that must be paired at a distance $>1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$\todo{It's not clear what "pairing" means when degrees differ.
+Thus it's confusing when it's used in the same sentence as a statement concerning differing degrees.
+If you want to use it as a concept, how about introducing it at the beginning of the section as a proof strategy?}
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
@@
 
+\section{Acknowledgements}
+Alex Gavruskin,
+Vladimir Minin,
+
+This work was funded by National Science Foundation award 1223057.
+
 \bibliographystyle{unsrt}

### 2015-02-19 AlexsNotes.tex
Trivial notes on ranked NNI.

\documentclass[12pt]{article}

\usepackage{amsmath,amsthm}
\usepackage{todonotes}
\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
\usepackage{cite}
\usepackage{enumerate}

\newtheorem{lemma}{Lemma}

%\newcommand{\deg}{\mathrm{deg}}

\begin{document}

We consider the ranked-NNI graph (the distance on ranked topologies inherited from $\tau$-space).

\begin{lemma}
Let $T$ be a ranked tree on $n>2$ leaves. Then $$n-1\leq \deg(T)\leq2(n-2).$$ 
\end{lemma}

\proof
Obvious. 
\endproof

\begin{lemma}
\begin{enumerate}[(1)]
\item $\frac{\deg(T)}{deg(R)}>\frac12.$
\item $deg(T)-deg(R)> n-3.$
\end{enumerate}
\end{lemma}

\proof
Follows from previous Lemma. 
\endproof

\begin{lemma}
$d_{\tau}(T,R) = 1 \Rightarrow |N_1(T)\cap N_1(R)|\in\{0,1\}.$
\end{lemma}

\proof
Obvious.
\endproof

\end{document}

### 2015-02-19 curvature.tex
handled most of the todos

@@
 	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
-	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ \todo{Um, why can't this intersection be empty?}.
+	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$.
+	Note that this intersection is nonempty, as the fact that $T$ and $S$ differ implies that $R$ contains at most $n-2$ of the leaves, so there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$.
+\todo{Um, why can't this intersection be empty?
+EM: reason added.
+This is not strictly necessary to get an upper bound but might be interesting.}
 	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
@@
 	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
-	By Lemma~ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
+	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
 	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
@@
 Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \kappa_n(T_n, S_n) = 0$ for \todo{TBA} random walk on the SPR graph.
+Then $\lim_{n \rightarrow \infty} \kappa_n(T_n, S_n) = 0$ for the uniform random walk on the SPR graph.
 \end{proposition}
@@
 To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
-\todo[inline]{Do you think that we should talk about $3 \OhOf{kn}$, or just talk about $\OhOf{kn}$?}
 First we claim that $W_{1,n}$ is bounded above by $(|N(T_n)|+\OhOf{kn})/|N(T_n)|$ by exhibiting a mass transport program satisfying that bound.
@@
 We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $|N(T_n)| < |N(S_n)|$ we can only account for at most $|N(T_n)| / |N(S_n)|$ of the mass directly and may have to move the $(|N(S_n)| - |N(T_n)|) / |N(S_n)|$ remainder to trees a distance at most 3.
-Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + 3\OhOf{kn}) / |N(S_n)| = (|N(S_n)| + 2\OhOf{kn}) / |N(S_n)|$.
+Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + \OhOf{kn}) / |N(S_n)| = (|N(S_n)| + \OhOf{kn}) / |N(S_n)|$.
 We again bound $W_{1,n}$ from below with $(|N(T_n)| - \OhOf{n}) / |N(T_n)|$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
@@
 	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
-\todo{lazy random walk is $\frac{6n-15}{3n^2 - 13n + 14}$.
-CW: I'm a little confused here.
-Doesn't the curvature of the lazy random walk depend on how lazy it is?}
 \end{lemma}
@@
 	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
-	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that must be paired at a distance $>1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$\todo{It's not clear what "pairing" means when degrees differ.
+	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that cannot be paired (in the sense of Corollary \ref{cor:paired_neighbors}) at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
+\todo{It's not clear what "pairing" means when degrees differ.
 Thus it's confusing when it's used in the same sentence as a statement concerning differing degrees.
-If you want to use it as a concept, how about introducing it at the beginning of the section as a proof strategy?}
+If you want to use it as a concept, how about introducing it at the beginning of the section as a proof strategy?
+EM: is this a bit better?
+I tried to clarify that we mean a Corollary \ref{cor:paired_neighbors} pairing, which does not care about degree.}
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
@@
 Exhaustive tests with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees\todo{cite one of the Ollivier papers}.
+Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of \cite{Ollivier2009-bw}).
 This directly leads to the following corollary:
@@
 We identify these candidates by iterating over increasingly large $k$-neighbourhoods of $T_1$ and $T_2$, and removing duplicate topologies through hashing.
-\todo{We might be able to do better by using ``edge protection'' to avoid duplicates.
-I need to check how accurately neighbourhood sizes follow the naive $\OhOf(n^{2k})$ bound.}
 
@@
 $\OhOf( n^{2k} M)$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.
-\todo{We could probably do something smarter by starting from trees that are 0, 1, etc SPRs from $T_1$.}
 

### 2015-02-19 curvature.tex
distance-based curvature bound. max curvature conjecture.

@@
 \newtheorem{definition}[theorem]{Definition}
+\newtheorem{conjecture}[theorem]{Conjecture}
 
@@
 
-Next we bound the maximum curvature of two adjacent trees.
+Next we note a rough bound on the curvature of two trees with respect to their distance.
+
+\begin{lemma}
+	\label{lem:curvature_distance_bound}
+	Let $T$ and $S$ be two trees. Then:
+	$$ \frac{-2}{\dspr{T,S}} \le \kappa_n(T,S) \le \frac{2}{\dspr{T,S}} .$$
+\end{lemma}
+\begin{proof}
+	Observe that the distance between neighbors of $S$ and $T$, is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
+	For the curvature upper bound, we then have $\kappa_n(T,S) \le 1 - \frac{\dspr{T,S} + 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
+	The lower bound follows similarly. 
+\end{proof}
+
+Next we obtain a tighter bound on the maximum curvature of two adjacent trees.
 \begin{lemma}
@@
 This bound is tight and has been verified computationally for $n \le 7$.
+
+It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees. 
+Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
+However, our experiments with $n \le 7$ suggest that curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
+Indeed, for all but $n = 5$ the maximum curvature is obtained by pairs of trees at maximum distance (for $n = 5$, the maximum curvature, between trees at distance 2, exceeds $2/3$, the maximum curvature obtainable by maximally distant pairs).
+Moreover, nearly all of the neighbors of maximally distant pairs approach each either.
+We thus conjecture the following:
+\begin{conjecture}
+	Let $k_n$ be the maximum curvature of the uniform random walk on trees of $n$ leaves.
+	Then:
+	\begin{enumerate}
+		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)}$, for all $n \ge 6$.
+		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)}$
+	\end{enumerate}
+\end{conjecture}
+Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
+However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.\todo{this belongs here if we can prove it.
+If, not then we should consider moving it and discuss this more in the conclusion as an area for future work.}
 
@@
 	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
-	$$\kappa(T,S) - \frac{1}{3\dspr{T,S}}
+	$$\kappa_n(T,S) - \frac{1}{3\dspr{T,S}}
 	\le \kappa_{MH}(T,S)
-	\le \kappa(T,S) + \frac{1}{3\dspr{T,S}}, \text{ and }$$
-	$$\kappa(T,S) - \frac{1}{6}
+	\le \kappa_n(T,S) + \frac{1}{3\dspr{T,S}}, \text{ and }$$
+	$$\kappa_n(T,S) - \frac{1}{6}
 	\le \kappa_{MH}(T,S)
-	\le \kappa(T,S) + \frac{1}{6}.$$
+	\le \kappa_n(T,S) + \frac{1}{6}.$$
 \end{lemma}
@@
 		\kappa_{MH}(T,S) & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
-		& \ge \kappa(T,S) - \frac{1}{3\dspr{T,S}}
+		& \ge \kappa_n(T,S) - \frac{1}{3\dspr{T,S}}
 	\end{align*}
@@
 	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
-	We thus obtain the lower bound of $\kappa(T,S) - \frac{1}{6}$ in this case.
+	We thus obtain the lower bound of $\kappa_n(T,S) - \frac{1}{6}$ in this case.
 

### 2015-02-20 curvature.tex
Most of the way through.

@@
 In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and make connections to access times for those random walks.
-For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic\todo{Shall we use uniform or isotropic?
-The former is more intuitive, but the latter seems more specific to me.} random walk and MCMC sampling from the uniform prior on trees.
+For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
@@
 Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ vertices in the SPR graph be $q$ and $p$ respectively.
-Then at least $o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$ trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
+\todo{We have three systems of notation for the degree of trees: $\deg(T)$, this $q$ and $p$, and below $|N(T)|$.
+Please make this consistent, and make it clear what we actually mean by the degree of a tree.}
+Then at least
+$$o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$$
+trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
 \end{corollary}
@@
 	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
-	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$.
-	Note that this intersection is nonempty, as the fact that $T$ and $S$ differ implies that $R$ contains at most $n-2$ of the leaves, so there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$.
-\todo{Um, why can't this intersection be empty?
-EM: reason added.
-This is not strictly necessary to get an upper bound but might be interesting.}
+	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
 	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
@@
 \begin{proof}
-	Observe that the distance between neighbors of $S$ and $T$, is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
+	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
 	For the curvature upper bound, we then have $\kappa_n(T,S) \le 1 - \frac{\dspr{T,S} + 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
@@
 The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
+\todo{In view of Lemma~\ref{lem:shared_neighbors} do you need the equal size statement?}
 By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
@@
 Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
-However, our experiments with $n \le 7$ suggest that curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
+However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
 Indeed, for all but $n = 5$ the maximum curvature is obtained by pairs of trees at maximum distance (for $n = 5$, the maximum curvature, between trees at distance 2, exceeds $2/3$, the maximum curvature obtainable by maximally distant pairs).
@@
 Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.\todo{this belongs here if we can prove it.
-If, not then we should consider moving it and discuss this more in the conclusion as an area for future work.}
-
+However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
+\todo{this belongs here if we can prove it. If, not then we should consider moving it and discuss this more in the conclusion as an area for future work.}
+\todo{Do you have some key lemmas that could be tested computationally to isolate the problem?}
 Next, we bound the minimum curvature of two adjacent trees.
@@
 \begin{proof}
-	Let $T$ and $S$ be two adjacent trees.
-	Because pairs of neighbors of adjacent trees are at most distance 3 apart, we can bound the mass transport cost from above by $1 + 2m$, where $m$ is the maximum proportion of mass that can not be moved between adjacent trees.
+    In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
+    \todo{Please make it crystal clear that these two criteria can be optimized simultaneously.}
+    To bound the former effect, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
+    We can bound the mass transport cost from above by $1 + 2m$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
 	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
-	The mass transport cost is maximized (and therefore curvature minimized) by a combination of two effects: trees that cannot be paired (in the sense of Corollary \ref{cor:paired_neighbors}) at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
-\todo{It's not clear what "pairing" means when degrees differ.
-Thus it's confusing when it's used in the same sentence as a statement concerning differing degrees.
-If you want to use it as a concept, how about introducing it at the beginning of the section as a proof strategy?
-EM: is this a bit better?
-I tried to clarify that we mean a Corollary \ref{cor:paired_neighbors} pairing, which does not care about degree.}
+
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
@@
 We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
-Exhaustive tests with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
 Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of \cite{Ollivier2009-bw}).
-This directly leads to the following corollary:
+This directly leads to the following Corollary:
 
@@
 
-We next bound the effect of laziness on curvature. Let $\kappa_{l,p}$ be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$.
-For example, $\kappa_{l,\frac{1}{4}}$ moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
+We next bound the effect of laziness on curvature.
+Let $\kappa_{l,p}$\todo{conflicts with original definition of curvature $\kappa_n$.
+Perhaps $\kappa^{(p)}_n$?} be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$.
+For example, $\kappa_{l,\frac{1}{4}}$\todo{$\kappa$ does not move} moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
 \todo{should be defined earlier, particularly the part where the curvature is divided by $p$.}
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-We now bound the difference between the curvature of uniform random walks $\kappa$, and that of MH random walks, $\kappa_{MH}$.
+We now bound the difference between the curvature of uniform random walks $\kappa$, and that of MH random walks, $\kappa_{MH}$. \todo{I suggest $\kappa^{\mathrm{MH}}_n$}.
 

### 2015-02-20 curvature.tex
curvature notation

@@
 \newattribute{dspr}{d_{\mathrm{SPR}}}
-\newattribute{degree}{\operatorname{deg}}
+\newparentheses{degree}{|N(}{)|}
+
+%---[ curvature ]---------------------------------------------------------------
+\newcommand{\curvature}[2][]{%
+    \ifthenelse{\equal{#1}{}}%
+		{\kappa_n(#2)}%
+		{\kappa_n(#1;#2)}%
+}
 
@@
 \begin{equation}
-\kappa(x, y) := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
+\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
 \label{eq:curvatureDef}
@@
 
+Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
+We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
+Note that this concept of degree is unrelated to the degree of vertices of $T$.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 \label{cor:paired_neighbors}
-Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, let the degrees of $T$ and $S$ vertices in the SPR graph be $q$ and $p$ respectively.
-\todo{We have three systems of notation for the degree of trees: $\deg(T)$, this $q$ and $p$, and below $|N(T)|$.
-Please make this consistent, and make it clear what we actually mean by the degree of a tree.}
-Then at least
-$$o := q - 2kb - 2(j-1) = p - 2ka - 2(i-1)$$
+Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
+$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
 trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
-Let $N(U)$ be the set of neighbors of a tree $U$ (this does not include $U$).
 
@@
 
+We first consider properties of the uniform random walk on the rSPR graph.
+Assume that the random walk begins at a tree $T$.
+The uniform random walk selects a tree uniformly at random from $N(T)$.
+We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.
+Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(x, y)}$.
+For the uniform random walk, $m_T$ is simply the set of neighbors of $T$ with each assigned a mass of $\frac{1}{\degree{T}}$.
+
 \begin{proposition}
@@
 Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \kappa_n(T_n, S_n) = 0$ for the uniform random walk on the SPR graph.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the SPR graph.
 \end{proposition}
@@
 To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
-First we claim that $W_{1,n}$ is bounded above by $(|N(T_n)|+\OhOf{kn})/|N(T_n)|$ by exhibiting a mass transport program satisfying that bound.
+First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
 Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
@@
 There are $\OhOf{kn}$ trees unmatched by this pairing, and we can pair each of them arbitrarily with another tree of distance at most 3.
-Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + \OhOf{kn}) / |N(T_n)|$.
+Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.
 
 A lower bound is also available because we can't do better than distance 1 for all trees except for shared neighbors, of which there are $\OhOf{n}$ by Lemma~\ref{lem:shared_neighbors}.
-By ignoring these trees we get a lower bound of $(|N(T_n)| - (\OhOf{n}))/|N(T_n)|$ for $W_{1,n}$.
+By ignoring these trees we get a lower bound of $(\degree{T_n} - (\OhOf{n}))/\degree{T_n}$ for $W_{1,n}$.
 
-The desired control of $W_{1,n}$ is thus obtained because $|N(T_n)|$ is quadratic in $n$.
+The desired control of $W_{1,n}$ is thus obtained because $\degree{T_n}$ is quadratic in $n$.
 
 Now we prove the proposition when the number of neighbors differ.
-Assume without loss of generality that $|N(T_n)| < |N(S_n)|$.
-By Lemma~\ref{lem:degree_change}, $|N(S_n)| - |N(T_n)| = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
-Thus, $|N(S_n)| - |N(T_n)| = \OhOf{kn}$.
-We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $|N(T_n)| < |N(S_n)|$ we can only account for at most $|N(T_n)| / |N(S_n)|$ of the mass directly and may have to move the $(|N(S_n)| - |N(T_n)|) / |N(S_n)|$ remainder to trees a distance at most 3.
-Thus, $W_{1,n}$ is bounded above by $(|N(T_n)| + \OhOf{kn}) / |N(S_n)| = (|N(S_n)| + \OhOf{kn}) / |N(S_n)|$.
-We again bound $W_{1,n}$ from below with $(|N(T_n)| - \OhOf{n}) / |N(T_n)|$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
-The proposition again follows because $|N(T_n)|$ is quadratic in $n$.
+Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
+By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
+Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
+We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
+Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
+We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
+The proposition again follows because $\degree{T_n}$ is quadratic in $n$.
 
@@
 	Let $T$ and $S$ be two trees. Then:
-	$$ \frac{-2}{\dspr{T,S}} \le \kappa_n(T,S) \le \frac{2}{\dspr{T,S}} .$$
+	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
 \end{lemma}
@@
 	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
-	For the curvature upper bound, we then have $\kappa_n(T,S) \le 1 - \frac{\dspr{T,S} + 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
+	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} + 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
 	The lower bound follows similarly.
@@
 However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
-Indeed, for all but $n = 5$ the maximum curvature is obtained by pairs of trees at maximum distance (for $n = 5$, the maximum curvature, between trees at distance 2, exceeds $2/3$, the maximum curvature obtainable by maximally distant pairs).
-Moreover, nearly all of the neighbors of maximally distant pairs approach each either.
+Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
+Moreover, nearly all of the neighbors of these pairs approach each either.
 We thus conjecture the following:
@@
 	\begin{enumerate}
-		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)}$, for all $n \ge 6$.
-		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)}$
+		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
+		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$
 	\end{enumerate}
@@
 However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
-\todo{this belongs here if we can prove it. If, not then we should consider moving it and discuss this more in the conclusion as an area for future work.}
 \todo{Do you have some key lemmas that could be tested computationally to isolate the problem?}
@@
     In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
-    \todo{Please make it crystal clear that these two criteria can be optimized simultaneously.}
-    To bound the former effect, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
+		As we will show, these effects can be optimized simultaneously.
+    To bound these effects, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
     We can bound the mass transport cost from above by $1 + 2m$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
@@
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
-	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees.
-	Then, let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
+	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
+	We now construct a pair of trees that maximizes both effects.
+	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
 	$T$ has degree at most $3.5n^2 -15n + 16$.
 	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
-	Then the maximum $m$ is:
+	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
+	Thus, the maximum $m$ is:
 	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
@@
 We next bound the effect of laziness on curvature.
-Let $\kappa_{l,p}$\todo{conflicts with original definition of curvature $\kappa_n$.
-Perhaps $\kappa^{(p)}_n$?} be the curvature of a lazy random walk which moves with probability $p$ and remains stationary with probability $1 - p$.
-For example, $\kappa_{l,\frac{1}{4}}$\todo{$\kappa$ does not move} moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
-\todo{should be defined earlier, particularly the part where the curvature is divided by $p$.}
+The lazy random walk again selects a tree uniformly at random, as with the uniform random walk, but only accepts the move with probability $p$.
+We denote the curvature of the lazy random walk between two trees $T$ and $S$ by $\curvature[\ell_p]{T,S}$.
+For example, $\curvature[\ell_{\frac{1}{4}}]{T,S}$ describes the curvature of the lazy random walk that moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
+For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
+Note that $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$. \todo{Do you have a good explanation for the lazy curvature definition?}
 
@@
 	\begin{enumerate}
-		\item	$\kappa_{l,p}(T,S) = \kappa_{l,0}(T,S)$, if $\dspr{T,S} > 1$, and
-		\item	$\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,S) \le \kappa_{l,0}(T,S) + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\curvature[\ell_p]{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
+		\item	$\curvature{T,S} \le \curvature[\ell_p]{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
 	\end{enumerate}
@@
 \begin{proof}
-	We first prove the lower bound, that is $\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,s)$.
+	We first prove the lower bound, that is $\curvature{T,S} \le \curvature[\ell_p]{T,S}$.
 	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
-	Recall that $\kappa_{l,0}(T,S) = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\kappa_{l,p}(T,S) = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
+	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[\ell_p]{T,S} = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
 	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
@@
 	\begin{align*}
-		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
+		\curvature[\ell_p]{T,S} &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
 		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
@@
 		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
-		&\ge \kappa_{l,0}(T,S)
+		&\ge \curvature{T,S}
 	\end{align*}
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-We now bound the difference between the curvature of uniform random walks $\kappa$, and that of MH random walks, $\kappa_{MH}$. \todo{I suggest $\kappa^{\mathrm{MH}}_n$}.
+We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\mathrm{MH}]{S,T}$.
 
@@
 	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
-	$$\kappa_n(T,S) - \frac{1}{3\dspr{T,S}}
-	\le \kappa_{MH}(T,S)
-	\le \kappa_n(T,S) + \frac{1}{3\dspr{T,S}}, \text{ and }$$
-	$$\kappa_n(T,S) - \frac{1}{6}
-	\le \kappa_{MH}(T,S)
-	\le \kappa_n(T,S) + \frac{1}{6}.$$
+	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
+	\le \curvature[MH]{T,S}
+	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
+	$$\curvature{T,S} - \frac{1}{6}
+	\le \curvature[MH]{T,S}
+	\le \curvature{T,S} + \frac{1}{6}.$$
 \end{lemma}
@@
 	\begin{align*}
-		\kappa_{MH}(T,S) & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
-		& \ge \kappa_n(T,S) - \frac{1}{3\dspr{T,S}}
+		\curvature[\mathrm{MH}]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
+		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
 	\end{align*}
@@
 	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
-	We thus obtain the lower bound of $\kappa_n(T,S) - \frac{1}{6}$ in this case.
+	We thus obtain the lower bound of $\curvature{T,S} - \frac{1}{6}$ in this case.
 

### 2015-02-20 curvature.tex
more detail in the adjacent tree max curvature proof

@@
 \begin{proof}
-The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap, are of equal size, and all other tree pairs are at distance 1.
-\todo{In view of Lemma~\ref{lem:shared_neighbors} do you need the equal size statement?}
+The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
 By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
-The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}$.
-This is maximized when $\degree{T}$ is as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
+The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
+The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
+This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.
+
+The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}= \frac{6n-17}{3n^2-13n+14}$.
 \end{proof}

### 2015-02-20 curvature.tex
OhOf brackets. clarified that lazy curvature comment was aimed at matsen

@@
 For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-Note that $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$. \todo{Do you have a good explanation for the lazy curvature definition?}
+Note that $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$. \todo{EM: Do you have a good explanation for the lazy curvature definition?}
 
@@
 The size of $N_k(T_1)$ is $\OhOf{n^{2k}}$, so this takes
-$\OhOf( n^{2k} M)$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.
+$\OhOf{ n^{2k} M}$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.
 

### 2015-02-20 curvature.tex
Typo.
exposition between degree and neighborhood lemmas

@@
 
+We can improve these bounds in the case of adjacent trees.
+To do so, we require the following lemma that characterizes exactly how the degree of a tree changes after an rSPR operation.
+
 \begin{lemma}
@@
 \end{proof}
+
+Moreover, we can use these ideas to determine the number of SPR moves that are, in some respects, independent of a given SPR move.
+That is, for two trees $S$ and $T$ differing by a single SPR move, we wish to know the number of SPR moves that are applicable to both trees rather than unique to one of the trees.
+To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
+The number of such ``squares'' involving two adjacent trees will play a key rule in our bounds on curvature, as they push the curvature of those trees towards 0.
 
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
+The number of such "triangles" involving two adjacent trees has a key role in determining whether the curvature of two adjacent trees is positive or negative.
 
@@
 Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(x, y)}$.
-For the uniform random walk, $m_T$ is simply the set of neighbors of $T$ with each assigned a mass of $\frac{1}{\degree{T}}$.
+For the uniform random walk, $m_T$ is simply the set of neighbors of $T$ each assigned a mass of $\frac{1}{\degree{T}}$.
 

### 2015-02-21 curvature.tex
justification for lazy curvature definition

@@
 	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
-	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} + 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
+	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
 	The lower bound follows similarly.
@@
 For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-Note that $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$. \todo{EM: Do you have a good explanation for the lazy curvature definition?}
+Note that Ollivier defined $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$~\cite{Ollivier2009-bw}.
+This definition of curvature is invariant of $p$ for small enough $p$ and was used by Ollivier to avoid parity problems on graphs where the uniform random walk is periodic.
+As we now prove, the curvature of the lazy random walk is closely related to that of the uniform random walk.
 

### 2015-02-24 curvature.tex
Typo.
Tiny comment.

@@
 We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.
-Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(x, y)}$.
-For the uniform random walk, $m_T$ is simply the set of neighbors of $T$ each assigned a mass of $\frac{1}{\degree{T}}$.
+Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
+For the uniform random walk, $m_T$ is  simply the set
+\todo{AlexG: Technically, $m_T$ is a measure on the set of vertices of the rSPR.}
+of neighbors of $T$ each assigned a mass of $\frac{1}{\degree{T}}$.
 

### 2015-02-24 curvature.tex
fixed T and S swap in lem:degree_change

@@
 \label{lem:degree_change}
-Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
 Let $L$ be the $\lca{U,V}$ in $T$.

### 2015-02-24 curvature.tex
Definition of m_T-- thanks @alexgavruskin.
cc @cwhidden

@@
 Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
-For the uniform random walk, $m_T$ is  simply the set
-\todo{AlexG: Technically, $m_T$ is a measure on the set of vertices of the rSPR.}
-of neighbors of $T$ each assigned a mass of $\frac{1}{\degree{T}}$.
+For the uniform random walk, $m_T$ is simply the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of the neighbors of $T$.
 
@@
 However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
-\todo{Do you have some key lemmas that could be tested computationally to isolate the problem?}
 Next, we bound the minimum curvature of two adjacent trees.

### 2015-03-03 curvature.tex
minor typos. balanced tree and LCA defs

@@
 A \emph{ladder tree} is a tree such that every internal vertex has a leaf as a direct descendant.
-LCA of subtrees.
+A \emph{balanced tree} is a tree such that the sum of the depths of internal vertices is minimum over all trees with the same number of leaves.
+The \emph{least common ancestor} (LCA) of a set $R$ of two or more vertices is the unique vertex that is an ancestor of each vertex $r \in R$ and at maximum depth.
+Similarly, the LCA of two or more subtrees is the LCA of their parent vertices.
 
@@
 Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
-We pair the these trees in the mass transport.
+We pair these trees in the mass transport.
 There are $\OhOf{kn}$ trees unmatched by this pairing, and we can pair each of them arbitrarily with another tree of distance at most 3.

### 2015-03-06 curvature.tex
started drafting computational parts of section 3

@@
 \section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
+
+\subsection{Computing curvature values}
+We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
+We first used \dots to compute the rSPR treespaces for four to seven leaves.
+This required \dots time.
+We then used \dots \todo{ILP, citation, github reference} to compute curvatures for given pairs of trees.
+This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
+We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
+We thus directly computed curvature values for one representative pair from each such equivalance class, or tanglegram. \todo{cite} \todo{indicate that a paper on tanglegrams is in preparation}
+
+\subsection{Simulating random walks on the rSPR graph}
+We next explored the correspondence between curvature and graph effects.
+To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
+The uniform random walk moves to a neighboring vertex in the graph uniformly at random.
+However, this makes the walk more likely to sample higher degree vertices.
+In contrast, the Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.
+
+To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be precomputed.
+Naively, this would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree and then pick one uniformly at random.
+To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor can be selected by its index in $\OhOf{n}$ time.
+Song \todo{cite} developed a recursive formula for the degree of a tree and an explicit equation that works in terms of the distribution of heights of each node.
+This formula does not group SPR moves that move a particularly subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
+
+Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
+Recall that a tree with $n$ leaves has $2n-1$ nodes.
+Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
+However, some of these moves will result in the same neighboring tree as other SPR moves.
+In particular:
+\begin{enumerate}
+\item Moving $R$ to its sibling results in the same tree, not a neighboring tree,
+\item Moving $R$ to its parent results in the same tree,
+\item Moving $R$ to its grandparent is the same as moving its aunt to its sibling
+\item Moving $R$ to its aunt is the same as moving its aunt to R.
+\end{enumerate}
+We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
+In this manner, we achieve an alternative solution for computing the degree of a tree:
+\todo{equation}
+In particular, this formulation groups moves that move the same subtree.
+We can thus apply the following algorithm to select a neighbor uniformly at random:
+
+\begin{enumerate}
+\item	compute the degree
+\item pick a random number $r$ from $1$ to the degree
+\item uniquely order the tree by the smallest alphanumeric label of descendants
+\item preorder the nodes of $R$
+\item for each tree node $R$ in order and while $r > 0$
+	\begin{enumerate}
+		\item decrease $r$ by $2n-1-x-4$ or $2n-1-x-2$
+		\item if $r < 0$, pick the $\abs{r}$ valid move of $R$ in order
+	\end{enumerate}
+\item apply the corresponding move and return the corresponding tree
+\end{enumerate}
+
+This requires linear time and space.
+
+\subsection{Mean access time computation}
+
+We then computed mean access times (MAT) between pairs of trees in the MH random walks---the mean number of iterations required to move from one tree to the other. \todo{cite}
+MAT values have been productively applied to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes . \todo{cite sprmix}
+We used the same sprspace software to compute MAT values for our simulated random walks.
+
+\subsection{Results}
 

### 2015-03-06 tree-fun.py
deleted:    tree-fun.py

@@ -1,217 +1,0 @@
-from sage.all import Graph, matrix, SymmetricGroup
-from sage.plot.graphics import GraphicsArray
-
-
-def plot_tree(t):
-    return t.plot(layout='tree', tree_root=0, tree_orientation="down")
-
-
-def plot_tree_list(l):
-    return GraphicsArray([plot_tree(t) for t in l])
-
-
-def _enumerate_rooted_trees(n_leaves, start_internal):
-    """
-    Tree enumeration with internal nodes starting at some value.
-    """
-    assert(n_leaves > 0)
-    if n_leaves == 1:
-        g = Graph()
-        g.add_vertices([0, 1])
-        g.add_edge(0, 1)
-        return [g]
-    else:
-        l = []
-        for g in _enumerate_rooted_trees(n_leaves-1, start_internal+1):
-            for (u, v, _) in g.edges():
-                h = g.copy()
-                h.delete_edge(u, v)
-                new_leaf = h.add_vertex()
-                # One more internal node, starting with start_internal.
-                h.add_vertex(start_internal)
-                h.add_edges([(u, start_internal),
-                             (v, start_internal),
-                             (new_leaf, start_internal)])
-                l.append(h)
-        return l
-
-
-def enumerate_rooted_trees(n_leaves):
-    """
-    Construct all the rooted (with leaf 0), bifurcating, leaf-labeled
-    phylogenetic trees.
-    """
-    return _enumerate_rooted_trees(n_leaves, n_leaves + 1)
-
-
-def indexed_tree_list(to):
-    """
-    Return a list of trees up to `to-1` such that the ith element in the list
-    is a list of trees with i leaves.
-    """
-    return [[]] + [enumerate_rooted_trees(i) for i in range(1, to)]
-
-
-def tree_reduce(t, f_internal, f_leaf):
-    """
-    Assume that t is rooted at 0 and recur down through the rest of the tree
-    using the supplied functions.
-    """
-    # Imagine arrow pointing from src to dst. That is where the subtree starts.
-    def aux(src, dst):
-        n = t.neighbors(dst)
-        n.remove(src)
-        if n == []:  # Leaf.
-            return f_leaf(dst)
-        else:  # Internal node.
-            [left, right] = n
-            return f_internal(aux(dst, left), aux(dst, right))
-    [internal_root] = t.neighbors(0)
-    return aux(0, internal_root)
-
-
-def to_newick(t):
-    """
[…]

### 2015-03-06 sbatch-ricci.sh
Cleaning out old running scripts.

@@ -1,54 +1,0 @@
-#!/bin/sh
-set -eu
-
-if [ $# -ne 2 ]
-then
-  echo "usage: $0 walk n_leaves"
-  exit
-fi
-
-walk=$1
-n_leaves=$2
-
-if [ $walk != "lurw" ] && [ $walk != "upmh" ]
-then
-  echo '`walk` must be either "lurw" or "upmh"'
-  exit
-fi
-
-mkdir -p results-rspr/$walk
-
-split_size=50
-slurm_config="-c 4 -n 1"
-
-# Prepare vars and split file.
-out_path="results-rspr/$walk/ricci$n_leaves.mat"
-splitargs="-a 4 -l $split_size tangle/rooted-symmetric/tangle$n_leaves.idx"
-files=$(split $splitargs --filter="echo \$FILE")
-# Apparently split doesn't actually split when given the --filter flag, so we run it again.
-split $splitargs
-
-# Run!
-for i in $files
-do
-    export SLURM_JOB_NAME=ricci-$i
-    cmd="./ricci-tangle.py --walk $walk --graph rspr/graph_$n_leaves.sobj --out $out_path-$i $i"
-    echo $cmd
-    echo "#!/bin/sh" > $i.sh
-    echo $cmd >> $i.sh
-    sbatch $slurm_config $i.sh
-done
-
-# Wait until done.
-while test 0 -ne $(squeue -u matsen -h | wc -l)
-do
-    squeue -h -u matsen | wc -l
-    sleep 5
-done
-
-rm -f $out_path
-touch $out_path
-for i in $files
-do
-    cat $out_path-$i >> $out_path
-done

### 2015-03-06 simple-ricci.sh

@@ -1,24 +1,0 @@
-#!/bin/sh
-
-if [ $# -ne 2 ]
-then
-  echo "usage: $0 walk n_leaves"
-  exit
-fi
-
-walk=$1
-n_leaves=$2
-
-if [ $walk != "lurw" ] && [ $walk != "upmh" ]
-then
-  echo '`walk` must be either "lurw" or "upmh"'
-  exit
-fi
-
-mkdir -p results-rspr/$walk
-
-
-idx_path="tangle/rooted-symmetric/tangle$n_leaves.idx"
-out_path="results-rspr/$walk/ricci$n_leaves.mat"
-
-./ricci-tangle.py --walk $walk --graph rspr/graph_$n_leaves.sobj --out $out_path $idx_path

### 2015-03-06 scripts/load-deps.py
We have a tangle counter that is doing a thing.

@@ -7,7 +7,13 @@ wd = os.path.dirname(os.path.realpath(__file__))
 sources = [wd+'/../gricci/'+f for f in [
     'all-hail-sage/nb-fun.py',
     'all-hail-sage/tree-fun.py',
-    'code.py']]
+    'code.py']] + [
+    wd+'/../tangle/tangle-fun.py'
+    ]
 
 for source in sources:
     load(source)
+
+gap.eval("""
+Read("{}/../tangle/tangle-fun.py");
+""".format(wd))

### 2015-03-06 tangle-counter.py

@@ -1,0 +1,75 @@
+def trees_shapes_autos_dn_ttsn(n):
+    trees = enumerate_rooted_trees(n)
+    # So that we can recognize trees after acting on t2 by mu^{-1}.
+    # `dn` is short for a dictionary keyed on Newick strings.
+    dn_trees = {to_newick(trees[i]): i for i in range(len(trees))}
+    shapes = []
+    # Using Newick representation of shape is much faster than
+    # searching for isomorphic trees.
+    dn_shapes = OrderedDict()
+    tree_to_shape_num = {}
+    for i in range(len(trees)):
+        newick_shape_i = to_newick_shape(trees[i])
+        if newick_shape_i in dn_shapes:
+            dn_shapes[newick_shape_i] = dn_shapes[newick_shape_i] + [i]
+        else:
+            shapes.append(trees[i])
+            dn_shapes[newick_shape_i] = [i]
+        tree_to_shape_num[i] = dn_shapes.keys().index(newick_shape_i)
+    shape_autos = [leaf_autom_group(s) for s in shapes]
+    return (trees, shapes, shape_autos, dn_trees, tree_to_shape_num)
+
+
+def newick_factorization_dict(n):
+    trees = enumerate_rooted_trees(n)
+    fS = sg.SymmetricGroup(n)
+    dn_trees = {to_newick(trees[i]): i for i in range(len(trees))}
+    factorization_d = OrderedDict()
+    shapes = []
+    dn_shapes = OrderedDict()
+    for i in range(len(trees)):
+        newick_shape_i = to_newick_shape(trees[i])
+        if newick_shape_i in dn_shapes:
+            dn_shapes[newick_shape_i] = dn_shapes[newick_shape_i] + [i]
+            # This is the _tree_ index of the shape.
+            idx_tree = dn_shapes[newick_shape_i][0]
+            (check, cert) = rooted_is_isomorphic(trees[idx_tree], trees[i])
+            assert(check)
+            # cert is a dictionary mapping all nodes to each other.
+            # Recall that 0 is the root node, and nodes > n are internal nodes.
+            isom = fS(Permutation(cert.values()[1:n+1]))
+        else:
+            shapes.append(trees[i])
+            dn_shapes[newick_shape_i] = [i]
+            idx_tree = i
+            isom = fS.identity()
+        tree_to_shape = {}
+        factorization_d[to_newick(trees[i])] = (idx_tree, isom)
[…]

### 2015-03-06 curvature.tex
Tidying.
fleshed out SPR graph construction

@@
 \usepackage{cite}
+\usepackage{enumitem}
 
@@
 
+\subsection{Computing rSPR treespaces}
+
+We computed the rSPR treespaces for four to seven leaves.
+This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) to $\OhOf{mn^3}$.
+In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
+Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance.
+In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees).
+Thus, we simply compared each of the $m$ trees to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
+As with our curvature computations, however, an $\OhOf{m^2}$ factor where $m = (2n-3)!!$ would require an enormous amount of computation.
+
+To quickly compute dense SPR graphs (those containing a significant portion of the full treespace) we avoid pairwise computations by successively adding trees to the graph:
+\begin{enumerate}[label={\arabic*}.]
+	\item Let $G$ be an empty graph
+	\item For each of the $m$ trees in a specified but arbitrary order,
+		\begin{enumerate}
+			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index
+			\item For each of the $\OhOf{n^2}$ neighbors of $T$
+				\begin{enumerate}
+					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index
+				\end{enumerate}
+		\end{enumerate}
+\end{enumerate}
+
+\begin{theorem}
+	The subset of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
+\end{theorem}
+\begin{proof}
+	We implement the graph as an adjacency list with integer-labelled vertices.
+	As described above, the integer labels are simply the order of the input trees.
+	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.
+	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$).
+	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
+	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
+	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick representation of the tree.
+	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
+	This requires only $\OhOf{n}$-time for each tree (i.e.
+a total of $\OhOf{mn^3}$-time) using a standard nodes and pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to realistic phylogenetic trees).
+	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
+	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
+\end{proof}
+This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}, which outputs an edge list format graph suitable for input to other software.
+\todo{github url}
+Our fast graph construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
+Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm requires only 303.45 seconds to construct it.
+Constructing the 8-taxon tree space is not feasible for the previous method in a reasonable amount of time.
+Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.
+
 \subsection{Computing curvature values}
 We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
-We first used \dots to compute the rSPR treespaces for four to seven leaves.
-This required \dots time.
 We then used \dots \todo{ILP, citation, github reference} to compute curvatures for given pairs of trees.
-This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
+This latter step would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
@@
 \subsection{Simulating random walks on the rSPR graph}
+\label{sec:random_walks}
 We next explored the correspondence between curvature and graph effects.
@@
 
-\begin{enumerate}
+\begin{enumerate}[label={\arabic*}.]
 \item	compute the degree
@@
 This requires linear time and space.
+TODO: proof
+TODO: corollary that we can explore the neighbors in $O(n^3)$-time.
 
@@
 Similarly, we lose $2kb$ moves in the second case.
-In the third case, $R's$ parent subtree may now make $2(i-1)$ moves into $U$.
+In the third case, $R$'s parent subtree may now make $2(i-1)$ moves into $U$.
 Similarly, we lose $2(j-1)$ moves in the fourth case.
@@
 	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
-	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T's$ neighbors to $S$.
+	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T$'s neighbors to $S$.
 	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
@@
 
+SAGE and GAP4 software
+
 This work was funded by National Science Foundation award 1223057.

### 2015-03-09 scripts/tangle-time-count.py
Adding tangle time counting to SConstruct.

@@ -1,10 +1,4 @@
 #!/usr/bin/env sage
-'''
-Suggested use:
-./access-time-count.py --tp \
-    "$(sed -n 60p ../../tangle/rooted-symmetric/tangle5.idx)" \
-    <mcmc-trace-files>
-'''
 
 import argparse
 import collections
@@ -15,8 +9,8 @@ from sage.all import load, RR
 load(os.path.dirname(os.path.realpath(__file__))+'/load-deps.py')
 load(os.path.dirname(os.path.realpath(__file__))+'/tangle-counter.py')
 
-parser = argparse.ArgumentParser(description='Find commute times',
-                                 prog='access-time-count.py')
+parser = argparse.ArgumentParser(description='Find commute times per tangle',
+                                 prog='tangle-time-count.py')
 
 parser.add_argument('tracefiles', nargs='+')
 parser.add_argument(
@@ -29,7 +23,7 @@ args = parser.parse_args()
 
 tc = TangleCounter(int(args.n))
 
-with open(args.o, 'w') as fout:
+with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
     for trace in args.tracefiles:
         print trace
         # Forget if we saw it in another trace file.

### 2015-03-10 scripts/tangle-time-count.py
Tidying up.

@@ -1,11 +1,9 @@
 #!/usr/bin/env sage
 
 import argparse
-import collections
 import gzip
 import os
-import re
-from sage.all import load, RR
+from sage.all import load
 load(os.path.dirname(os.path.realpath(__file__))+'/load-deps.py')
 load(os.path.dirname(os.path.realpath(__file__))+'/tangle-counter.py')
 
@@ -32,6 +30,8 @@ with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
             for l in fin:
                 (idx, curr) = l.split()
                 idx = int(idx)
+                if 0 == idx % 100:
+                    print str(idx)
                 for (prev, when) in last_seen.items():
                     if when is not None:
                         # idx - last_seen[prev] is the number of steps
@@ -42,6 +42,8 @@ with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
                 last_seen[curr] = idx
             for (t1_idx, t2_idx), l in tc.count_d.items():
                 for (coset, counter) in l:
-                    newicks = to_newick_pair(tc.trees[t1_idx], tc.trees[t2_idx], coset)
+                    newicks = to_newick_pair(
+                        tc.trees[t1_idx], tc.trees[t2_idx], coset)
                     for (time, count) in counter.items():
-                        fout.write('\t'.join([newicks, str(time), str(count)])+'\n')
+                        fout.write(
+                            '\t'.join([newicks, str(time), str(count)])+'\n')

### 2015-03-10 curvature.tex
A sub-pass through tex. cc @cwhidden.

@@
 \newcommand{\TT}{\mathcal T}
+\newcommand{\MH}{\operatorname{MH}}
 
@@
 We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
-We then used \dots \todo{ILP, citation, github reference} to compute curvatures for given pairs of trees.
-This latter step would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
+We first used \todo{\dots} to compute the rSPR treespaces for four to seven leaves.
+This required \dots time.
+We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
+This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-We thus directly computed curvature values for one representative pair from each such equivalance class, or tanglegram. \todo{cite} \todo{indicate that a paper on tanglegrams is in preparation}
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} (enumeration methods described in a manuscript in preparation).
 
@@
 However, this makes the walk more likely to sample higher degree vertices.
-In contrast, the Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
 The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.
 
-To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be precomputed.
-Naively, this would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree and then pick one uniformly at random.
-To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor can be selected by its index in $\OhOf{n}$ time.
-Song \todo{cite} developed a recursive formula for the degree of a tree and an explicit equation that works in terms of the distribution of heights of each node.
-This formula does not group SPR moves that move a particularly subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
+To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
+A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
+To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor as described in the next paragraph.
+Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
+We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
 
 Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
-Recall that a tree with $n$ leaves has $2n-1$ nodes.
+Recall that a tree with $n$ leaves has $2n-1$ total nodes.
 Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
 However, some of these moves will result in the same neighboring tree as other SPR moves.
-In particular:
+In particular, where ``sibling edge'' means the edge connecting the sibling subtree to the rest of the tree, etc.:
 \begin{enumerate}
-\item Moving $R$ to its sibling results in the same tree, not a neighboring tree,
-\item Moving $R$ to its parent results in the same tree,
-\item Moving $R$ to its grandparent is the same as moving its aunt to its sibling
-\item Moving $R$ to its aunt is the same as moving its aunt to R.
+\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
+\item Moving $R$ to its parent edge results in the same tree,
+\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge
+\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
 \end{enumerate}
+\todo{How do we know that there aren't any other symmetries?}
 We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
-In this manner, we achieve an alternative solution for computing the degree of a tree:
+In this manner, we achieve an alternative solution for computing the degree of a tree in the SPR graph:
 \todo{equation}
@@
 \item uniquely order the tree by the smallest alphanumeric label of descendants
-\item preorder the nodes of $R$
+\item preorder the nodes of $R$ \todo{preorder how?}
 \item for each tree node $R$ in order and while $r > 0$
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\mathrm{MH}]{S,T}$.
+We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\MH]{S,T}$.
 
@@
 	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
-	\le \curvature[MH]{T,S}
+	\le \curvature[\MH]{T,S}
 	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
 	$$\curvature{T,S} - \frac{1}{6}
-	\le \curvature[MH]{T,S}
+	\le \curvature[\MH]{T,S}
 	\le \curvature{T,S} + \frac{1}{6}.$$
@@
 	\begin{align*}
-		\curvature[\mathrm{MH}]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
+		\curvature[\MH]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
 		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
@@
 SAGE and GAP4 software
+Alexander Hulpke
 

### 2015-03-10 curvature.tex
consistent usage of vertex for SPR graph vertices and node for tree vertices

@@
 SPR graph.
+For clarity, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
+We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices. \todo{Not sure if this is better here or in the preliminaries. Revisit this later}
 For MCMC moves, one would like to ensure good mixing.
@@
 The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
-A \emph{ladder tree} is a tree such that every internal vertex has a leaf as a direct descendant.
-A \emph{balanced tree} is a tree such that the sum of the depths of internal vertices is minimum over all trees with the same number of leaves.
-The \emph{least common ancestor} (LCA) of a set $R$ of two or more vertices is the unique vertex that is an ancestor of each vertex $r \in R$ and at maximum depth.
-Similarly, the LCA of two or more subtrees is the LCA of their parent vertices.
+A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
+A \emph{balanced tree} is a tree such that the sum of the depths of internal nodes is minimum over all trees with the same number of leaves.
+The \emph{least common ancestor} (LCA) of a set $R$ of two or more nodes is the unique node that is an ancestor of each node $r \in R$ and at maximum depth.
+Similarly, the LCA of two or more subtrees is the LCA of their parent nodes.
 
@@
 We computed the rSPR treespaces for four to seven leaves.
-This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) to $\OhOf{mn^3}$.
+This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case).
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
@@
 In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees).
-Thus, we simply compared each of the $m$ trees to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
-As with our curvature computations, however, an $\OhOf{m^2}$ factor where $m = (2n-3)!!$ would require an enormous amount of computation.
+We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
+However, applying this method to complete SPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, would require an enormous amount of computation.
 
@@
 \begin{theorem}
-	The subset of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
+	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
 \end{theorem}
@@
 Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate vertices on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
-Similarly, let $b$ be the number of intermediate vertices on the path from $V$ to $L$ in $T$, excluding endpoints.
+Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
+Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
 Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
@@
 The set of permissible SPR moves changes in four different ways due to the movement of $R$:
-(i) subtrees that include vertices on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent vertex,
-(ii) subtrees that include vertices on the path from $V$ to $L$ may no longer be moved into $R$ and its parent vertex,
+(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
+(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
 (iii) $R$'s parent subtree may now be moved into $U$, and
@@
 \begin{proof}
-By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target vertices.
+By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.

### 2015-03-10 curvature.tex
a few tweaks in response to ec79c6dcdf72aa0f15c582a6eb49e3c9be33734b

@@
 \subsection{Computing rSPR treespaces}
+\label{sec:computing_treespace}
 
@@
 Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm requires only 303.45 seconds to construct it.
-Constructing the 8-taxon tree space is not feasible for the previous method in a reasonable amount of time.
+Constructing the 8-taxon tree space is not feasible using the previous method in a reasonable amount of time.
+\todo{Going on 92 hours at the moment.
+I'll give it a bit longer and then just say that we terminated our attempt after X hours}
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.
@@
 We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
-We first used \todo{\dots} to compute the rSPR treespaces for four to seven leaves.
-This required \dots time.
+We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
@@
 A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
-To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor as described in the next paragraph.
+To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
 Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.

### 2015-03-10 curvature.tex
dropped k-tubes

@@
 To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
-A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
+A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
 To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
@@
 
-\section{$k$-tubes}
-
-The curvature with respect to two topologies, $T_1$ and $T_2$, is primarily determined by a subset of the SPR graph in the localized neighborhood of these topologies.
-We first require some notation to define these subsets.
-Let $\TT$ be the full set of topologies with a given number of leaves.
-Denote the set of trees with SPR distance at most $k$ from a given tree $T$ by $N_k(T)$.
-These trees are the $k$-neighbors of $T$.
-That is, $$ N_k(T) = \set{t \in \TT \mid \dspr{T, t} \le k} .$$
-
-We now define the $k$-tube of two trees $T_1$ and $T_2$ to be the union of their $k$-neighbors and \emph{every tree on a shortest path between a $k$-neighbor of $T_1$ and a $k$-neighbor of $T_2$}.
-More formally, the $k$-tube for $T_1$ and $T_2$ is:
-\[ \set{t \in \TT \mid \dspr{t, t_1} + \dspr{t, t_2} = \dspr{t_1, t_2},\ \exists_{t_1 \in N_k(T_1)},\ \exists_{t_2 \in N_k(T_2)} } .\]
-We call $T_1$ and $T_2$ the centers of the $k$-tube.
-The distance between $k$-neighbors of different central trees is bounded by the maximum distance from each $k$-neighbor to its respective center and the maximum distance between the centers:
-
-\begin{lemma}
-	\label{lem:neighbor_distance}
-	$\dspr{t_1, t_2} \le \dspr{T_1, T_2} + 2k,\ \forall_{t_1 \in N_k(T_1)},\ \forall_{t_2 \in N_k(T_2)}.$
-\end{lemma}
-
-To efficiently compute a $k$-tube, we first bound the maximum distance from members of the $k$-tube to the closest central tree.
-\begin{lemma}
-	\label{lem:k_tube_distance}
-	Any tree $t$ of the $k$-tube of two trees $T_1$ and $T_2$ satisfies:
-	\[ \min\left({\dspr{t, T_1}, \dspr{t, T_2}}\right) \le \floor{\dspr{T_1, T_2}/2} + 2k .\]
-\end{lemma}
-\begin{proof}
-Without loss of generality, assume that $t$ is on a shortest path between two trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$, and is either closer to $t_1$ than $t_2$ or equidistant from both.
-By Lemma~\ref{lem:neighbor_distance} there is a path of length at most $\dspr{T_1, T_2} + 2k$ between $t_1$ and $t_2$.
-Then $\dspr{t, t_1} \le \floor{(\dspr{T_1, T_2} + 2k) / 2}$.
-
-Along with the fact that $\dspr{T_1, t_1} \le k$, we have that
-\begin{align*}
-\dspr{t, T_1} &\le \dspr{t, t_1} + \dspr{t_1, T_1} \\
-&\le \floor{(\dspr{T_1, T_2} + 2k) / 2} + k \\
-&\le \floor{\dspr{T_1, T_2} / 2} + 2k.
-\end{align*}
-
-\end{proof}
-We can thus restrict our search for members of the $k$-tube to topologies of distance at most $\floor{\dspr{T_1, T_2}/2} + 2k$ from the central trees.
-We call this set of trees $k$-tube candidates.
-Note that there are $\OhOf{n^{2(\floor{\dspr{T_1, T_2}/2} + 2k)}}$ $k$-tube candidates.
-We identify these candidates by iterating over increasingly large $k$-neighbourhoods of $T_1$ and $T_2$, and removing duplicate topologies through hashing.
-
-In the second part of our procedure for computing a $k$-tube, we identify $k$-tube candidates that are not on the shortest path between two $k$-neighbors of the central trees.
-We first compute shortest path lengths from each member of $N_k(T_1)$ to each $k$-tube candidate using breadth-first search (BFS).
-For each BFS traversal tree, we backtrack from members of $N_k(T_2)$, identifying topologies in the $k$-tube.
-We can use unique tree indices here, rather than hashing, to avoid duplicates.
-The size of $N_k(T_1)$ is $\OhOf{n^{2k}}$, so this takes
-$\OhOf{ n^{2k} M}$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.
-
-This procedure identifies exactly the trees in the $k$-tube, so we have the following lemma:
-\begin{lemma}
-	\label{lem:k_tube_correctness}
-	This procedure correctly identifies the $k$-tube.
-\end{lemma}
-\begin{proof}
-	First, suppose that this procedure identifies a tree $x$ that is not in the $k$-tube.
-	Assume, without loss of generality, that the BFS tree $B$ from tree $t_1 \in N_k(T_1)$ transitioned from $x$ to a $k$-tube tree $t$.
-	Moreover, there must be a path in $B$ from $t$ to a tree $t_2 \in N_k(T_2)$ such that $t$ is on a shortest path from $t_1$ to $t_2$.
-	Then, by the properties of BFS, $x$ is on this shortest path and belongs in the $k$-tube.
-
-	Now, suppose that this procedure does not identify a tree $x$ that belongs in the $k$-tube.
-	$x$ must be on a shortest path between some trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$.
-	Then, by the properties of BFS, $x$ would have been on the path from $t_1$ to $t_2$ in the $t_1$ BFS tree.
-\end{proof}
-
 \section{Conclusion and future work}

### 2015-03-11 scripts/double-coset-counter.py
A fast double coset counter. First commit.

@@ -1,0 +1,15 @@
+gap.eval("""
+Read("scripts/double-coset-counter.g");
+""")
+
+
+# Groups and cosets
+
+def new_dc_counter(g, u, v):
+    return gap.function_call('NewDCCounter', [gap(g), gap(u), gap(v)])
+
+def add_dc_counter(dcc, r, time):
+    return gap.function_call('AddDCCounter', [dcc, gap(r), time])
+
+def show_dc_counter(dcc):
+    return gap.function_call('ShowDCCounter', dcc)

### 2015-03-11 scripts/double-coset-counter.py
Converting to sage-digestible table.

@@ -11,5 +11,5 @@ def new_dc_counter(g, u, v):
 def add_dc_counter(dcc, r, time):
     return gap.function_call('AddDCCounter', [dcc, gap(r), time])
 
-def show_dc_counter(dcc):
-    return gap.function_call('ShowDCCounter', dcc)
+def dc_counter_table(dcc):
+    return gap.function_call('DCCounterTable', dcc)

### 2015-03-11 scripts/double-coset-counter.py
2x coset counting integrated into timing.

@@ -1,15 +1,20 @@
-gap.eval("""
-Read("scripts/double-coset-counter.g");
-""")
-
-
 # Groups and cosets
+
+wd = os.path.dirname(os.path.realpath(__file__))
+gap.eval("""
+Read("{}/double-coset-counter.g");
+""".format(wd))
 
 def new_dc_counter(g, u, v):
     return gap.function_call('NewDCCounter', [gap(g), gap(u), gap(v)])
 
-def add_dc_counter(dcc, r, time):
-    return gap.function_call('AddDCCounter', [dcc, gap(r), time])
+def new_dc_counter_exemplar(g, coset):
+    return gap.function_call('NewDCCounterExemplar', [gap(g), coset])
+
+def add_dc_counter(dcc, coset, time):
+    return gap.function_call('AddDCCounter', [dcc, coset, time])
 
 def dc_counter_table(dcc):
-    return gap.function_call('DCCounterTable', dcc)
+    tab = gap.function_call('DCCounterTable', dcc)
+    return [[gap.eval(coset_str), counts]
+        for (coset_str, counts) in tab.sage()]

### 2015-03-11 scripts/load-deps.py

@@ -9,7 +9,8 @@ sources = [wd+'/../gricci/'+f for f in [
     'all-hail-sage/nb-fun.py',
     'all-hail-sage/tree-fun.py',
     'code.py']] + [
-    wd+'/../tangle/tangle-fun.py'
+    wd+'/../tangle/tangle-fun.py',
+    wd+'/../scripts/double-coset-counter.py'
     ]
 
 for source in sources:
@@ -17,4 +18,5 @@ for source in sources:
 
 gap.eval("""
 Read("{}/../tangle/tangle-fun.g");
-""".format(wd))
+Read("{}/double-coset-counter.g");
+""".format(wd, wd))

### 2015-03-11 scripts/tangle-counter.py

@@ -55,12 +55,16 @@ class TangleCounter:
         self.nfd = newick_factorization_dict(n)
         (self.trees, _, self.shape_autos, _, self.ttsn) = trees_shapes_autos_dn_ttsn(n)
         self.count_d = defaultdict(list)
+        self.fS = sg.SymmetricGroup(n)
 
     # Define an "ntangle" to be a triple of (idx_tree1, idx_tree2, coset)
     # WRT the `sigma_1 * inverse(sigma_2)` term below,
-    # recall that the element in the middle of the double coset gets applied to the first tree to get the right tangle.
-    # sigma_1 maps us from leaves of first tree to "middle" leaves, and sigma_2 maps us from "middle" leaves to leaves of second tree.
-    # Thus sigma_1 * inverse(sigma_2) gives us the map to apply to the first tree to get correct tangle when also using identity map for second tree.
+    # recall that the element in the middle of the double coset gets applied to
+    # the first tree to get the right tangle.
+    # sigma_1 maps us from leaves of first tree to "middle" leaves, and sigma_2
+    # maps us from "middle" leaves to leaves of second tree.
+    # Thus sigma_1 * inverse(sigma_2) gives us the map to apply to the first
+    # tree to get correct tangle when also using identity map for second tree.
     def newick_pair_to_ntangle(self, n1, n2):
         (idx_tree1, sigma_1) = self.nfd[n1]
         (idx_tree2, sigma_2) = self.nfd[n2]
@@ -77,12 +81,9 @@ class TangleCounter:
 
     def add_newick_pair_observation(self, n1, n2, time):
         (idxs, coset) = self.newick_pair_to_ntangle(n1, n2)
-        if idxs in self.count_d:
-            for known_coset, count in self.count_d[idxs]:
-                if coset == known_coset:
-                    count[time] += 1
-                    break
-            else:  # Not a known coset.
-                self.count_d[idxs].append((coset, Counter({time: 1})))
-        else:  # Not a known pair of indices.
-            self.count_d[idxs].append((coset, Counter({time: 1})))
+        if idxs not in self.count_d:
+            self.count_d[idxs] = new_dc_counter_exemplar(self.fS, coset)
+        add_dc_counter(self.count_d[idxs], coset, time)
+
+    def get_counts(self, t1_idx, t2_idx):
+        return dc_counter_table(self.count_d[t1_idx, t2_idx])

### 2015-03-11 scripts/tangle-time-count.py

@@ -3,9 +3,10 @@
 import argparse
 import gzip
 import os
-from sage.all import load
-load(os.path.dirname(os.path.realpath(__file__))+'/load-deps.py')
-load(os.path.dirname(os.path.realpath(__file__))+'/tangle-counter.py')
+from sage.all import load, gap
+wd = os.path.dirname(os.path.realpath(__file__))
+load(wd+'/load-deps.py')
+load(wd+'/tangle-counter.py')
 
 parser = argparse.ArgumentParser(description='Find commute times per tangle',
                                  prog='tangle-time-count.py')
@@ -40,10 +41,10 @@ with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
                         tc.add_newick_pair_observation(
                             prev, curr, idx - last_seen[prev])
                 last_seen[curr] = idx
-            for (t1_idx, t2_idx), l in tc.count_d.items():
-                for (coset, counter) in l:
+            for (t1_idx, t2_idx) in tc.count_d.keys():
+                for (coset, counts) in tc.get_counts(t1_idx, t2_idx):
                     newicks = to_newick_pair(
                         tc.trees[t1_idx], tc.trees[t2_idx], coset)
-                    for (time, count) in counter.items():
+                    for time, count in enumerate(counts):
                         fout.write(
                             '\t'.join([newicks, str(time), str(count)])+'\n')

### 2015-03-12 curvature.tex
SPR symmetry lemma

@@
 \usepackage{enumitem}
+\usepackage{url}
 
@@
 \end{proof}
-This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}, which outputs an edge list format graph suitable for input to other software.
-\todo{github url}
+This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
 Our fast graph construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
-Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm requires only 303.45 seconds to construct it.
-Constructing the 8-taxon tree space is not feasible using the previous method in a reasonable amount of time.
-\todo{Going on 92 hours at the moment.
-I'll give it a bit longer and then just say that we terminated our attempt after X hours}
+Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it.
+Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and that method is infeasible for construcing larger tree spaces.
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.
@@
 
-To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
+To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
 A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
@@
 However, some of these moves will result in the same neighboring tree as other SPR moves.
-In particular, where ``sibling edge'' means the edge connecting the sibling subtree to the rest of the tree, etc.:
+In particular, where a node's edge means the edge connecting the subtree rooted at that node to the rest of the tree,
 \begin{enumerate}
@@
 \end{enumerate}
-\todo{How do we know that there aren't any other symmetries?}
 We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
+Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
 In this manner, we achieve an alternative solution for computing the degree of a tree in the SPR graph:
-\todo{equation}
+
+\begin{lemma}
+	For a tree $T$ with $n$ leaves,
+	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
+	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
+	$$\size{N(T,u)} = \begin{cases}
+		2n - 1 - x - 4 &\text{if depth($u$) $ > 1$, } \\
+		2n - 1 - x - 2 &\text{if depth($u$) $ = 1$ } \\
+		0 &\text{if depth($u$) $ = 0$ } \\
+	\end{cases}.$$
+	Moreover, each tree of $N(T,u)$ can be obtained by an SPR operation that moves the subtree rooted at $u$.
+\end{lemma}
+\begin{proof}
+	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
+	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
+	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
+	Call these $U$ and $V$, respectively.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$.
+	Then the SPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is SPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
+	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
+(iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e.
+iv above), a contradiction.
+	Therefore the claim holds.
+\end{proof}
+
 In particular, this formulation groups moves that move the same subtree.

### 2015-03-12 curvature.tex
proof that we can select a random SPR neighbor in linear time

@@
 \begin{theorem}
+	\label{thm:construct_graph}
 	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
@@
 However, some of these moves will result in the same neighboring tree as other SPR moves.
-In particular, where a node's edge means the edge connecting the subtree rooted at that node to the rest of the tree,
+In particular, where a node's edge means the edge connecting the subtree rooted at that node to the rest of the tree:
 \begin{enumerate}
@@
 \begin{lemma}
+	\label{lem:compute_degree}
 	For a tree $T$ with $n$ leaves,
@@
 In particular, this formulation groups moves that move the same subtree.
-We can thus apply the following algorithm to select a neighbor uniformly at random:
+We can thus apply the following algorithm to select a neighbor uniformly at random for a tree $T$:
 
 \begin{enumerate}[label={\arabic*}.]
-\item	compute the degree
-\item pick a random number $r$ from $1$ to the degree
-\item uniquely order the tree by the smallest alphanumeric label of descendants
-\item preorder the nodes of $R$ \todo{preorder how?}
-\item for each tree node $R$ in order and while $r > 0$
+	\item	compute the degree of $T$, $\degree{T}$.
+	\item pick a random integer $r$ in the range $[1,\degree{T}]$
+	\item label each node $u$ of $T$ by its preorder number and compute the number of nodes in the subtree rooted at each $u$.
+\item for each tree node $u$ and while $r > 0$
 	\begin{enumerate}
-		\item decrease $r$ by $2n-1-x-4$ or $2n-1-x-2$
-		\item if $r < 0$, pick the $\abs{r}$ valid move of $R$ in order
+		\item decrease $r$ by $\size{N(T,u)}$
+		\item if $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop
 	\end{enumerate}
-\item apply the corresponding move and return the corresponding tree
+\item return the neighbor $S$
 \end{enumerate}
 
-This requires linear time and space.
-TODO: proof
-TODO: corollary that we can explore the neighbors in $O(n^3)$-time.
+\begin{lemma}
+	\label{lem:select_random_neighbor}
+	Given a tree $T$, an SPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOF{n}$ space.
+\end{lemma}
+\begin{proof}
+	We apply the above procedure.
+	We use a standard nodes and pointer representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
+	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
+	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted $u$.
+	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in array indexed by preorder number.
+	Both of these traversals require $\OhOf{n}$-time.
+	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
+	Moreover, the tree $S$ can be found by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding SPR destination.
+	Finally, we require linear time to apply the chosen SPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
+	Thus, the for loop requires linear time.
+	By Lemma~\ref{lem:compute_degree} the chosen tree is an SPR neighbor of $T$ and is chosen uniformly at random.
+	Therefore, the procedure uses linear time and space and selects an SPR neighbor of $T$ uniformly at random.
+\end{proof}
+
+Observe that this procedure can be easily adapted to explore the full neighborhood of a tree in $\OhOf{n^3}$ time, which we use for Theorem~\ref{thm:construct_graph}.
+We thus have the following corollary:
+
+\begin{corollary}
+	Given a tree $T$, the SPR neighbors of $T$ can be enumerated in $\OhOf{n^3}$-time.
+\end{corollary}
 
@@
 	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
-	By Lemma~\ref{ }\todo{need unique move lemma}, each of the subtrees is assigned to at most $2n-6$ unique moves.
+	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
 	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).

### 2015-03-12 curvature.tex
fleshing out the intro

@@
 Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
-The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution (cite{Felsenstein}).
+The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution~\cite{felsenstein1981evolutionary}.
 The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
@@
 % Here we focus on the discrete search part.
-There are a number of such means of doing so.
-SPR.
-SPR graph.
-For clarity, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
-We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices. \todo{Not sure if this is better here or in the preliminaries. Revisit this later}
-For MCMC moves, one would like to ensure good mixing.
-For ML moves, we are concerned about local optima.
-We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
-We use "random walks" as being a general notion of movement around a graph, including MCMC.
+There are a number of means of doing so, most of which are based on \emph{splits}---partitions of leaves induced by edges in the trees---such as the Robinson Foulds distance~\cite{robinson81}.
+However, phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves~\cite{hein96} or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
 
-Researchers have noticed SPR graph effects in MCMC.
-\cite{Mossel2005-ly,Mossel2006-fo}
-\cite{Ronquist2006-fv}
-\cite{Stefankovic2011-hu}
-We recently showed that graph structure has an effect of MCMC mixing using real data and programs.
-\cite{Whidden2015-yi}
-Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves \cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object called the compact real tree which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
+It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
+We will call these "graph effects"---one can't easily move from one tree to another due to the structure of the graph.
+Researchers have noticed SPR graph effects in MCMC~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
+We recently showed that graph structure has a significant effect on MCMC mixing using real data and programs~\cite{Whidden2015-yi}.
+Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object called the compact real tree which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
 
 ML also has graph effects, which manifest as local maxima.
-There have been investigation of local maxima, but concern branch lengths.
-\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}
-The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another \cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
+There have been investigations of local maxima, but mainly from the perspective of branch lengths rather than topology~\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}.
+The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another~\cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
 We are not aware of any general conclusions concerning hill-climbing on the SPR or related graphs that has come from this work.
@@
 
-Recently, a new approach to quantifying properties of discrete and continuous spaces with respect to random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
+Recently, a new approach to quantifying properties of discrete and continuous spaces with respect to random walks has been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
 They define a notion of curvature on a discrete or continuous case.
-Formalizes idea of the extent to which random walking brings points together, and is generalization of the notion of Ricci curvature on manifolds.
+This formalizes the notion of to what extent random walking brings points together, and is a generalization of the notion of Ricci curvature on manifolds.
 
 In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and make connections to access times for those random walks.
+We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC and ML searches.
 For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
+We present a fast new algorithm for computing SPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
+As the the full rSPR treespace contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of treespace.
+By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
+By carefully examining the overlap in SPR moves, we present a new method for computing the degree of a tree in the SPR graph that allows one to select an SPR neighbor uniformly at random in linear-time without explicitly generating the graph.
+Using this method to simulate these random walks, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
-By exploiting symmetries in the rSPR graph, we are able to calculate all of the curvatures for pairs of trees with up to seven leaves.
-Using simulation of these processes, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
+In particular, we present a full characterization of the change in SPR degree that occurs from a given SPR move and find that, contrary to previous assumptions, SPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
 
@@
 Similarly, the LCA of two or more subtrees is the LCA of their parent nodes.
+
+For clarity, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
+We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices. \todo{Not sure if this is better here or in the preliminaries. Revisit this later}
+
+Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
+We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
+Note that this concept of degree is unrelated to the degree of vertices of $T$.
 
@@
 We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
+We then used linear programming to compute curvatures for given pairs of trees~\cite{gricci}.
 This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
@@
 Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
-We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
+We note that the recursive formula of Song~\cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
 
@@
 	\label{lem:select_random_neighbor}
-	Given a tree $T$, an SPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOF{n}$ space.
+	Given a tree $T$, an SPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
 \end{lemma}
@@
 	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
-	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted $u$.
+	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
 	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in array indexed by preorder number.
@@
 
+We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
+We sampled \ldots \todo{sampling and runtime information}.
+
 \subsection{Mean access time computation}
 
-We then computed mean access times (MAT) between pairs of trees in the MH random walks---the mean number of iterations required to move from one tree to the other. \todo{cite}
-MAT values have been productively applied to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes . \todo{cite sprmix}
-We used the same sprspace software to compute MAT values for our simulated random walks.
+We then computed mean access times (MAT) between pairs of trees in the MH random walks---the mean number of iterations required to move from one tree to the other~\cite{lovasz1993random}.
+MAT values have been productively applied to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi}.
+We used the same \texttt{sprspace}~\cite{sprspace} software to compute MAT values for our simulated random walks.\todo{runtime information}
 
@@
 
-Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
-We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
-Note that this concept of degree is unrelated to the degree of vertices of $T$.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 
-\begin{lemma}[{Song \cite{Song2003-gf}}]
+\begin{lemma}[{Song~\cite{Song2003-gf}}]
     \pushQED{\qed}
@@
 Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of \cite{Ollivier2009-bw}).
+Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
 This directly leads to the following Corollary:
@@
 
-Ranked trees \cite{Song2006-xe}
-Partitions \cite{Gusfield2002-il}
+Ranked trees~\cite{Song2006-xe}
+Partitions~\cite{Gusfield2002-il}
 

### 2015-03-13 scripts/double-coset-counter.py
Standardizing double cosets.

@@ -5,6 +5,9 @@ gap.eval("""
 Read("{}/double-coset-counter.g");
 """.format(wd))
 
+def standardize_double_coset(dc):
+    return gap.function_call('StandardizeDoubleCoset', dc)
+
 def new_dc_counter(g, u, v):
     return gap.function_call('NewDCCounter', [gap(g), gap(u), gap(v)])
 

### 2015-03-13 scripts/tangle-time-count.py

@@ -3,7 +3,7 @@
 import argparse
 import gzip
 import os
-from sage.all import load, gap
+from sage.all import load
 wd = os.path.dirname(os.path.realpath(__file__))
 load(wd+'/load-deps.py')
 load(wd+'/tangle-counter.py')
@@ -44,7 +44,9 @@ with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
             for (t1_idx, t2_idx) in tc.count_d.keys():
                 for (coset, counts) in tc.get_counts(t1_idx, t2_idx):
                     newicks = to_newick_pair(
-                        tc.trees[t1_idx], tc.trees[t2_idx], coset)
+                        tc.trees[t1_idx],
+                        tc.trees[t2_idx],
+                        standardize_double_coset(coset))
                     for time, count in enumerate(counts):
                         fout.write(
                             '\t'.join([newicks, str(time), str(count)])+'\n')

### 2015-03-13 scripts/double-coset-counter.py
Starting to standardize tangles.

@@ -5,9 +5,6 @@ gap.eval("""
 Read("{}/double-coset-counter.g");
 """.format(wd))
 
-def standardize_double_coset(dc):
-    return gap.function_call('StandardizeDoubleCoset', dc)
-
 def new_dc_counter(g, u, v):
     return gap.function_call('NewDCCounter', [gap(g), gap(u), gap(v)])
 

### 2015-03-13 scripts/tangle-time-count.py

@@ -43,10 +43,13 @@ with gzip.GzipFile(args.o, mode='wb', mtime=0.) as fout:
                 last_seen[curr] = idx
             for (t1_idx, t2_idx) in tc.count_d.keys():
                 for (coset, counts) in tc.get_counts(t1_idx, t2_idx):
-                    newicks = to_newick_pair(
-                        tc.trees[t1_idx],
-                        tc.trees[t2_idx],
-                        standardize_double_coset(coset))
+                    x = standardize_tangle(
+                        tc.trees[t1_idx], tc.trees[t2_idx], coset)
+                    newicks = to_newick_pair(*x)
                     for time, count in enumerate(counts):
                         fout.write(
-                            '\t'.join([newicks, str(time), str(count)])+'\n')
+                            '\t'.join(
+                                [str(item) for item in
+                                 [t1_idx, t2_idx, "".join(str(x[2]).split()),
+                                 newicks, time, count]])
+                            + '\n')

### 2015-03-14 curvature.tex
should probably use IEEE bibtex format
cite SPR distance papers

@@
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
-Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance.
-In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees).
+Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
+In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast,Whidden2015-yi}.
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
@@
 Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it.
-Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and that method is infeasible for construcing larger tree spaces.
+Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and that method is infeasible for constructing larger tree spaces.
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{IEEEtran}
 \bibliography{curvature}

### 2015-03-18 scripts/plot-access.py
Adding distribution plots too. Ugly!

@@ -26,9 +26,6 @@ parser.add_argument('curvatures', type=str, help='Tangle curvatures.')
 parser.add_argument('--prefix', required=True,
                     type=str, help='Prefix of plot paths for saving.')
 
-parser.add_argument('--which', choices=['hexbin', 'scatter'],
-                    help='What type of plot is desired.')
-
 args = parser.parse_args()
 for p in [args.access, args.tangles, args.degrees, args.curvatures]:
     assert(os.path.exists(p))
@@ -48,25 +45,37 @@ catted['mean_access'] = \
               for (tangle_num, adf) in access.groupby('tangle'))
 catted['tangle'] = catted.index
 
-df_deg_grouped = catted.groupby(['t1_deg','t2_deg', 'dist'])
 t1_degs = catted['t1_deg'].unique()
 t2_degs = catted['t2_deg'].unique()
 
-fig, axs = plt.subplots(
-    len(t1_degs), len(t2_degs),
-    sharex=True,
-    sharey=True,
-    figsize=(4*len(t1_degs), 4*len(t1_degs)),
-    squeeze=False)  # Always return an array of axs, even if 1x1.
+def subplots():
+    return plt.subplots(
+        len(t1_degs), len(t2_degs),
+        sharex=True,
+        sharey=True,
+        figsize=(4*len(t1_degs), 4*len(t1_degs)),
+        squeeze=False)  # Always return an array of axs, even if 1x1.
+e_fig, e_axs = subplots()
+d_fig, d_axs = subplots()
 
-for (t1_deg, t2_deg, dist), group in df_deg_grouped:
+for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist']):
     i = np.where(t1_degs == t1_deg)[0][0]
     j = np.where(t2_degs == t2_deg)[0][0]
     p = group['mean_access'].plot(
-        ax=axs[i][j], kind='hist', stacked=True, bins=25)
+        ax=e_axs[i][j], kind='hist', stacked=True, bins=25)
     p.grid(b=None)
+
+    norm = mpl.colors.Normalize(group['kappa'].min(), group['kappa'].max())
+    m = mpl.cm.ScalarMappable(norm=norm, cmap=mpl.cm.winter)
+
+    for (idx, _) in group.iterrows():
+        kappa = group['kappa'][idx]
+        sub_d = access[access['tangle'] == idx]
[…]

### 2015-03-18 scripts/plot-access.py
SConstruct niceties.

@@ -28,7 +28,9 @@ parser.add_argument('--prefix', required=True,
 
 args = parser.parse_args()
 for p in [args.access, args.tangles, args.degrees, args.curvatures]:
-    assert(os.path.exists(p))
+    if not os.path.exists(p):
+        error = "Can't find "+p+'.'
+        raise ValueError(error)
 
 access = pd.read_table(gzip.open(args.access),
                         names=['tangle', 'time', 'count'], header=False)

### 2015-03-18 scripts/plot-access.py
Tidier plots.

@@ -60,6 +60,9 @@ def subplots():
 e_fig, e_axs = subplots()
 d_fig, d_axs = subplots()
 
+cmaps = map(mpl.cm.get_cmap, ['spring', 'summer', 'autumn', 'winter', 'cool',
+                              'copper', 'gray', 'hot', 'pink'])
+
 for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist']):
     i = np.where(t1_degs == t1_deg)[0][0]
     j = np.where(t2_degs == t2_deg)[0][0]
@@ -68,12 +71,15 @@ for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist'])
     p.grid(b=None)
 
     norm = mpl.colors.Normalize(group['kappa'].min(), group['kappa'].max())
-    m = mpl.cm.ScalarMappable(norm=norm, cmap=mpl.cm.winter)
+    m = mpl.cm.ScalarMappable(norm=norm, cmap=cmaps[dist])
 
     for (idx, _) in group.iterrows():
         kappa = group['kappa'][idx]
+        if np.isnan(kappa):
+            continue
         sub_d = access[access['tangle'] == idx]
-        d_axs[i][j].plot(sub_d['time'], sub_d['count'], color=m.cmap(kappa), label=kappa)
+        d_axs[i][j].plot(sub_d['time'], sub_d['count'],
+                         color=m.cmap(kappa), label=kappa)
         d_axs[i][j].legend()
 
 sns.despine()

### 2015-03-19 scripts/plot-access.py
More deuglifying.

@@ -25,6 +25,10 @@ parser.add_argument('degrees', type=str, help='Degrees file.')
 parser.add_argument('curvatures', type=str, help='Tangle curvatures.')
 parser.add_argument('--prefix', required=True,
                     type=str, help='Prefix of plot paths for saving.')
+parser.add_argument('--maxtime',
+                    type=int, help='Maximum time for distribution plotting.')
+parser.add_argument('--dist',
+                    type=int, help='Restrict to this distance.')
 
 args = parser.parse_args()
 for p in [args.access, args.tangles, args.degrees, args.curvatures]:
@@ -64,12 +68,18 @@ cmaps = map(mpl.cm.get_cmap, ['spring', 'summer', 'autumn', 'winter', 'cool',
                               'copper', 'gray', 'hot', 'pink'])
 
 for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist']):
+    if args.dist and dist != args.dist:
+        continue
+    title_str = 'deg1 = {},  deg2 = {}'.format(t1_deg, t2_deg)
+
     i = np.where(t1_degs == t1_deg)[0][0]
     j = np.where(t2_degs == t2_deg)[0][0]
+    e_axs[i][j].set_title(title_str, weight=500)
     p = group['mean_access'].plot(
         ax=e_axs[i][j], kind='hist', stacked=True, bins=25)
     p.grid(b=None)
 
+    d_axs[i][j].set_title(title_str, weight=500)
     norm = mpl.colors.Normalize(group['kappa'].min(), group['kappa'].max())
     m = mpl.cm.ScalarMappable(norm=norm, cmap=cmaps[dist])
 
@@ -78,12 +88,14 @@ for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist'])
         if np.isnan(kappa):
             continue
         sub_d = access[access['tangle'] == idx]
+        if args.maxtime:
+            sub_d = sub_d[sub_d['time'] < args.maxtime]
         d_axs[i][j].plot(sub_d['time'], sub_d['count'],
                          color=m.cmap(kappa), label=kappa)
-        d_axs[i][j].legend()
+        # d_axs[i][j].legend()
 
-sns.despine()
-# sns.despine(offset=10, trim=True)
+sns.despine(offset=4, trim=True)
+plt.tight_layout()
 
 e_fig.savefig(args.prefix+'-expected.svg')
 d_fig.savefig(args.prefix+'-distribution.svg')

### 2015-03-19 scripts/plot-access.py
Using env.Command.

@@ -74,12 +74,12 @@ for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist'])
 
     i = np.where(t1_degs == t1_deg)[0][0]
     j = np.where(t2_degs == t2_deg)[0][0]
-    e_axs[i][j].set_title(title_str, weight=500)
+    e_axs[i][j].set_title(title_str, weight=600)
     p = group['mean_access'].plot(
         ax=e_axs[i][j], kind='hist', stacked=True, bins=25)
     p.grid(b=None)
 
-    d_axs[i][j].set_title(title_str, weight=500)
+    d_axs[i][j].set_title(title_str, weight=600)
     norm = mpl.colors.Normalize(group['kappa'].min(), group['kappa'].max())
     m = mpl.cm.ScalarMappable(norm=norm, cmap=cmaps[dist])
 

### 2015-03-19 curvature.tex
title page, refs on new page

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm}
+\usepackage[foot]{amsaddr}
 \usepackage{ifthen}
@@
 \usepackage{url}
+%\usepackage[margin=1in]{geometry}
 
@@
 
+% uncomment to remove the proofs for length estimates
+%\usepackage{environ}
+%\NewEnviron{killcontents}{}
+%\let\proof\killcontents
+%\let\endproof\endkillcontents
+
 \begin{document}
 \title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
-\author[Whidden]{Christopher Whidden}
-\author[Matsen]{Frederick A Matsen IV}
-\address{}
+\author[Whidden and Matsen]{Christopher Whidden$^1$ and Frederick A Matsen IV$^1$}
+\address{$^1$Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, 98199}
+\email{cwhidden@fredhutch.org}
+\email{matsen@fredhutch.org}
 \thanks{}
@@
 \maketitle
+\pagebreak
 
@@
 To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
-The uniform random walk moves to a neighboring vertex in the graph uniformly at random.
+The uniform random walk moves from the current vertex to one of its neighbors uniformly at random.
 However, this makes the walk more likely to sample higher degree vertices.
@@
 To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
-A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
+A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree---and then one could be picked uniformly at random.
 To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
@@
 	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
-	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in array indexed by preorder number.
+	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
 	Both of these traversals require $\OhOf{n}$-time.
@@
 
+\pagebreak
+
 \bibliographystyle{IEEEtran}

### 2015-03-19 curvature.tex
preliminaries

@@
 The SPR graph is very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
-\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rSPR graph, and corresponding bounds on degree.
+\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
 \cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
@@
 \section{Preliminaries}
-\begin{equation}
-\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
-\label{eq:curvatureDef}
-\end{equation}
 
-Rooted $X$-trees.
-Leaf restriction notation with $|$.
-We are talking about rooted, and would like to make it clear that this is not the most realistic graph to consider.
-Our motivations for considering this graph are because it is understandable, and some programs do rSPR-type moves.
-Note that in rSPR you can move subtrees to the root.
-Parent and parent subtree.
-Define adjacency of subtrees.
+We follow the definitions and notation from~\cite{bordewich05,whidden2013hybridization, Whidden2015-yi}.
+A (rooted binary phylogenetic) $X$-tree is a rooted tree $T$ whose nodes have zero or two children.
+Leaves of $T$ are bijectively labelled with the members of a label set $X$.
+As in~\cite{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$.
+We generally use $n$ to refer to the number of leaves in an $X$-tree.
+For a subset $V$ of $X$, $T(V)$ is the smallest subtree of $T$ that connects all nodes in $V$.
+The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children.
+For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
+\todo{Figure of restriction and SPR operation}
 
-For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
 A \emph{parent (sub)tree} of a subtree $U$ is the smallest subtree strictly containing $U$.
@@
 The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
+\todo{Generally we add a $\rho$ label and root edge to make rSPR work formally.
+I've put this in the definition so lets see if this confuses anything.}
 A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
@@
 
+A \emph{subtree prune-and-regraft} (SPR) operation on an $X$-tree $T$ cuts an edge $e = (x, p_x)$ where $p_x$ denotes the parent of node $x$.
+$T$ is divided into two subtrees $T_x$ and $T_{p_x}$ containing $x$ and $p_x$, respectively.
+Then the operation adds a new node $p'_x$ to $T_{p_x}$ by subdividing an edge of $T_{p_x}$ and adding a new edge $(x, p'_x)$, making $x$ a child of $p'_x$.
+Finally, $p_x$ is suppressed. \todo{figure}
+Observe that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
+
+SPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of SPR operations required to transform an $X$-tree $T_1$ into $T_2$. \todo{ref figure}
+Moreover, SPR operations naturally give rise to a graph on the set of $X$-trees.
+Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
+Then the SPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
+Observe that distances in this graph are given by $\dspr{\cdot,\cdot}$.
+
 For clarity, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
-We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices. \todo{Not sure if this is better here or in the preliminaries. Revisit this later}
-
 Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
 We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
-Note that this concept of degree is unrelated to the degree of vertices of $T$.
+We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices.
+
+Ollivier-Ricci curvature provides an intuitive formalization for quantifying the difficulty of moving between two points of a graph during a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
+Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
+The $L^1$ transportation, or Wasserstein distance between $m_x$ and $m_y$ is the minimum amount of "work" required to move $m_x$ to $m_y$ following the graph, that is
+ $$ W_1(m_x, m_y) = \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
+where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of trees in $V \times V$. \todo{I tried to keep this simple, feel free to change.}
+
+The curvature of $x$ and $y$ is then defined as:
+\begin{equation}
+\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
+\label{eq:curvatureDef}
+\end{equation}
+Positive curvature implies that $m_x$ and $m_y$ are "closer" than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are actually more distant than $x$ and $y$.
+Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
 
@@
 We computed the rSPR treespaces for four to seven leaves.
-This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case).
+This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
 Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
-In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast,Whidden2015-yi}.
+In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
-However, applying this method to complete SPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, would require an enormous amount of computation.
+However, applying this method to complete SPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, requires an enormous amount of computation.
 
-To quickly compute dense SPR graphs (those containing a significant portion of the full treespace) we avoid pairwise computations by successively adding trees to the graph:
+To quickly compute dense SPR graphs (those containing a significant portion of the full treespace) we can avoid pairwise computations by successively adding trees to the graph:
 \begin{enumerate}[label={\arabic*}.]

### 2015-03-20 curvature.tex
IEEEtran style. Shifted margins left to allow for notes on the right side

@@
-\documentclass{amsart}
+\documentclass[11pt,onecolumn,conference]{IEEEtran}
 
 \usepackage{amsmath,amsthm}
-\usepackage[foot]{amsaddr}
 \usepackage{ifthen}
 \usepackage{todonotes}
-\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
+%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
+\usepackage[right]{showlabels}
 \usepackage{cite}
@@
 \usepackage{url}
+% shift for todonotes
+\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
+% regular margins
 %\usepackage[margin=1in]{geometry}
@@
 \begin{document}
-\title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
-\author[Whidden and Matsen]{Christopher Whidden$^1$ and Frederick A Matsen IV$^1$}
-\address{$^1$Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, 98199}
-\email{cwhidden@fredhutch.org}
-\email{matsen@fredhutch.org}
-\thanks{}
+\title{Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-prune-regraft Graph}
 
-\date{\today}
+\author{
+	\IEEEauthorblockN{Christopher Whidden}
+	\IEEEauthorblockA{Program in Computational Biology\\
+	Fred Hutchinson Cancer Research Center\\
+	Seattle, WA 98199\\
+	Email: cwhidden@fredhutch.org}
+	\and
+	\IEEEauthorblockN{Frederick A Matsen IV}
+	\IEEEauthorblockA{Program in Computational Biology\\
+	Fred Hutchinson Cancer Research Center\\
+	Seattle, WA 98199\\
+	Email: matsen@fredhutch.org}
+}
+
+\maketitle
+\IEEEpeerreviewmaketitle
 
@@
 
-\maketitle
 \pagebreak

### 2015-03-20 curvature.tex
Figures: X-trees, SPR operations, and SPR neighborhood

@@
 \usepackage{url}
+\usepackage{subfigure}
+\usepackage[nomarkers,nofiglist]{endfloat}
+\let\MYoriglatexcaption\caption
+\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
 % shift for todonotes
@@
 
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:x-tree}]{\includegraphics[scale=1.25]{figs/x-tree}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:subtree}]{\includegraphics[scale=1.25]{figs/subtree}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:induced}]{\includegraphics[scale=1.25]{figs/contracted-subtree}}
+	\hspace*{\stretch{1}}
+	\\
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:spr}]{\includegraphics[scale=1.25]{figs/spr}}
+	\hspace*{\stretch{1}}
+	\\
+	\hspace*{\stretch{1}}
+	\hspace*{\stretch{1}}
+
+	\caption{(a) An $X$-tree $T$.
+		(b) $T(V)$, where $V = \set{1,2,5}$.
+		(c) $T|V$.
+		(d) An SPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
+	}
+	\label{fig:trees}
+\end{figure}
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\includegraphics[scale=1.25]{figs/two-spr}
+	\hspace*{\stretch{1}}
+	\caption{Two trees differing by two SPR operations that move the grey subtrees.}
+	\label{fig:two-spr}
+\end{figure}
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\includegraphics[width=0.6\textwidth]{figs/neighborhood}
+	\hspace*{\stretch{1}}
+	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections between neighbors.}
+	\label{fig:neighborhood}
+\end{figure}
+
 We follow the definitions and notation from~\cite{bordewich05,whidden2013hybridization, Whidden2015-yi}.
@@
 Leaves of $T$ are bijectively labelled with the members of a label set $X$.
-As in~\cite{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$.
+As in~\cite{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$ (Fig.~\ref{fig:x-tree}).
 We generally use $n$ to refer to the number of leaves in an $X$-tree.
-For a subset $V$ of $X$, $T(V)$ is the smallest subtree of $T$ that connects all nodes in $V$.
-The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children.
+For a subset $V$ of $X$, $T(V)$ is the smallest subtree of $T$ that connects all nodes in $V$ (Fig.~\ref{fig:subtree}).
+The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children (Fig.~\ref{fig:induced}).
 For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
@@
 Then the operation adds a new node $p'_x$ to $T_{p_x}$ by subdividing an edge of $T_{p_x}$ and adding a new edge $(x, p'_x)$, making $x$ a child of $p'_x$.
-Finally, $p_x$ is suppressed. \todo{figure}
-Observe that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
+Finally, $p_x$ is suppressed.
+See Figure~\ref{fig:spr} for an example.
+Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
 
-SPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of SPR operations required to transform an $X$-tree $T_1$ into $T_2$. \todo{ref figure}
+SPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of SPR operations required to transform an $X$-tree $T_1$ into $T_2$.
+For example, the trees in Figure~\ref{fig:two-spr} are separated by two SPR operations.
 Moreover, SPR operations naturally give rise to a graph on the set of $X$-trees.
@@
 Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
-We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
+For example, the tree with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
 We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices.

### 2015-03-22 curvature.tex
Access times symmetries via tangles.

@@
 \begin{abstract}
-Statistical phylogenetics inference methods use tree rearrangement operations to perform either local hill-climbing search or Markov chain Monte Carlo across tree topologies.
+Statistical phylogenetics inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
 The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
-Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the graph.
+Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the SPR graph.
 Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
-In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
+In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
 This notion of curvature is especially relevant to stochastic search on the graph.
 By proving theorems and direct calculation, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
-This indicates significant structure of the graph, and a greater understanding of this structure could ultimately lead to improved strategies for tree search.
-
-We find that indeed there is
+This indicates significant structure of the graph; a greater understanding of this structure could ultimately lead to improved strategies for tree search.
 \end{abstract}
@@
 In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
-% Here we focus on the discrete search part.
-There are a number of means of doing so, most of which are based on \emph{splits}---partitions of leaves induced by edges in the trees---such as the Robinson Foulds distance~\cite{robinson81}.
-However, phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves~\cite{hein96} or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves~\cite{hein96} or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
@@
 We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then used linear programming to compute curvatures for given pairs of trees~\cite{gricci}.
+We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK \todo{GLPK} solver; code can be found in \cite{gricci}.
+
 This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} (enumeration methods described in a manuscript in preparation).
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found in \cite{tangle}.
 
@@
 
-\subsection{Mean access time computation}
+\cite{GAP4}
 
-We then computed mean access times (MAT) between pairs of trees in the MH random walks---the mean number of iterations required to move from one tree to the other~\cite{lovasz1993random}.
-MAT values have been productively applied to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi}.
-We used the same \texttt{sprspace}~\cite{sprspace} software to compute MAT values for our simulated random walks.\todo{runtime information}
+\subsection{Access time computation}
+
+The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
+In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks---the mean number of iterations required to move from one tree to the other.
+We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
+
+Here, to gain more insight, we used simulation to approximate the entire access time distribution.
+As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
+Here again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
+Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
+For example, because the pairs
+$\{(((1,2),3),4) ((1,2),(3,4))\}$ and
+$\{(((1,4),3),4) ((1,4),(2,3))\}$
+are the same after relabeling, their access times will have identical distributions.
+Thus to calculate the empirical distributions of access times we aggregate all access times for the same tangle.
 

### 2015-03-23 curvature.tex
endfloat apparently doesn't like figures defined in commands. Removing a spurious <<<<<<< HEAD

@@
 
+%---[ figures ]-----------------------------------------------------------------
+
+%\newcommand{\FIGspr}{\
+%\begin{figure}
+%\begin{center}
+%  \includegraphics[width=4in]{figures/spr.pdf}
+%\end{center}
+%\caption{\
+%  A rooted subtree-prune-regraft (rSPR) move~\cite{hein96}.
+%}
+%\label{FIGspr}
+%\end{figure}
+%}
+
 % uncomment to remove the proofs for length estimates
@@
 In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
-Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves~\cite{hein96} or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves (Figure~\ref{FIGspr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
+
+%\FIGspr
+\begin{figure}
+\begin{center}
+  \includegraphics[width=4in]{figures/spr.pdf}
+\end{center}
+\caption{\
+  A rooted subtree-prune-regraft (rSPR) move~\cite{hein96}.
+}
+\label{FIGspr}
+\end{figure}
 
@@
 
-The SPR graph is very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
+The SPR graph is thus very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
 \cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
@@
 They define a notion of curvature on a discrete or continuous case.
-This formalizes the notion of to what extent random walking brings points together, and is a generalization of the notion of Ricci curvature on manifolds.
+This formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
+Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$.
 
-In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and make connections to access times for those random walks.
-We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC and ML searches.
+In this paper, we investigate curvature of the rSPR graph with respect to two random walks and make connections to access times for those random walks.
+We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC and ML searches \todo{Not sure what random walks have to do with ML searches}.
 For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
 We present a fast new algorithm for computing SPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
-As the the full rSPR treespace contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of treespace.
+As the full rSPR treespace contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of treespace.
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
@@
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
-In particular, we present a full characterization of the change in SPR degree that occurs from a given SPR move and find that, contrary to previous assumptions, SPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+In particular, we present a full characterization of the change in SPR degree that occurs from a given SPR move and find that, contrary to previous assumptions\todo{This is going to need a citation or be taken out}, SPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
 
@@
 		(d) An SPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
-	}
 	\label{fig:trees}
 \end{figure}
+
 \begin{figure}
@@
 \end{figure}
+
 \begin{figure}
@@
 I've put this in the definition so lets see if this confuses anything.}
-A \emph{ladder tree} is a tree such that every internal node has a leaf as a direct descendant.
+A \emph{ladder tree} (also known as a \emph{caterpillar tree}) is a tree such that every internal node has a leaf as a direct descendant.
 A \emph{balanced tree} is a tree such that the sum of the depths of internal nodes is minimum over all trees with the same number of leaves.
@@
 See Figure~\ref{fig:spr} for an example.
-Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
+Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.\todo{Erm? I thought that we weren't going to have a root edge.}
 
@@
 Moreover, SPR operations naturally give rise to a graph on the set of $X$-trees.
-Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
+Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.\todo{$\rho$ again?}
 Then the SPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
@@
 
-For clarity, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
+To avoid confusion between the two types of graph structures considered here, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
 Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
 For example, the tree with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
-We assume that all trees are bifurcating, and use degree to refer only to the degree of SPR graph vertices.
+We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
+Because we assume that all trees are bifurcating, we use degree to refer only to the degree of SPR graph vertices.
 
-Ollivier-Ricci curvature provides an intuitive formalization for quantifying the difficulty of moving between two points of a graph during a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
+Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The $L^1$ transportation, or Wasserstein distance between $m_x$ and $m_y$ is the minimum amount of "work" required to move $m_x$ to $m_y$ following the graph, that is
- $$ W_1(m_x, m_y) = \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
-where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of trees in $V \times V$. \todo{I tried to keep this simple, feel free to change.}
+The $L^1$ transportation, or Wasserstein, distance between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
+ $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
+where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$. \todo{FAM: cites.}
 
@@
 \end{equation}
-Positive curvature implies that $m_x$ and $m_y$ are "closer" than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are actually more distant than $x$ and $y$.
+Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average more distant than $x$ and $y$.
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
@@
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found in \cite{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
 
@@
 For example, because the pairs
-$\{(((1,2),3),4) ((1,2),(3,4))\}$ and
-$\{(((1,4),3),4) ((1,4),(2,3))\}$
+$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
+$\{(((1,4),3),4), ((1,4),(2,3))\}$
 are the same after relabeling, their access times will have identical distributions.
-Thus to calculate the empirical distributions of access times we aggregate all access times for the same tangle.
+Thus to calculate the empirical distributions of access times we aggregate all access times for the same tangle \todo{cite software}.
+
+The full analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool; instructions are in the repository README file.
 

### 2015-03-23 curvature.tex
Illustration of SPR degree change lemma

@@
 To do so, we require the following lemma that characterizes exactly how the degree of a tree changes after an rSPR operation.
+See Figure~\ref{fig:spr-degree-change} for an illustration of the Lemma.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\includegraphics[width=\textwidth]{figs/spr-degree-change}
+	\hspace*{\stretch{1}}
+	\caption{An SPR move labelled as in Lemma~\ref{lem:degree_change}.
+Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the SPR degree by $2\left(k(a-b) + i - j\right)$.}
+	\label{fig:spr-degree-change}
+\end{figure}
 

### 2015-03-24 curvature.tex
missing citations and other minor todos

@@
 
-%\newcommand{\FIGspr}{\
-%\begin{figure}
-%\begin{center}
-%  \includegraphics[width=4in]{figures/spr.pdf}
-%\end{center}
-%\caption{\
-%  A rooted subtree-prune-regraft (rSPR) move~\cite{hein96}.
-%}
-%\label{FIGspr}
-%\end{figure}
-%}
-
 % uncomment to remove the proofs for length estimates
@@
 This notion of curvature is especially relevant to stochastic search on the graph.
-By proving theorems and direct calculation, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
+By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
 This indicates significant structure of the graph; a greater understanding of this structure could ultimately lead to improved strategies for tree search.
@@
 In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
-Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves (Figure~\ref{FIGspr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
-
-%\FIGspr
-\begin{figure}
-\begin{center}
-  \includegraphics[width=4in]{figures/spr.pdf}
-\end{center}
-\caption{\
-  A rooted subtree-prune-regraft (rSPR) move~\cite{hein96}.
-}
-\label{FIGspr}
-\end{figure}
 
@@
 In this paper, we investigate curvature of the rSPR graph with respect to two random walks and make connections to access times for those random walks.
-We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC and ML searches \todo{Not sure what random walks have to do with ML searches}.
+We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC searches.
 For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
@@
 For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
-\todo{Figure of restriction and SPR operation}
 
@@
 A \emph{parent edge} of a subtree $U$ is the edge connecting $U$ to the rest of the tree.
-The \emph{internal edges} of a tree are the edges that do not contact a leaf (we do not consider a rooted tree to have a ``root edge'', so trees have a degree two root).
-\todo{Generally we add a $\rho$ label and root edge to make rSPR work formally.
-I've put this in the definition so lets see if this confuses anything.}
+The \emph{internal edges} of a tree are the edges that do not contact a leaf or $\rho$.
 A \emph{ladder tree} (also known as a \emph{caterpillar tree}) is a tree such that every internal node has a leaf as a direct descendant.
@@
 See Figure~\ref{fig:spr} for an example.
-Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.\todo{Erm? I thought that we weren't going to have a root edge.}
+Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
 
@@
 Moreover, SPR operations naturally give rise to a graph on the set of $X$-trees.
-Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.\todo{$\rho$ again?}
+Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
 Then the SPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
@@
 Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
-For example, the tree with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
+For example, the tree $T$ with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
 We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
@@
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The $L^1$ transportation, or Wasserstein, distance between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
+The earth movers distance~\cite{rubner2000earth} (equivalently $L^1$ transportation, or Wasserstein distance~\cite{wasserstein}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
  $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
-where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$. \todo{FAM: cites.}
+where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.
 
@@
 We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK \todo{GLPK} solver; code can be found in \cite{gricci}.
+We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci}.
 
@@
 
-After the chain has achieved stationarity, one sees a
+After the chain has achieved stationarity, one sees a \ldots~\todo{results}.
 

### 2015-03-25 curvature.tex
Some reasonable kappa-access plots.

@@
 
-%---[ figures ]-----------------------------------------------------------------
-
 % uncomment to remove the proofs for length estimates
@@
 \begin{document}
-\title{Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-prune-regraft Graph}
+\title{Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph}
 
@@
 
-After the chain has achieved stationarity, one sees a \ldots~\todo{results}.
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
+	\hspace*{\stretch{1}}
+    \caption{Access times for degree 24 trees and $\kappa$.}
+	\label{fig:kappa-access}
+\end{figure}
 

### 2015-03-25 scripts/plot.py
More plot tidying.

@@ -51,17 +51,20 @@ elif args.which == 'scatter':
             y='dist_jitter',
             alpha=0.4)
     else:
+        # Easy way to re-label colorbar. :)
+        df['average degree'] = df['avg_deg']
         p = df.plot(
             kind='scatter',
             x='kappa_R',
             y='dist_jitter',
             alpha=0.4,
-            c='avg_deg',
+            c='average degree',
             cmap=plt.get_cmap('Blues'))
 
 p.grid(b=None)
-sns.despine()
-# sns.despine(offset=10, trim=True)
+sns.despine(offset=10)
+p.set_xlabel('kappa')
+p.set_ylabel('distance (jittered)')
 
 if args.d3:
     from mpld3 import fig_to_html, plugins

### 2015-03-27 curvature.tex
minor tweaks

@@
 Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
-Recall that a tree with $n$ leaves has $2n-1$ total nodes.
+Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
 Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
@@
 \end{enumerate}
-We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
+We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
 Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
@@
 		2n - 1 - x - 2 &\text{if depth($u$) $ = 1$ } \\
-		0 &\text{if depth($u$) $ = 0$ } \\
+		0 &\text{if depth($u$) $ \le 0$ } \\
 	\end{cases}.$$
@@
 	Call these $U$ and $V$, respectively.
-	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$.
-	Then the SPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is SPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an SPR operation.
+	Then the SPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, SPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
 	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
@@
 	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
-	Moreover, the tree $S$ can be found by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding SPR destination.
+	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding SPR destination.
 	Finally, we require linear time to apply the chosen SPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
@@
 
+Having explored curvature values empirically for SPR graphs with 3--7 leaves, we now consider them theoretically.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 
-Next we obtain a tighter bound on the maximum curvature of two adjacent trees.
+We can obtain a tighter bound on the maximum curvature of two adjacent trees.
 \begin{lemma}
@@
 This definition of curvature is invariant of $p$ for small enough $p$ and was used by Ollivier to avoid parity problems on graphs where the uniform random walk is periodic.
-As we now prove, the curvature of the lazy random walk is closely related to that of the uniform random walk.
+As we now prove, the curvature of the lazy random walk under this definition is closely related to that of the uniform random walk.
 

### 2015-03-27 curvature.tex
Happy with plotting and regression.

@@
 \usepackage{amsmath,amsthm}
+\usepackage{csvsimple}
 \usepackage{ifthen}
@@
 
+Now
+
 \begin{figure}
@@
 \end{figure}
+
+\begin{table}
+\centering
+\caption{Regression showing factors determining mean access time for rSPR.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
+\label{TABregressionMAT}
+\end{table}
+
+\begin{table}
+\centering
+\caption{Regression showing factors determining $\delta_1$ for rSPR.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
+\label{TABregressionDelta}
+\end{table}
 

### 2015-03-27 scripts/plot-access.py

@@ -75,8 +75,14 @@ for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist'])
     i = np.where(t1_degs == t1_deg)[0][0]
     j = np.where(t2_degs == t2_deg)[0][0]
     e_axs[i][j].set_title(title_str, weight=600)
-    p = group['mean_access'].plot(
-        ax=e_axs[i][j], kind='hist', stacked=True, bins=25)
+
+    binwidth=1
+    data = group['mean_access']
+    bins = 10
+#     if len(data) > 1:
+#         bins = np.arange(min(data), max(data) + binwidth, binwidth)
+    p = data.plot(
+        ax=e_axs[i][j], kind='hist', stacked=True, bins=bins)
     p.grid(b=None)
 
     d_axs[i][j].set_title(title_str, weight=600)

### 2015-03-27 curvature.tex
scatter svgs

@@
 	Fred Hutchinson Cancer Research Center\\
-	Seattle, WA 98199\\
+	Seattle, WA 98109\\
 	Email: cwhidden@fredhutch.org}
@@
 	Fred Hutchinson Cancer Research Center\\
-	Seattle, WA 98199\\
+	Seattle, WA 98109\\
 	Email: matsen@fredhutch.org}
@@
 \begin{abstract}
-Statistical phylogenetics inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
+Statistical phylogenetic inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
 The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
@@
 By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
-This indicates significant structure of the graph; a greater understanding of this structure could ultimately lead to improved strategies for tree search.
+We confirm using simulation that degree and curvature have the expected impact on mean access time distributions.
+This indicates significant structure of the graph beyond that which which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
 \end{abstract}

### 2015-03-27 prefigs/get-figs.sh

@@ -1,0 +1,5 @@
+for graph in nni rspr; do
+    for i in ../analysis/6-walks-50K/$graph/upmh/scatter*svg; do
+        cp $i $graph-$(basename $i)
+    done
+done

### 2015-03-27 prefigs/get-figs.sh
Better plot names.

@@ -1,5 +1,5 @@
 for graph in nni rspr; do
     for i in ../analysis/6-walks-50K/$graph/upmh/scatter*svg; do
-        cp $i $graph-$(basename $i)
+        cp $i $graph-upmh-$(basename $i)
     done
 done

### 2015-03-27 curvature.tex
Adding scatterplots of curvature.

@@
 \newattribute{dspr}{d_{\mathrm{SPR}}}
+\newattribute{dnni}{d_{\mathrm{NNI}}}
 \newparentheses{degree}{|N(}{)|}
@@
 \subsection{Computing curvature values}
-We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
-We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci}.
@@
 \subsection{Access time computation}
-
 The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
@@
 
+\subsection{Computational investigation}
+We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work as described before. \todo{Update once 7 taxon runs have finished}.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
+	\hspace*{\stretch{2}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{1}}
+    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+	\label{fig:rspr-scatter}
+\end{figure}
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter5}}
+	\hspace*{\stretch{2}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter6}}
+	\hspace*{\stretch{1}}
+    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dnni{T_1,T_2}$ for the NNI graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+	\label{fig:nni-scatter}
+\end{figure}
+
 \section{Conclusion and future work}

### 2015-03-28 curvature.tex
A full draft! (I think)

@@
 %\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
-\usepackage[right]{showlabels}
+%\usepackage[right]{showlabels}
 \usepackage{cite}
@@
 \pagebreak
-
-% % Make a todo table of contents.
-% \makeatletter
-% \providecommand\@dotsep{5}
-% \makeatother
-% \listoftodos\relax
 
@@
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The earth movers distance~\cite{rubner2000earth} (equivalently $L^1$ transportation, or Wasserstein distance~\cite{wasserstein}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
+The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance'') between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
  $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
@@
 
-\subsection{Access time computation}
+\subsection{Access time simulation}
 The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
@@
 are the same after relabeling, their access times will have identical distributions.
-Thus to calculate the empirical distributions of access times we aggregate all access times for the same tangle \todo{cite software}.
+Thus to calculate the empirical distributions of access times we aggregate all access times for the same tanglegram \cite{gricci, tangle}.
 
-The full analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool; instructions are in the repository README file.
+\begin{table}
+\centering
+\caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
+\label{tab:regressionMAT}
+\end{table}
 
-\subsection{Results}
-
-Now
+We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\deg(T_1)$ and $\deg(T_2)$ (Table~\ref{tab:regressionMAT}).
+Furthermore, by plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
+We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
+For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access}) is the count for time 2 minus the count for time 3.
+Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
+This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the likelihood of a random walk going from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
+On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
 
@@
 	\hspace*{\stretch{1}}
-    \caption{Access times for degree 24 trees and $\kappa$.}
+    \caption{\
+        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.\
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows curvature $\kappa(\MH; \cdot, \cdot)$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        }
 	\label{fig:kappa-access}
@@
 \centering
-\caption{Regression showing factors determining mean access time for rSPR.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
-\label{TABregressionMAT}
+\caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
+\label{tab:regressionDelta}
 \end{table}
 
-\begin{table}
-\centering
-\caption{Regression showing factors determining $\delta_1$ for rSPR.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
-\label{TABregressionDelta}
-\end{table}
+The full analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool; instructions are in the repository README file.
 
 \section{Rooted SPR Neighborhoods}
-
-Having explored curvature values empirically for SPR graphs with 3--7 leaves, we now consider them theoretically.
+Having made the connection between curvature values and access times on the SPR graphs, we now consider them theoretically.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
-
 \end{proof}
 
-\subsection{Computational investigation}
-We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work as described before. \todo{Update once 7 taxon runs have finished}.
+\subsection{Computational investigation and validation}
+We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work with direct calculation as described before.
+\todo{Update once 7 taxon runs have finished}
+
+Indeed we see that for rSPR ... (Figure~\ref{fig:rspr-scatter}). \todo{Chris: can you fill this in?}
 
@@
 
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter5}}
-	\hspace*{\stretch{2}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter6}}
-	\hspace*{\stretch{1}}
-    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dnni{T_1,T_2}$ for the NNI graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
-	\label{fig:nni-scatter}
-\end{figure}
+\section{Conclusion and future work}
+In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
+We done so by developing theoretical and computational frameworks to compute Ricci-Ollivier curvature of the rSPR graph.
+We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
 
-\section{Conclusion and future work}
-This isn't quite the right graph to be looking at (rooted).
+We emphasize that the influence of curvature is small in this setting of equiprobable trees (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution is the predominant factor determining hitting times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets with hundreds or thousands of trees~\cite{Whidden2015-yi}.
+Thus, our next step will be to investigate curvature of MCMC with a nontrivial likelihood function, which will significantly reduce the effective size of the posterior distribution, and may lead to ``bottlenecks'' like those we have observed in real data.
+We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
 
-Want to investigate curvature of MCMC steps with a nontrivial likelihood function.
-MCMC does both branch length and topology changes, thus this isn't the real thing.
-Nevertheless, graph effects still matter.
-\cite{Stefankovic2011-hu} take a chain where the likelihoods are the likelihoods maximized across branch length.
-
-Ranked trees~\cite{Song2006-xe}
-Partitions~\cite{Gusfield2002-il}
+In addition, we will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
+We will also explore random walks ranked trees~\cite{Song2006-xe} and models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
+Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.
 
 \section{Acknowledgements}
-Alex Gavruskin,
-Vladimir Minin,
-
-SAGE and GAP4 software
-Alexander Hulpke
-
+The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
+They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
 This work was funded by National Science Foundation award 1223057.

### 2015-03-28 curvature.tex
curvature distribution observations and a pass through

@@
 \usepackage{subfigure}
-\usepackage[nomarkers,nofiglist]{endfloat}
+\usepackage[nomarkers,nofiglist,notablist]{endfloat}
 \let\MYoriglatexcaption\caption
@@
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance'') between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
+The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
  $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
@@
 We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
+
+We saw a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
+Curvature values tended to increase with increasing SPR distance.
+In addition, their variance decreased with increasing distance.
+Minimum curvature values for a given treespace were observed between neighboring trees, while maximum curvature values were observed between trees at maximum distance or one rSPR move closer than the maximum.
+This suggests that the increased difficulty of moving between trees in a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR treespaces.
+
+Larger treespaces tended to have smaller curvature values.
+Indeed, the 7-leaf treespace contained adjacent pairs of trees with negative curvature.
+Such pairs are possible bottlenecks in phylogenetic searches which may be exarcebated by likelihood or branch length constraints.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
+	\hspace*{\stretch{2}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{1}}
+    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+	\label{fig:rspr-scatter}
+\end{figure}
 
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}. 
-We sampled \ldots \todo{sampling and runtime information}.
-
-\cite{GAP4}
+We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
+This required XXX and YYY time respectively.~\todo{EM: can you fill in the time this required?}
 
@@
 \section{Rooted SPR Neighborhoods}
-Having made the connection between curvature values and access times on the SPR graphs, we now consider them theoretically.
+Having made the connection between curvature values and access times on the SPR graphs, we now consider curvature theoretically.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 
-\subsection{Computational investigation and validation}
-We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work with direct calculation as described before.
-\todo{Update once 7 taxon runs have finished}
-
-Indeed we see that for rSPR ... (Figure~\ref{fig:rspr-scatter}). \todo{Chris: can you fill this in?}
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
-	\hspace*{\stretch{2}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
-	\hspace*{\stretch{1}}
-    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
-	\label{fig:rspr-scatter}
-\end{figure}
-
 \section{Conclusion and future work}
 In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
-We done so by developing theoretical and computational frameworks to compute Ricci-Ollivier curvature of the rSPR graph.
+We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
 We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree and its change in depth and that rSPR treespace tends toward flatness with respect to SPR moves that move small subtrees.
 

### 2015-03-28 curvature.tex
move tanglegram example to first usage

@@
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
+For example, because the pairs
+$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
+$\{(((1,4),3),4), ((1,4),(2,3))\}$
+are the same after relabeling, their curvatures are the same.
 We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
@@
 Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
-For example, because the pairs
-$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
-$\{(((1,4),3),4), ((1,4),(2,3))\}$
-are the same after relabeling, their access times will have identical distributions.
-Thus to calculate the empirical distributions of access times we aggregate all access times for the same tanglegram \cite{gricci, tangle}.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram \cite{gricci, tangle}.
 

### 2015-03-28 curvature.tex
use rSPR consistently

@@
 For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
-We present a fast new algorithm for computing SPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
+We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
 As the full rSPR treespace contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of treespace.
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
-By carefully examining the overlap in SPR moves, we present a new method for computing the degree of a tree in the SPR graph that allows one to select an SPR neighbor uniformly at random in linear-time without explicitly generating the graph.
+By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
 Using this method to simulate these random walks, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
-In particular, we present a full characterization of the change in SPR degree that occurs from a given SPR move and find that, contrary to previous assumptions\todo{This is going to need a citation or be taken out}, SPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that, contrary to previous assumptions\todo{This is going to need a citation or be taken out}, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
 
@@
 		(c) $T|V$.
-		(d) An SPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
+		(d) An rSPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
 	\label{fig:trees}
@@
 	\hspace*{\stretch{1}}
-	\caption{Two trees differing by two SPR operations that move the grey subtrees.}
+	\caption{Two trees differing by two rSPR operations that move the grey subtrees.}
 	\label{fig:two-spr}
@@
 
-A \emph{subtree prune-and-regraft} (SPR) operation on an $X$-tree $T$ cuts an edge $e = (x, p_x)$ where $p_x$ denotes the parent of node $x$.
+A (rooted) \emph{subtree prune-and-regraft} (rSPR) operation on an $X$-tree $T$ cuts an edge $e = (x, p_x)$ where $p_x$ denotes the parent of node $x$.
 $T$ is divided into two subtrees $T_x$ and $T_{p_x}$ containing $x$ and $p_x$, respectively.
@@
 See Figure~\ref{fig:spr} for an example.
-Note that the inclusion of $\rho$ allows for SPR moves that move subtrees to the root of the tree.
+Note that the inclusion of $\rho$ allows for rSPR moves that move subtrees to the root of the tree.
 
-SPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of SPR operations required to transform an $X$-tree $T_1$ into $T_2$.
-For example, the trees in Figure~\ref{fig:two-spr} are separated by two SPR operations.
-Moreover, SPR operations naturally give rise to a graph on the set of $X$-trees.
+rSPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
+For example, the trees in Figure~\ref{fig:two-spr} are separated by two rSPR operations.
+Moreover, rSPR operations naturally give rise to a graph on the set of $X$-trees.
 Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
-Then the SPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
+Then the rSPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
 Observe that distances in this graph are given by $\dspr{\cdot,\cdot}$.
 
-To avoid confusion between the two types of graph structures considered here, we refer to vertices of the SPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
-Let $N(T)$ be the set of SPR neighbors of a tree $T$ (this does not include $T$).
+To avoid confusion between the two types of graph structures considered here, we refer to vertices of the rSPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
+Let $N(T)$ be the set of rSPR neighbors of a tree $T$ (this does not include $T$).
 For example, the tree $T$ with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
-We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single SPR operation.
-Because we assume that all trees are bifurcating, we use degree to refer only to the degree of SPR graph vertices.
+We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single rSPR operation.
+Because we assume that all trees are bifurcating, we use degree to refer only to the degree of rSPR graph vertices.
 
@@
 We computed the rSPR treespaces for four to seven leaves.
-This required a fast new algorithm for SPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
+This required a fast new algorithm for rSPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
 Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
-In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the SPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
+In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
-However, applying this method to complete SPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, requires an enormous amount of computation.
+However, applying this method to complete rSPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, requires an enormous amount of computation.
 
-To quickly compute dense SPR graphs (those containing a significant portion of the full treespace) we can avoid pairwise computations by successively adding trees to the graph:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full treespace) we can avoid pairwise computations by successively adding trees to the graph:
 \begin{enumerate}[label={\arabic*}.]
@@
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-For example, because the pairs
+For example, the pairs
 $\{(((1,2),3),4), ((1,2),(3,4))\}$ and
 $\{(((1,4),3),4), ((1,4),(2,3))\}$
-are the same after relabeling, their curvatures are the same.
+are the same after relabeling, so their curvatures are the same.
 We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
@@
 We saw a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
-Curvature values tended to increase with increasing SPR distance.
+Curvature values tended to increase with increasing rSPR distance.
 In addition, their variance decreased with increasing distance.
-Minimum curvature values for a given treespace were observed between neighboring trees, while maximum curvature values were observed between trees at maximum distance or one rSPR move closer than the maximum.
+Neighboring trees had minimum curvature values for a given treespace, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
 This suggests that the increased difficulty of moving between trees in a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR treespaces.
@@
 
-To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
+To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing rSPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
 A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree---and then one could be picked uniformly at random.
-To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
+To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of rSPR moves with a one-to-one correspondence to rSPR neighbors, as described in the next paragraph.
 Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
-We note that the recursive formula of Song~\cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
+We note that the recursive formula of Song~\cite{Song2003-gf} for the degree of a tree does not group rSPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific rSPR neighbor by index.
 
-Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
+Instead, we consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
 Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
 Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
-However, some of these moves will result in the same neighboring tree as other SPR moves.
+However, some of these moves will result in the same neighboring tree as other rSPR moves.
 In particular, where a node's edge means the edge connecting the subtree rooted at that node to the rest of the tree:
@@
 Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
-In this manner, we achieve an alternative solution for computing the degree of a tree in the SPR graph:
+In this manner, we achieve an alternative solution for computing the degree of a tree in the rSPR graph:
 
@@
 	\end{cases}.$$
-	Moreover, each tree of $N(T,u)$ can be obtained by an SPR operation that moves the subtree rooted at $u$.
+	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation that moves the subtree rooted at $u$.
 \end{lemma}
@@
 	Call these $U$ and $V$, respectively.
-	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an SPR operation.
-	Then the SPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, SPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
+	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
 	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
@@
 	\label{lem:select_random_neighbor}
-	Given a tree $T$, an SPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
+	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
 \end{lemma}
@@
 	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
-	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding SPR destination.
-	Finally, we require linear time to apply the chosen SPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
+	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
+	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
 	Thus, the for loop requires linear time.
-	By Lemma~\ref{lem:compute_degree} the chosen tree is an SPR neighbor of $T$ and is chosen uniformly at random.
-	Therefore, the procedure uses linear time and space and selects an SPR neighbor of $T$ uniformly at random.
+	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
+	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
 \end{proof}
@@
 \begin{corollary}
-	Given a tree $T$, the SPR neighbors of $T$ can be enumerated in $\OhOf{n^3}$-time.
+	Given a tree $T$, the rSPR neighbors of $T$ can be enumerated in $\OhOf{n^3}$-time.
 \end{corollary}
@@
 
-We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\deg(T_1)$ and $\deg(T_2)$ (Table~\ref{tab:regressionMAT}).
-Furthermore, by plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
+We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
+Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
 We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
@@
 \section{Rooted SPR Neighborhoods}
-Having made the connection between curvature values and access times on the SPR graphs, we now consider curvature theoretically.
+Having made the connection between curvature values and access times on the rSPR graphs, we now consider curvature theoretically.
 In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
@@
 
-We now bound the relative and exact difference in SPR degree between two trees with $n$ leaves.
+We now bound the relative and exact difference in rSPR degree between two trees with $n$ leaves.
 
@@
 	\hspace*{\stretch{1}}
-	\caption{An SPR move labelled as in Lemma~\ref{lem:degree_change}.
-Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the SPR degree by $2\left(k(a-b) + i - j\right)$.}
+	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
+Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
 	\label{fig:spr-degree-change}
@@
 \begin{proof}
-The set of permissible SPR moves changes in four different ways due to the movement of $R$:
+The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
 (i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
@@
 (iv) $R$'s parent subtree may no longer be moved into $V$.
-No additional moves are introduced or blocked by the original SPR operation on $R$.
+No additional moves are introduced or blocked by the original rSPR operation on $R$.
 
@@
 
-The difference in SPR degree is thus $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
+The difference in rSPR degree is thus $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
 \end{proof}
 
-Moreover, we can use these ideas to determine the number of SPR moves that are, in some respects, independent of a given SPR move.
-That is, for two trees $S$ and $T$ differing by a single SPR move, we wish to know the number of SPR moves that are applicable to both trees rather than unique to one of the trees.
+Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
+That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
 To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
@@
 \begin{proof}
-By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ SPR moves can be applied to $T$ and $S$ with the same source and target nodes.
+By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
@@
 \begin{proof}
-	$T$ and $S$ differ by one SPR move that moves a subtree $R$.
+	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
 	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
@@
 Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the SPR graph.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
 \end{proposition}
@@
 First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
-Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one SPR move apart as per Corollary~\ref{cor:paired_neighbors}.
+Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
 We pair these trees in the mass transport.
@@
 Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the SPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
+Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
 This directly leads to the following Corollary:
@@
 We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree and its change in depth and that rSPR treespace tends toward flatness with respect to SPR moves that move small subtrees.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree and its change in depth and that rSPR treespace tends toward flatness with respect to rSPR moves that move small subtrees.
 

### 2015-03-30 curvature.tex
Textual changes on the plane.

@@
 Researchers have noticed SPR graph effects in MCMC~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
-We recently showed that graph structure has a significant effect on MCMC mixing using real data and programs~\cite{Whidden2015-yi}.
-Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object called the compact real tree which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
+We recently showed that graph structure has a significant effect on MCMC mixing applying a popular inference program to real data~\cite{Whidden2015-yi}.
+Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
 
-ML also has graph effects, which manifest as local maxima.
-There have been investigations of local maxima, but mainly from the perspective of branch lengths rather than topology~\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}.
+Phylogenetic maximum-likelihood (ML) inference is also impacted by graph structure, which can manifest itself as local maxima.
+There have been investigations of local maxima, though mainly from the perspective of branch lengths rather than topology~\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}.
 The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another~\cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
@@
 
-The SPR graph is thus very important in phylogenetic inference, and yet still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
+Thus the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures; given this importance still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
 \cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
@@
 $$
-Despite the centrality of the SPR graph to phylogenetic inference, we are not aware of any other work investigating its properties.
+We are not aware of any other work investigating large-scale properties of the SPR graph.
 
-Recently, a new approach to quantifying properties of discrete and continuous spaces with respect to random walks has been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
-They define a notion of curvature on a discrete or continuous case.
-This formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
-Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$.
+A new approach to quantifying properties of random walks on discrete and continuous spaces has recently been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
+They define a notion of curvature which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
+Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$, which includes biased walks such as MCMC.
 
-In this paper, we investigate curvature of the rSPR graph with respect to two random walks and make connections to access times for those random walks.
-We use random walks on the rSPR graph as being a general notion of movement around a graph, modeling phylogenetic MCMC searches.
-For this first effort we investigate random walks defined in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
+In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
+For this first effort we investigate random walks defined only in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
 We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
-As the full rSPR treespace contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of treespace.
+As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of graph.
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
@@
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
-In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that, contrary to previous assumptions\todo{This is going to need a citation or be taken out}, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that modeling phylogenetic MCMC search.
 
@@
 We follow the definitions and notation from~\cite{bordewich05,whidden2013hybridization, Whidden2015-yi}.
-A (rooted binary phylogenetic) $X$-tree is a rooted tree $T$ whose nodes have zero or two children.
-Leaves of $T$ are bijectively labelled with the members of a label set $X$.
+A (rooted binary phylogenetic) $X$-tree is a rooted tree $T$ whose nodes have zero or two children such that the leaves of $T$ are bijectively labelled with the members of a label set $X$.
 As in~\cite{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$ (Fig.~\ref{fig:x-tree}).
@@
 The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children (Fig.~\ref{fig:induced}).
-For the rest of the paper, \textbf{we will assume that all phylogenetic trees are rooted,} and thus that tree inclusion is rooted tree inclusion.
+For the rest of the paper, \textbf{we will assume that all phylogenetic trees are binary and rooted,} and thus that tree inclusion is rooted tree inclusion.
 
@@
 Then the operation adds a new node $p'_x$ to $T_{p_x}$ by subdividing an edge of $T_{p_x}$ and adding a new edge $(x, p'_x)$, making $x$ a child of $p'_x$.
-Finally, $p_x$ is suppressed.
+Finally, $p_x$ is suppressed, joining the two edges on either side of that node.
 See Figure~\ref{fig:spr} for an example.
@@
 For example, the trees in Figure~\ref{fig:two-spr} are separated by two rSPR operations.
-Moreover, rSPR operations naturally give rise to a graph on the set of $X$-trees.
+Moreover, rSPR operations naturally give rise to a graph on the set of $X$-trees for which this distance is simply the shortest-path graph distance.
 Let $\mathcal{T}_n$ be the set of trees with $n$ leaves and label set $X = \set{1, 2, \ldots n, \rho}$.
 Then the rSPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
-Observe that distances in this graph are given by $\dspr{\cdot,\cdot}$.
 
-To avoid confusion between the two types of graph structures considered here, we refer to vertices of the rSPR graph as \emph{vertices} and vertices of individual trees as \emph{nodes}.
+To avoid confusion between the two types of graph structures considered here, we refer to vertices of the rSPR graph as \emph{vertices} and vertices of individual trees (i.e.\ leaves and internal nodes) as \emph{nodes}.
 Let $N(T)$ be the set of rSPR neighbors of a tree $T$ (this does not include $T$).
@@
 
-Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
+Ollivier-Ricci curvature provides a rigorous yet intuitive formalization of the shape of a graph with respect to a random walk (or indeed the shape of a variety of metric spaces, see~\cite{Ollivier2009-bw}).
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
+The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
  $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
@@
 \end{equation}
-Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average more distant than $x$ and $y$.
+Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
@@
 
-\subsection{Computing rSPR treespaces}
+\subsection{Computing rSPR graphs}
 \label{sec:computing_treespace}
 
-We computed the rSPR treespaces for four to seven leaves.
-This required a fast new algorithm for rSPR graph construction, reducing the time required to do so from $\OhOf{m^2 n}$ to $\OhOf{mn^3}$, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
+We computed the rSPR graph explicitly for trees on four to seven leaves.
+This required a fast new algorithm for rSPR graph construction, as previous algorithms required $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
+Here we reduce that time to $\OhOf{mn^3}$.
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
@@
 In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
-We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2n^2}$-time in the unrooted case).
+We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
 However, applying this method to complete rSPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, requires an enormous amount of computation.
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
 
-We saw a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
-Curvature values tended to increase with increasing rSPR distance.
-In addition, their variance decreased with increasing distance.
-Neighboring trees had minimum curvature values for a given treespace, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees in a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR treespaces.
+We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
+Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
+Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
 
@@
 Indeed, the 7-leaf treespace contained adjacent pairs of trees with negative curvature.
-Such pairs are possible bottlenecks in phylogenetic searches which may be exarcebated by likelihood or branch length constraints.
+Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
 
@@
 To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
-The uniform random walk moves from the current vertex to one of its neighbors uniformly at random.
-However, this makes the walk more likely to sample higher degree vertices.
+The uniform random walk moves from the current vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
 In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
@@
 To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing rSPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
-A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree---and then one could be picked uniformly at random.
+A na\"ive approach would require $\OhOf{n^3}$ time: $\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree so that one could be picked uniformly at random.
 To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of rSPR moves with a one-to-one correspondence to rSPR neighbors, as described in the next paragraph.
@@
 The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
-In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks---the mean number of iterations required to move from one tree to the other.
+In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks: the mean number of iterations required to move from one tree to the other.
 We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
@@
 As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
-Here again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
+Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
 Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
-To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram \cite{gricci, tangle}.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using our group-theoretic methods \cite{tangle}.
 
@@
 Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
-This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the likelihood of a random walk going from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
+This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
 On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
@@
     \caption{\
-        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.\
+        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
         Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows curvature $\kappa(\MH; \cdot, \cdot)$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
 
-The full analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool; instructions are in the repository README file.
+The analysis generating the can be reproduced using the SCons (\url{http://scons.org/}) build tool and an IPython notebook; instructions are in the repository README file.
 
@@
 Having made the connection between curvature values and access times on the rSPR graphs, we now consider curvature theoretically.
-In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
-Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.
+We begin by bounding differences between degrees, and then continue by enumerating squares and triangles in the rSPR graph.
 
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such "triangles" involving two adjacent trees has a key role in determining whether the curvature of two adjacent trees is positive or negative.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether the curvature of two adjacent trees is positive or negative.
 
@@
 
-We first consider properties of the uniform random walk on the rSPR graph.
+We first consider properties of the uniform (a.k.a.\ isotropic) random walk on the rSPR graph.
 Assume that the random walk begins at a tree $T$.
@@
 \section{Conclusion and future work}
-In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
+In summary, we have learned characteristics of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond previous work concerning graph diameter and vertex degree.
 We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
 We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree and its change in depth and that rSPR treespace tends toward flatness with respect to rSPR moves that move small subtrees.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR treespace tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
 
-We emphasize that the influence of curvature is small in this setting of equiprobable trees (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution is the predominant factor determining hitting times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+We emphasize that the influence of curvature is small in this setting of equiprobable trees (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
 However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets with hundreds or thousands of trees~\cite{Whidden2015-yi}.
@@
 
-In addition, we will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
+We will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
 We will also explore random walks ranked trees~\cite{Song2006-xe} and models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''

### 2015-03-30 curvature.tex
Finishing eliminating treespace.
cc @cwhidden-- LMK if you disagree.

@@
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full treespace) we can avoid pairwise computations by successively adding trees to the graph:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph:
 \begin{enumerate}[label={\arabic*}.]
@@
 Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and that method is infeasible for constructing larger tree spaces.
-Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.
+Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
 
@@
 
-Larger treespaces tended to have smaller curvature values.
-Indeed, the 7-leaf treespace contained adjacent pairs of trees with negative curvature.
+Larger rSPR graphs tended to have smaller curvature values.
+Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
 Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
@@
 We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR treespace tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
 

### 2015-03-30 curvature.tex
A few minor tweaks, and notes for @cwhidden.

@@
 
-Thus the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures; given this importance still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
+Thus although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
 \cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
@@
 They define a notion of curvature which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
-Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$, which includes biased walks such as MCMC.
+Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying a couple of reasonable assumptions, which includes biased walks such as MCMC.
 
@@
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
-In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that modeling phylogenetic MCMC search.
+In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
 
@@
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by simply successively adding trees to the graph:
 \begin{enumerate}[label={\arabic*}.]
@@
 	As described above, the integer labels are simply the order of the input trees.
-	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.
-	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$).
+	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.\todo{"in ascending order to the end of the vertex list": confused!}
+	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$)\todo{sorry, but I don't understand this $\log n$ term from the explanation}.
 	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
 	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick representation of the tree.
+	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
 	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
 	This requires only $\OhOf{n}$-time for each tree (i.e.
-a total of $\OhOf{mn^3}$-time) using a standard nodes and pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to realistic phylogenetic trees).
+a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
 	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
@@
 This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
-Our fast graph construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
-Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it.
-Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and that method is infeasible for constructing larger tree spaces.
+The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
+Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
+Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and is thus method is infeasible for constructing larger rSPR tree graphs.
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.

### 2015-03-30 curvature.tex
One big rearrangement-suggestion for @cwhidden.
Along with some smaller ones.

@@
 
+\todo{Chris, especially given that we are at a place that is not known for its CS, I might suggest putting your results concerning curvature first to emphasize to reviewers that this work is not trivial.
+That's what they will be scanning for after you convince them that the area is interesting.}
+
 \subsection{Computing rSPR graphs}
@@
 Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
-Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and is thus method is infeasible for constructing larger rSPR tree graphs.
+Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
@@
 
-Instead, we consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
+Instead\todo{does this seem like a strange comparison to you?}, we consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
 Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
@@
 However, some of these moves will result in the same neighboring tree as other rSPR moves.
-In particular, where a node's edge means the edge connecting the subtree rooted at that node to the rest of the tree:
+In particular, where we call the edge connecting the subtree rooted at that node to the rest of the tree the ``node's edge'', we have:
 \begin{enumerate}
@@
 \item Moving $R$ to its parent edge results in the same tree,
-\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge
+\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge,
 \item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
 \end{enumerate}
+\todo{Is it clear that this is an exhaustive list?}
 We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).

### 2015-03-30 curvature.tex
minor typos

@@
 We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
-As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of graph.
+As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph.
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
@@
 
-The analysis generating the can be reproduced using the SCons (\url{http://scons.org/}) build tool and an IPython notebook; instructions are in the repository README file.
+The analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool and an IPython notebook; instructions are in the repository README file.
 

### 2015-03-30 curvature.tex
removed more usages of tree space rather than SPR graph

@@
 This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
-The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
-Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
-Constructing the 8-taxon tree space using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
+The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
+Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
+Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
 Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
@@
 \subsection{Computing curvature values}
-To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci}.
@@
 We will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
-We will also explore random walks ranked trees~\cite{Song2006-xe} and models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
+We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
 Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.

### 2015-03-30 curvature.tex
more graph algorithm explanation and citations

@@
 
-Thus although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
+Although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
 \cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
@@
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by simply successively adding trees to the graph:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph and efficiently identifying their neighbors:
 \begin{enumerate}[label={\arabic*}.]
@@
 \begin{proof}
-	We implement the graph as an adjacency list with integer-labelled vertices.
+	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
+red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
 	As described above, the integer labels are simply the order of the input trees.
-	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.\todo{"in ascending order to the end of the vertex list": confused!}
-	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$)\todo{sorry, but I don't understand this $\log n$ term from the explanation}.
+	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
+	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
 	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
 	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
+	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
 	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
@@
 	We apply the above procedure.
-	We use a standard nodes and pointer representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
+	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
 	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.

### 2015-03-30 curvature.tex
A few more twiddles.

@@
 
-The analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool and an IPython notebook; instructions are in the repository README file.
+The analysis generating the can be reproduced by simply invoking the SCons (\url{http://scons.org/}) build tool and evaluating cells in an IPython notebook; instructions are in the repository README file.
 
@@
 
-We now bound the relative and exact difference in rSPR degree between two trees with $n$ leaves.
-
+We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
 \begin{lemma}
@@
 
-The difference in rSPR degree is thus $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
+Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
 \end{proof}
@@
 \section{Curvature}
-For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term.
+For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
 
@@
 The proposition again follows because $\degree{T_n}$ is quadratic in $n$.
-
 \end{proof}
 
-Next we note a rough bound on the curvature of two trees with respect to their distance.
+Next we note a rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
 
@@
 
-We can obtain a tighter bound on the maximum curvature of two adjacent trees.
 \begin{lemma}
@@
 We next bound the effect of laziness on curvature.
-The lazy random walk again selects a tree uniformly at random, as with the uniform random walk, but only accepts the move with probability $p$.
+The lazy random walk again selects a neighbor uniformly at random, as with the uniform random walk, but only travels to the neighbor with probability $p$.
 We denote the curvature of the lazy random walk between two trees $T$ and $S$ by $\curvature[\ell_p]{T,S}$.
@@
 	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[\ell_p]{T,S} = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
-	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
+	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
 	Then:

### 2015-03-30 curvature.tex
a few edits in response to comments

@@
 
-\todo{Chris, especially given that we are at a place that is not known for its CS, I might suggest putting your results concerning curvature first to emphasize to reviewers that this work is not trivial.
-That's what they will be scanning for after you convince them that the area is interesting.}
-
 \subsection{Computing rSPR graphs}
@@
 
-Instead\todo{does this seem like a strange comparison to you?}, we consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
+We consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
 Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
@@
 \end{enumerate}
-\todo{Is it clear that this is an exhaustive list?}
+We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive.
 We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
@@
 
-The analysis generating the can be reproduced by simply invoking the SCons (\url{http://scons.org/}) build tool and evaluating cells in an IPython notebook; instructions are in the repository README file.
+The analysis can be reproduced by invoking the SCons (\url{http://scons.org/}) build tool and evaluating cells in an IPython notebook; instructions are in the repository README file.
 

### 2015-03-30 curvature.tex
Citing Romon.

@@
 To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci}.
+We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
 
@@
 
-Next we note a rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
+Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
 
@@
 
-We continue to explore the effect of biased random walks by considering the curvature of Metropolis Hastings random walks.
-The Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+Next we consider the curvature of the Metropolis-Hastings (MH) random walk sampling the uniform distribution on trees, which we will simply call the MH random walk.
+This random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.

### 2015-03-30 curvature.tex
Citing Jost2013.
@cwhidden, please check. I believe this is what you meant in responding
to

https://matsengrp.slack.com/archives/arbornauts/p1424389874000089

Is that right?
introduce ric() for asymptotic curvature

@@
     \ifthenelse{\equal{#1}{}}%
-		{\kappa_n(#2)}%
-		{\kappa_n(#1;#2)}%
+		{\kappa(#2)}%
+		{\kappa_{#1}(#2)}%
+}
+\newcommand{\ric}[1]{
+	ric(#1)%
 }
@@
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+Finally, we prove that the two competing notions of Ricci-Ollivier curvature (coarse and asymptotic) are closely related when applied to the rSPR graph, differing only by a small factor inversely proportional to the rSPR degrees of the given trees.
 In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
@@
 
-Ollivier-Ricci curvature provides a rigorous yet intuitive formalization of the shape of a graph with respect to a random walk (or indeed the shape of a variety of metric spaces, see~\cite{Ollivier2009-bw}).
+Ricci-Ollivier curvature provides a rigorous yet intuitive formalization of the shape of a graph with respect to a random walk (or indeed the shape of a variety of metric spaces, see~\cite{Ollivier2009-bw}).
 Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
@@
 
-The curvature of $x$ and $y$ is then defined as:
+The (coarse) Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
 \begin{equation}
@@
 \end{equation}
+We use $\curvature{MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
 Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
+
+Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the (asymptotic) Olivier-Ricci curvature.
+The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
+We denote the coarse curvature of the lazy random walk between two trees $x$ and $y$ by $\curvature[p]{x,y}$.
+For example, $\curvature[\frac{1}{4}]{x,y}$ describes the curvature of the lazy random walk that follows the given distribution with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
+For the lazy uniform random walk, the mass assignment $m^p_x$ is $x \cup N(x)$, with each neighbor assigned mass $\frac{p}{\degree{x}}$ and $x$ assigned the remaining $1 - p$ mass.
+The (asymptotic) Ricci-Ollivier curvature of $x$ and $y$ is then:
+\begin{equation}
+\ric{x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{x,y}}{p}
+\label{eq:curvatureDef}
+\end{equation}
+This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
+Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
+As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent trees and are equal for nonadjacent trees.
+We define $\ric{MH; x, y}$ analogously to $\curvature{MH; x, y}$.
 
@@
 To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then compute curvatures for given pairs of trees directly, by using linear programming to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
+We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
 
@@
 	\hspace*{\stretch{1}}
-    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+    \caption{Scatterplot of $\ric{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
 	\label{fig:rspr-scatter}
@@
         Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows curvature $\kappa(\MH; \cdot, \cdot)$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows asymptotic curvature $\ric{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
         }
@@
 \section{Curvature}
-For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
+For the purposes of this paper, $\kappa$ is the coarse Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
 
@@
 Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
+We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.
 
-We next bound the effect of laziness on curvature.
-The lazy random walk again selects a neighbor uniformly at random, as with the uniform random walk, but only travels to the neighbor with probability $p$.
-We denote the curvature of the lazy random walk between two trees $T$ and $S$ by $\curvature[\ell_p]{T,S}$.
-For example, $\curvature[\ell_{\frac{1}{4}}]{T,S}$ describes the curvature of the lazy random walk that moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
-For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-Note that Ollivier defined $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$~\cite{Ollivier2009-bw}.
-This definition of curvature is invariant of $p$ for small enough $p$ and was used by Ollivier to avoid parity problems on graphs where the uniform random walk is periodic.
-As we now prove, the curvature of the lazy random walk under this definition is closely related to that of the uniform random walk.
+We next bound the difference between the coarse and asymptotic curvatures.
+Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
+For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
+The asympototic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
+As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
 
 \begin{lemma}
+\label{lem:asymptotic}
 	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
@@
 	\begin{enumerate}
-		\item	$\curvature[\ell_p]{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
-		\item	$\curvature{T,S} \le \curvature[\ell_p]{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
+		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\ric{MH; T,S} = \curvature{MH; T,S}$, if $\dspr{T,S} > 1$, and
+		\item	$\curvature{MH; T,S} \le \ric{MH; T,S} \le \curvature{MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
 	\end{enumerate}
@@
 \begin{proof}
-	We first prove the lower bound, that is $\curvature{T,S} \le \curvature[\ell_p]{T,S}$.
+	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
 	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
-	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[\ell_p]{T,S} = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
+	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
 	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
@@
 	\begin{align*}
-		\curvature[\ell_p]{T,S} &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
+		\curvature[p]{T,S} /p &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
 		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
@@
 	\end{align*}
+The lower bound in the MH case follows analogously.
 
@@
 	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
+Note that the MH distribution similarly assigns at most $\frac{1}{\degree{X}$ to each neighbor $X$ of the given tree.
+The upper bound in the MH case thus follows as well.
 \end{proof}
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature[\MH]{S,T}$.
+We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature{\MH; S,T}$.
 
@@
 	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
-	\le \curvature[\MH]{T,S}
+	\le \curvature{\MH; T,S}
 	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
 	$$\curvature{T,S} - \frac{1}{6}
-	\le \curvature[\MH]{T,S}
+	\le \curvature{\MH; T,S}
 	\le \curvature{T,S} + \frac{1}{6}.$$
@@
 	\begin{align*}
-		\curvature[\MH]{T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
+		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
 		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
@@
 Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Finally, we proved that the two competing definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
 

### 2015-03-30 curvature.tex
drop time sentence for sampling random walks
whoops - broken missing bracket

@@
 We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
-This required XXX and YYY time respectively.~\todo{EM: can you fill in the time this required?}
 
@@
 	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
-Note that the MH distribution similarly assigns at most $\frac{1}{\degree{X}$ to each neighbor $X$ of the given tree.
+Note that the MH distribution similarly assigns at most $\frac{1}{\degree{X}}$ to each neighbor $X$ of the given tree.
 The upper bound in the MH case thus follows as well.

### 2015-03-31 scripts/plot.py
More bigger fonts; colorbar behaving.

@@ -10,9 +10,6 @@ import pickle
 import seaborn as sns
 
 sns.set_style('ticks')
-mpl.rcParams.update({
-    'font.size': 32, 'font.family': 'Lato',
-    'font.weight': 600, 'axes.labelweight': 600})
 
 parser = argparse.ArgumentParser(
     description='plot curvature results',
@@ -35,6 +32,11 @@ df = pickle.load(gzip.open(args.results_path, 'rb'))
 df['dist_jitter'] = df['dist'] + np.random.normal(0, 0.1, size=len(df))
 df['info'] = \
     df['t1_nwk']+'<br/>'+df['t2_nwk']+'<br/>avg_deg: '+map(str, df['avg_deg'])
+
+mpl.rcParams.update({
+    'font.size': 22, 'axes.labelsize': 16, 'xtick.labelsize':14, 'ytick.labelsize':14,
+    'font.family': 'Lato',
+    'font.weight': 600, 'axes.labelweight': 600})
 
 if args.which == 'hexbin':
     p = df.plot(
@@ -65,6 +67,10 @@ p.grid(b=None)
 sns.despine(offset=10)
 p.set_xlabel('kappa')
 p.set_ylabel('distance (jittered)')
+cbar = p.get_figure().get_children()[-1]
+cbar.tick_params(left=False, right=True, labelleft=False, labelright=True)
+cbar.get_yaxis().set_major_locator(mpl.ticker.MultipleLocator(0.25))
+cbar.get_yaxis().labelpad = 25
 
 if args.d3:
     from mpld3 import fig_to_html, plugins

### 2015-03-31 curvature.tex
More bigger fonts; colorbar behaving.
Even bigger fonts! And I follow my own suggestions.

@@
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
-Finally, we prove that the two competing notions of Ricci-Ollivier curvature (coarse and asymptotic) are closely related when applied to the rSPR graph, differing only by a small factor inversely proportional to the rSPR degrees of the given trees.
+Finally, we prove that the two related notions of Ricci-Ollivier curvature (coarse and asymptotic) are close, and in fact identical in the limit of many taxa, when applied to the rSPR graph.
 In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
+
+The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
 
@@
 
-The (coarse) Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
+The \emph{coarse} Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
 \begin{equation}
@@
 
-Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the (asymptotic) Olivier-Ricci curvature.
+Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic} Olivier-Ricci curvature.
 The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
-We denote the coarse curvature of the lazy random walk between two trees $x$ and $y$ by $\curvature[p]{x,y}$.
+We denote the coarse curvature of the lazy random walk between two vertices $x$ and $y$ by $\curvature[p]{x,y}$.
 For example, $\curvature[\frac{1}{4}]{x,y}$ describes the curvature of the lazy random walk that follows the given distribution with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
@@
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
-As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent trees and are equal for nonadjacent trees.
+As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
 We define $\ric{MH; x, y}$ analogously to $\curvature{MH; x, y}$.
@@
 Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
-Finally, we proved that the two competing definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
+Finally, we proved that the coarse and asymptotic definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
 

### 2015-03-31 scripts/plot.py

@@ -34,7 +34,7 @@ df['info'] = \
     df['t1_nwk']+'<br/>'+df['t2_nwk']+'<br/>avg_deg: '+map(str, df['avg_deg'])
 
 mpl.rcParams.update({
-    'font.size': 22, 'axes.labelsize': 16, 'xtick.labelsize':14, 'ytick.labelsize':14,
+    'font.size': 26, 'axes.labelsize': 26, 'xtick.labelsize':20, 'ytick.labelsize':20,
     'font.family': 'Lato',
     'font.weight': 600, 'axes.labelweight': 600})
 

### 2015-03-31 curvature.tex
Title change + fixes.

@@
 \begin{document}
-\title{Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph}
+\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph}
 
@@
 The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
-Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of the SPR graph.
+Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of this \emph{SPR graph}.
 Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
-In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the recently-developed notion of Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
-This notion of curvature is especially relevant to stochastic search on the graph.
+In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
 By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
-We confirm using simulation that degree and curvature have the expected impact on mean access time distributions.
-This indicates significant structure of the graph beyond that which which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
+We confirm using simulation that degree and curvature have the expected impact on mean access time distributions, demonstrating relevance of these curvature results to stochastic tree search.
+This indicates significant structure of the rSPR graph beyond that which which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
 \end{abstract}
@@
 \end{equation}
-We use $\curvature{MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
+We use $\curvature{\MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
 Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
@@
 As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
-We define $\ric{MH; x, y}$ analogously to $\curvature{MH; x, y}$.
+We define $\ric{\MH; x, y}$ analogously to $\curvature{\MH; x, y}$.
 
@@
 
-The analysis can be reproduced by invoking the SCons (\url{http://scons.org/}) build tool and evaluating cells in an IPython notebook; instructions are in the repository README file.
+The analysis can be reproduced by invoking the SCons (\url{http://scons.org/}) build tool and running the cells in an IPython notebook; instructions are in the repository README file.
 
@@
 \section{Conclusion and future work}
-In summary, we have learned characteristics of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond previous work concerning graph diameter and vertex degree.
-We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
-We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
-Finally, we proved that the coarse and asymptotic definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
+In summary, we have advanced understanding of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond what was previously known concerning graph diameter and vertex degree.
+We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
+We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Finally, we proved that the coarse and asymptotic definitions of Ollivier-Ricci curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
 
-We emphasize that the influence of curvature is small in this setting of equiprobable trees (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
-However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets with hundreds or thousands of trees~\cite{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with a nontrivial likelihood function, which will significantly reduce the effective size of the posterior distribution, and may lead to ``bottlenecks'' like those we have observed in real data.
-We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
+In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
+Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will significantly reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to ``bottlenecks'' like those we have observed in real data.
+In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
+Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
+% We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
 
-We will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
 We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''

### 2015-03-31 curvature.tex
A howler.

@@
 The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
- $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
+ $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{z,w} \xi(z,w),$$
 where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.

### 2015-03-31 curvature.tex
Non-controversial revisions.

@@
 
-Ricci-Ollivier curvature provides a rigorous yet intuitive formalization of the shape of a graph with respect to a random walk (or indeed the shape of a variety of metric spaces, see~\cite{Ollivier2009-bw}).
-Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
+Ricci-Ollivier curvature provides a rigorous yet intuitive formalization of the shape of a metric space with respect to a random walk.
+For the purposes of this paper, we will specialize to that space being a graph equipped with the shortest-path distance, and we will describe curvature intuitively.
+For a more rigorous presentation in the more general setting of a Polish metric space, see~\cite{Ollivier2009-bw} or the survey~\cite{Ollivier2010-ao}.
+
+Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after a fixed period of time starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
 The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
- $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{z,w} \xi(z,w),$$
-where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.
+ $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} \dspr{z,w} \xi(z,w),$$
+where $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second).
 
-The \emph{coarse} Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
+The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
 \begin{equation}
@@
 
-Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic} Olivier-Ricci curvature.
+Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
 The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
@@
 For the lazy uniform random walk, the mass assignment $m^p_x$ is $x \cup N(x)$, with each neighbor assigned mass $\frac{p}{\degree{x}}$ and $x$ assigned the remaining $1 - p$ mass.
-The (asymptotic) Ricci-Ollivier curvature of $x$ and $y$ is then:
+The asymptotic Ricci-Ollivier curvature of $x$ and $y$ is then:
 \begin{equation}
@@
 Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
-Finally, we proved that the coarse and asymptotic definitions of Ollivier-Ricci curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
+Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
 

### 2015-03-31 curvature.tex
Making definitions more rigoro and clearer.

@@
 \newcommand{\ric}[1]{
-	ric(#1)%
+	\operatorname{ric}(#1)%
 }
@@
 Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after a fixed period of time starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
+The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} \dspr{z,w} \xi(z,w),$$
@@
 \begin{equation}
-\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
+\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
 \label{eq:curvatureDef}
 \end{equation}
-We use $\curvature{\MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
+For the purposes of this paper, we will use $\curvature{x,y}$ to denote the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
 Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
@@
 
-Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
-The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
-We denote the coarse curvature of the lazy random walk between two vertices $x$ and $y$ by $\curvature[p]{x,y}$.
-For example, $\curvature[\frac{1}{4}]{x,y}$ describes the curvature of the lazy random walk that follows the given distribution with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
-For the lazy uniform random walk, the mass assignment $m^p_x$ is $x \cup N(x)$, with each neighbor assigned mass $\frac{p}{\degree{x}}$ and $x$ assigned the remaining $1 - p$ mass.
+Lin et al.~\cite{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
+The lazy random walk selects a neighbor according to a given distribution $m$ as before, but only travels to the neighbor with probability $p$.
+Thus the lazy mass assignment $m^p_x$ assigns $p\, m_x$ mass to each neighbor of $x$ and the remaining $1 - p$ mass to $x$.
+We denote the coarse curvature of the $p$-lazy random walk between two vertices $x$ and $y$ with respect to a random walk $m$ by $\curvature[p]{m; x,y}$.
+For example, $\curvature[\frac{1}{4}]{m; x,y}$ describes the curvature of the lazy random walk that follows the given random walk $m$ with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
 The asymptotic Ricci-Ollivier curvature of $x$ and $y$ is then:
 \begin{equation}
-\ric{x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{x,y}}{p}
+\ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p}
 \label{eq:curvatureDef}
 \end{equation}
+As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
 This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
@@
 As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
-We define $\ric{\MH; x, y}$ analogously to $\curvature{\MH; x, y}$.
 

### 2015-03-31 curvature.tex
Coarse, not ric.
Mostly typography.

@@
 	\hspace*{\stretch{1}}
-	\caption{Two trees differing by two rSPR operations that move the grey subtrees.}
+	\caption{\
+    Two rSPR operations, each of which moves one grey subtree.
+    The leftmost and rightmost trees are rSPR distance two apart.}
 	\label{fig:two-spr}
@@
 \ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p}
-\label{eq:curvatureDef}
+\label{eq:asympCurvatureDef}
 \end{equation}
@@
 	\hspace*{\stretch{1}}
-    \caption{Scatterplot of $\ric{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+    \caption{Scatterplot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+Color displays the average degree of $T_1$ and $T_2$.
+Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
 	\label{fig:rspr-scatter}
@@
 		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
-		\item	$\ric{MH; T,S} = \curvature{MH; T,S}$, if $\dspr{T,S} > 1$, and
-		\item	$\curvature{MH; T,S} \le \ric{MH; T,S} \le \curvature{MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and
+		\item	$\curvature{\MH; T,S} \le \ric{\MH; T,S} \le \curvature{\MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
 	\end{enumerate}
@@
 	\begin{align*}
-		\curvature[p]{T,S} /p &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
-		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
-		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
-		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
-		&\ge \curvature{T,S}
+		\curvature[p]{T,S} /p &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
+		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
+		&= \left(\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}\right) \\
+		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
+		&= \curvature{T,S}
 	\end{align*}

### 2015-03-31 curvature.tex
one more curvature in place of ric, a missing the

@@
         Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows asymptotic curvature $\ric{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows asymptotic curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
         }
@@
 We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
 Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.

### 2015-03-31 curvature.tex
consistent enumerate punctation, closes #15

@@
 \begin{enumerate}[label={\arabic*}.]
-	\item Let $G$ be an empty graph
-	\item For each of the $m$ trees in a specified but arbitrary order,
+	\item Let $G$ be an empty graph.
+	\item For each of the $m$ trees in a specified but arbitrary order:
 		\begin{enumerate}
-			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index
-			\item For each of the $\OhOf{n^2}$ neighbors of $T$
+			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index.
+			\item For each of the $\OhOf{n^2}$ neighbors of $T$:
 				\begin{enumerate}
-					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index
+					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index.
 				\end{enumerate}
@@
 \item Moving $R$ to its parent edge results in the same tree,
-\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge,
+\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge, and
 \item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
@@
 \begin{enumerate}[label={\arabic*}.]
-	\item	compute the degree of $T$, $\degree{T}$.
-	\item pick a random integer $r$ in the range $[1,\degree{T}]$
-	\item label each node $u$ of $T$ by its preorder number and compute the number of nodes in the subtree rooted at each $u$.
-\item for each tree node $u$ and while $r > 0$
+	\item	Compute the degree of $T$, $\degree{T}$.
+	\item Pick a random integer $r$ in the range $[1,\degree{T}]$.
+	\item Label each node $u$ of $T$ by its preorder number and compute the number of nodes in the subtree rooted at each $u$.
+\item For each tree node $u$ and while $r > 0$:
 	\begin{enumerate}
-		\item decrease $r$ by $\size{N(T,u)}$
-		\item if $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop
+		\item Decrease $r$ by $\size{N(T,u)}$.
+		\item If $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop.
 	\end{enumerate}
-\item return the neighbor $S$
+\item Return the neighbor $S$.
 \end{enumerate}
@@
 	\begin{enumerate}
-		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
+		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
 		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
@@
 	\begin{enumerate}
-		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
-		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$
+		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
+		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
 	\end{enumerate}
@@
 	\begin{enumerate}
-		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$, and
-		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
+		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$,
 		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and

### 2015-03-31 curvature.tex
slaying an extra which

@@
 We confirm using simulation that degree and curvature have the expected impact on mean access time distributions, demonstrating relevance of these curvature results to stochastic tree search.
-This indicates significant structure of the rSPR graph beyond that which which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
+This indicates significant structure of the rSPR graph beyond that which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
 \end{abstract}

### 2015-03-31 curvature.tex
MH curvatures are coarse

@@
         Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows asymptotic curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
         }

### 2015-03-31 curvature.tex
use d() for graph shortest-path distance

@@
 The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
-$$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} \dspr{z,w} \xi(z,w),$$
-where $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second).
+$$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),$$
+where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second).
 

### 2015-03-31 curvature.tex
short note on correctness of the graph construction algorithm

@@
 \begin{proof}
+	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
+
 	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.

### 2015-03-31 prefigs/get-figs.sh
Adding the 7 taxon scatter.

@@ -1,5 +1,3 @@
-for graph in nni rspr; do
-    for i in ../analysis/6-walks-50K/$graph/upmh/scatter*svg; do
-        cp $i $graph-upmh-$(basename $i)
-    done
+for i in ../analysis/6-walks-50K/rspr/upmh/scatter*svg ../analysis/7-walks-5K/rspr/upmh/scatter7.svg; do
+    cp $i rspr-upmh-$(basename $i)
 done

### 2015-03-31 curvature.tex
Adding the 7 taxon scatter.
Cleaning out todonotes.

@@
 \usepackage{ifthen}
-\usepackage{todonotes}
 %\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
@@
 \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-% shift for todonotes
-\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
+%\usepackage{todonotes}
+%% shift for todonotes
+%\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
 % regular margins
-%\usepackage[margin=1in]{geometry}
+\usepackage[margin=1in]{geometry}
 
@@
 
-We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
+We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
 Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
@@
 	\hspace*{\stretch{1}}
-	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter5}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
 	\hspace*{\stretch{2}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
 	\hspace*{\stretch{1}}

### 2015-03-31 curvature.tex
Spelling pass.

@@
 	\hspace*{\stretch{1}}
-    \caption{Scatterplot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
 Color displays the average degree of $T_1$ and $T_2$.
-Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
+Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
 	\label{fig:rspr-scatter}
@@
 	We apply the above procedure.
-	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
+	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
 	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
 	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
-	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
+	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
 	Both of these traversals require $\OhOf{n}$-time.
@@
 For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-The asympototic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
+The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
 As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
@@
 
-	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
+	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additional tree, as $T$ and $S$ are adjacent.
 	We thus obtain the lower bound of $\curvature{T,S} - \frac{1}{6}$ in this case.

### 2015-03-31 curvature.tex
Adding in graphicx to fix figures.
read-aloud fixes

@@
 \usepackage{subfigure}
+\usepackage{graphicx}
 \usepackage[nomarkers,nofiglist,notablist]{endfloat}
@@
 In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
-For this first effort we investigate random walks defined only in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
+For this first effort we investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.
 We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
@@
 $\{(((1,2),3),4), ((1,2),(3,4))\}$ and
-$\{(((1,4),3),4), ((1,4),(2,3))\}$
+$\{(((1,4),2),3), ((1,4),(2,3))\}$
 are the same after relabeling, so their curvatures are the same.
@@
 To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
-The number of such ``squares'' involving two adjacent trees will play a key rule in our bounds on curvature, as they push the curvature of those trees towards 0.
+The number of such ``squares'' involving two adjacent trees will play a key role in our bounds on curvature, as they push the curvature of those trees towards 0.
 
@@
 	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
-	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
+	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
 	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.

### 2015-03-31 curvature.tex
Fixing things around lem:asymptotic.

@@
 		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
-		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$,
-		\item	$\ric{\MH; T,S} = \curvature{\MH; T,S}$, if $\dspr{T,S} > 1$, and
-		\item	$\curvature{\MH; T,S} \le \ric{\MH; T,S} \le \curvature{\MH; T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
 	\end{enumerate}
@@
 	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
-	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
+	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
 	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
@@
 		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
-		&= \left(\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}\right) \\
+		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
 		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
-		&= \curvature{T,S}
+		&= \curvature{T,S}.
 	\end{align*}
-The lower bound in the MH case follows analogously.
 
-	For the upper bound, we observe that $$W_1'(T,S) \ge p(W_1(T,S) + (1-p)\left(\dspr{T,S}\right) - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $\frac{1}{\max(\degree{T},\degree{S})}$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
+	For the upper bound, we observe that $$W_1'(T,S) \ge p W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $\frac{1}{\max(\degree{T},\degree{S})}$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
 	The upper bound then follows analogously to the lower bound.
 	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
-Note that the MH distribution similarly assigns at most $\frac{1}{\degree{X}}$ to each neighbor $X$ of the given tree.
-The upper bound in the MH case thus follows as well.
 \end{proof}
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-We now bound the difference between the curvature of uniform random walks $\curvature{S,T}$, and that of MH random walks, $\curvature{\MH; S,T}$.
+The same statement and proof of Lemma~\ref{lem:asymptotic} works with $\curvature{T,S}$ replaced by the MH random walk curvature $\curvature{\MH; T,S}$.
+We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the MH random walk $\curvature{\MH; T,S}$.
 

### 2015-03-31 curvature.tex
Clarifications in the proof of another lemma.

@@
 	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
-	Thus, the hastings ratio is always $\ge \frac{5}{6}$.
+	Thus, the Hastings ratio is always $\ge \frac{5}{6}$.
 	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
-	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
-	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T$'s neighbors to $S$.
+	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
+	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
 	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
@@
 \section{Conclusion and future work}
-In summary, we have advanced understanding of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond what was previously known concerning graph diameter and vertex degree.
+In summary, we have advanced understanding of the phylogenetic rSPR graph substantially beyond what was previously known, which concerned graph diameter and vertex degree.
 We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.

### 2015-03-31 curvature.tex
Fixing lem:asymptotic reference.

@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-The same statement and proof of Lemma~\ref{lem:asymptotic} works with $\curvature{T,S}$ replaced by the MH random walk curvature $\curvature{\MH; T,S}$.
+The same statement and proof of Lemma~\ref{lem:asymptotic} works with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH random walk curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
 We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the MH random walk $\curvature{\MH; T,S}$.

### 2015-03-31 curvature.tex
A few more formatting capers.

@@
 	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
-	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
-	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
+	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p$.
+	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p) \, \dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
 	Then:
@@
 
-	For the upper bound, we observe that $$W_1'(T,S) \ge p W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $\frac{1}{\max(\degree{T},\degree{S})}$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
+	For the upper bound, we observe that $$W_1'(T,S) \ge p W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
 	The upper bound then follows analogously to the lower bound.

### 2015-03-31 curvature.tex
Clarifying definition of lazy random walk.

@@
 Lin et al.~\cite{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
-The lazy random walk selects a neighbor according to a given distribution $m$ as before, but only travels to the neighbor with probability $p$.
-Thus the lazy mass assignment $m^p_x$ assigns $p\, m_x$ mass to each neighbor of $x$ and the remaining $1 - p$ mass to $x$.
+The lazy random walk only travels according to $m_x$ with probability $p$ and otherwise stays put.
+Thus the lazy mass assignment $m^p_x$ is the sum of $p\, m_x$ and a point mass of $1 - p$ on $x$.
 We denote the coarse curvature of the $p$-lazy random walk between two vertices $x$ and $y$ with respect to a random walk $m$ by $\curvature[p]{m; x,y}$.

### 2015-03-31 curvature.tex
clarify mass transport in MH lemma

--- 2015-03-31 curvature.tex
@@
 	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
+	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
 	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
+	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
+	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions 
+	$\xi(z,w) \max\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right) - \xi'(z,w)$,
+	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
+	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
 	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.

### 2015-04-01 curvature.tex
hide all but 3 proofs for the extended abstract. The optional argument to cuttable can be used for replacement text.

@@
 
-\newcommand{\cuttable}[1]{#1} % In case we need to be succinct for limits.
+\newcommand{\cuttable}[2][]{% In case we need to be succinct for limits.
+    \ifthenelse{\equal{#1}{}}%
+		{}%
+		{#1}%
+% replace with these lines for the full version.
+%		{#2}%
+%		{#2}%
+}
 
@@
 \end{proof}
+
 This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
@@
 \end{lemma}
+\cuttable{
 \begin{proof}
@@
 \end{proof}
+}
 
@@
 \end{corollary}
+\cuttable{
 \begin{proof}
@@
 
+\cuttable{
 \begin{proof}
@@
 \end{proposition}
+\cuttable{
 \begin{proof}
@@
 \end{proof}
+}
 This bound is tight and has been verified computationally for $n \le 7$.

### 2015-04-01 curvature.tex
cut down extended abstract to 10 pages

@@
 \author{
-	\IEEEauthorblockN{Christopher Whidden}
+	\IEEEauthorblockN{Chris Whidden}
 	\IEEEauthorblockA{Program in Computational Biology\\
@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
+We omit most proofs which are unnecessary to understand the paper, and which can be found in~\cite{curvature_arxiv}.
 
@@
 The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children (Fig.~\ref{fig:induced}).
-For the rest of the paper, \textbf{we will assume that all phylogenetic trees are binary and rooted,} and thus that tree inclusion is rooted tree inclusion.
+For the rest of the paper, \textbf{we will assume that all phylogenetic trees are binary and rooted}\cuttable{, and thus that tree inclusion is rooted tree inclusion}.
 
@@
 Then the operation adds a new node $p'_x$ to $T_{p_x}$ by subdividing an edge of $T_{p_x}$ and adding a new edge $(x, p'_x)$, making $x$ a child of $p'_x$.
-Finally, $p_x$ is suppressed, joining the two edges on either side of that node.
+Finally, $p_x$ is suppressed\cuttable{, joining the two edges on either side of that node}.
 See Figure~\ref{fig:spr} for an example.
@@
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),$$
-where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second).
+where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ \cuttable{(more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second)}.
 
@@
 For the purposes of this paper, we will use $\curvature{x,y}$ to denote the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
-Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
-Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
+Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are \cuttable{on average }``more distant'' than $x$ and $y$.
+Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with \cuttable[a]{the given} random walk.
 
@@
 We denote the coarse curvature of the $p$-lazy random walk between two vertices $x$ and $y$ with respect to a random walk $m$ by $\curvature[p]{m; x,y}$.
-For example, $\curvature[\frac{1}{4}]{m; x,y}$ describes the curvature of the lazy random walk that follows the given random walk $m$ with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
+For example, $\curvature[1/4]{m; x,y}$ describes the curvature of the lazy random walk that follows the given random walk $m$ with probability $1/4$ and remains stationary with probability $3/4$.
 The asymptotic Ricci-Ollivier curvature of $x$ and $y$ is then:
@@
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph and efficiently identifying their neighbors:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we avoid pairwise computations by successively adding trees to the graph and \cuttable{efficiently }identifying their neighbors:
 \begin{enumerate}[label={\arabic*}.]
@@
 \subsection{Computing curvature values}
-To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
@@
 	\end{cases}.$$
-	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation that moves the subtree rooted at $u$.
+	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation moving the subtree rooted at $u$.
 \end{lemma}
@@
 	\label{lem:select_random_neighbor}
-	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
+	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time\cuttable{ using $\OhOf{n}$ space}.
 \end{lemma}
@@
 	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
-	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
+	Therefore, the procedure uses linear time and space and \cuttable[is correct]{selects an rSPR neighbor of $T$ uniformly at random}.
 \end{proof}
@@
 \begin{corollary}
-	Given a tree $T$, the rSPR neighbors of $T$ can be enumerated in $\OhOf{n^3}$-time.
+	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
 \end{corollary}
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
-We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
+We sampled \cuttable{a }200,000\cuttable{-iteration} random walk\cuttable[ iterations]{} on the 4-leaf rSPR graph and \cuttable{a }50,000-iteration\cuttable[s]{ random walk} on the 5-leaf rSPR graph.
 
@@
 Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
-To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using our group-theoretic methods \cite{tangle}.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using \cuttable{our }group-theoretic methods \cite{tangle}.
 
@@
 We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
-For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access}) is the count for time 2 minus the count for time 3.
+For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines\cuttable{ in Fig.~\ref{fig:kappa-access}}) is the count for time 2 minus the count for time 3.
 Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
 This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
-On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
+On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$\cuttable{ and instead achieves stationarity}, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
 
@@
 \begin{lemma}[{Song~\cite{Song2003-gf}}]
-    \pushQED{\qed}
+\cuttable{    \pushQED{\qed}}
 	\label{lem:degree_extremes}
@@
 	\end{enumerate}
-    \popQED
+\cuttable{    \popQED}
 \end{lemma}
@@
 	\begin{enumerate}
-		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
+		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
 		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
@@
 We can improve these bounds in the case of adjacent trees.
-To do so, we require the following lemma that characterizes exactly how the degree of a tree changes after an rSPR operation.
+To do so, we \cuttable{require the following lemma that }characterize\cuttable{s} exactly how the degree of a tree changes after an rSPR operation.
 See Figure~\ref{fig:spr-degree-change} for an illustration of the Lemma.
@@
 (iv) $R$'s parent subtree may no longer be moved into $V$.
-No additional moves are introduced or blocked by the original rSPR operation on $R$.
+No additional moves are introduced or blocked\cuttable{ by the original rSPR operation on $R$}.
 
-Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges (recall that we are excluding any ``root edge'' in these calculations).
+Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges\cuttable{(recall that we are excluding any ``root edge'' in these calculations)}.
 In the first case there are $a$ subtrees that can now be moved onto the $2k$ edges in $R$ (including its newly introduced parent edge and one of the newly subdivided root edges of $V$) for a total gain of $2ka$ distinct moves.
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such ``triangles'' involving two adjacent trees has a key role in determining whether the curvature of two adjacent trees is positive or negative.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether \cuttable[their]{the} curvature \cuttable{of two adjacent trees }is positive or negative.
 
@@
 The uniform random walk selects a tree uniformly at random from $N(T)$.
-We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.
-Recall that $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
-For the uniform random walk, $m_T$ is simply the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of the neighbors of $T$.
+\cuttable{We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
+Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
+For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
 
@@
 Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
 \end{proposition}
@@
 \begin{conjecture}
-	Let $k_n$ be the maximum curvature of the uniform random walk on trees of $n$ leaves.
+	Let $k_n$ be the maximum curvature for uniform random walks on trees of $n$ leaves.
 	Then:
@@
 Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
+\cuttable[W]{However, w}e suspect that this will require a \cuttable[new]{greater} understanding of the distribution of tree neighborhoods with respect to one another\cuttable{ than is currently known}.
 Next, we bound the minimum curvature of two adjacent trees.
@@
 \begin{lemma}
-	The curvature of the uniform random walk between two adjacent trees with $n$ leaves is at least
+	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
 	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
@@
 
-Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
-We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.
+Note that this bound is not tight (at least for small $n$)\cuttable{ as it is rarely necessary to transport mass the maximum distance between unpaired trees}.
+\cuttable{We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.}
 
@@
 The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
-As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
+\cuttable[T]{As we now prove, t}hese two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$\cuttable[:]{.}
 
@@
 
-Next we consider the curvature of the Metropolis-Hastings (MH) random walk sampling the uniform distribution on trees, which we will simply call the MH random walk.
-This random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
-The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-The same statement and proof of Lemma~\ref{lem:asymptotic} works with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH random walk curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
-We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the MH random walk $\curvature{\MH; T,S}$.
+We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
+Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
 
@@
 	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }$$
-	$$\curvature{T,S} - \frac{1}{6}
+	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
+	$$\curvature{T,S} - 1/6
 	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + \frac{1}{6}.$$
+	\le \curvature{T,S} + 1/6$$
 \end{lemma}
@@
 However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will significantly reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to ``bottlenecks'' like those we have observed in real data.
-In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
-Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to ``bottlenecks'' like those we have observed in real data.
+\cuttable[T]{In those cases t}he curvature between \cuttable{two }trees at either end of a bottleneck will describe how difficult it is to traverse\cuttable{ the bottleneck}.
+\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
 % We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
 
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees'', and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
 We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
 We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
-Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.
+Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
 

### 2015-04-02 curvature.tex

@@
 The uniform random walk selects a tree uniformly at random from $N(T)$.
-\cuttable{We denote the curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
+\cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
 Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
@@
 We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
-Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
 The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
@@
 
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees'', and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
 We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.

### 2015-04-02 curvature.tex
a few last minor tweaks. Managed to squeeze in: Our results follow from the lemmas of Section~\ref{sec:neighborhoods}, in the curvature intro paragraph.

@@
 In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
-Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree prune-and-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
@@
 
-A (rooted) \emph{subtree prune-and-regraft} (rSPR) operation on an $X$-tree $T$ cuts an edge $e = (x, p_x)$ where $p_x$ denotes the parent of node $x$.
+A (rooted) \emph{subtree-prune-regraft} (rSPR) operation on an $X$-tree $T$ cuts an edge $e = (x, p_x)$ where $p_x$ denotes the parent of node $x$.
 $T$ is divided into two subtrees $T_x$ and $T_{p_x}$ containing $x$ and $p_x$, respectively.
@@
 Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
-Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
-Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
+Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
+Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
 
@@
 
-This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
+This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
@@
 \section{Rooted SPR Neighborhoods}
+\label{sec:neighborhoods}
 Having made the connection between curvature values and access times on the rSPR graphs, we now consider curvature theoretically.
@@
 
-We first consider properties of the uniform (a.k.a.\ isotropic) random walk on the rSPR graph.
-Assume that the random walk begins at a tree $T$.
-The uniform random walk selects a tree uniformly at random from $N(T)$.
+We first consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the rSPR graph.
+Recall that the uniform random walk begins at a tree $T$ and selects a tree uniformly at random from $N(T)$.
 \cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
@@
 For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
+Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
 

### 2015-04-02 curvature.tex
might be good to actually updat the arXiv link before submission
changed a selects to moves to, removed doubled reference used to check arXiv fields

@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
-We omit most proofs which are unnecessary to understand the paper, and which can be found in~\cite{curvature_arxiv}.
+\cuttable[We omit most proofs which are unnecessary to understand the paper, and which can be found in~\cite{curvature_arxiv}.]{}
 
@@
 We first consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the rSPR graph.
-Recall that the uniform random walk begins at a tree $T$ and selects a tree uniformly at random from $N(T)$.
+Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
 \cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}

### 2015-06-25 curvature.tex
first stab at abstract and intro revisions

@@
 \begin{abstract}
-Statistical phylogenetic inference methods use tree rearrangement operations to perform either hill-climbing local search or Markov chain Monte Carlo across tree topologies.
-The canonical class of such moves are the subtree-prune-regraft (SPR) moves that remove a subtree and reattach it somewhere else via the cut edge of the subtree.
-Phylogenetic trees and such moves naturally form the vertices and edges of a graph, such that tree search algorithms perform a (potentially stochastic) traversal of this \emph{SPR graph}.
-Despite the centrality of such graphs in phylogenetic inference, rather little is known about their large-scale properties.
-In this paper we learn about the rooted-tree version of the graph, known as the rSPR graph, by calculating the Ricci-Ollivier curvature for pairs of vertices in the rSPR graph with respect to two simple random walks on the rSPR graph.
-By proving theorems and direct calculation with novel algorithms, we find a remarkable diversity of different curvatures on the rSPR graph for pairs of vertices separated by the same distance.
-We confirm using simulation that degree and curvature have the expected impact on mean access time distributions, demonstrating relevance of these curvature results to stochastic tree search.
-This indicates significant structure of the rSPR graph beyond that which was previously understood in terms of pairwise distances and vertex degrees; a greater understanding of curvature could ultimately lead to improved strategies for tree search.
+Statistical phylogenetic inference methods use subtree-prune-regraft (SPR) tree rearrangement operations to perform Markov chain Monte Carlo across tree topologies.
+These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of  multi-modal posterior distributions induced by such parameters.
+To begin bridging the gap between these models, we developed a framework for computing the Ricci-Ollivier curvature, a formalism of the difficulty in moving between graph nodes via stochastic traversal, of the \emph{rSPR graph} induced by the simple uniform case on rooted trees.
+To do so, we required fast algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+
+We confirmed using simulation that mean access time distributions depend on degree and curvature, demonstrating the relevance of these curvature results to stochastic tree search.
+Surprisingly, we observed negative curvature for some tree pairs, indicating tree space bottlenecks even in this simplest case.
+Finally, we developed a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree, and used this to bound the earth mover's distance between tree neighborhoods and thereby obtain bounds on curvature.
+Our results suggest that bottlenecks are caused by rapid degree changes, which have little effect on mixing times in this well connected space but may be more problematic in real posterior distributions.
+
 \end{abstract}
@@
 Within this framework researchers typically choose either a Bayesian approach, in which an algorithm (typically Markov chain Monte Carlo or MCMC) approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which an algorithm searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
+We focus on the Bayesian approach in this paper, but note that our observations will likely have ramifications for the ML approach as well.
 
-In order to get accurate such estimates, MCMC samplers must sufficiently explore the collection of trees, or heuristic local search algorithms must find a way to the ML tree, avoiding and escaping local peaks.
-In both settings, one needs a means of describing the trees that are ``near'' to the current tree.
-Phylogenetic search algorithms typically explore the collection of trees by rearranging subtrees with subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+In order to get accurate such estimates, MCMC samplers must sufficiently explore the collection of trees.
+Phylogenetic search algorithms typically attempt to do so through a combination of modifications to the continuous parameters and tree topology.
+Topology changes have been identified as the main limiting factor of Bayesian MCMC algorithms~\cite{lakner2008efficiency,hohna2012guided}, as other parameters cannot be accurately estimated if the topology distribution is not accurately sampled.
+Commonly used phylogenetics software packages such as MrBayes~\cite{Ronquist2012-hi} and BEAST~\cite{bouckaert2014beast} rearrange subtrees via subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
@@
 We will call these "graph effects"---one can't easily move from one tree to another due to the structure of the graph.
-Researchers have noticed SPR graph effects in MCMC~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
-We recently showed that graph structure has a significant effect on MCMC mixing applying a popular inference program to real data~\cite{Whidden2015-yi}.
+Researchers have identified theoretical SPR graph effects in MCMC previously~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
 Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
-
-Phylogenetic maximum-likelihood (ML) inference is also impacted by graph structure, which can manifest itself as local maxima.
-There have been investigations of local maxima, though mainly from the perspective of branch lengths rather than topology~\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}.
-The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another~\cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
-We are not aware of any general conclusions concerning hill-climbing on the SPR or related graphs that has come from this work.
+In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
+However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations made it difficult to study such effects.
+By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by "bottlenecks" of specific classes of SPR moves.
 
@@
 $$
-We are not aware of any other work investigating large-scale properties of the SPR graph.
+We are not aware of any work investigating properties of the SPR graph beyond local neighborhoods of trees.
+Again, this stems primarily from a lack of sufficient computational tools.
+Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
+Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\cite{whidden2013hybridization,Whidden2015-yi} that have begun to allow such investigation.
 
-A new approach to quantifying properties of random walks on discrete and continuous spaces has recently been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
+A new approach to quantifying properties of random walks has recently been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
 They define a notion of curvature which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
+For discrete graphs, this is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying a couple of reasonable assumptions, which includes biased walks such as MCMC.
@@
 For this first effort we investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.
+Our algorithms and methods can be easily extended to study more complicated distributions with non-uniform topology probabilities.
+
+We required several new computational tools. 
 We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
-As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph.
+As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph (or, as we do here, the full graph for small numbers of leaves).
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
 By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
-Using this method to simulate these random walks, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
+This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not sample SPR moves uniformly.
+
+Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
+Moreover, we found that rSPR graphs with 6 or more leaves have tree pairs with negative curvature, corresponding to bottlenecks.
+This is a surprising result, as the Markov chain in this simple case is known to mix rapidly~\cite{spade2014note}.
+
+In response, we considered properties of the graph analytically.
+By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
-Finally, we prove that the two related notions of Ricci-Ollivier curvature (coarse and asymptotic) are close, and in fact identical in the limit of many taxa, when applied to the rSPR graph.
+Pairs of trees separated by such moves correspond to the "bottlenecks" identified in our simulation results.
+These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
+However, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}. 
 In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

### 2015-06-25 curvature.tex
a few simplifications in the preliminaries

@@
 Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-Moreover, we found that rSPR graphs with 6 or more leaves have tree pairs with negative curvature, corresponding to bottlenecks.
+Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to bottlenecks.
 This is a surprising result, as the Markov chain in this simple case is known to mix rapidly~\cite{spade2014note}.
@@
 Ricci-Ollivier curvature provides a rigorous yet intuitive formalization of the shape of a metric space with respect to a random walk.
-For the purposes of this paper, we will specialize to that space being a graph equipped with the shortest-path distance, and we will describe curvature intuitively.
+For the purposes of this paper, we will specialize to that space being a graph equipped with the shortest-path distance.
 For a more rigorous presentation in the more general setting of a Polish metric space, see~\cite{Ollivier2009-bw} or the survey~\cite{Ollivier2010-ao}.
 
-Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after a fixed period of time starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
+Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after one iteration of MCMC, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
 The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),$$
-where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$ \cuttable{(more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second)}.
+where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$\cuttable{ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second)}.
 
@@
 \end{equation}
-For the purposes of this paper, we will use $\curvature{x,y}$ to denote the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
-Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are \cuttable{on average }``more distant'' than $x$ and $y$.
+For the purposes of this paper, we will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
+Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are \cuttable{on average }``more distant'' than $x$ and $y$.
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with \cuttable[a]{the given} random walk.
@@
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
-As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
+As we prove in Lemma~\ref{lem:asymptotic}, the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
 

### 2015-07-01 curvature.tex
started reorganizing the computational results.

@@
 \usepackage{graphicx}
+\usepackage{todonotes}
 \usepackage[nomarkers,nofiglist,notablist]{endfloat}
@@
 
-\section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
+\section{Efficient algorithms for computing and sampling the rSPR graph}
 
-\subsection{Computing rSPR graphs}
+\subsection{Computing the rSPR graph on $n$ leaves in $\OhOf{mn^3}$-time}
 \label{sec:computing_treespace}
 
-We computed the rSPR graph explicitly for trees on four to seven leaves.
-This required a fast new algorithm for rSPR graph construction, as previous algorithms required $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph ($(2n-3)!!$ in this case) and $n$ the number of leaves.
+To study the full rSPR graph for a fixed number of leaves it is necessary to have an efficient means of constructing it.
+The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
 Here we reduce that time to $\OhOf{mn^3}$.
+Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement.
+
 In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
@@
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
-However, applying this method to complete rSPR graphs, with the $\OhOf{m^2}$ factor where $m = (2n-3)!!$, requires an enormous amount of computation.
+However, this method is infeasible when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we avoid pairwise computations by successively adding trees to the graph and \cuttable{efficiently }identifying their neighbors:
+To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we avoid pairwise computations by successively adding trees to the graph and \cuttable{efficiently }identifying their neighbors: \todo{make this more concrete and CSy, push the proofs to the supplemental material.}
 \begin{enumerate}[label={\arabic*}.]
@@
 Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
-
-\subsection{Computing curvature values}
-To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
-
-This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
-We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-For example, the pairs
-$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
-$\{(((1,4),2),3), ((1,4),(2,3))\}$
-are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
-
-We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
-Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
-Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
-
-Larger rSPR graphs tended to have smaller curvature values.
-Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
-Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
-	\hspace*{\stretch{2}}
-	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
-	\hspace*{\stretch{1}}
-    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
-Color displays the average degree of $T_1$ and $T_2$.
-Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
-	\label{fig:rspr-scatter}
-\end{figure}
 
@@
 We sampled \cuttable{a }200,000\cuttable{-iteration} random walk\cuttable[ iterations]{} on the 4-leaf rSPR graph and \cuttable{a }50,000-iteration\cuttable[s]{ random walk} on the 5-leaf rSPR graph.
+
+\section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
+
+\subsection{Computing curvature values}
+To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
+
+This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
+We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
+For example, the pairs
+$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
+$\{(((1,4),2),3), ((1,4),(2,3))\}$
+are the same after relabeling, so their curvatures are the same.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
+
+We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
+Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
+Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
+
+Larger rSPR graphs tended to have smaller curvature values.
+Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
+Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{2}}
+	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
+	\hspace*{\stretch{1}}
+    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+Color displays the average degree of $T_1$ and $T_2$.
+Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
+	\label{fig:rspr-scatter}
+\end{figure}
 
@@
 
-\begin{proposition}
+\begin{theorem}
 Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
@@
 Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
-\end{proposition}
+\end{theorem}
 \cuttable{
 \begin{proof}
-Because $d(T_n, S_n) = 1$, we will prove the proposition by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.
+Because $d(T_n, S_n) = 1$, we will prove the theorem by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.
 
-To start we demonstrate the proposition in the case that $T_n$ and $S_n$ have the same number of neighbors.
+To start we demonstrate the theorem in the case that $T_n$ and $S_n$ have the same number of neighbors.
 First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
@@
 
-Now we prove the proposition when the number of neighbors differ.
+Now we prove the theorem when the number of neighbors differ.
 Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
@@
 We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
-The proposition again follows because $\degree{T_n}$ is quadratic in $n$.
+The theorem again follows because $\degree{T_n}$ is quadratic in $n$.
 \end{proof}

### 2015-07-01 curvature.tex
graph construction revisions

@@
 
-\section{Efficient algorithms for computing and sampling the rSPR graph}
+\section{Efficient algorithms for computing and sampling rSPR graphs}
 
-\subsection{Computing the rSPR graph on $n$ leaves in $\OhOf{mn^3}$-time}
+\subsection{Computing the rSPR graph of $m$ trees with $n$ leaves in $\OhOf{mn^3}$-time}
 \label{sec:computing_treespace}
@@
 Here we reduce that time to $\OhOf{mn^3}$.
-Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement.
+Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement in practice, as we demonstrate below.
 
@@
 
-To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we avoid pairwise computations by successively adding trees to the graph and \cuttable{efficiently }identifying their neighbors: \todo{make this more concrete and CSy, push the proofs to the supplemental material.}
+The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
+The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of each tree as a string.
+These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
+We construct a mapping from each tree $T_i$ to its order index in this list $i$.
+We then begin with an empty graph $G$.
+For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
+This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
+We use the tree to index mappings to determine whether these trees already reside in the tree and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
+The high-level steps are as follows, and we show in Theorem~\ref{thm:construct_graph} that this algorithm is correct and can be implemented to run in the stated time.
+
+\textsc{Construct-rSPR-Graph($\mathcal{T}$)}
 \begin{enumerate}[label={\arabic*}.]
 	\item Let $G$ be an empty graph.
-	\item For each of the $m$ trees in a specified but arbitrary order:
+	\item Let $M$ be a mapping from trees to integers.
+	\item Let $i = 0$.
+	\item For each of the $m$ trees:
 		\begin{enumerate}
-			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index.
-			\item For each of the $\OhOf{n^2}$ neighbors of $T$:
+			\item Add a vertex $i$ to $G$ representing the current tree $T_i$.
+			\item Add $T_i \rightarrow i$ to $M$.
+			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}:
 				\begin{enumerate}
-					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index.
+					\item If the current neighbor $T_j$ is in $M$ then add an edge $(i,M[T_j])$ to $G$.
 				\end{enumerate}
+		\item $i = i + 1$.
 		\end{enumerate}
@@
 	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
-	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
+	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
 	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
-	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
+	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using Newick representation.
 	This requires only $\OhOf{n}$-time for each tree (i.e.
 a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
-	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
-	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
+	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
+	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
 \end{proof}
 
-This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
+We implemented this procedure in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
 The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
@@
 Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
-Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.
+Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}, as the algorithm can construct rSPR graphs for any given subset of trees.
 
@@
 \begin{corollary}
+\label{cor:enumerate_neighbors}
 	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.

### 2015-07-02 curvature.tex
minor revisions to random walk generation

@@
 \label{sec:random_walks}
-We next explored the correspondence between curvature and graph effects.
-To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
 The uniform random walk moves from the current vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
@@
 \end{enumerate}
-We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive.
+We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
 We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
@@
 (iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e.
-iv above), a contradiction.
+(iv) above), a contradiction.
 	Therefore the claim holds.
@@
 
-In particular, this formulation groups moves that move the same subtree.
+In particular, this formulation implies a total ordering of rSPR moves such thatevery move moving the same subtree $R$ forms a contiguous subsequence.
 We can thus apply the following algorithm to select a neighbor uniformly at random for a tree $T$:
 
+\vspace{1em}
+\textsc{Select-rSPR-Neighbor($T$)}
 \begin{enumerate}[label={\arabic*}.]
-	\item	Compute the degree of $T$, $\degree{T}$.
+	\item	Compute the degree of $T$, $\degree{T}$ using Lemma~\ref{lem:compute_degree}.
 	\item Pick a random integer $r$ in the range $[1,\degree{T}]$.
@@
 Observe that this procedure can be easily adapted to explore the full neighborhood of a tree in $\OhOf{n^3}$ time, which we use for Theorem~\ref{thm:construct_graph}.
+We call the resulting procedure \textsc{Enumerate-rSPR-Neighbors($T$)}.
 We thus have the following corollary:

### 2015-07-02 curvature.tex
pass through degree and curvature sections

@@
 
-In particular, this formulation implies a total ordering of rSPR moves such thatevery move moving the same subtree $R$ forms a contiguous subsequence.
+In particular, this formulation implies a total ordering of rSPR moves such that every move moving the same subtree $R$ forms a contiguous subsequence.
 We can thus apply the following algorithm to select a neighbor uniformly at random for a tree $T$:
@@
 \label{sec:neighborhoods}
-Having made the connection between curvature values and access times on the rSPR graphs, we now consider curvature theoretically.
-We begin by bounding differences between degrees, and then continue by enumerating squares and triangles in the rSPR graph.
+Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
+We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call "squares" and "triangles" in the rSPR graph.
+Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.
 
@@
 \begin{lemma}
+\label{lem:curvature_lower_bound}
 	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
@@
 \cuttable{We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.}
+However, we note that the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
 
@@
 However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to ``bottlenecks'' like those we have observed in real data.
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to significant ``bottlenecks'' like those we have observed in real data.
 \cuttable[T]{In those cases t}he curvature between \cuttable{two }trees at either end of a bottleneck will describe how difficult it is to traverse\cuttable{ the bottleneck}.

### 2015-07-02 curvature.tex
moved proofs to an appendix

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{thmtools}
+\usepackage{thm-restate}
 
@@
 
-% uncomment to remove the proofs for length estimates
-%\usepackage{environ}
-%\NewEnviron{killcontents}{}
-%\let\proof\killcontents
-%\let\endproof\endkillcontents
-
 \begin{document}
@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
-\cuttable[We omit most proofs which are unnecessary to understand the paper, and which can be found in~\cite{curvature_arxiv}.]{}
+\cuttable[Proofs of our theorems and lemmas can be found in the appendix.]
 
@@
 
-\begin{theorem}
+%\begin{theorem}
+%	\label{thm:construct_graph}
+%	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
+%\end{theorem}
+\begin{restatable}{theorem}{constructgraph}
 	\label{thm:construct_graph}
 	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
-\end{theorem}
-\begin{proof}
-	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
-
-	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
-red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
-	As described above, the integer labels are simply the order of the input trees.
-	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
-	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
-	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
-	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using Newick representation.
-	This requires only $\OhOf{n}$-time for each tree (i.e.
-a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
-	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
-	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
-\end{proof}
+\end{restatable}
 
@@
 
-\begin{lemma}
+\begin{restatable}{lemma}{computedegree}
 	\label{lem:compute_degree}
@@
 	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation moving the subtree rooted at $u$.
-\end{lemma}
-\cuttable{
-\begin{proof}
-	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
-	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
-	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
-	Call these $U$ and $V$, respectively.
-	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
-	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
-	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
-(iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e.
-(iv) above), a contradiction.
-	Therefore the claim holds.
-\end{proof}
-}
+\end{restatable}
 
@@
 
-\begin{lemma}
+\begin{restatable}{lemma}{selectrandomneighbor}
 	\label{lem:select_random_neighbor}
 	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time\cuttable{ using $\OhOf{n}$ space}.
+\end{restatable}
+
+Observe that this procedure can be easily adapted to explore the full neighborhood of a tree in $\OhOf{n^3}$ time, which we use for Theorem~\ref{thm:construct_graph}.
+We call the resulting procedure \textsc{Enumerate-rSPR-Neighbors($T$)}.
+We thus have the following corollary:
+
+\begin{corollary}
+\label{cor:enumerate_neighbors}
+	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
+\end{corollary}
+
+We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
+We sampled \cuttable{a }200,000\cuttable{-iteration} random walk\cuttable[ iterations]{} on the 4-leaf rSPR graph and \cuttable{a }50,000-iteration\cuttable[s]{ random walk} on the 5-leaf rSPR graph.
+
+\section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
+
+\subsection{Computing curvature values}
+To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
+
+This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
+We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
+For example, the pairs
+$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
+$\{(((1,4),2),3), ((1,4),(2,3))\}$
+are the same after relabeling, so their curvatures are the same.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
+
+We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
+Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
+Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
+
+Larger rSPR graphs tended to have smaller curvature values.
+Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
+Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{2}}
+	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
+	\hspace*{\stretch{1}}
+    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+Color displays the average degree of $T_1$ and $T_2$.
+Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
+	\label{fig:rspr-scatter}
+\end{figure}
+
+\subsection{Access time simulation}
+The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
+In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks: the mean number of iterations required to move from one tree to the other.
+We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
+
+Here, to gain more insight, we used simulation to approximate the entire access time distribution.
+As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
+Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
+Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using \cuttable{our }group-theoretic methods \cite{tangle}.
+
+\begin{table}
+\centering
+\caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
+\label{tab:regressionMAT}
+\end{table}
+
+We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
+Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
+We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
+For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines\cuttable{ in Fig.~\ref{fig:kappa-access}}) is the count for time 2 minus the count for time 3.
+Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
+This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
+On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$\cuttable{ and instead achieves stationarity}, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
+	\hspace*{\stretch{1}}
+    \caption{\
+        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        }
+	\label{fig:kappa-access}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
+\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
+\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
+\label{tab:regressionDelta}
+\end{table}
+
+The analysis can be reproduced by invoking the SCons (\url{http://scons.org/}) build tool and running the cells in an IPython notebook; instructions are in the repository README file.
+
+\section{Rooted SPR Neighborhoods}
+\label{sec:neighborhoods}
+Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
+We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call "squares" and "triangles" in the rSPR graph.
+Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.
+
+\begin{lemma}[{Song~\cite{Song2003-gf}}]
+\cuttable{    \pushQED{\qed}}
+	\label{lem:degree_extremes}
+	Let $T$ be a rooted phylogenetic tree with $n$ leaves. Then
+	\begin{enumerate}
+		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree,
+		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
+		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
+	\end{enumerate}
+\cuttable{    \popQED}
 \end{lemma}
+
+We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
+\begin{restatable}{lemma}{degreemaxdelta}
+	\label{lem:degree_max_delta}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
+	Then:
+	\begin{enumerate}
+		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
+		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
+	\end{enumerate}
+\end{restatable}
+
+We can improve these bounds in the case of adjacent trees.
+To do so, we \cuttable{require the following lemma that }characterize\cuttable{s} exactly how the degree of a tree changes after an rSPR operation.
+See Figure~\ref{fig:spr-degree-change} for an illustration of the Lemma.
+
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\includegraphics[width=\textwidth]{figs/spr-degree-change}
+	\hspace*{\stretch{1}}
+	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
+Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
+	\label{fig:spr-degree-change}
+\end{figure}
+
+\begin{restatable}{lemma}{degreechange}
+\label{lem:degree_change}
+Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+Let $L$ be the $\lca{U,V}$ in $T$.
+Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
+Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
+Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
+Then the degrees of $T$ and $S$ differ by:
+$$2\left(k(a-b) + i - j\right).$$
+
+\end{restatable}
+
+Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
+That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
+To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
+The number of such ``squares'' involving two adjacent trees will play a key role in our bounds on curvature, as they push the curvature of those trees towards 0.
+
+\begin{restatable}{corollary}{pairedneighbors}
+\label{cor:paired_neighbors}
+Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
+$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
+trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
+\end{restatable}
+
+We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
+\begin{restatable}{lemma}{degreemaxdeltaadjacent}
+	\label{lem:degree_max_delta_adjacent}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
+	Then:
+	\begin{enumerate}
+		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
+		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
+		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
+	\end{enumerate}
+\end{restatable}
+
+Next, we bound the number of neighbors shared by two adjacent trees.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether \cuttable[their]{the} curvature \cuttable{of two adjacent trees }is positive or negative.
+
+\begin{restatable}{lemma}{sharedneighbors}
+	\label{lem:shared_neighbors}
+Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
+Then $\size{N(T) \cap N(S)} \le 6n - 17$.
+\end{restatable}
+
+\section{Curvature}
+For the purposes of this paper, $\kappa$ is the coarse Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
+
+We first consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the rSPR graph.
+Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
+\cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
+Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
+For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
+Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
+
+\begin{restatable}{theorem}{flatcurvature}
+\label{thm:flat_curvature}
+Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
+Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
+\end{restatable}
+
+Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
+
+\begin{restatable}{lemma}{curvaturedistancebound}
+	\label{lem:curvature_distance_bound}
+	Let $T$ and $S$ be two trees. Then:
+	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
+\end{restatable}
+
+\begin{restatable}{lemma}{maxadjacentcurvature}
+	\label{lem:max_adjacent_curvature}
+	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
+\end{restatable}
+This bound is tight and has been verified computationally for $n \le 7$.
+
+It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
+Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
+However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
+Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
+Moreover, nearly all of the neighbors of these pairs approach each either.
+We thus conjecture the following:
+\begin{conjecture}
+	Let $k_n$ be the maximum curvature for uniform random walks on trees of $n$ leaves.
+	Then:
+	\begin{enumerate}
+		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
+		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
+	\end{enumerate}
+\end{conjecture}
+Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
+\cuttable[W]{However, w}e suspect that this will require a \cuttable[new]{greater} understanding of the distribution of tree neighborhoods with respect to one another\cuttable{ than is currently known}.
+Next, we bound the minimum curvature of two adjacent trees.
+
+\begin{restatable}{lemma}{curvaturelowerbound}
+\label{lem:curvature_lower_bound}
+	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
+	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
+\end{restatable}
+
+We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
+Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
+This directly leads to the following Corollary:
+
+\begin{corollary}
+	The curvature of the uniform random walk between two phylogenetic trees is at least $-\frac{2}{5}$.
+\end{corollary}
+
+Note that this bound is not tight (at least for small $n$)\cuttable{ as it is rarely necessary to transport mass the maximum distance between unpaired trees}.
+\cuttable{We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.}
+However, we note that the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
+
+We next bound the difference between the coarse and asymptotic curvatures.
+Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
+For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
+The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
+\cuttable[T]{As we now prove, t}hese two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$\cuttable[:]{.}
+
+\begin{restatable}{lemma}{asymptotic}
+\label{lem:asymptotic}
+	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
+	Then:
+	\begin{enumerate}
+		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
+		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+	\end{enumerate}
+\end{restatable}
+
+We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
+Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
+
+\begin{restatable}{lemma}{mhcurvature}
+	\label{lem:mh_curvature}
+	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
+	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
+	\le \curvature{\MH; T,S}
+	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
+	$$\curvature{T,S} - 1/6
+	\le \curvature{\MH; T,S}
+	\le \curvature{T,S} + 1/6$$
+\end{restatable}
+
+\section{Conclusion and future work}
+In summary, we have advanced understanding of the phylogenetic rSPR graph substantially beyond what was previously known, which concerned graph diameter and vertex degree.
+We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
+We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
+
+In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
+Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to significant ``bottlenecks'' like those we have observed in real data.
+\cuttable[T]{In those cases t}he curvature between \cuttable{two }trees at either end of a bottleneck will describe how difficult it is to traverse\cuttable{ the bottleneck}.
+\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
+% We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
+
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
+We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
+Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
+
+\section{Acknowledgements}
+The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
+They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
+This work was funded by National Science Foundation award 1223057.
+
+\pagebreak
+
+\bibliographystyle{IEEEtran}
+\bibliography{curvature}
+
+\pagebreak
+
+\appendices
+
+\section{Supplementary Proofs}
+
+\constructgraph*
+\begin{proof}
+	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
+
+	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
+red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
+	As described above, the integer labels are simply the order of the input trees.
+	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
+	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
+	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
+	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
+	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using Newick representation.
+	This requires only $\OhOf{n}$-time for each tree (i.e.
+a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
+	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
+	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
+\end{proof}
+
+\computedegree*
+\begin{proof}
+	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
+	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
+	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
+	Call these $U$ and $V$, respectively.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
+	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
+	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
+(iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e.
+(iv) above), a contradiction.
+	Therefore the claim holds.
+\end{proof}
+
+\selectrandomneighbor*
 \begin{proof}
@@
 
-Observe that this procedure can be easily adapted to explore the full neighborhood of a tree in $\OhOf{n^3}$ time, which we use for Theorem~\ref{thm:construct_graph}.
-We call the resulting procedure \textsc{Enumerate-rSPR-Neighbors($T$)}.
-We thus have the following corollary:
-
-\begin{corollary}
-\label{cor:enumerate_neighbors}
-	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
-\end{corollary}
-
-We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
-We sampled \cuttable{a }200,000\cuttable{-iteration} random walk\cuttable[ iterations]{} on the 4-leaf rSPR graph and \cuttable{a }50,000-iteration\cuttable[s]{ random walk} on the 5-leaf rSPR graph.
-
-\section{Access times of random walks on the rSPR graph can be understood using distance, degree, and curvature}
-
-\subsection{Computing curvature values}
-To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
-
-This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
-We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
-For example, the pairs
-$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
-$\{(((1,4),2),3), ((1,4),(2,3))\}$
-are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
-
-We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
-Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
-Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
-
-Larger rSPR graphs tended to have smaller curvature values.
-Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
-Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
-	\hspace*{\stretch{2}}
-	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
-	\hspace*{\stretch{1}}
-    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
-Color displays the average degree of $T_1$ and $T_2$.
-Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
-	\label{fig:rspr-scatter}
-\end{figure}
-
-\subsection{Access time simulation}
-The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
-In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks: the mean number of iterations required to move from one tree to the other.
-We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
-
-Here, to gain more insight, we used simulation to approximate the entire access time distribution.
-As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
-Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
-Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
-To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using \cuttable{our }group-theoretic methods \cite{tangle}.
-
-\begin{table}
-\centering
-\caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
-\label{tab:regressionMAT}
-\end{table}
-
-We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
-Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
-We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
-For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines\cuttable{ in Fig.~\ref{fig:kappa-access}}) is the count for time 2 minus the count for time 3.
-Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
-This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
-On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$\cuttable{ and instead achieves stationarity}, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
-	\hspace*{\stretch{1}}
-    \caption{\
-        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
-        }
-	\label{fig:kappa-access}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
-\label{tab:regressionDelta}
-\end{table}
-
-The analysis can be reproduced by invoking the SCons (\url{http://scons.org/}) build tool and running the cells in an IPython notebook; instructions are in the repository README file.
-
-\section{Rooted SPR Neighborhoods}
-\label{sec:neighborhoods}
-Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
-We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call "squares" and "triangles" in the rSPR graph.
-Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.
-
-\begin{lemma}[{Song~\cite{Song2003-gf}}]
-\cuttable{    \pushQED{\qed}}
-	\label{lem:degree_extremes}
-	Let $T$ be a rooted phylogenetic tree with $n$ leaves. Then
-	\begin{enumerate}
-		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree,
-		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
-		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
-	\end{enumerate}
-\cuttable{    \popQED}
-\end{lemma}
-
-We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
-\begin{lemma}
-	\label{lem:degree_max_delta}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
-	Then:
-	\begin{enumerate}
-		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
-		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
-	\end{enumerate}
-\end{lemma}
-\cuttable{
+\degreemaxdelta*
 \begin{proof}
@@
 \end{proof}
-}
-
-We can improve these bounds in the case of adjacent trees.
-To do so, we \cuttable{require the following lemma that }characterize\cuttable{s} exactly how the degree of a tree changes after an rSPR operation.
-See Figure~\ref{fig:spr-degree-change} for an illustration of the Lemma.
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\includegraphics[width=\textwidth]{figs/spr-degree-change}
-	\hspace*{\stretch{1}}
-	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
-Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
-	\label{fig:spr-degree-change}
-\end{figure}
-
-\begin{lemma}
-\label{lem:degree_change}
-Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
-Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
-Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
-Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
-Then the degrees of $T$ and $S$ differ by:
-$$2\left(k(a-b) + i - j\right).$$
-
-\end{lemma}
+
+\degreechange*
 \begin{proof}
@@
 
-Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
-That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
-To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
-The number of such ``squares'' involving two adjacent trees will play a key role in our bounds on curvature, as they push the curvature of those trees towards 0.
-
-\begin{corollary}
-\label{cor:paired_neighbors}
-Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
-$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
-trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
-\end{corollary}
-\cuttable{
+\pairedneighbors*
 \begin{proof}
@@
 \end{proof}
-}
-
-We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
-\begin{lemma}
-	\label{lem:degree_max_delta_adjacent}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
-	Then:
-	\begin{enumerate}
-		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
-		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
-		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
-	\end{enumerate}
-\end{lemma}
-\cuttable{
+
+\degreemaxdeltaadjacent
 \begin{proof}
@@
 \end{proof}
-}
-
-Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such ``triangles'' involving two adjacent trees has a key role in determining whether \cuttable[their]{the} curvature \cuttable{of two adjacent trees }is positive or negative.
-
-\begin{lemma}
-	\label{lem:shared_neighbors}
-Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
-Then $\size{N(T) \cap N(S)} \le 6n - 17$.
-\end{lemma}
-
-\cuttable{
+
+\sharedneighbors*
 \begin{proof}
@@
 \end{proof}
-}
-
-\section{Curvature}
-For the purposes of this paper, $\kappa$ is the coarse Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
-
-We first consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the rSPR graph.
-Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
-\cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
-Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
-For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
-Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
-
-\begin{theorem}
-Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
-Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
-\end{theorem}
-\cuttable{
+
+\flatcurvature*
 \begin{proof}
@@
 \end{proof}
-}
-
-Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
-
-\begin{lemma}
-	\label{lem:curvature_distance_bound}
-	Let $T$ and $S$ be two trees. Then:
-	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
-\end{lemma}
-\cuttable{
+
+\curvaturedistancebound
 \begin{proof}
@@
 \end{proof}
-}
-
-\begin{lemma}
-	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
-\end{lemma}
-\cuttable{
+
+\maxadjacentcurvature*
 \begin{proof}
@@
 \end{proof}
-}
-This bound is tight and has been verified computationally for $n \le 7$.
-
-It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
-Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
-However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
-Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
-Moreover, nearly all of the neighbors of these pairs approach each either.
-We thus conjecture the following:
-\begin{conjecture}
-	Let $k_n$ be the maximum curvature for uniform random walks on trees of $n$ leaves.
-	Then:
-	\begin{enumerate}
-		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
-		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
-	\end{enumerate}
-\end{conjecture}
-Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-\cuttable[W]{However, w}e suspect that this will require a \cuttable[new]{greater} understanding of the distribution of tree neighborhoods with respect to one another\cuttable{ than is currently known}.
-Next, we bound the minimum curvature of two adjacent trees.
-
-\begin{lemma}
-\label{lem:curvature_lower_bound}
-	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
-	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
-\end{lemma}
-\cuttable{
+
+\curvaturelowerbound*
 \begin{proof}
@@
 \end{proof}
-}
-
-We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
-Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
-This directly leads to the following Corollary:
-
-\begin{corollary}
-	The curvature of the uniform random walk between two phylogenetic trees is at least $-\frac{2}{5}$.
-\end{corollary}
-
-Note that this bound is not tight (at least for small $n$)\cuttable{ as it is rarely necessary to transport mass the maximum distance between unpaired trees}.
-\cuttable{We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.}
-However, we note that the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
-
-We next bound the difference between the coarse and asymptotic curvatures.
-Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
-For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
-\cuttable[T]{As we now prove, t}hese two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$\cuttable[:]{.}
-
-\begin{lemma}
-\label{lem:asymptotic}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
-	Then:
-	\begin{enumerate}
-		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
-		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
-	\end{enumerate}
-\end{lemma}
-
-\cuttable{
+
+\asymptotic*
 \begin{proof}
@@
 \end{proof}
-}
-
-We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
-Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
-The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
-
-\begin{lemma}
-	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
-	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
-	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
-	$$\curvature{T,S} - 1/6
-	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + 1/6$$
-\end{lemma}
-\cuttable{
+
+\mhcurvature*
 \begin{proof}
@@
 \end{proof}
-}
-
-\section{Conclusion and future work}
-In summary, we have advanced understanding of the phylogenetic rSPR graph substantially beyond what was previously known, which concerned graph diameter and vertex degree.
-We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
-We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
-Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
-
-In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
-Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
-However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to significant ``bottlenecks'' like those we have observed in real data.
-\cuttable[T]{In those cases t}he curvature between \cuttable{two }trees at either end of a bottleneck will describe how difficult it is to traverse\cuttable{ the bottleneck}.
-\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
-% We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
-
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
-We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
-We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
-Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
-
-\section{Acknowledgements}
-The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
-They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
-This work was funded by National Science Foundation award 1223057.
-
-\pagebreak
-
-\bibliographystyle{IEEEtran}
-\bibliography{curvature}
+
 \end{document}

### 2015-07-06 curvature.tex
minor edits in intro

@@
 \section{Introduction}
-Molecular phylogenetic methods, which reconstruct evolutionary trees from DNA or RNA data, are of fundamental importance to modern biology.
+Molecular phylogenetic methods reconstruct evolutionary trees from DNA or RNA data and are of fundamental importance to modern biology.
 Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
@@
 Thus likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
-Within this framework researchers typically choose either a Bayesian approach, in which an algorithm (typically Markov chain Monte Carlo or MCMC) approximates the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, in which an algorithm searches across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
-We focus on the Bayesian approach in this paper, but note that our observations will likely have ramifications for the ML approach as well.
-
-In order to get accurate such estimates, MCMC samplers must sufficiently explore the collection of trees.
+Within this framework researchers typically choose either a Bayesian approach, using an algorithm such as Markov chain Monte Carlo (MCMC) to approximate the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, searching across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
+We focus on the Bayesian approach in this paper, noting that our observations will likely also have ramifications for ML.
+
+In order to estimate these distributions accurately, MCMC samplers must sufficiently explore the collection of trees.
 Phylogenetic search algorithms typically attempt to do so through a combination of modifications to the continuous parameters and tree topology.
@@
 In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
-However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations made it difficult to study such effects.
+However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations has made it difficult to study graph effects.
 By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by "bottlenecks" of specific classes of SPR moves.
@@
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
-Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\cite{whidden2013hybridization,Whidden2015-yi} that have begun to allow such investigation.
-
-A new approach to quantifying properties of random walks has recently been pioneered by Ollivier and colleagues~\cite{Ollivier2009-bw,Joulin2010-jg}.
-They define a notion of curvature which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
+Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\cite{whidden2013hybridization,Whidden2015-yi} and begun to allow such investigation.
+
+Ollivier and colleagues recently pioneered a new approach to quantifying properties of random walks~\cite{Ollivier2009-bw,Joulin2010-jg}.
+They define a notion of \emph{curvature} which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
 For discrete graphs, this is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
@@
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
-In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
+In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
 Pairs of trees separated by such moves correspond to the "bottlenecks" identified in our simulation results.

### 2015-07-06 curvature.tex
back down to 10 pages

@@
 	\item Let $i = 0$.
-	\item For each of the $m$ trees:
+	\item For each of the $m$ trees: \vspace{-0.2em}
 		\begin{enumerate}
@@
 			\item Add $T_i \rightarrow i$ to $M$.
-			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}:
+			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}: \vspace{-1.4em}
 				\begin{enumerate}
@@
 Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
-Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}, as the algorithm can construct rSPR graphs for any given subset of trees.
+Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
 
@@
 \label{sec:random_walks}
-The uniform random walk moves from the current vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
+The uniform random walk moves from one vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
 In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
@@
 We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
-We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
-Call the portion of the neighborhood of a tree $T$ that can be obtained by moving a subtree $R$ rooted at a node $u$, and that is assigned to $u$, $N(T,u)$.
-In this manner, we achieve an alternative solution for computing the degree of a tree in the rSPR graph:
+We can thus assign $(2n-1-x)-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $(2n-1-x)-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
+Let $N(T,u)$ denote the neighbors of $T$ assigned to node $u$, obtained by moving the subtree $R$ rooted at $u$.
+We thus achieve a new method for computing the neighborhood size:
 
@@
 	$$\size{N(T,u)} = \begin{cases}
-		2n - 1 - x - 4 &\text{if depth($u$) $ > 1$, } \\
-		2n - 1 - x - 2 &\text{if depth($u$) $ = 1$ } \\
+		2n - x - 5 &\text{if depth($u$) $ > 1$, } \\
+		2n - x - 3 &\text{if depth($u$) $ = 1$ } \\
 		0 &\text{if depth($u$) $ \le 0$ } \\
 	\end{cases}.$$
-	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation moving the subtree rooted at $u$.
 \end{restatable}
@@
 The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
-In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks: the mean number of iterations required to move from one tree to the other.
+In previous work, we computed mean access times (MAT) between pairs of trees in  MCMC random walks: the mean number of iterations required to move from one tree to the other.
 We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
@@
 Here, to gain more insight, we used simulation to approximate the entire access time distribution.
-As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
 Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
-Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
+Thus rather than enumerate access times between trees, which would have required a tremendous amount of memory and computational power to obtain accurate estimates, we enumerate times between pairs of trees in a tanglegram.
 To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using \cuttable{our }group-theoretic methods \cite{tangle}.
@@
 	\label{lem:degree_extremes}
-	Let $T$ be a rooted phylogenetic tree with $n$ leaves. Then
+	Let $T$ be a tree with $n$ leaves. Then
 	\begin{enumerate}
@@
 	\label{lem:degree_max_delta}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
+	Let $T$ and $S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
 	Then:
@@
 \label{lem:degree_change}
-Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
 Let $L$ be the $\lca{U,V}$ in $T$.
@@
 	\label{lem:degree_max_delta_adjacent}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
+	Let $T$ and $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
 	Then:
@@
 \section{Curvature}
-For the purposes of this paper, $\kappa$ is the coarse Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.
-
-We first consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the rSPR graph.
+We now consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the $n$-leaf rSPR graph.
 Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
 \cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
-Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$.
+Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
 For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
@@
 Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
-Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
+Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
 Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
@@
 	\label{lem:max_adjacent_curvature}
-	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
+	The maximum curvature between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
 \end{restatable}
@@
 \begin{conjecture}
-	Let $k_n$ be the maximum curvature for uniform random walks on trees of $n$ leaves.
+	Let $k_n$ be the maximum curvature of the uniform random walk on $n$-leaf trees.
 	Then:
@@
 \begin{corollary}
-	The curvature of the uniform random walk between two phylogenetic trees is at least $-\frac{2}{5}$.
+	The curvature of the uniform random walk between two trees is at least $-\frac{2}{5}$.
 \end{corollary}
@@
 \label{lem:asymptotic}
-	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
+	Let $T$ and $S$ be trees with $n$ leaves.
 	Then:
@@
 	\label{lem:mh_curvature}
-	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
+	Let $T$ and $S$ be trees with $n$ leaves. Then:
 	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}

### 2015-07-07 curvature.tex
minor edits

@@
 By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
-This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not sample SPR moves uniformly.
+This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not propose SPR moves uniformly.
 
@@
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
-However, this method is infeasible when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.
+However, this method is impractical when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.
 
@@
 
-We now bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
+Finally, we bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
 Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.

### 2015-07-07 curvature.tex
Tidying up quotes (cc @cwhidden)

@@
 It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
-We will call these "graph effects"---one can't easily move from one tree to another due to the structure of the graph.
+We will call these ``graph effects"---one can't easily move from one tree to another due to the structure of the graph.
 Researchers have identified theoretical SPR graph effects in MCMC previously~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
@@
 However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations has made it difficult to study graph effects.
-By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by "bottlenecks" of specific classes of SPR moves.
+By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by ``bottlenecks" of specific classes of SPR moves.
 
@@
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
-Pairs of trees separated by such moves correspond to the "bottlenecks" identified in our simulation results.
+Pairs of trees separated by such moves correspond to the ``bottlenecks" identified in our simulation results.
 These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
@@
 Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
-We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call "squares" and "triangles" in the rSPR graph.
+We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call ``squares" and ``triangles" in the rSPR graph.
 Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.

### 2015-07-07 curvature.tex
Reworked abstract.

@@
 \begin{abstract}
-Statistical phylogenetic inference methods use subtree-prune-regraft (SPR) tree rearrangement operations to perform Markov chain Monte Carlo across tree topologies.
+Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
 These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions induced by such parameters.
-To begin bridging the gap between these models, we developed a framework for computing the Ricci-Ollivier curvature, a formalism of the difficulty in moving between graph nodes via stochastic traversal, of the \emph{rSPR graph} induced by the simple uniform case on rooted trees.
-To do so, we required fast algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
-
-We confirmed using simulation that mean access time distributions depend on degree and curvature, demonstrating the relevance of these curvature results to stochastic tree search.
-Surprisingly, we observed negative curvature for some tree pairs, indicating tree space bottlenecks even in this simplest case.
-Finally, we developed a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree, and used this to bound the earth mover's distance between tree neighborhoods and thereby obtain bounds on curvature.
-Our results suggest that bottlenecks are caused by rapid degree changes, which have little effect on mixing times in this well connected space but may be more problematic in real posterior distributions.
+The structure of the underlying graph is an important determinant of the mixing properties of MCMC, and as such we are motivated to characterize the structure of this graph in greater detail.
+To do so, we calculated the Ricci-Ollivier curvature of the \emph{rSPR graph} for all pairs of rooted trees up to seven leaves with respect to two simple random walks.
+We found that the curvature varies substantially across pairs of trees in the graph separated by the same distance, indicating significant structure to the graph beyond that seen by pairwise distance.
+This required fast algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+We also developed a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
+This gave bounds on the curvature, as well as a flatness-in-the-limit theorem for pairs of trees differentiated by moving an asymptotically small subtree.
+We also confirmed using simulation that mean access time distributions depend on distance, degree, and curvature, confirming the relevance of these curvature results to stochastic tree search.
 

### 2015-07-07 curvature.tex
Incorporating Chris' comments and then some.

@@
 These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions induced by such parameters.
-The structure of the underlying graph is an important determinant of the mixing properties of MCMC, and as such we are motivated to characterize the structure of this graph in greater detail.
+The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC.
+Indeed, topology changes, and SPR rearrangements in particular, have been identified as a primary limiting factor of phylogenetic MCMC algorithms.
+As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the \emph{rSPR graph} in greater detail.
 To do so, we calculated the Ricci-Ollivier curvature of the \emph{rSPR graph} for all pairs of rooted trees up to seven leaves with respect to two simple random walks.
@@
 The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
-Thus likelihood-based phylogenetics gives an optimality criterion that comes from both branch length and topology.
-Within this framework researchers typically choose either a Bayesian approach, using an algorithm such as Markov chain Monte Carlo (MCMC) to approximate the posterior distribution of trees and their associated parameters, or a maximum likelihood (ML) approach, searching across tree topologies and continuous parameters in order to find the combination with the highest likelihood.
-We focus on the Bayesian approach in this paper, noting that our observations will likely also have ramifications for ML.
+It is now common for researchers to approximate the posterior distribution of trees and their associated parameters in a Bayesian setting using an algorithm such as Markov chain Monte Carlo (MCMC).
 
@@
 Commonly used phylogenetics software packages such as MrBayes~\cite{Ronquist2012-hi} and BEAST~\cite{bouckaert2014beast} rearrange subtrees via subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
-Thus, phylogenetic searches can be viewed as traversing the graph of phylogenetic trees induced by SPR adjacencies---the \emph{SPR graph}.
+Thus, phylogenetic searches can be viewed as traversing the \emph{SPR graph}: the graph with phylogenetic trees as vertices and SPR adjacencies as edges.
 
 It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
-We will call these ``graph effects"---one can't easily move from one tree to another due to the structure of the graph.
-Researchers have identified theoretical SPR graph effects in MCMC previously~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv,Stefankovic2011-hu}.
-Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves\cite{Evans2006-xh,Athreya2014-de}.
+We will call these ``graph effects", whereby one can't easily move from one tree to another due to the structure of the graph.
+Researchers have previously identified SPR graph effects in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}, contrasting with fast mixing identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu}.
+Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
 In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
@@
 
-Although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
-\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
-\cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
+Although the SPR graph is thus very important in determining the success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself.
+\cite{Song2003-gf}~developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
+\cite{Ding2011-bj}~showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
 $$ n - 2\ceil{\sqrt{n}} + 1
@@
 $$
-We are not aware of any work investigating properties of the SPR graph beyond local neighborhoods of trees.
-Again, this stems primarily from a lack of sufficient computational tools.
+We are not aware of any further work investigating properties of the SPR graph, which this stems primarily from a lack of sufficient computational tools.
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
@@
 
-Ollivier and colleagues recently pioneered a new approach to quantifying properties of random walks~\cite{Ollivier2009-bw,Joulin2010-jg}.
-They define a notion of \emph{curvature} which formalizes the notion of to what extent random walking brings points together, and is a generalization of the classical notion of Ricci curvature on manifolds.
+Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\cite{Ollivier2009-bw,Joulin2010-jg}.
+In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of \emph{curvature} formalizes the notion of to what extent random walking brings points together.
+Applying the framework to Brownian motion on a manifold returns us back of the classical definition of Ricci curvature.
 For discrete graphs, this is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.

### 2015-07-07 curvature.tex
Clearer "in this paper" sentence and present tense.

@@
 As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the \emph{rSPR graph} in greater detail.
-To do so, we calculated the Ricci-Ollivier curvature of the \emph{rSPR graph} for all pairs of rooted trees up to seven leaves with respect to two simple random walks.
-We found that the curvature varies substantially across pairs of trees in the graph separated by the same distance, indicating significant structure to the graph beyond that seen by pairwise distance.
-This required fast algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
-We also developed a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
-This gave bounds on the curvature, as well as a flatness-in-the-limit theorem for pairs of trees differentiated by moving an asymptotically small subtree.
-We also confirmed using simulation that mean access time distributions depend on distance, degree, and curvature, confirming the relevance of these curvature results to stochastic tree search.
-
+
+In this paper, we investigate the rSPR graph in a new way: by calculating Ricci-Ollivier curvature with respect to two simple random walks.
+We exactly calculate the curvature for all pairs of rooted trees up to seven leaves, finding that the curvature varies substantially across pairs of trees in the graph separated by the same distance, thus indicating significant structure to the graph beyond that seen by pairwise distance.
+This calculation requires fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+We also develop a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
+This gives bounds on the curvature, as well as a flatness-in-the-limit theorem for pairs of trees differentiated by moving an asymptotically small subtree.
+We also confirm using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
 \end{abstract}

### 2015-07-07 curvature.tex
a stab at revising the end of the abstract

--- 2015-07-07 curvature.tex
@@
 As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the \emph{rSPR graph} in greater detail.
-To do so, we calculated the Ricci-Ollivier curvature of the \emph{rSPR graph} for all pairs of rooted trees up to seven leaves with respect to two simple random walks.
-We found that the curvature varies substantially across pairs of trees in the graph separated by the same distance, indicating significant structure to the graph beyond that seen by pairwise distance.
-This required fast algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
-We also developed a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
-This gave bounds on the curvature, as well as a flatness-in-the-limit theorem for pairs of trees differentiated by moving an asymptotically small subtree.
-We also confirmed using simulation that mean access time distributions depend on distance, degree, and curvature, confirming the relevance of these curvature results to stochastic tree search.
+
+In this paper, we investigate the rSPR graph in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
+We confirmed using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
+These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+We then develop a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
+This gives bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
+However, we find that large topology changes induce paths that are difficult to traverse stochastically, and which may be diagnosed by pairs of trees with negative curvature.
+Our results suggest that, although these difficult to traverse regions do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
 

### 2015-07-08 curvature.tex
Some slight tweaks to the intro.

@@
 Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\cite{Ollivier2009-bw,Joulin2010-jg}.
-In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of \emph{curvature} formalizes the notion of to what extent random walking brings points together.
-Applying the framework to Brownian motion on a manifold returns us back of the classical definition of Ricci curvature.
-For discrete graphs, this is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
+In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings points together.
+Applying the framework to Brownian motion on a manifold gives back of the classical definition of Ricci curvature.
+For discrete graphs, curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying a couple of reasonable assumptions, which includes biased walks such as MCMC.
@@
 In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
-For this first effort we investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.
-Our algorithms and methods can be easily extended to study more complicated distributions with non-uniform topology probabilities.
+Our explicit focus here is to investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.
+In future work, we will extend these methods to study more complicated distributions with non-uniform topology probabilities.
 
@@
 Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to bottlenecks.
+Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to paths that are difficult to traverse stochastically.
 This is a surprising result, as the Markov chain in this simple case is known to mix rapidly~\cite{spade2014note}.
@@
 In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
-Pairs of trees separated by such moves correspond to the ``bottlenecks" identified in our simulation results.
+Pairs of trees separated by such moves correspond to the pairs with negative curvature identified in our simulation results.
 These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
@@
 \begin{equation}
-\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
+\curvature{m; x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}.
 \label{eq:curvatureDef}
 \end{equation}
-For the purposes of this paper, we will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
-Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are \cuttable{on average }``more distant'' than $x$ and $y$.
+For the purposes of this paper, ``curvature'' without further specification will refer to \eqref{eq:curvatureDef}.
+We will use $\curvature{x,y}$ to denote the curvature of the simple (uniform choice of neighbor) random walk, and use $\curvature{\MH; x,y}$ to indicate curvature with respect to the Metropolis-Hastings random walk sampling the uniform distribution (described in detail in Section~\ref{sec:random_walks}).
+Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with \cuttable[a]{the given} random walk.
@@
 As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
-This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
+This definition of curvature is invariant of $p$ for small $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).

### 2015-07-08 curvature.tex
removed two extraneous words

@@
 $$
-We are not aware of any further work investigating properties of the SPR graph, which this stems primarily from a lack of sufficient computational tools.
+We are not aware of any further work investigating properties of the SPR graph, which stems primarily from a lack of sufficient computational tools.
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
@@
 In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings points together.
-Applying the framework to Brownian motion on a manifold gives back of the classical definition of Ricci curvature.
+Applying the framework to Brownian motion on a manifold gives back the classical definition of Ricci curvature.
 For discrete graphs, curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.

### 2015-07-08 curvature.tex
Pass through diffs.

@@
 By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
-This is a major improvement over the sampling methods used in current software programs such as MrBayes which do not propose SPR moves uniformly.
+This stands in contrast to the sampling methods used in current software such as MrBayes, which do not propose SPR moves uniformly.
 
@@
 
-Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after one iteration of MCMC, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
+Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after one step of the random walk, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
 The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
@@
 We construct a mapping from each tree $T_i$ to its order index in this list $i$.
-We then begin with an empty graph $G$.
+Begin with an empty graph $G$.
 For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
 This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
-We use the tree to index mappings to determine whether these trees already reside in the tree and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
+We use the tree to index mappings to determine whether these trees are already vertices of the graph and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
 The high-level steps are as follows, and we show in Theorem~\ref{thm:construct_graph} that this algorithm is correct and can be implemented to run in the stated time.
@@
 Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
-Neighboring trees had minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR graph.
+Neighboring trees achieved minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR graph.
 
@@
 Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
-Such pairs are possible bottlenecks in phylogenetic searches which may be exacerbated by likelihood or branch length constraints.
+Such pairs indicate difficult paths for phylogenetic searches, which may be exacerbated by likelihood or branch length constraints.
 

### 2015-07-08 curvature.tex
fix multiply defined labels
switched away from old IEEEtran template, revised a bit to maintain space limit

@@
-\documentclass[11pt,onecolumn,conference]{IEEEtran}
-
+%\documentclass[11pt,onecolumn,conference]{IEEEtran}
+\documentclass[11pt]{amsart}
+
+\usepackage[foot]{amsaddr}
 \usepackage{amsmath,amsthm}
@@
 
-\author{
-	\IEEEauthorblockN{Chris Whidden}
-	\IEEEauthorblockA{Program in Computational Biology\\
-	Fred Hutchinson Cancer Research Center\\
-	Seattle, WA 98109\\
-	Email: cwhidden@fredhutch.org}
-	\and
-	\IEEEauthorblockN{Frederick A Matsen IV}
-	\IEEEauthorblockA{Program in Computational Biology\\
-	Fred Hutchinson Cancer Research Center\\
-	Seattle, WA 98109\\
-	Email: matsen@fredhutch.org}
-}
-
-\maketitle
-\IEEEpeerreviewmaketitle
+\author{Chris Whidden}
+\address{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA 98109}
+\email{cwhidden@fredhutch.org}
+
+\author{Frederick A Matsen IV}
+%\address{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA 98109}
+\email{matsen@fredhutch.org}
 
@@
 
+\maketitle
+
 \pagebreak
@@
 The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
-It is now common for researchers to approximate the posterior distribution of trees and their associated parameters in a Bayesian setting using an algorithm such as Markov chain Monte Carlo (MCMC).
+It is now common for researchers to approximate the posterior distribution of trees and their associated parameters in a Bayesian setting using Markov chain Monte Carlo (MCMC).
 
@@
 For discrete graphs, curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
-Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying a couple of reasonable assumptions, which includes biased walks such as MCMC.
-
-In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
+Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying reasonable assumptions, which includes biased walks such as MCMC.
+
+In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (i.e. hitting times) for those random walks.
 Our explicit focus here is to investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.
@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
-\cuttable[Proofs of our theorems and lemmas can be found in the appendix.]
+\cuttable[Figures, tables, and proofs of our theorems and lemmas can be found in the appendix.]
 
@@
 See Figure~\ref{fig:spr} for an example.
-Note that the inclusion of $\rho$ allows for rSPR moves that move subtrees to the root of the tree.
+The inclusion of $\rho$ allows for rSPR moves which move subtrees to the root of the tree.
 
@@
 We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single rSPR operation.
-Because we assume that all trees are bifurcating, we use degree to refer only to the degree of rSPR graph vertices.
+We assume that all trees are bifurcating, and thus use degree to refer only to the degree of rSPR graph vertices.
 
@@
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
-As we prove in Lemma~\ref{lem:asymptotic}, the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
+As we prove in Lemma~\ref{lem:asymptotic}, the notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
 
@@
 We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
-We can thus assign $(2n-1-x)-4$ moves to any node at a depth greater than 1 (from the original non-$\rho$ root) and $(2n-1-x)-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
+We assign $(2n-1-x)-2$ moves to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ moves to each other non-root node.
 Let $N(T,u)$ denote the neighbors of $T$ assigned to node $u$, obtained by moving the subtree $R$ rooted at $u$.
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code can be found online \cite{tangle}.
-
-We find a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code is at \cite{tangle}.
+
+We find a wide variation in curvature among tanglegrams (Figure~\ref{fig:rspr-scatter}).
 Curvature values tended to increase with increasing rSPR distance, and their variance decreased with increasing distance.
 Neighboring trees achieved minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR graph.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR~graph.
 
@@
 In previous work, we computed mean access times (MAT) between pairs of trees in  MCMC random walks: the mean number of iterations required to move from one tree to the other.
-We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.
+We applied this work to demonstrate graph effects of real MCMC posteriors sampled with MrBayes~\cite{Whidden2015-yi} using \texttt{sprspace}~\cite{sprspace}.
 
@@
 	\label{lem:degree_max_delta}
-	Let $T$ and $S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
+	Let $T$,$S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
 	Then:
@@
 To do so, we \cuttable{require the following lemma that }characterize\cuttable{s} exactly how the degree of a tree changes after an rSPR operation.
-See Figure~\ref{fig:spr-degree-change} for an illustration of the Lemma.
+See Figure~\ref{fig:spr-degree-change} for an illustration.
 
@@
 That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
-To formalize this concept, we consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
-The number of such ``squares'' involving two adjacent trees will play a key role in our bounds on curvature, as they push the curvature of those trees towards 0.
+To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
+The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they push the curvature of those trees towards 0.
 
@@
 
-We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
+We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
 \begin{restatable}{lemma}{degreemaxdeltaadjacent}
 	\label{lem:degree_max_delta_adjacent}
-	Let $T$ and $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
+	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
 	Then:
@@
 	\label{lem:shared_neighbors}
-Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
+Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
 Then $\size{N(T) \cap N(S)} \le 6n - 17$.
@@
 \begin{conjecture}
-	Let $k_n$ be the maximum curvature of the uniform random walk on $n$-leaf trees.
+	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
 	Then:
@@
 \label{lem:curvature_lower_bound}
-	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
+	The curvature between adjacent trees with $n$ leaves is at least
 	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
@@
 \begin{corollary}
-	The curvature of the uniform random walk between two trees is at least $-\frac{2}{5}$.
+	The curvature between two trees is at least $-\frac{2}{5}$.
 \end{corollary}
@@
 We next bound the difference between the coarse and asymptotic curvatures.
-Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
+Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that \cuttable{remains at a given tree with probability $1-p$ and }moves with probability $p$.
 For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
@@
 
-\appendices
+\appendix
 
@@
 
-\degreemaxdeltaadjacent
+\degreemaxdeltaadjacent*
 \begin{proof}
@@
 
-\curvaturedistancebound
+\curvaturedistancebound*
 \begin{proof}

### 2015-07-24 curvature.tex
Abstract fixes.

@@
 Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
-These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions induced by such parameters.
+These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions for real data induced by such parameters.
 The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC.
 Indeed, topology changes, and SPR rearrangements in particular, have been identified as a primary limiting factor of phylogenetic MCMC algorithms.
-As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the \emph{rSPR graph} in greater detail.
+As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the underlying \emph{rSPR graph} in greater detail.
 
 In this paper, we investigate the rSPR graph in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-We confirmed using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
-These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+We confirm using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
+Curvature calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
 We then develop a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.

### 2015-07-24 curvature.tex
Adding 7 taxa to tables.

@@
 \caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
+\csvautotabular{tables/mean-access.csv}}
 \label{tab:regressionMAT}
@@
 \caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
-\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
-\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
+\csvautotabular{tables/delta1.csv}}
 \label{tab:regressionDelta}

### 2015-07-24 tables/gather.sh

@@ -1,0 +1,15 @@
+set -eu
+
+tmp=$(mktemp)
+tmp2=$(mktemp)
+
+for c in delta1 mean-access; do
+    csvcut -c1 ../analysis/5-walks-200K/rspr/$c*csv | sed 's#""#variable#' > $c.csv
+    for w in ../analysis/5-walks-200K ../analysis/6-walks-50K ../analysis/7-walks-5K; do
+        ntaxa=$(echo $w | sed -e "s/-.*//" -e "s#.*/##")
+        csvcut -c3 $w/rspr/$c*csv | sed "s/p-value/$ntaxa taxa/" > $tmp
+        paste -d, $c.csv $tmp > $tmp2
+        cp $tmp2 $c.csv
+    done
+done
+

### 2015-07-24 curvature.tex
Better explanation of regression p-values.

@@
 \centering
-\caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
+\caption{p-values for ordinary least squares linear multiple regression of rSPR mean access time against degree and distance (two-tailed $t$-test of regression coefficient).}
 \csvautotabular{tables/mean-access.csv}}
@@
 \centering
-\caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
+\caption{p-values for ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$ (two-tailed $t$-test of regression coefficient).}
 \csvautotabular{tables/delta1.csv}}

### 2015-07-24 curvature.tex
Fix from @cwhidden

@@
 We confirm using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
-Curvature calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
+These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
 We then develop a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.

### 2015-07-24 curvature.tex
removed graph effect terminology

@@
 Indeed, topology changes, and SPR rearrangements in particular, have been identified as a primary limiting factor of phylogenetic MCMC algorithms.
-As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the underlying \emph{rSPR graph} in greater detail.
+As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the underlying \emph{rSPR graph} in greater~detail.
 
@@
 It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
-We will call these ``graph effects", whereby one can't easily move from one tree to another due to the structure of the graph.
-Researchers have previously identified SPR graph effects in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}, contrasting with fast mixing identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu}.
+Researchers have previously identified slow mixing in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}, contrasting with fast mixing identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu}.
 Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
-In particular, the simple case of sampling the uniform topology distribution without continuous parameters via SPR operations is known to mix quickly~\cite{spade2014note}.
-However, despite indications of graph effects on real data~\cite{beiko2006searching, lakner2008efficiency} the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations has made it difficult to study graph effects.
+In particular, the simple case of sampling the uniform topology distribution without continuous parameters via (slightly modified) SPR operations is known to mix quickly~\cite{spade2014note}.
+Studies on real data~\cite{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
+Previously, the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations made it difficult to determine the cause of these difficulties.
 By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by ``bottlenecks" of specific classes of SPR moves.
@@
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
-In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees may have greater graph effects than those that move small subtrees.
+In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which modify large subtrees are less likely to be explored during these random walks.
 Pairs of trees separated by such moves correspond to the pairs with negative curvature identified in our simulation results.
@@
 
-In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify graph effects.
+In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify local maxima.
 Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
@@
 In previous work, we computed mean access times (MAT) between pairs of trees in  MCMC random walks: the mean number of iterations required to move from one tree to the other.
-We applied this work to demonstrate graph effects of real MCMC posteriors sampled with MrBayes~\cite{Whidden2015-yi} using \texttt{sprspace}~\cite{sprspace}.
+We applied this work to demonstrate the influence of SPR graph structure on real MCMC posteriors sampled with MrBayes~\cite{Whidden2015-yi} using \texttt{sprspace}~\cite{sprspace}.
 
@@
 \centering
-\caption{p-values for ordinary least squares linear multiple regression of rSPR mean access time against degree and distance (two-tailed $t$-test of regression coefficient).}
+\caption{p-values for ordinary least squares linear multiple regression of rSPR mean access time against degree and distance (two-tailed $t$-test of regression coefficient).
+The p-values for 7 taxa are smaller than the machine precision used to calculate them.}
 \csvautotabular{tables/mean-access.csv}}
@@
 Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
-However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
 Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to significant ``bottlenecks'' like those we have observed in real data.

### 2015-07-25 curvature.tex
A few tweaks.

@@
 This gives bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
-However, we find that large topology changes induce paths that are difficult to traverse stochastically, and which may be diagnosed by pairs of trees with negative curvature.
-Our results suggest that, although these difficult to traverse regions do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
+However, we find that large topology changes (i.e. moving a large subtree) gives pairs of trees with negative curvature.
+Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
 
@@
 Molecular phylogenetic methods reconstruct evolutionary trees from DNA or RNA data and are of fundamental importance to modern biology.
-Statistical phylogenetics forms the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
+Statistical phylogenetics is the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
 The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution~\cite{felsenstein1981evolutionary}.
@@
 
-In order to estimate these distributions accurately, MCMC samplers must sufficiently explore the collection of trees.
+In order to estimate these distributions accurately, MCMC samplers must sufficiently explore the set of trees.
 Phylogenetic search algorithms typically attempt to do so through a combination of modifications to the continuous parameters and tree topology.
@@
 It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
-Researchers have previously identified slow mixing in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}, contrasting with fast mixing identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu}.
-Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
-In particular, the simple case of sampling the uniform topology distribution without continuous parameters via (slightly modified) SPR operations is known to mix quickly~\cite{spade2014note}.
+Researchers have previously identified slow mixing in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}.
+On the other hand, fast mixing has been identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu} or with no data~\cite{spade2014note}.
+% Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
 Studies on real data~\cite{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
@@
 $$
-We are not aware of any further work investigating properties of the SPR graph, which stems primarily from a lack of sufficient computational tools.
+We are not aware of any further work investigating properties of the SPR graph, which may be due to its complexity.
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
@@
 In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings points together.
-Applying the framework to Brownian motion on a manifold gives back the classical definition of Ricci curvature.
-For discrete graphs, curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices and the distance between the vertices.
+Applying the framework to Brownian motion on a manifold returns the classical definition of Ricci curvature.
+Curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices given by a random walk and the distance between the vertices.
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying reasonable assumptions, which includes biased walks such as MCMC.
 
-In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (i.e. hitting times) for those random walks.
+In this paper, we investigate curvature of the rSPR graph with respect to two random walks and compare those results to access times (i.e. hitting times) for those random walks.
 Our explicit focus here is to investigate random walks defined only in terms of the graph itself: the uniform random walk and MCMC sampling from the uniform prior on trees.

### 2015-07-27 curvature.tex
minor fix to make a proof more standalone. removed extra braces from the table changes

--- 2015-07-24 curvature.tex
@@
 The p-values for 7 taxa are smaller than the machine precision used to calculate them.}
-\csvautotabular{tables/mean-access.csv}}
+\csvautotabular{tables/mean-access.csv}
 \label{tab:regressionMAT}
@@
 \caption{p-values for ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$ (two-tailed $t$-test of regression coefficient).}
-\csvautotabular{tables/delta1.csv}}
+\csvautotabular{tables/delta1.csv}
 \label{tab:regressionDelta}
@@
 	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
-	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e.
-(iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e.
-(iv) above), a contradiction.
+	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (move type (iii)) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (move type (iv)), a contradiction.
 	Therefore the claim holds.

### 2015-07-27 curvature.tex
My proposed edits for intro.

@@
 Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to paths that are difficult to traverse stochastically.
-This is a surprising result, as the Markov chain in this simple case is known to mix rapidly~\cite{spade2014note}.
-
-In response, we considered properties of the graph analytically.
+Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to paths that are difficult to traverse stochastically in the period before reaching stationarity.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
@@
 Pairs of trees separated by such moves correspond to the pairs with negative curvature identified in our simulation results.
-These pairs occur infrequently in these well-connected graphs, thus allowing the Markov chain to mix rapidly.
-However, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}.
+These pairs occur infrequently in these well-connected graphs, however, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}.
 In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

### 2015-07-28 curvature.tex
Mutually agreeable sentence for negative curvature.

@@
 Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-Moreover, we found that rSPR graphs with 7 or more leaves have tree pairs with negative curvature, corresponding to paths that are difficult to traverse stochastically in the period before reaching stationarity.
+Moreover, we found that rSPR graphs for trees with 7 or more leaves have tree pairs with negative curvature, corresponding to direct paths that are difficult to traverse stochastically.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.

### 2015-07-29 curvature.tex
Some topic sentence smoothing.

@@
 
-To study the full rSPR graph for a fixed number of leaves it is necessary to have an efficient means of constructing it.
+It is necessary to have an efficient means of constructing the full rSPR graph for a fixed number of leaves in order to study it.
 The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
@@
 \section{Conclusion and future work}
-In summary, we have advanced understanding of the phylogenetic rSPR graph substantially beyond what was previously known, which concerned graph diameter and vertex degree.
+In summary, we have gone beyond graph diameter and vertex degree to substantially advance understanding of the phylogenetic rSPR graph.
 We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.

### 2015-07-30 curvature.tex
fix max curvature conjecture with asymptotically equal to
cite other works using ricci curvature of graphs
rephrasing thanks to @matsen
A little tightening and clarifying.

@@
 These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions for real data induced by such parameters.
-The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC.
-Indeed, topology changes, and SPR rearrangements in particular, have been identified as a primary limiting factor of phylogenetic MCMC algorithms.
-As mixing properties of phylogenetic MCMC algorithms with a nontrivial likelihood function are difficult to characterize analytically, we were motivated to take the different strategy of characterizing the properties of the underlying \emph{rSPR graph} in greater~detail.
+The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating study of the underlying \emph{rSPR graph} in greater~detail.
 
@@
 Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
-
+This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of random search on the rSPR graph.
 \end{abstract}
@@
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying reasonable assumptions, which includes biased walks such as MCMC.
+This approach has been useful for determining properties of a wide variety of graphs including the internet topology~\cite{ni2015ricci} and cancer networks~\cite{sandhu2015graph}.
 
@@
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),$$
-where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of mass transport programs moving the mass in $m_x$ to the configuration described by $m_y$\cuttable{ (more formally, a density on $V \times V$ which is $m_x$ after projecting on the first component and $m_y$ after projecting on the second)}.
+where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of densities on $V \times V$ that are $m_x$ after projecting on the first component and $m_y$ after projecting on the second.
 
@@
 		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
-		\item $\lim_{n\rightarrow\infty} k_n  = \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
+		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
 	\end{enumerate}

### 2015-10-22 ltexpprt.tex
added SIAM ltexpprt style file and example/documentation

%%%%%%%%%%%%%%%%%%%%%%%%%%  ltexpprt.tex  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is ltexpprt.tex, an example file for use with the SIAM LaTeX2E
% Preprint Series macros. It is designed to provide double-column output.
% Please take the time to read the following comments, as they document
% how to use these macros. This file can be composed and printed out for
% use as sample output.

% Any comments or questions regarding these macros should be directed to:
%
%                 Donna Witzleben
%                 SIAM
%                 3600 University City Science Center
%                 Philadelphia, PA 19104-2688
%                 USA
%                 Telephone: (215) 382-9800
%                 Fax: (215) 386-7999
%                 e-mail: witzleben@siam.org

% This file is to be used as an example for style only. It should not be read
% for content.

%%%%%%%%%%%%%%% PLEASE NOTE THE FOLLOWING STYLE RESTRICTIONS %%%%%%%%%%%%%%%

%%  1. There are no new tags.  Existing LaTeX tags have been formatted to match
%%     the Preprint series style.
%%
%%  2. You must use \cite in the text to mark your reference citations and
%%     \bibitem in the listing of references at the end of your chapter. See
%%     the examples in the following file. If you are using BibTeX, please
%%     supply the bst file with the manuscript file.
%%
%%  3. This macro is set up for two levels of headings (\section and
%%     \subsection). The macro will automatically number the headings for you.
%%
%%  5. No running heads are to be used for this volume.
%%
%%  6. Theorems, Lemmas, Definitions, etc. are to be double numbered,
%%     indicating the section and the occurence of that element
%%     within that section. (For example, the first theorem in the second
%%     section would be numbered 2.1. The macro will
%%     automatically do the numbering for you.
%%
%%  7. Figures, equations, and tables must be single-numbered.
%%     Use existing LaTeX tags for these elements.
%%     Numbering will be done automatically.
%%
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[twoside,leqno,twocolumn]{article}
\usepackage{ltexpprt}

\begin{document}

%\setcounter{chapter}{2} % If you are doing your chapter as chapter one,
%\setcounter{section}{3} % comment these two lines out.

\title{\Large SIAM/ACM Preprint Series Macros for
Use With LaTeX\thanks{Supported by GSF grants ABC123, DEF456, and GHI789.}}
\author{Corey Gray\thanks{Society for Industrial and Applied Mathematics.} \\
\and
Tricia Manning\thanks{Society for Industrial and Applied Mathematics.}}
\date{}

\maketitle

%\pagenumbering{arabic}
%\setcounter{page}{1}%Leave this line commented out.

\begin{abstract} \small\baselineskip=9pt This is the text of my abstract. It is a brief
description of my
paper, outlining the purposes and goals I am trying to address.\end{abstract}

\section{Problem Specification.}In this paper, we consider the solution of the $N \times
N$ linear
system
\begin{equation} \label{e1.1}
A x = b
\end{equation}
where $A$ is large, sparse, symmetric, and positive definite.  We consider
the direct solution of (\ref{e1.1}) by means of general sparse Gaussian
elimination.  In such a procedure, we find a permutation matrix $P$, and
compute the decomposition
\[
P A P^{t} = L D L^{t}
\]
where $L$ is unit lower triangular and $D$ is diagonal.

\section{Design Considerations.}Several good ordering algorithms (nested dissection and
minimum degree)
are available for computing $P$  \cite{GEORGELIU}, \cite{ROSE72}.
Since our interest here does not
focus directly on the ordering, we assume for convenience that $P=I$,
or that $A$ has been preordered to reflect an appropriate choice of $P$.

Our purpose here is to examine the nonnumerical complexity of the
sparse elimination algorithm given in  \cite{BANKSMITH}.
As was shown there, a general sparse elimination scheme based on the
bordering algorithm requires less storage for pointers and
row/column indices than more traditional implementations of general
sparse elimination.  This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.

\begin{theorem} The method  was extended to three
dimensions. For the standard multigrid
coarsening
(in which, for a given grid, the next coarser grid has $1/8$
as many points), anisotropic problems require plane
relaxation to
obtain a good smoothing factor.\end{theorem}

Our purpose here is to examine the nonnumerical complexity of the
sparse elimination algorithm given in  \cite{BANKSMITH}.
As was shown there, a general sparse elimination scheme based on the
bordering algorithm requires less storage for pointers and
row/column indices than more traditional implementations of general
sparse elimination.  This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.
Several good ordering algorithms (nested dissection and minimum degree)
are available for computing $P$  \cite{GEORGELIU}, \cite{ROSE72}.
Since our interest here does not
focus directly on the ordering, we assume for convenience that $P=I$,
or that $A$ has been preordered to reflect an appropriate choice of $P$.

\begin{proof} In this paper we consider two methods. The first method
is
basically the method considered with two differences:
first, we perform plane relaxation by a two-dimensional
multigrid method, and second, we use a slightly different
choice of
interpolation operator, which improves performance
for nearly singular problems. In the second method coarsening
is done by successively coarsening in each of the three
independent variables and then ignoring the intermediate
grids; this artifice simplifies coding considerably.
\end{proof}

Our purpose here is to examine the nonnumerical complexity of the
sparse elimination algorithm given in  \cite{BANKSMITH}.
As was shown there, a general sparse elimination scheme based on the
bordering algorithm requires less storage for pointers and
row/column indices than more traditional implementations of general
sparse elimination.  This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.

\begin{Definition}{\rm We describe the two methods in \S 1.2. In \S\ 1.3. we
discuss
some remaining details.}
\end{Definition}

Our purpose here is to examine the nonnumerical complexity of the
sparse elimination algorithm given in  \cite{BANKSMITH}.
As was shown there, a general sparse elimination scheme based on the
bordering algorithm requires less storage for pointers and
row/column indices than more traditional implementations of general
sparse elimination.  This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.
Several good ordering algorithms (nested dissection and minimum degree)
are available for computing $P$  \cite{GEORGELIU}, \cite{ROSE72}.
Since our interest here does not
focus directly on the ordering, we assume for convenience that $P=I$,
or that $A$ has been preordered to reflect an appropriate choice of $P$.

Our purpose here is to examine the nonnumerical complexity of the
sparse elimination algorithm given in  \cite{BANKSMITH}.
As was shown there, a general sparse elimination scheme based on the
bordering algorithm requires less storage for pointers and
row/column indices than more traditional implementations of general
sparse elimination.

\begin{lemma} We discuss first the choice for $I_{k-1}^k$
which is a generalization. We assume that $G^{k-1}$ is
obtained
from $G^k$
by standard coarsening; that is, if $G^k$ is a tensor product
grid $G_{x}^k \times G_{y}^k \times G_{z}^k$,
$G^{k-1}=G_{x}^{k-1} \times G_{y}^{k-1} \times G_{z}^{k-1}$,
where $G_{x}^{k-1}$ is obtained by deleting every other grid
point of $G_x^k$ and similarly for $G_{y}^k$ and $G_{z}^k$.
\end{lemma}

To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6]. In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].

In \S 1.2, we review the bordering algorithm, and introduce
the sorting and intersection problems that arise in the
sparse formulation of the algorithm.
In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].

For the old approach, we show that the
complexity of the intersection problem is $O(n^{3})$, the same
as the complexity of the numerical computations.  For the
new approach, the complexity of the second part is reduced to
$O(n^{2} (\log n)^{2})$.

To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6]. In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].
This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.
To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
\cite{EISENSTAT} - \cite{LIU2}, \cite{ROSE76},  \cite{SCHREIBER}.

\subsection{Robustness.}We do not
attempt to present an overview
here, but rather attempt to focus on those results that
are relevant to our particular algorithm.
This section assumes prior knowledge of the role of graph theory
in sparse Gaussian elimination; surveys of this role are
available in \cite{ROSE72} and \cite{GEORGELIU}. More general
discussions of elimination trees are given in
\cite{LAW} - \cite{LIU2}, \cite{SCHREIBER}.
Thus, at the $k$th stage, the bordering algorithm consists of
solving the lower triangular system
\begin{equation} \label{1.2}
 L_{k-1}v = c
\end{equation}
and setting
\begin{eqnarray}
\ell &=& D^{-1}_{k-1}v , \\
\delta &=& \alpha - \ell^{t} v .
\end{eqnarray}

\begin{figure}
\vspace{14pc}
\caption{This is a figure 1.1.}
\end{figure}

\section{Robustness.} We do not
attempt to present an overview
here, but rather attempt to focus on those results that
are relevant to our particular algorithm.

\subsection{Versatility.}The special
structure of this problem allows us to make exact estimates of
the complexity.  For the old approach, we show that the
complexity of the intersection problem is $O(n^{3})$, the same
as the complexity of the numerical computations
\cite{GEORGELIU}, \cite{ROSEWHITTEN}.  For the
new approach, the complexity of the second part is reduced to
$O(n^{2} (\log n)^{2})$.

To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6]. In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].

In \S 1.2, we review the bordering algorithm, and introduce
the sorting and intersection problems that arise in the
sparse formulation of the algorithm.
In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].

For the old approach, we show that the
complexity of the intersection problem is $O(n^{3})$, the same
as the complexity of the numerical computations.  For the
new approach, the complexity of the second part is reduced to
$O(n^{2} (\log n)^{2})$.

To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6]. In \S 1.3., we analyze the complexity of the old and new
approaches to the intersection problem for the special case of
an $n \times n$ grid ordered by nested dissection. The special
structure of this problem allows us to make exact estimates of
the complexity. To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
[4] - [10], [5], [6].
This is accomplished by exploiting the m-tree,
a particular spanning tree for the graph of the filled-in matrix.
To our knowledge, the m-tree previously has not been applied in this
fashion to the numerical factorization, but it has been used,
directly or indirectly, in several optimal order algorithms for
computing the fill-in during the symbolic factorization phase
\cite{EISENSTAT} - \cite{LIU2}, \cite{ROSE76},  \cite{SCHREIBER}.

\begin{thebibliography}{99}

%\bibitem{GUIDE}
%R.~E. Bank, {\em PLTMG  users' guide, edition 5.0}, tech. report,
%  Department of Mathematics, University of California, San Diego, CA, 1988.

%\bibitem{HBMG}
%R.~E. Bank, T.~F. Dupont, and H.~Yserentant, {\em The hierarchical basis
%  multigrid method}, Numer. Math., 52 (1988), pp.~427--458.

\bibitem{BANKSMITH}
R.~E. Bank and R.~K. Smith, {\em General sparse elimination requires no
  permanent integer storage}, SIAM J. Sci. Stat. Comput., 8 (1987),
  pp.~574--584.

\bibitem{EISENSTAT}
S.~C. Eisenstat, M.~C. Gursky, M.~Schultz, and A.~Sherman, {\em
  Algorithms and data structures for sparse symmetric gaussian elimination},
  SIAM J. Sci. Stat. Comput., 2 (1982), pp.~225--237.

\bibitem{GEORGELIU}
A.~George and J.~Liu, {\em Computer Solution of Large Sparse Positive
  Definite Systems}, Prentice Hall, Englewood Cliffs, NJ, 1981.

\bibitem{LAW}
K.~H. Law and S.~J. Fenves, {\em A node addition model for symbolic
  factorization}, ACM TOMS, 12 (1986), pp.~37--50.

\bibitem{LIU}
J.~W.~H. Liu, {\em A compact row storage scheme for cholesky factors
  using elimination trees}, ACM TOMS, 12 (1986), pp.~127--148.

\bibitem{LIU2}
\sameauthor , {\em The role of
  elimination trees in sparse factorization}, Tech. Report CS-87-12,Department
  of Computer Science, York University, Ontario, Canada, 1987.

\bibitem{ROSE72}
D.~J. Rose, {\em A graph theoretic study of the numeric solution of
  sparse positive definite systems}, in Graph Theory and Computing, Academic  Press, New
York, 1972.

\bibitem{ROSE76}
D.~J. Rose, R.~E. Tarjan, and G.~S. Lueker, {\em Algorithmic aspects of
  vertex elimination on graphs}, SIAM J. Comput., 5 (1976), pp.~226--283.

\bibitem{ROSEWHITTEN}
D.~J. Rose and G.~F. Whitten, {\em A recursive analysis of disection
  strategies}, in Sparse Matrix Computations, Academic Press, New York, 1976.

\bibitem{SCHREIBER}
R.~Schrieber, {\em A new implementation of sparse gaussian elimination},
  ACM TOMS, 8 (1982), pp.~256--276.

\end{thebibliography}
\end{document}

% End of ltexpprt.tex

### 2015-10-22 curvature.tex
curvature formatting, still need to fix some overlap and integrate figures

@@
 %\documentclass[11pt,onecolumn,conference]{IEEEtran}
-\documentclass[11pt]{amsart}
-
-\usepackage[foot]{amsaddr}
-\usepackage{amsmath,amsthm}
+\documentclass[10pt,twoside,leqno,twocolumn]{article}
+\usepackage{ltexpprt}
+
+%\usepackage[foot]{amsaddr}
+\usepackage{amsmath}
 \usepackage{csvsimple}
@@
 \usepackage{enumitem}
-\usepackage{url}
 \usepackage{subfigure}
@@
 % regular margins
-\usepackage[margin=1in]{geometry}
 \usepackage{thmtools}
@@
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{conjecture}[theorem]{Conjecture}
+\newtheorem{thm}{Theorem}[section]
+\newtheorem{lem}[thm]{Lemma}
+\newtheorem{cor}[thm]{Corollary}
+\newtheorem{conj}[thm]{Conjecture}
+\renewcommand{\theequation}{\arabic{equation}}
 
@@
 \begin{document}
-\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph}
-
-\author{Chris Whidden}
-\address{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA 98109}
-\email{cwhidden@fredhutch.org}
-
-\author{Frederick A Matsen IV}
-%\address{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA 98109}
-\email{matsen@fredhutch.org}
+
+\setlist{noitemsep}
+
+\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\thanks{This work was funded by National Science Foundation award 1223057.
+Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.}}
+
+\author{Chris Whidden\thanks{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, USA 98109. \set{cwhidden,matsen}@fredhutch.org} \\
+\and
+Frederick A Matsen IV\footnotemark[2]}
+\date{}
+
+\maketitle
 
@@
 This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of random search on the rSPR graph.
+
 \end{abstract}
-
-\maketitle
-
-\pagebreak
 
@@
 \cite{Ding2011-bj}~showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
-$$ n - 2\ceil{\sqrt{n}} + 1
+\begin{equation}
+n - 2\ceil{\sqrt{n}} + 1
 \le \Delta_{\text{uSPR}}(n)
 \le n - 3 - \floorV{\frac{\sqrt{n - 2} - 1}{2}}.
-$$
+\end{equation}
 We are not aware of any further work investigating properties of the SPR graph, which may be due to its complexity.
@@
 The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
-$$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),$$
+\begin{equation}
+W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum_{\{z,w\} \subset V} d(z,w) \xi(z,w),
+\end{equation}
 where $d(z,w)$ is the graph shortest-path distance ($\dspr{z,w}$ in our case) and $\Pi(m_x, m_y)$ is the set of densities on $V \times V$ that are $m_x$ after projecting on the first component and $m_y$ after projecting on the second.
@@
 
-\subsection{Computing the rSPR graph of $m$ trees with $n$ leaves in $\OhOf{mn^3}$-time}
+\subsection{Computing the rSPR graph of $m$ trees with $n$ leaves in $\OhOf{mn^3}$-time.}
 \label{sec:computing_treespace}
@@
 
+\vspace{1em}
 \textsc{Construct-rSPR-Graph($\mathcal{T}$)}
@@
 			\item Add $T_i \rightarrow i$ to $M$.
-			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}: \vspace{-1.4em}
+			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}:
 				\begin{enumerate}
@@
 
-%\begin{theorem}
+%\begin{thm}
 %	\label{thm:construct_graph}
 %	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
-%\end{theorem}
-\begin{restatable}{theorem}{constructgraph}
+%\end{thm}
+\begin{restatable}{thm}{constructgraph}
 	\label{thm:construct_graph}
@@
 
-\subsection{Simulating random walks on the rSPR graph}
+\subsection{Simulating random walks on the rSPR graph.}
 \label{sec:random_walks}
@@
 
-\begin{restatable}{lemma}{computedegree}
+\begin{restatable}{lem}{computedegree}
 	\label{lem:compute_degree}
@@
 
-\begin{restatable}{lemma}{selectrandomneighbor}
+\begin{restatable}{lem}{selectrandomneighbor}
 	\label{lem:select_random_neighbor}
@@
 
-\begin{corollary}
+\begin{cor}
 \label{cor:enumerate_neighbors}
 	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
-\end{corollary}
+\end{cor}
 
@@
 
-\subsection{Computing curvature values}
+\subsection{Computing curvature values.}
 To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
@@
 
-\subsection{Access time simulation}
+\subsection{Access time simulation.}
 The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
@@
 
-\begin{lemma}[{Song~\cite{Song2003-gf}}]
+\begin{lem}[{Song~\cite{Song2003-gf}}]
 \cuttable{    \pushQED{\qed}}
@@
 \cuttable{    \popQED}
-\end{lemma}
+\end{lem}
 
 We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
-\begin{restatable}{lemma}{degreemaxdelta}
+\begin{restatable}{lem}{degreemaxdelta}
 	\label{lem:degree_max_delta}
@@
 
-\begin{restatable}{lemma}{degreechange}
+\begin{restatable}{lem}{degreechange}
 \label{lem:degree_change}
@@
 
-\begin{restatable}{corollary}{pairedneighbors}
+\begin{restatable}{cor}{pairedneighbors}
 \label{cor:paired_neighbors}
@@
 We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
-\begin{restatable}{lemma}{degreemaxdeltaadjacent}
+\begin{restatable}{lem}{degreemaxdeltaadjacent}
 	\label{lem:degree_max_delta_adjacent}
@@
 
-\begin{restatable}{lemma}{sharedneighbors}
+\begin{restatable}{lem}{sharedneighbors}
 	\label{lem:shared_neighbors}
@@
 
-\begin{restatable}{theorem}{flatcurvature}
+\begin{restatable}{thm}{flatcurvature}
 \label{thm:flat_curvature}
@@
 
-\begin{restatable}{lemma}{curvaturedistancebound}
+\begin{restatable}{lem}{curvaturedistancebound}
 	\label{lem:curvature_distance_bound}
@@
 
-\begin{restatable}{lemma}{maxadjacentcurvature}
+\begin{restatable}{lem}{maxadjacentcurvature}
 	\label{lem:max_adjacent_curvature}
@@
 
-\begin{restatable}{lemma}{curvaturelowerbound}
+\begin{restatable}{lem}{curvaturelowerbound}
 \label{lem:curvature_lower_bound}
@@
 
-\begin{corollary}
+\begin{cor}
 	The curvature between two trees is at least $-\frac{2}{5}$.
-\end{corollary}
+\end{cor}
 
@@
 
-\begin{restatable}{lemma}{asymptotic}
+\begin{restatable}{lem}{asymptotic}
 \label{lem:asymptotic}
@@
 
-\begin{restatable}{lemma}{mhcurvature}
+\begin{restatable}{lem}{mhcurvature}
 	\label{lem:mh_curvature}
@@
 They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
-This work was funded by National Science Foundation award 1223057.
 

### 2015-10-23 curvature.tex
figures in main text, equations no longer overflow

@@
 \usepackage{todonotes}
-\usepackage[nomarkers,nofiglist,notablist]{endfloat}
+\usepackage{url}
+%\usepackage[nomarkers,nofiglist,notablist]{endfloat}
 \let\MYoriglatexcaption\caption
@@
 
+\setlist{noitemsep}
+
 \begin{document}
-
-\setlist{noitemsep}
 
@@
 
+\begin{figure*}[t]
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:x-tree}]{\includegraphics[scale=1.25]{figs/x-tree}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:subtree}]{\includegraphics[scale=1.25]{figs/subtree}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:induced}]{\includegraphics[scale=1.25]{figs/contracted-subtree}}
+	\hspace*{\stretch{1}}
+	\\
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:spr}]{\includegraphics[scale=1.25]{figs/spr}}
+	\hspace*{\stretch{1}}
+	\\
+	\hspace*{\stretch{1}}
+	\hspace*{\stretch{1}}
+
+	\caption{(a) An $X$-tree $T$.
+		(b) $T(V)$, where $V = \set{1,2,5}$.
+		(c) $T|V$.
+		(d) An rSPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
+	\label{fig:trees}
+\end{figure*}
+
 Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
@@
 
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:x-tree}]{\includegraphics[scale=1.25]{figs/x-tree}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:subtree}]{\includegraphics[scale=1.25]{figs/subtree}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:induced}]{\includegraphics[scale=1.25]{figs/contracted-subtree}}
-	\hspace*{\stretch{1}}
-	\\
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:spr}]{\includegraphics[scale=1.25]{figs/spr}}
-	\hspace*{\stretch{1}}
-	\\
-	\hspace*{\stretch{1}}
-	\hspace*{\stretch{1}}
-
-	\caption{(a) An $X$-tree $T$.
-		(b) $T(V)$, where $V = \set{1,2,5}$.
-		(c) $T|V$.
-		(d) An rSPR operation transforms $T$ into a new tree $T'$ by \emph{pruning} a \emph{subtree} and \emph{regrafting} it in another location.}
-	\label{fig:trees}
-\end{figure}
-
-\begin{figure}
+\begin{figure*}
 	\hspace*{\stretch{1}}
@@
 	\label{fig:two-spr}
-\end{figure}
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\includegraphics[width=0.6\textwidth]{figs/neighborhood}
-	\hspace*{\stretch{1}}
-	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections between neighbors.}
-	\label{fig:neighborhood}
-\end{figure}
+\end{figure*}
 
@@
 The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
+\vspace{-0.5em}
 \begin{equation}
@@
 The \emph{coarse Ricci-Ollivier curvature} of $x$ and $y$ is then defined as:
+\vspace{-0.5em}
 \begin{equation}
@@
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with \cuttable[a]{the given} random walk.
+\begin{figure}[!h]
+	\hspace*{\stretch{1}}
+	\includegraphics[width=0.5\textwidth]{figs/neighborhood}
+	\hspace*{\stretch{1}}
+	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections between neighbors.}
+	\label{fig:neighborhood}
+\end{figure}
+\vspace{0.1em}
+
+\pagebreak
 
@@
 
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{2}}
+	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
+	\hspace*{\stretch{1}}
+    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+Color displays the average degree of $T_1$ and $T_2$.
+Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
+	\label{fig:rspr-scatter}
+\end{figure*}
+
 \begin{restatable}{lem}{selectrandomneighbor}
@@
 
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
 	\hspace*{\stretch{2}}
-	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
-	\hspace*{\stretch{1}}
-    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
-Color displays the average degree of $T_1$ and $T_2$.
-Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
-	\label{fig:rspr-scatter}
-\end{figure}
+	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
+	\hspace*{\stretch{1}}
+    \caption{\
+        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        }
+	\label{fig:kappa-access}
+\end{figure*}
 
@@
 
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
-	\hspace*{\stretch{1}}
-    \caption{\
-        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
-        }
-	\label{fig:kappa-access}
-\end{figure}
-
 \begin{table}
@@
 \label{tab:regressionDelta}
+\vspace{2em}
 \end{table}
+
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\includegraphics[width=0.8\textwidth]{figs/spr-degree-change}
+	\hspace*{\stretch{1}}
+	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
+Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
+	\label{fig:spr-degree-change}
+\end{figure*}
 
@@
 	\label{lem:degree_extremes}
-	Let $T$ be a tree with $n$ leaves. Then
+	For a tree $T$ with $n$ leaves:
 	\begin{enumerate}
@@
 See Figure~\ref{fig:spr-degree-change} for an illustration.
-
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\includegraphics[width=\textwidth]{figs/spr-degree-change}
-	\hspace*{\stretch{1}}
-	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
-Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
-	\label{fig:spr-degree-change}
-\end{figure}
 
@@
 We thus conjecture the following:
-\begin{conjecture}
+\begin{conj}
 	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
@@
 	\end{enumerate}
-\end{conjecture}
+\end{conj}
 Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
@@
 	Let $T$ and $S$ be trees with $n$ leaves. Then:
-	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
-	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
+	\begin{align*}
+	\curvature{T,S} - \frac{1}{3\dspr{T,S}}
+	&\le \curvature{\MH; T,S} \\
+	\curvature{\MH; T,S}
+	&\le \curvature{T,S} + \frac{1}{3\dspr{T,S}},\text{ and }
+\end{align*}
 	$$\curvature{T,S} - 1/6
@@
 
-\pagebreak
-
-\bibliographystyle{IEEEtran}
+\bibliographystyle{siam}
 \bibliography{curvature}
-
-\pagebreak
 
@@
 	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
+	Let ${\Delta}N = \degree{S} - \degree{T}$.
 	If we move the additional leaf, we have:
-$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}$$
+	\begin{align*}
+		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
+		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}
+	\end{align*}
 like before.
 Similarly, if we do not move the additional leaf, we also have:
-$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
+\begin{align*}
+	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
+&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
+\end{align*}
 proving (i).
@@
 	\begin{align*}
-		\curvature[p]{T,S} /p &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
+		\frac{\curvature[p]{T,S}}{p} &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
 		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
@@
 
-	For the upper bound, we observe that $$W_1'(T,S) \ge p W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
+	For the upper bound, we observe that $W_1'(T,S) \ge$
+	\begin{align*}
+		p &W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},
+	\end{align*}
+	as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
 	The upper bound then follows analogously to the lower bound.

### 2015-10-27 curvature.tex
addressed reviewer 1's minor comments
addressed reviewer 2's minor comments

@@
 
+% number of paired neighbors
+\newcommand{\overlap}{%
+	\gamma
+}
+
 \begin{document}
@@
 Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
-These methods are known to mix quickly in the absence of sequence data and other parameters such as branch lengths, but may become stuck in the local optima of multi-modal posterior distributions for real data induced by such parameters.
+These methods are known to mix quickly when sampling from the simple uniform distribution of trees but may become stuck in the local optima of multi-modal posterior distributions for real data induced by non-uniform likelihoods.
 The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating study of the underlying \emph{rSPR graph} in greater~detail.
@@
 Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
-This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of random search on the rSPR graph.
+This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of sampling the rSPR graph.
 
@@
 Researchers have previously identified slow mixing in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}.
-On the other hand, fast mixing has been identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu} or with no data~\cite{spade2014note}.
+On the other hand, fast mixing has been identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu} or with a uniform distribution~\cite{spade2014note}.
 % Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
@@
 We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
-As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph (or, as we do here, the full graph for small numbers of leaves).
+As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!
+= 3 \cdot 5 \cdot \ldots \cdot (2n-3)$ trees, this is a significant improvement in practice for exploring large subsets of the graph (or, as we do here, the full graph for small numbers of leaves).
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
@@
 \begin{equation}
-\ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p}
+\ric{m; x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{m; x,y}}{p} .
 \label{eq:asympCurvatureDef}
@@
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
-As we prove in Lemma~\ref{lem:asymptotic}, the notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
+As we prove in Lemma~\ref{lem:asymptotic}, the notions of coarse and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
 
@@
 
-It is necessary to have an efficient means of constructing the full rSPR graph for a fixed number of leaves in order to study it.
+It is necessary to have an efficient method of constructing the full rSPR graph for a fixed number of leaves in order to study it.
 The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
@@
 Begin with an empty graph $G$.
-For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
+For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} below to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
 This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
@@
 Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
-$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
+$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
 trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
@@
 	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + 1/6$$
+	\le \curvature{T,S} + 1/6 .$$ 
 \end{restatable}
@@
 		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
-		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\}
+		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\}
 	\end{align*}
@@
 	\begin{align*}
-		\cuttable{\degree{S} - \degree{T}
-		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
-		- (3n^2 - 13n + 14) \\}
-		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
+		{\Delta}N &= \degree{S} - \degree{T} \\
+		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
+		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
+		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
 		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
-		=\ &n^2 - 5n + 6.
+		&=\ n^2 - 5n + 6.
 	\end{align*}
@@
 \begin{proof}
-By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
+By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
@@
 		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
-		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}
+		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
 	\end{align*}
@@
 First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
-Let $(T'_n, S'_n)$ be any of the $o$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
+Let $(T'_n, S'_n)$ be any of the $\overlap$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
 We pair these trees in the mass transport.
@@
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
-	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
+	By Corollary~\ref{cor:paired_neighbors}, there are at most $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
 	We now construct a pair of trees that maximizes both effects.
@@
 		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
-		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
+		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}.
 	\end{align*}

### 2015-11-03 curvature.tex
integrate cuttables into main text

@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
-\cuttable[Figures, tables, and proofs of our theorems and lemmas can be found in the appendix.]
 
@@
 The $V$-tree induced by $T$ is the smallest tree $T|V$ that can be obtained from $T(V)$ by suppressing unlabelled nodes with fewer than two children (Fig.~\ref{fig:induced}).
-For the rest of the paper, \textbf{we will assume that all phylogenetic trees are binary and rooted}\cuttable{, and thus that tree inclusion is rooted tree inclusion}.
+For the rest of the paper, \textbf{we will assume that all phylogenetic trees are binary and rooted}, and that tree inclusion is rooted tree inclusion.
 
@@
 Then the operation adds a new node $p'_x$ to $T_{p_x}$ by subdividing an edge of $T_{p_x}$ and adding a new edge $(x, p'_x)$, making $x$ a child of $p'_x$.
-Finally, $p_x$ is suppressed\cuttable{, joining the two edges on either side of that node}.
+Finally, $p_x$ is suppressed, joining the two edges on either side of that node.
 See Figure~\ref{fig:spr} for an example.
@@
 Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
-Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with \cuttable[a]{the given} random walk.
+Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
 \begin{figure}[!h]
@@
 	\label{lem:select_random_neighbor}
-	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time\cuttable{ using $\OhOf{n}$ space}.
+	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
 \end{restatable}
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
-We sampled \cuttable{a }200,000\cuttable{-iteration} random walk\cuttable[ iterations]{} on the 4-leaf rSPR graph and \cuttable{a }50,000-iteration\cuttable[s]{ random walk} on the 5-leaf rSPR graph.
+We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
 
@@
 \subsection{Computing curvature values.}
-To compute curvature values, we first \cuttable[computed]{used \texttt{dense\_spr\_graph} to compute} the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
@@
 
-Here, to gain more insight, we used simulation to approximate the entire access time distribution.
-Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
-Thus rather than enumerate access times between trees, which would have required a tremendous amount of memory and computational power to obtain accurate estimates, we enumerate times between pairs of trees in a tanglegram.
-To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using \cuttable{our }group-theoretic methods \cite{tangle}.
-
 \begin{table}
@@
 
-We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
-Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
-We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
-For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines\cuttable{ in Fig.~\ref{fig:kappa-access}}) is the count for time 2 minus the count for time 3.
-Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
-This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
-On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$\cuttable{ and instead achieves stationarity}, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
+Here, to gain more insight, we used simulation to approximate the entire access time distribution.
+Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
+Thus rather than enumerate access times between trees, which would have required a tremendous amount of memory and computational power to obtain accurate estimates, we enumerate times between pairs of trees in a tanglegram.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using our group-theoretic methods \cite{tangle}.
 
@@
 \end{table}
+
+We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
+Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
+We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
+For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access}) is the count for time 2 minus the count for time 3.
+Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
+This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
+On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
 
@@
 We can improve these bounds in the case of adjacent trees.
-To do so, we \cuttable{require the following lemma that }characterize\cuttable{s} exactly how the degree of a tree changes after an rSPR operation.
+To do so, we require the following lemma that characterizes how the degree of a tree changes after an rSPR operation.
 See Figure~\ref{fig:spr-degree-change} for an illustration.
@@
 Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such ``triangles'' involving two adjacent trees has a key role in determining whether \cuttable[their]{the} curvature \cuttable{of two adjacent trees }is positive or negative.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive or negative.
 
@@
 \section{Curvature}
-We now consider properties of the uniform \cuttable{(a.k.a.\ isotropic)} random walk on the $n$-leaf rSPR graph.
+We now consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.
 Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
-\cuttable{We denote the coarse curvature of the uniform random walk between two trees $T$ and $S$ with $\curvature{T,S}$.}
-Recall that \cuttable[the coarse uniform random walk curvature ]{}$\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
-For the uniform random walk, $m_T$ \cuttable[assigns]{is simply the probability measure assigning} a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
+Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
+For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
 Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
@@
 Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
 \end{restatable}
@@
 Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-\cuttable[W]{However, w}e suspect that this will require a \cuttable[new]{greater} understanding of the distribution of tree neighborhoods with respect to one another\cuttable{ than is currently known}.
+However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
 Next, we bound the minimum curvature of two adjacent trees.
@@
 
-Note that this bound is not tight (at least for small $n$)\cuttable{ as it is rarely necessary to transport mass the maximum distance between unpaired trees}.
-\cuttable{We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.}
-However, we note that the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
+Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
+We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.
+However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
 
 We next bound the difference between the coarse and asymptotic curvatures.
-Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that \cuttable{remains at a given tree with probability $1-p$ and }moves with probability $p$.
+Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and moves with probability $p$.
 For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
 The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
-\cuttable[T]{As we now prove, t}hese two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$\cuttable[:]{.}
+As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
 
@@
 Finally, we bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
-Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, \cuttable{which in this case is }$\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
-The mass distribution \cuttable{for the MH random walk }thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-Note that \cuttable{the same statement and proof of }Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
+Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, which in this case is $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+Note that the same statement and proof of Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
 
@@
 However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will \cuttable{significantly }reduce the posterior distribution to a more realistic effective size, and \cuttable{in certain cases }will lead to significant ``bottlenecks'' like those we have observed in real data.
-\cuttable[T]{In those cases t}he curvature between \cuttable{two }trees at either end of a bottleneck will describe how difficult it is to traverse\cuttable{ the bottleneck}.
-\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
+In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
+%\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
 % We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
 
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, \cuttable[many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.]{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
-We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
-We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
-Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.
+%{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+%We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
+%We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
+%Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
 
@@
 	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
-	Therefore, the procedure uses linear time and space and \cuttable[is correct]{selects an rSPR neighbor of $T$ uniformly at random}.
+	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
 \end{proof}
@@
 		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
-		\cuttable{= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\}
+		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
 		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
-		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\}
+		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
 	\end{align*}
@@
 		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
-		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
+		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
 		&=\ n^2 - 5n + 6.
@@
 (iv) $R$'s parent subtree may no longer be moved into $V$.
-No additional moves are introduced or blocked\cuttable{ by the original rSPR operation on $R$}.
-
-Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges\cuttable{(recall that we are excluding any ``root edge'' in these calculations)}.
+No additional moves are introduced or blocked by the original rSPR operation on $R$.
+
+Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges(recall that we are excluding any ``root edge'' in these calculations).
 In the first case there are $a$ subtrees that can now be moved onto the $2k$ edges in $R$ (including its newly introduced parent edge and one of the newly subdivided root edges of $V$) for a total gain of $2ka$ distinct moves.

### 2015-11-04 curvature.tex
down to 15 pages for ANALCO proceedings

@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
+Selected proofs of our theorems and lemmas can be found in the appendix.
+Proofs omitted for space can be found in~\cite{curvature_arxiv}.
 
@@
 Positive curvature implies that the neighborhoods $m_x$ and $m_y$ are closer in transportation distance than point masses at $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are more distant than point masses at $x$ and $y$.
-Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.
+Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
 \begin{figure}[!h]
 	\hspace*{\stretch{1}}
-	\includegraphics[width=0.5\textwidth]{figs/neighborhood}
-	\hspace*{\stretch{1}}
-	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections between neighbors.}
+	\includegraphics[width=0.43\textwidth]{figs/neighborhood}
+	\hspace*{\stretch{1}}
+	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
 	\label{fig:neighborhood}
@@
 
+\footnotesize
 \bibliographystyle{siam}
 \bibliography{curvature}
+\normalsize
 
@@
 
-\section{Supplementary Proofs}
+\section{Selected Proofs}
 
@@
 
+\cuttable{
 \degreemaxdelta*
@@
 \end{proof}
-
+}
+
+\cuttable{
 \degreechange*
@@
 \end{proof}
-
+}
+
+\cuttable{
 \pairedneighbors*
@@
 	Let ${\Delta}N = \degree{S} - \degree{T}$.
+	\cuttable[
+		In both cases we have
+		$${\Delta}N \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
+	]{
 	If we move the additional leaf, we have:
@@
 \end{align*}
+}
 proving (i).
@@
 
+\cuttable{
 \curvaturedistancebound*
@@
 \end{proof}
-
+}
+
+\cuttable{
 \maxadjacentcurvature*
@@
 		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
-		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
+		\cuttable{		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\}
 		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
@@
 
+\cuttable{
 \mhcurvature*

### 2016-03-08 curvature.tex
preprint url
TCS format, started moving proofs inline

@@
-%\documentclass[11pt,onecolumn,conference]{IEEEtran}
-\documentclass[10pt,twoside,leqno,twocolumn]{article}
-\usepackage{ltexpprt}
+\documentclass[]{elsarticle}
 
 %\usepackage[foot]{amsaddr}
-\usepackage{amsmath}
+\usepackage{amsmath, amsthm}
 \usepackage{csvsimple}
@@
 
-\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\thanks{This work was funded by National Science Foundation award 1223057.
-Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.}}
-
-\author{Chris Whidden\thanks{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, USA 98109. \set{cwhidden,matsen}@fredhutch.org} \\
-\and
-Frederick A Matsen IV\footnotemark[2]}
-\date{}
-
-\maketitle
+\journal{Theoretical Computer Science}
+
+\begin{frontmatter}
+
+%\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\thanks{
+	\title{Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}}
+
+	\tnotetext[t1]{This work was funded by National Science Foundation award 1223057.} 
+	\tnotetext[t2]{A preliminary version of this work appeared in the 2016 Proceedings of the Thirteenth Workshop on Analytic Algorithmics and Combinatorics (ANALCO).}
+
+	\author{Chris Whidden\corref{cor1}\fnref{fn1}}
+\ead{cwhidden@fredhutch.org}
+
+\author{Frederick A. Matsen IV\corref{cor2}}
+\ead{matsen@fredhutch.org}
+
+\cortext[cor1]{Corresponding author}
+\fntext[fn1]{Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.}
+
+\address{Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, USA 98109}
 
@@
 This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of sampling the rSPR graph.
-
 \end{abstract}
+
+\begin{keyword}
+keywords \sep go \sep here
+\end{keyword}
+
+\end{frontmatter}
 
@@
 Statistical phylogenetics is the currently most popular means of reconstructing phylogenetic trees, in which the tree is viewed as an unknown parameter in a likelihood-based statistical inference problem.
-The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution~\cite{felsenstein1981evolutionary}.
+The likelihood function in this setting is the likelihood of generating the observed sequences via a continuous time Markov chain (CTMC) evolving down the tree starting from a sequence assumed to be sampled from the stationary distribution~\citep{felsenstein1981evolutionary}.
 The lengths of the branches of the phylogenetic tree give the ``time'' parameter in the CTMC, where the generated sequence accrues mutations, typically in an IID manner across sites.
@@
 Phylogenetic search algorithms typically attempt to do so through a combination of modifications to the continuous parameters and tree topology.
-Topology changes have been identified as the main limiting factor of Bayesian MCMC algorithms~\cite{lakner2008efficiency,hohna2012guided}, as other parameters cannot be accurately estimated if the topology distribution is not accurately sampled.
-Commonly used phylogenetics software packages such as MrBayes~\cite{Ronquist2012-hi} and BEAST~\cite{bouckaert2014beast} rearrange subtrees via subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\cite{robinson1971comparison}.
+Topology changes have been identified as the main limiting factor of Bayesian MCMC algorithms~\citep{lakner2008efficiency,hohna2012guided}, as other parameters cannot be accurately estimated if the topology distribution is not accurately sampled.
+Commonly used phylogenetics software packages such as MrBayes~\citep{Ronquist2012-hi} and BEAST~\citep{bouckaert2014beast} rearrange subtrees via subtree-prune-regraft (SPR) moves (Figure~\ref{fig:spr}) or the subset of SPR moves called nearest neighbor interchanges (NNI)~\citep{robinson1971comparison}.
 Thus, phylogenetic searches can be viewed as traversing the \emph{SPR graph}: the graph with phylogenetic trees as vertices and SPR adjacencies as edges.
@@
 It has become increasingly clear that the structure of the SPR graph plays an important role in determining the accuracy of tree searches.
-Researchers have previously identified slow mixing in MCMC with pathological data~\cite{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}.
-On the other hand, fast mixing has been identified with exceptionally well-behaved data~\cite{Stefankovic2011-hu} or with a uniform distribution~\cite{spade2014note}.
-% Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\cite{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\cite{Evans2006-xh,Athreya2014-de}.
-Studies on real data~\cite{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
+Researchers have previously identified slow mixing in MCMC with pathological data~\citep{Mossel2005-ly,Mossel2006-fo,Ronquist2006-fv}.
+On the other hand, fast mixing has been identified with exceptionally well-behaved data~\citep{Stefankovic2011-hu} or with a uniform distribution~\citep{spade2014note}.
+% Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\citep{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\citep{Evans2006-xh,Athreya2014-de}.
+Studies on real data~\citep{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
 Previously, the lack of sufficient computational tools for examining phylogenetic posteriors in terms of SPR operations made it difficult to determine the cause of these difficulties.
-By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\cite{Whidden2015-yi}, and that multimodal posteriors are common and separated by ``bottlenecks" of specific classes of SPR moves.
+By developing the first such tools, we recently showed that graph structure has a significant effect on MCMC mixing with MrBayes applied to real data~\citep{Whidden2015-yi}, and that multimodal posteriors are common and separated by ``bottlenecks" of specific classes of SPR moves.
 
 Although the SPR graph is thus very important in determining the success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself.
-\cite{Song2003-gf}~developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
-\cite{Ding2011-bj}~showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
+\citep{Song2003-gf}~developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
+\citep{Ding2011-bj}~showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
 \begin{equation}
@@
 We are not aware of any further work investigating properties of the SPR graph, which may be due to its complexity.
-Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\cite{bordewich05,hickey2008sdc}.
-Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\cite{whidden2013hybridization,Whidden2015-yi} and begun to allow such investigation.
-
-Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\cite{Ollivier2009-bw,Joulin2010-jg}.
+Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\citep{bordewich05,hickey2008sdc}.
+Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\citep{whidden2013hybridization,Whidden2015-yi} and begun to allow such investigation.
+
+Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\citep{Ollivier2009-bw,Joulin2010-jg}.
 In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings points together.
 Applying the framework to Brownian motion on a manifold returns the classical definition of Ricci curvature.
-Curvature is determined by the ratio of the earth mover's distance~\cite{rubner2000earth} between neighborhoods of a pair of vertices given by a random walk and the distance between the vertices.
+Curvature is determined by the ratio of the earth mover's distance~\citep{rubner2000earth} between neighborhoods of a pair of vertices given by a random walk and the distance between the vertices.
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying reasonable assumptions, which includes biased walks such as MCMC.
-This approach has been useful for determining properties of a wide variety of graphs including the internet topology~\cite{ni2015ricci} and cancer networks~\cite{sandhu2015graph}.
+This approach has been useful for determining properties of a wide variety of graphs including the internet topology~\citep{ni2015ricci} and cancer networks~\citep{sandhu2015graph}.
 
@@
 Pairs of trees separated by such moves correspond to the pairs with negative curvature identified in our simulation results.
-These pairs occur infrequently in these well-connected graphs, however, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\cite{Whidden2015-yi}.
+These pairs occur infrequently in these well-connected graphs, however, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\citep{Whidden2015-yi}.
 In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
@@
 Selected proofs of our theorems and lemmas can be found in the appendix.
-Proofs omitted for space can be found in~\cite{curvature_arxiv}.
+Proofs omitted for space can be found in~\citep{curvature_arxiv}.
 
@@
 
-We follow the definitions and notation from~\cite{bordewich05,whidden2013hybridization, Whidden2015-yi}.
+We follow the definitions and notation from~\citep{bordewich05,whidden2013hybridization, Whidden2015-yi}.
 A (rooted binary phylogenetic) $X$-tree is a rooted tree $T$ whose nodes have zero or two children such that the leaves of $T$ are bijectively labelled with the members of a label set $X$.
-As in~\cite{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$ (Fig.~\ref{fig:x-tree}).
+As in~\citep{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$ (Fig.~\ref{fig:x-tree}).
 We generally use $n$ to refer to the number of leaves in an $X$-tree.
@@
 For the purposes of this paper, we will specialize to that space being a graph equipped with the shortest-path distance.
-For a more rigorous presentation in the more general setting of a Polish metric space, see~\cite{Ollivier2009-bw} or the survey~\cite{Ollivier2010-ao}.
+For a more rigorous presentation in the more general setting of a Polish metric space, see~\citep{Ollivier2009-bw} or the survey~\citep{Ollivier2010-ao}.
 
 Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after one step of the random walk, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The transportation distance~\cite{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
+The transportation distance~\citep{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\citep{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
 \vspace{-0.5em}
@@
 
-Lin et al.~\cite{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
+Lin et al.~\citep{lin2011ricci} defined a variant definition of curvature in terms of lazy random walks which Loisel and Romon~\citep{Loisel2014-hu} dubbed the \emph{asymptotic Ricci-Olivier curvature}.
 The lazy random walk only travels according to $m_x$ with probability $p$ and otherwise stays put.
@@
 As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
-This definition of curvature is invariant of $p$ for small $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
+This definition of curvature is invariant of $p$ for small $p$~\citep{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
 Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
@@
 It is necessary to have an efficient method of constructing the full rSPR graph for a fixed number of leaves in order to study it.
-The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
+The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\citep{Whidden2015-yi}.
 Here we reduce that time to $\OhOf{mn^3}$.
@@
 
-In previous work~\cite{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify local maxima.
-Although the SPR distance (rooted and unrooted) is NP-hard to compute~\cite{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\cite{bordewich05}.
-In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\cite{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
+In previous work~\citep{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify local maxima.
+Although the SPR distance (rooted and unrooted) is NP-hard to compute~\citep{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05}.
+In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\citep{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
 We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
@@
 The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
-The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of each tree as a string.
+The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{wiki:newick} representation of each tree as a string.
 These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
@@
 
-%\begin{thm}
-%	\label{thm:construct_graph}
-%	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
-%\end{thm}
-\begin{restatable}{thm}{constructgraph}
+\begin{thm}
 	\label{thm:construct_graph}
 	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
-\end{restatable}
-
-We implemented this procedure in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
+\end{thm}
+\begin{proof}
+	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
+
+	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
+red-black trees~\citep{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
+	As described above, the integer labels are simply the order of the input trees.
+	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
+	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
+	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
+	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
+	We store the tree to index mappings for current vertices of $G$ in a trie~\citep{fredkin1960trie} using Newick representation.
+	This requires only $\OhOf{n}$-time for each tree (i.e.
+a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
+	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
+	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
+\end{proof}
+
+We implemented this procedure in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\citep{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
 The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
@@
 Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
-Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
+Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\citep{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
 
@@
 Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
-We note that the recursive formula of Song~\cite{Song2003-gf} for the degree of a tree does not group rSPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific rSPR neighbor by index.
+We note that the recursive formula of Song~\citep{Song2003-gf} for the degree of a tree does not group rSPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific rSPR neighbor by index.
 
@@
 
-\begin{restatable}{lem}{computedegree}
+\begin{lem}
 	\label{lem:compute_degree}
@@
 	\end{cases}.$$
-\end{restatable}
+\end{lem}
+\begin{proof}
+	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
+	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
+	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
+	Call these $U$ and $V$, respectively.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
+	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
+	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (move type (iii)) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (move type (iv)), a contradiction.
+	Thus the claim holds.
+\end{proof}
 
@@
 
-\begin{restatable}{lem}{selectrandomneighbor}
+\begin{lem}
 	\label{lem:select_random_neighbor}
 	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
-\end{restatable}
+\end{lem}
+\begin{proof}
+	We apply the above procedure.
+	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
+	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
+	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
+	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
+	Both of these traversals require $\OhOf{n}$-time.
+	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
+	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
+	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
+	Thus, the for loop requires linear time.
+	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
+	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
+\end{proof}
 
@@
 
-We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
+We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\citep{random_spr_walk}.
 We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
@@
 To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then computed curvatures for given pairs of trees directly, by using linear programming~\cite{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \cite{SAGE} front-end to the GLPK~\cite{glpk} solver; code can be found in \cite{gricci} which grew from the code described in \cite{Loisel2014-hu}.
+We then computed curvatures for given pairs of trees directly, by using linear programming~\citep{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \citep{SAGE} front-end to the GLPK~\citep{glpk} solver; code can be found in \citep{gricci} which grew from the code described in \citep{Loisel2014-hu}.
 
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \cite{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \cite{SAGE} and GAP4 \cite{GAP4} code is at \cite{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
 
@@
 \subsection{Access time simulation.}
-The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
+The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\citep{lovasz1993random}; we were interested in the connection between curvature and access time.
 In previous work, we computed mean access times (MAT) between pairs of trees in  MCMC random walks: the mean number of iterations required to move from one tree to the other.
-We applied this work to demonstrate the influence of SPR graph structure on real MCMC posteriors sampled with MrBayes~\cite{Whidden2015-yi} using \texttt{sprspace}~\cite{sprspace}.
+We applied this work to demonstrate the influence of SPR graph structure on real MCMC posteriors sampled with MrBayes~\citep{Whidden2015-yi} using \texttt{sprspace}~\citep{sprspace}.
 
@@
 Thus rather than enumerate access times between trees, which would have required a tremendous amount of memory and computational power to obtain accurate estimates, we enumerate times between pairs of trees in a tanglegram.
-To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using our group-theoretic methods \cite{tangle}.
+To calculate the empirical distributions of access times we aggregate all access times for the same tanglegram using our group-theoretic methods \citep{tangle}.
 
@@
 
-\begin{lem}[{Song~\cite{Song2003-gf}}]
+\begin{lem}[{Song~\citep{Song2003-gf}}]
 \cuttable{    \pushQED{\qed}}
@@
 Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\cite{Ollivier2009-bw}).
+Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\citep{Ollivier2009-bw}).
 This directly leads to the following Corollary:
@@
 Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
-We also note that the lower bounds in this section do not follow from the more general setting described in \cite{Jost2013-ce}.
+We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
 However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
@@
 Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
-However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\cite{Whidden2015-yi}.
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
 Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
@@
 %\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
-% We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.
-
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\cite{Song2006-xe}, graphs of BEAST~\cite{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as trees.
+% We will likely follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.
+
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\citep{Song2006-xe}, graphs of BEAST~\citep{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as trees.
 %{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
 %We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
-%We will also explore random walks on ranked trees~\cite{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
-%Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
+%We will also explore random walks on ranked trees~\citep{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \citep{Drummond2012-ek} that infer rooted ``time-trees.''
+%Random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
 
@@
 
-\footnotesize
-\bibliographystyle{siam}
+\bibliographystyle{elsarticle-num}
 \bibliography{curvature}
-\normalsize
 
@@
 \section{Selected Proofs}
-
-\constructgraph*
-\begin{proof}
-	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
-
-	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
-red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
-	As described above, the integer labels are simply the order of the input trees.
-	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
-	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
-	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
-	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using Newick representation.
-	This requires only $\OhOf{n}$-time for each tree (i.e.
-a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
-	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
-	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
-\end{proof}
-
-\computedegree*
-\begin{proof}
-	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
-	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
-	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
-	Call these $U$ and $V$, respectively.
-	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
-	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
-	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (move type (iii)) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (move type (iv)), a contradiction.
-	Therefore the claim holds.
-\end{proof}
-
-\selectrandomneighbor*
-\begin{proof}
-	We apply the above procedure.
-	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
-	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
-	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
-	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
-	Both of these traversals require $\OhOf{n}$-time.
-	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
-	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
-	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
-	Thus, the for loop requires linear time.
-	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
-	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
-\end{proof}
 

### 2016-03-09 curvature.tex
moved the remainder of proofs to the main text
integrate remaining 'cuttables' into text

@@
 The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
-Selected proofs of our theorems and lemmas can be found in the appendix.
-Proofs omitted for space can be found in~\citep{curvature_arxiv}.
 
@@
 \begin{lem}[{Song~\citep{Song2003-gf}}]
-\cuttable{    \pushQED{\qed}}
 	\label{lem:degree_extremes}
@@
 	\end{enumerate}
-\cuttable{    \popQED}
 \end{lem}
@@
 We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
-\begin{restatable}{lem}{degreemaxdelta}
+\begin{lem}
 	\label{lem:degree_max_delta}
@@
 	\end{enumerate}
-\end{restatable}
-
-We can improve these bounds in the case of adjacent trees.
-To do so, we require the following lemma that characterizes how the degree of a tree changes after an rSPR operation.
-See Figure~\ref{fig:spr-degree-change} for an illustration.
-
-\begin{restatable}{lem}{degreechange}
-\label{lem:degree_change}
-Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
-Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
-Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
-Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
-Then the degrees of $T$ and $S$ differ by:
-$$2\left(k(a-b) + i - j\right).$$
-
-\end{restatable}
-
-Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
-That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
-To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
-The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they push the curvature of those trees towards 0.
-
-\begin{restatable}{cor}{pairedneighbors}
-\label{cor:paired_neighbors}
-Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
-$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
-trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
-\end{restatable}
-
-We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
-\begin{restatable}{lem}{degreemaxdeltaadjacent}
-	\label{lem:degree_max_delta_adjacent}
-	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
-	Then:
-	\begin{enumerate}
-		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
-		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
-		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
-	\end{enumerate}
-\end{restatable}
-
-Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive or negative.
-
-\begin{restatable}{lem}{sharedneighbors}
-	\label{lem:shared_neighbors}
-Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
-Then $\size{N(T) \cap N(S)} \le 6n - 17$.
-\end{restatable}
-
-\section{Curvature}
-We now consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.
-Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
-Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
-For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
-Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
-
-\begin{restatable}{thm}{flatcurvature}
-\label{thm:flat_curvature}
-Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
-Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
-Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
-\end{restatable}
-
-Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
-
-\begin{restatable}{lem}{curvaturedistancebound}
-	\label{lem:curvature_distance_bound}
-	Let $T$ and $S$ be two trees. Then:
-	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
-\end{restatable}
-
-\begin{restatable}{lem}{maxadjacentcurvature}
-	\label{lem:max_adjacent_curvature}
-	The maximum curvature between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
-\end{restatable}
-This bound is tight and has been verified computationally for $n \le 7$.
-
-It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
-Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
-However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
-Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
-Moreover, nearly all of the neighbors of these pairs approach each either.
-We thus conjecture the following:
-\begin{conj}
-	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
-	Then:
-	\begin{enumerate}
-		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
-		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
-	\end{enumerate}
-\end{conj}
-Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
-Next, we bound the minimum curvature of two adjacent trees.
-
-\begin{restatable}{lem}{curvaturelowerbound}
-\label{lem:curvature_lower_bound}
-	The curvature between adjacent trees with $n$ leaves is at least
-	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
-\end{restatable}
-
-We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
-Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
-Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\citep{Ollivier2009-bw}).
-This directly leads to the following Corollary:
-
-\begin{cor}
-	The curvature between two trees is at least $-\frac{2}{5}$.
-\end{cor}
-
-Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
-We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
-However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
-
-We next bound the difference between the coarse and asymptotic curvatures.
-Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and moves with probability $p$.
-For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
-The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
-As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
-
-\begin{restatable}{lem}{asymptotic}
-\label{lem:asymptotic}
-	Let $T$ and $S$ be trees with $n$ leaves.
-	Then:
-	\begin{enumerate}
-		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
-		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
-	\end{enumerate}
-\end{restatable}
-
-Finally, we bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
-Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, which in this case is $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
-The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
-Note that the same statement and proof of Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
-
-\begin{restatable}{lem}{mhcurvature}
-	\label{lem:mh_curvature}
-	Let $T$ and $S$ be trees with $n$ leaves. Then:
-	\begin{align*}
-	\curvature{T,S} - \frac{1}{3\dspr{T,S}}
-	&\le \curvature{\MH; T,S} \\
-	\curvature{\MH; T,S}
-	&\le \curvature{T,S} + \frac{1}{3\dspr{T,S}},\text{ and }
-\end{align*}
-	$$\curvature{T,S} - 1/6
-	\le \curvature{\MH; T,S}
-	\le \curvature{T,S} + 1/6 .$$ 
-\end{restatable}
-
-\section{Conclusion and future work}
-In summary, we have gone beyond graph diameter and vertex degree to substantially advance understanding of the phylogenetic rSPR graph.
-We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
-We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
-Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
-
-In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
-Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
-However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
-Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
-In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
-%\cuttable{Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).}
-% We will likely follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.
-
-Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored including NNI graphs, unrooted SPR graphs, graphs of ranked trees~\citep{Song2006-xe}, graphs of BEAST~\citep{Drummond2012-ek} rooted ``time-trees,'' and random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as trees.
-%{we will extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
-%We will next consider the case of unrooted SPR, which is more commonly used in phylogenetic search.
-%We will also explore random walks on ranked trees~\citep{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \citep{Drummond2012-ek} that infer rooted ``time-trees.''
-%Random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.}
-
-\section{Acknowledgements}
-The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
-They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
-
-\bibliographystyle{elsarticle-num}
-\bibliography{curvature}
-
-\appendix
-
-\section{Selected Proofs}
-
-\cuttable{
-\degreemaxdelta*
+\end{lem}
 \begin{proof}
@@
 \end{proof}
-}
-
-\cuttable{
-\degreechange*
+
+We can improve these bounds in the case of adjacent trees.
+To do so, we require the following lemma that characterizes how the degree of a tree changes after an rSPR operation.
+See Figure~\ref{fig:spr-degree-change} for an illustration.
+
+\begin{lem}
+\label{lem:degree_change}
+Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
+Let $L$ be the $\lca{U,V}$ in $T$.
+Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
+Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
+Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
+Then the degrees of $T$ and $S$ differ by:
+$$2\left(k(a-b) + i - j\right).$$
+
+\end{lem}
 \begin{proof}
@@
 \end{proof}
-}
-
-\cuttable{
-\pairedneighbors*
+
+Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
+That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
+To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
+The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they push the curvature of those trees towards 0.
+
+\begin{cor}
+\label{cor:paired_neighbors}
+Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
+$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
+trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
+\end{cor}
 \begin{proof}
@@
 \end{proof}
-}
-
-\degreemaxdeltaadjacent*
+
+We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
+\begin{lem}
+	\label{lem:degree_max_delta_adjacent}
+	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
+	Then:
+	\begin{enumerate}
+		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
+		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
+		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
+	\end{enumerate}
+\end{lem}
 \begin{proof}
@@
 	Let ${\Delta}N = \degree{S} - \degree{T}$.
-	\cuttable[
-		In both cases we have
-		$${\Delta}N \le 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},$$
-	]{
 	If we move the additional leaf, we have:
@@
 \end{align*}
-}
 proving (i).
@@
 
-\sharedneighbors*
+Next, we bound the number of neighbors shared by two adjacent trees.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive or negative.
+
+\begin{lem}
+	\label{lem:shared_neighbors}
+Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
+Then $\size{N(T) \cap N(S)} \le 6n - 17$.
+\end{lem}
 \begin{proof}
@@
 
-\flatcurvature*
+\section{Curvature}
+We now consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.
+Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
+Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
+For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
+Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
+
+\begin{thm}
+\label{thm:flat_curvature}
+Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
+Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
+Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.
+\end{thm}
 \begin{proof}
@@
 
-\cuttable{
-\curvaturedistancebound*
+Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
+
+\begin{lem}
+	\label{lem:curvature_distance_bound}
+	Let $T$ and $S$ be two trees. Then:
+	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
+\end{lem}
 \begin{proof}
@@
 \end{proof}
-}
-
-\cuttable{
-\maxadjacentcurvature*
+
+\begin{lem}
+	\label{lem:max_adjacent_curvature}
+	The maximum curvature between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
+\end{lem}
 \begin{proof}
@@
 \end{proof}
-}
-
-\curvaturelowerbound*
+
+This bound is tight and has been verified computationally for $n \le 7$.
+
+It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
+Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
+However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
+Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
+Moreover, nearly all of the neighbors of these pairs approach each either.
+We thus conjecture the following:
+\begin{conj}
+	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
+	Then:
+	\begin{enumerate}
+		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
+		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
+	\end{enumerate}
+\end{conj}
+Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
+However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
+Next, we bound the minimum curvature of two adjacent trees.
+
+\begin{lem}
+\label{lem:curvature_lower_bound}
+	The curvature between adjacent trees with $n$ leaves is at least
+	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
+\end{lem}
 \begin{proof}
@@
 
-\asymptotic*
+We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
+Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\citep{Ollivier2009-bw}).
+This directly leads to the following Corollary:
+
+\begin{cor}
+	The curvature between two trees is at least $-\frac{2}{5}$.
+\end{cor}
+
+Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
+We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
+However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
+
+We next bound the difference between the coarse and asymptotic curvatures.
+Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and moves with probability $p$.
+For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
+The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
+As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.
+
+\begin{lem}
+\label{lem:asymptotic}
+	Let $T$ and $S$ be trees with $n$ leaves.
+	Then:
+	\begin{enumerate}
+		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
+		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
+	\end{enumerate}
+\end{lem}
 \begin{proof}
@@
 		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
-		\cuttable{		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\}
+		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
 		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
@@
 
-\cuttable{
-\mhcurvature*
+Finally, we bound the difference between the curvature of the uniform random walk $\curvature{T,S}$ and that of the Metropolis-Hastings (MH) random walk $\curvature{\MH; T,S}$.
+Recall that this random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, which in this case is $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
+The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+Note that the same statement and proof of Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
+
+\begin{lem}
+	\label{lem:mh_curvature}
+	Let $T$ and $S$ be trees with $n$ leaves. Then:
+	\begin{align*}
+	\curvature{T,S} - \frac{1}{3\dspr{T,S}}
+	&\le \curvature{\MH; T,S} \\
+	\curvature{\MH; T,S}
+	&\le \curvature{T,S} + \frac{1}{3\dspr{T,S}},\text{ and }
+\end{align*}
+	$$\curvature{T,S} - 1/6
+	\le \curvature{\MH; T,S}
+	\le \curvature{T,S} + 1/6 .$$ 
+\end{lem}
 \begin{proof}
@@
 \end{proof}
-}
+
+\section{Conclusion and future work}
+In summary, we have gone beyond graph diameter and vertex degree to substantially advance understanding of the phylogenetic rSPR graph.
+We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
+We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
+
+In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
+Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
+Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
+In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
+Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
+One avenue of exploration will be to follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.
+
+One could extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored.
+In particular, the case of unrooted SPR, which is also commonly used in phylogenetic search, is an obvious next step.
+One could also explore random walks on ranked trees~\citep{Song2006-xe} and graphical models of tree space relevant for phylogenetic algorithms such as BEAST \citep{Drummond2012-ek} that infer rooted ``time-trees.''
+Random walks on other discrete structures such as partitions~\citep{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.
+
+\section{Acknowledgements}
+The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
+They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
+A preliminary version of this work appeared in the 2016 Proceedings of the Thirteenth Workshop on Analytic Algorithmics and Combinatorics (ANALCO).
+The authors would like to thank the conference organizers, reviewers, and attendees for helpful comments and discussions that greatly improved this full version of our work.
+
+\bibliographystyle{elsarticle-num}
+\bibliography{curvature}
 

### 2016-03-10 curvature.tex
removed uneccesary page break
4 leaf treespace figure

@@
 Then the rSPR graph $G$ of $\mathcal{T}_n$ is the graph with vertex set $V(G) = \mathcal{T}_n$ and edge set $E(G) = \set{(T,S) \mid \dspr{T,S} = 1, T \in V, S \in V}$.
+Figure~\ref{fig:4-taxa-treespace} illustrates the rSPR graph of 4-leaf trees.
 
@@
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
-\begin{figure}[!h]
-	\hspace*{\stretch{1}}
-	\includegraphics[width=0.43\textwidth]{figs/neighborhood}
-	\hspace*{\stretch{1}}
-	\caption{The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
-	\label{fig:neighborhood}
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:4-taxa-treespace}]{\includegraphics[width=0.55\textwidth]{figs/4-taxa-treespace}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:neighborhood}]{\includegraphics[width=0.4\textwidth]{figs/neighborhood}}
+	\hspace*{\stretch{1}}
+	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
 \end{figure}
-\vspace{0.1em}
-
-\pagebreak
 

### 2016-03-10 curvature.tex
figure positioning and size

@@
 
+\begin{figure}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:4-taxa-treespace}]{\includegraphics[width=0.55\textwidth]{figs/4-taxa-treespace}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:neighborhood}]{\includegraphics[width=0.4\textwidth]{figs/neighborhood}}
+	\hspace*{\stretch{1}}
+	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
+\end{figure}
+
 rSPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
@@
 Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with a random walk.
-\begin{figure}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:4-taxa-treespace}]{\includegraphics[width=0.55\textwidth]{figs/4-taxa-treespace}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:neighborhood}]{\includegraphics[width=0.4\textwidth]{figs/neighborhood}}
-	\hspace*{\stretch{1}}
-	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
-\end{figure}
 
@@
 
-\begin{figure*}
-	\hspace*{\stretch{1}}
-	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
-	\hspace*{\stretch{2}}
-	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
-	\hspace*{\stretch{1}}
-    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
-Color displays the average degree of $T_1$ and $T_2$.
-Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
-	\label{fig:rspr-scatter}
-\end{figure*}
-
 \begin{lem}
@@
 
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\subfigure[6 taxa]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter6}}
+	\hspace*{\stretch{2}}
+	\subfigure[7 taxa]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter7}}
+	\hspace*{\stretch{1}}
+    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph.
+Color displays the average degree of $T_1$ and $T_2$.
+Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
+	\label{fig:rspr-scatter}
+\end{figure*}
+
 We find a wide variation in curvature among tanglegrams (Figure~\ref{fig:rspr-scatter}).
@@
 Such pairs indicate difficult paths for phylogenetic searches, which may be exacerbated by likelihood or branch length constraints.
-
-\begin{figure*}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
-	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
-	\hspace*{\stretch{1}}
-    \caption{\
-        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
-        }
-	\label{fig:kappa-access}
-\end{figure*}
 
@@
 
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/short-time-kappa-access}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/long-time-kappa-access}}
+	\hspace*{\stretch{1}}
+    \caption{\
+        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        }
+	\label{fig:kappa-access}
+\end{figure*}
+
 We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
@@
 On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
-
-\begin{figure*}
-	\hspace*{\stretch{1}}
-	\includegraphics[width=0.8\textwidth]{figs/spr-degree-change}
-	\hspace*{\stretch{1}}
-	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
-Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
-	\label{fig:spr-degree-change}
-\end{figure*}
 
@@
 \end{proof}
+
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\includegraphics[width=0.8\textwidth]{figs/spr-degree-change}
+	\hspace*{\stretch{1}}
+	\caption{An rSPR move labelled as in Lemma~\ref{lem:degree_change}.
+Moving the grey subtree $R$ from its position adjacent to $U$ in tree $T$ to its position adjacent to $V$ in tree $S$ changes the rSPR degree by $2\left(k(a-b) + i - j\right)$.}
+	\label{fig:spr-degree-change}
+\end{figure*}
 

### 2016-03-11 curvature.tex
a bit more detail in the algorithms section

@@
 This gives bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
-However, we find that large topology changes (i.e. moving a large subtree) gives pairs of trees with negative curvature.
+However, we find that large topology changes (i.e. moving a large subtree) give pairs of trees with negative curvature.
 Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
@@
 	\hspace*{\stretch{1}}
-	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of an $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
+	\caption{(a) The rSPR graph of $X$-trees with 4 leaves. (b) The neighborhood of a particular $X$-tree $T$ with 4 leaves, showing connections with $T$ and between neighbors.}
 \end{figure}
@@
 Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
-Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\citep{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
+Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets induced by MCMC credible sets similar to~\citep{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
 
@@
 
-In particular, this formulation implies a total ordering of rSPR moves such that every move moving the same subtree $R$ forms a contiguous subsequence.
-We can thus apply the following algorithm to select a neighbor uniformly at random for a tree $T$:
+This neighbor sequence implies a total ordering of rSPR moves such that every rSPR move which moves the same subtree $R$ forms a contiguous subsequence.
+We can use this fact to quickly select a neighbor uniformly at random for a tree $T$.
+Instead of directly computing each neighbor of $T$, we count the number of neighbors $\size{N(T,u)}$ in the sequence that can be reached by moving the subtree rooted at each node $u$ of $T$.
+We then use this list of neighbor counts to select the position of a uniformly random neighbor of $T$ in the total ordering and the subtree $R$ corresponding to that position.
+By generating only the neighbors of $T$ induced by moving the subtree $R$ we then quickly identify the specific neighbor of $T$.
+We thus achieve the following algorithm to select a neighbor uniformly at random for a tree $T$:
 
@@
 
+We now prove that this algorithm is correct and runs in linear time with respect to the number of leaves in the tree.
+
 \begin{lem}
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\citep{random_spr_walk}.
-We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
+We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.\todo{How many iterations for 6 and 7 leaves?
+Was this actually 5 and 6 leaves, not 4 and 5 leaves?}
 

### 2016-03-11 curvature.tex
more exposition in degree section, idea for bounding degree change with a small number of SPRs > 1.

@@
 Having made the connection between curvature values and access times on rSPR graphs, we now consider curvature theoretically.
+In this section we consider properties of neighborhoods of trees in rSPR graphs.
+We will use these properties to bound curvature values in Section~\ref{sec:curvature}.
+
 We begin by bounding differences between degrees, and then continue by considering features relevant to the earth mover's distance that we call ``squares" and ``triangles" in the rSPR graph.
 Many of our results in this section follow from a characterization of the change in degree and distribution of permissible rSPR moves after an rSPR move is applied.
+
+Our first lemma of this section comes from Song, who characterized the degrees of ladder trees and balanced trees, and showed that the degrees of all other trees fall between these two extremes.
+Note that every tree degree is quadratic with respect to the number of leaves in the tree.
 
@@
 
-We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
+We can apply Lemma~\ref{lem:degree_extremes} to bound the ratio and difference of rSPR degree between any two arbitrary trees with $n$ leaves:
+
 \begin{lem}
@@
 
-We can improve these bounds in the case of adjacent trees.
-To do so, we require the following lemma that characterizes how the degree of a tree changes after an rSPR operation.
+Pairs of trees that are close in the graph can not achieve these extreme degree differences.
+In our next lemma, we characterize how the degree of a tree changes after an rSPR operation.
+We show that the magnitude of the degree change depends solely on the size of the subtree to be moved, the difference between the depths of its original and new locations, and the size of its original and new neighboring subtrees.
+This relation allows us to improve the degree bounds between adjacent trees.
 See Figure~\ref{fig:spr-degree-change} for an illustration.
@@
 To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
-The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they push the curvature of those trees towards 0.
+The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they signify pairs of matched paths that help push the curvature of those trees towards 0.
 
@@
 
-We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
+We now use Lemma~\ref{lem:degree_change} to improve the degree change bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
+Surprisingly, we found that the degree change between two adjacent trees may be quadratic in $n$, that is of the same order as the degree itself.
+This maximum degree change occurs when a large subtree experiences a large depth change.\todo{Can we generalize this to more distant (but necessarily still close trees)?
+Idea: consider the moves induced by an AF, solve for max cumulative degree change.
+If so, add it and suggest a better version of this as future work in the conclusion.}
+
 \begin{lem}
@@
 
-Next, we bound the number of neighbors shared by two adjacent trees.
-The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive or negative.
+In our final lemma of this section, we bound the number of neighbors shared by two adjacent trees.
+The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive, as mass assigned to a shared neighbor may not require a mass transport cost.
+We show that the maximum number of shared neighbors is linear in the number of leaves.
+In other words, only a small proportion of any two neighborhoods are shared and the relative size of shared neighborhoods grows much more slowly than the quadratic scaling of total neighborhood sizes.
 
@@
 \section{Curvature}
+\label{sec:curvature}
 We now consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.

### 2016-03-11 curvature.tex
fixed random walk sampling iteration leaf numbers, added length of 7-leaf sampling run.

@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\citep{random_spr_walk}.
-We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.\todo{How many iterations for 6 and 7 leaves?
-Was this actually 5 and 6 leaves, not 4 and 5 leaves?}
+We sampled a 200,000-iteration random walk on the 5-leaf rSPR graph, a 50,000-iteration random walk on the 6-leaf rSPR graph, and a 5,000-iteration random walk on the 7-leaf rSPR graph.
 

### 2016-03-17 curvature.tex
more explanation between lemmas of the curvature theory section. Added a bit about or conjecture to the conclusions

@@
 \subsection{Computing curvature values.}
+\label{sec:results:curvature}
 To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
@@
 \label{sec:curvature}
-We now consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.
+In this section we apply the rSPR neighborhood properties we proved in Section~\ref{sec:neighborhoods} to derive bounds on curvature.
+In particular, we consider properties of the uniform (a.k.a.\ isotropic) random walk on the $n$-leaf rSPR graph.
+
 Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
@@
 For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
-Our results follow from the lemmas of Section~\ref{sec:neighborhoods}.
+
+In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the pruned and regrafted subtree remains fixed.
+To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
+The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
+
+This flatness-in-the-limit theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
 
@@
 
-Next we note a simple and rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.
+Next we note a simple and rough bound on the curvature of two trees with respect to their distance.
+This bound arises naturally in any metric space with integer distances.
 
@@
 \end{proof}
+
+We now obtain a tighter bound on the maximum curvature between two adjacent trees.
+We do so by considering trees with a maximum intersecting set of neighbors using Lemma~\ref{lem:shared_neighbors}.
 
@@
 However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
-Next, we bound the minimum curvature of two adjacent trees.
+
+Next, we bound the minimum curvature of two adjacent trees by bounding the amount of probability mass that cannot be moved between adjacent neighbors.
 
@@
 We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
-Complete enumeration with $n \le 7$ show that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Complete enumeration with $n \le 7$ shows that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
 Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\citep{Ollivier2009-bw}).
@@
 However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
+
+\begin{cor}
+For all $n \ge 7$, there exist two adjacent trees $T$ and $S$ with $n$ leaves such that $\curvature{T,S} < 0$
+\end{cor}
 
@@
 The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
+In Lemma~\ref{lem:mh_curvature} we bound the effect of this mass on curvature and thereby bound the difference in curvatures of the two random walks.
 Note that the same statement and proof of Lemma~\ref{lem:asymptotic} holds with $\curvature{T,S}$ and $\ric{T,S}$ replaced by the MH curvatures $\curvature{\MH; T,S}$ and $\ric{\MH; T,S}$, respectively.
@@
 
+Finally, we conjectured that the maximum curvature for an rSPR graph with $n$ leaves is inversely proportional to the diameter of the graph.
+We suspect that proving or disproving this conjecture will require a new understanding of the effect of multiple rSPR moves in tandem, just as many of our results came from a better understanding of the impact of a single rSPR move.
+Such an understanding may also lead to improved curvature bounds or a means for computing them more efficiently.
+
 \section{Acknowledgements}

### 2016-04-19 curvature.tex
new access time figs

@@
 	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/short-time-kappa-access}}
+	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-short-time-kappa-access}}
 	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/long-time-kappa-access}}
+	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
 	\hspace*{\stretch{1}}

### 2016-04-19 curvature.tex
new figures with captions

--- 2016-03-17 curvature.tex
@@
 \begin{figure*}
-	\hspace*{\stretch{1}}
-	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/short-time-kappa-access}}
+	\includegraphics[width=\textwidth]{figs/access5-distribution.png}
+
+	\caption{\
+        Distribution of rSPR $\MH$ access times for all pairs of 5-taxon trees, grouped by the degree of the trees.
+        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+	}
+	\label{fig:kappa-access-distribution}
+\end{figure*}
+
+\begin{figure*}
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:6t-short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-short-time-kappa-access}}
 	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/long-time-kappa-access}}
+	\subfigure[\label{fig:7t-short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/7t-short-time-kappa-access}}
+	\hspace*{\stretch{1}}
+
+	\hspace*{\stretch{1}}
+	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
+	\hspace*{\stretch{2}}
+	\subfigure[\label{fig:7t-short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
 	\hspace*{\stretch{1}}
     \caption{\
-        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.
+        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-taxon trees with degree 44 and (b) 7-taxon trees with degree 70.
+        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-taxon trees with degree 24 and (d) 6-taxon trees with degree 44.
+				In both cases we exclude pairs that are simple inclusions of pairs of trees with a smaller number of taxa.
 				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.

### 2016-04-20 curvature.tex
revised access time figures, correct text references

@@
 	\caption{\
-        Distribution of rSPR $\MH$ access times for all pairs of 5-taxon trees, grouped by the degree of the trees.
-        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-taxon trees, grouped by the degree of the trees.
+        Color signifies rSPR distance between the trees, with green, orange, blue, light blue, and brown signifying distances of 1, 2, 3, 4, and 5, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
 	}
@@
 	\hspace*{\stretch{2}}
-	\subfigure[\label{fig:7t-short-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
+	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
 	\hspace*{\stretch{1}}
@@
 We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
-Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
+Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:kappa-access-distribution}, Fig.~\ref{fig:6t-short-time-kappa-access}, and Fig.~\ref{fig:7t-short-time-kappa-access}).
 We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
-For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access}) is the count for time 2 minus the count for time 3.
+For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 2 minus the count for time 3.
 Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
 This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the propensity of a random walk to go from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
-On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).
+On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:5t-long-time-kappa-access} and Fig.~\ref{fig:6t-long-time-kappa-access}).
 

### 2016-04-21 curvature.tex
minor edits, tanglegram reference, keywords

@@
 \begin{keyword}
-keywords \sep go \sep here
+Markov chain Monte Carlo \sep phylogenetics \sep Ricci-Ollivier curvature \sep subtree-prune-regraft
 \end{keyword}
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in a manuscript in preparation, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in \citep{billey2016trees}, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
 
@@
 We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\degree{T_1}$ and $\degree{T_2}$ (Table~\ref{tab:regressionMAT}).
-Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:kappa-access-distribution}, Fig.~\ref{fig:6t-short-time-kappa-access}, and Fig.~\ref{fig:7t-short-time-kappa-access}).
+Furthermore, plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:kappa-access-distribution}).
+We saw a similar effect between pairs of trees with 5 (Fig.~\ref{fig:kappa-access-distribution}), 6 (Fig.~\ref{fig:6t-short-time-kappa-access}) and 7 (Fig.~\ref{fig:7t-short-time-kappa-access}) leaves.
 We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
@@
 Surprisingly, we found that the degree change between two adjacent trees may be quadratic in $n$, that is of the same order as the degree itself.
-This maximum degree change occurs when a large subtree experiences a large depth change.\todo{Can we generalize this to more distant (but necessarily still close trees)?
-Idea: consider the moves induced by an AF, solve for max cumulative degree change.
-If so, add it and suggest a better version of this as future work in the conclusion.}
+This maximum degree change occurs when a large subtree experiences a large depth change.
 
@@
 We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth, as well as that the rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
+Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth.
+As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.
 Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
@@
 In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
-Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
+Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:kappa-access-distribution}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:kappa-access}).
 However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.

### 2016-04-22 curvature.tex
match color scheme to figures

@@
         Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-taxon trees, grouped by the degree of the trees.
-        Color signifies rSPR distance between the trees, with green, orange, blue, light blue, and brown signifying distances of 1, 2, 3, 4, and 5, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
 	}
@@
 				In both cases we exclude pairs that are simple inclusions of pairs of trees with a smaller number of taxa.
-				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
         }

### 2016-04-22 curvature.tex
fixed overful hbox

@@
 
-rSPR operations give rise to a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
+rSPR operations induce a distance measure between $X$-trees: $\dspr{T_1, T_2}$ is the minimum number of rSPR operations required to transform an $X$-tree $T_1$ into $T_2$.
 For example, the trees in Figure~\ref{fig:two-spr} are separated by two rSPR operations.

### 2016-04-24 curvature.tex
Full pass excluding most proofs.

@@
 In this paper, we investigate the rSPR graph in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-We confirm using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
 These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
@@
 However, we find that large topology changes (i.e. moving a large subtree) give pairs of trees with negative curvature.
+We show using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
 Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
@@
 Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\citep{whidden2013hybridization,Whidden2015-yi} and begun to allow such investigation.
+%EM You don't want to cite your latest work on uSPR?
 
 Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\citep{Ollivier2009-bw,Joulin2010-jg}.
-In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings points together.
+In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings neighborhoods together.
 Applying the framework to Brownian motion on a manifold returns the classical definition of Ricci curvature.
@@
 See Figure~\ref{fig:spr} for an example.
-The inclusion of $\rho$ allows for rSPR moves which move subtrees to the root of the tree.
+The inclusion of $\rho$ allows for rSPR moves which move subtrees to the root of the tree, however, this definition does not allow arbitrary re-rooting of trees by moving $\rho$.
 
@@
 For example, the tree $T$ with 4 leaves in Figure~\ref{fig:neighborhood} has 10 neighbors.
-We say that the degree of $T$ is $\degree{T}$, that is, the number of trees which can be obtained from $T$ by a single rSPR operation.
+The degree of $T$, i.e.\ the number of trees which can be obtained from $T$ by a single rSPR operation, will be denoted $\degree{T}$.
 We assume that all trees are bifurcating, and thus use degree to refer only to the degree of rSPR graph vertices.
@@
 In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\citep{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
-We applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
+We previously applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
 However, this method is impractical when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.
@@
 \begin{proof}
-	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.
+	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that given the subgraph of vertices $1, 2,  \ldots i$ the procedure will construct the entire subgraph of vertices $1, 2, \ldots i+1$.
 
@@
 	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
-	We store the tree to index mappings for current vertices of $G$ in a trie~\citep{fredkin1960trie} using Newick representation.
+	We store the tree to index mappings for current vertices of $G$ in a trie~\citep{fredkin1960trie} using the Newick representation.
 	This requires only $\OhOf{n}$-time for each tree (i.e.
@@
 \end{enumerate}
-We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
+We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is, each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
 We assign $(2n-1-x)-2$ moves to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ moves to each other non-root node.
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in \citep{billey2016trees}, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in \citep{Matsen2015-yv} with a closed formula in \citep{billey2016trees}, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
 
@@
 This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR~graph.
-
-Larger rSPR graphs tended to have smaller curvature values.
+%EM Do we want to interpret the degree at all? It shows a nice pattern in the plot, but I don't see any description.
+
+Larger rSPR graphs tended to have pairs of trees with smaller curvature values.
 Indeed, the 7-leaf rSPR graph contained adjacent pairs of trees with negative curvature.
@@
 Here, to gain more insight, we used simulation to approximate the entire access time distribution.
-Again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
+Again we use the idea that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
 Thus rather than enumerate access times between trees, which would have required a tremendous amount of memory and computational power to obtain accurate estimates, we enumerate times between pairs of trees in a tanglegram.
@@
 
-Pairs of trees that are close in the graph can not achieve these extreme degree differences.
-In our next lemma, we characterize how the degree of a tree changes after an rSPR operation.
+In our next lemma, we characterize how the degree of a tree changes after an rSPR operation, showing that pairs of trees that are close in the graph can not achieve these extreme degree differences.
 We show that the magnitude of the degree change depends solely on the size of the subtree to be moved, the difference between the depths of its original and new locations, and the size of its original and new neighboring subtrees.
-This relation allows us to improve the degree bounds between adjacent trees.
-See Figure~\ref{fig:spr-degree-change} for an illustration.
+This relation allows us to improve the degree bounds between adjacent trees (Figure~\ref{fig:spr-degree-change}).
 
@@
 
-Moreover, we can use these ideas to determine the number of rSPR moves that are, in some respects, independent of a given rSPR move.
+Moreover, we can use these ideas to determine the number of rSPR moves that are, in a sense, independent of a given rSPR move.
 That is, for two trees $S$ and $T$ differing by a single rSPR move, we wish to know the number of rSPR moves that are applicable to both trees rather than unique to one of the trees.
-To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in S(T)$ such that $\dspr{T',S'} = 1$.
+To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in N(S)$ such that $\dspr{T',S'} = 1$.
+%EM don't we need to exclude the case of $T' = S$ and $S' = T$?
 The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they signify pairs of matched paths that help push the curvature of those trees towards 0.
@@
 $$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
+%EM "o" trees?
 trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
@@
 We now use Lemma~\ref{lem:degree_change} to improve the degree change bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
+%EM this following sentence stands in contrast to the statement about extreme degree changes above. Or am I confused?
 Surprisingly, we found that the degree change between two adjacent trees may be quadratic in $n$, that is of the same order as the degree itself.
@@
 
-In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the pruned and regrafted subtree remains fixed.
+In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the size of the subtree moved to transition between the two trees remains the same size.
 To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
@@
 This flatness-in-the-limit theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
+%EM If desired, could we push this statement a little bit by describing the expected size of subtrees between neighbors.
 
@@
 \end{proof}
-
+%EM do we actually show that the bound is tight?
 This bound is tight and has been verified computationally for $n \le 7$.
@@
 It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
-Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees as neighborhood effects decrease with respect to the increasing distance.
+Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees, because neighborhood effects decrease with respect to the increasing distance.
 However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
@@
     In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
-		As we will show, these effects can be optimized simultaneously.
+		As we will show, these effects can be maximized simultaneously.
+%EM Using m is a little confusing here, given how prominently m is used in the early parts of the paper for a different purpose.
     To bound these effects, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
@@
 We further observe that the limit of our curvature lower bound is $-\frac{2}{7}$.
-Complete enumeration with $n \le 7$ shows that no pair of trees have curvature less than $-\frac{2}{5}$ and our bound meets or exceeds this value for $n > 7$.
+Complete enumeration with $n \le 7$ shows that no pair of trees have curvature less than $-\frac{2}{5}$; our bound meets or exceeds this value for $n > 7$.
 Moreover, the rSPR distance is a metric, so this bounds the curvature for arbitrary pairs of trees (Proposition 19 of~\citep{Ollivier2009-bw}).
@@
 \begin{cor}
-For all $n \ge 7$, there exist two adjacent trees $T$ and $S$ with $n$ leaves such that $\curvature{T,S} < 0$
+For all $n \ge 7$, there exist two adjacent trees $T$ and $S$ with $n$ leaves such that $\curvature{T,S} < 0$.
 \end{cor}
@@
 
-One could extend this work in several directions in addition to adding a nontrivial likelihood function as just described.
+One could extend this work in several other directions.
 Now that we have established the foundations of using curvature to understand graphs relevant for phylogenetic inference, many graph structures remain to be explored.

### 2016-04-25 curvature.tex
reference edits, all but 3 comments addressed that need clarification

@@
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\citep{bordewich05,hickey2008sdc}.
-Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and efficient fixed-parameter algorithms have recently been developed~\citep{whidden2013hybridization,Whidden2015-yi} and begun to allow such investigation.
-%EM You don't want to cite your latest work on uSPR?
+Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and unrooted case~\citep{bonet2010complexity} and efficient fixed-parameter algorithms have recently been developed~\citep{Whidden2015-yi,whidden2013hybridization,whidden2016calculating} and begun to allow such investigation.
 
@@
 The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
-The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{wiki:newick} representation of each tree as a string.
+The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string.
 These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
@@
 We show that the magnitude of the degree change depends solely on the size of the subtree to be moved, the difference between the depths of its original and new locations, and the size of its original and new neighboring subtrees.
-This relation allows us to improve the degree bounds between adjacent trees (Figure~\ref{fig:spr-degree-change}).
+This relation will allow us to improve the degree bounds between adjacent trees.
+See Figure~\ref{fig:spr-degree-change} for an illustration of the lemma.
 
@@
 To formalize this concept, consider pairs of trees $T' \in N(T)$ and $S' \in N(S)$ such that $\dspr{T',S'} = 1$.
-%EM don't we need to exclude the case of $T' = S$ and $S' = T$?
 The number of such ``squares'' involving two adjacent trees will play a key role in our curvature bounds, as they signify pairs of matched paths that help push the curvature of those trees towards 0.
@@
 $$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
-%EM "o" trees?
-trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
+trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
 \end{cor}
@@
 \end{proof}
-%EM do we actually show that the bound is tight?
 This bound is tight and has been verified computationally for $n \le 7$.
@@
 		As we will show, these effects can be maximized simultaneously.
-%EM Using m is a little confusing here, given how prominently m is used in the early parts of the paper for a different purpose.
-    To bound these effects, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
-    We can bound the mass transport cost from above by $1 + 2m$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
-	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.
+    To bound these effects, let $\mu$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
+    We can bound the mass transport cost from above by $1 + 2\mu$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
+	This gives a lower bound of $1 - (1 + 2\mu) / 1) = -2\mu$ on the curvature.
 
@@
 	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
-	Thus, the maximum $m$ is:
+	Thus, the maximum $\mu$ is:
 	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$

### 2016-04-25 curvature.tex
revised degree change intro

@@
 
-In our next lemma, we characterize how the degree of a tree changes after an rSPR operation, showing that pairs of trees that are close in the graph can not achieve these extreme degree differences.
+These degree bounds apply only to arbitrary trees, and the maximum degree difference occurs between trees that differ by many SPR operations.
+In our next lemma, we characterize how the degree of a tree changes after an rSPR operation.
+This relation will allow us to improve the degree bounds between adjacent trees.
 We show that the magnitude of the degree change depends solely on the size of the subtree to be moved, the difference between the depths of its original and new locations, and the size of its original and new neighboring subtrees.
-This relation will allow us to improve the degree bounds between adjacent trees.
 See Figure~\ref{fig:spr-degree-change} for an illustration of the lemma.

### 2016-04-28 curvature.tex
two comments resolved through discussion, one todo
consider the distribution of SPR subtree sizes

--- 2016-04-25 curvature.tex
@@
 This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR~graph.
-%EM Do we want to interpret the degree at all? It shows a nice pattern in the plot, but I don't see any description.
 
@@
 
-In our next lemma, we characterize how the degree of a tree changes after an rSPR operation, showing that pairs of trees that are close in the graph can not achieve these extreme degree differences.
+The maximum degree difference from Lemma~\ref{lem:degree_max_delta} occurs between trees that differ by many SPR operations.
+In our next lemma, we characterize how the degree of a tree changes after a single rSPR operation.
+This relation will allow us to improve the degree bounds between adjacent trees.
 We show that the magnitude of the degree change depends solely on the size of the subtree to be moved, the difference between the depths of its original and new locations, and the size of its original and new neighboring subtrees.
-This relation will allow us to improve the degree bounds between adjacent trees.
 See Figure~\ref{fig:spr-degree-change} for an illustration of the lemma.
@@
 We now use Lemma~\ref{lem:degree_change} to improve the degree change bounds in Lemma~\ref{lem:degree_max_delta} for two adjacent trees.
-%EM this following sentence stands in contrast to the statement about extreme degree changes above. Or am I confused?
 Surprisingly, we found that the degree change between two adjacent trees may be quadratic in $n$, that is of the same order as the degree itself.
@@
 
-In our final lemma of this section, we bound the number of neighbors shared by two adjacent trees.
+In our next lemma of this section, we bound the number of neighbors shared by two adjacent trees.
 The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive, as mass assigned to a shared neighbor may not require a mass transport cost.
@@
 
+In our final two lemmas of this section, we consider the distribution of trees in a neighborhood with respect to the size of the subtrees moved by SPR moves.
+First, we bound the number of neighbors of a tree that are induced by SPR moves involving a subtree with three or more leaves.
+In particular, we show that the majority of neighbors are induced by small subtree moves (one or two leaves).
+Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR move depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
+	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} SPR move is one that moves a small subtree.
+
+\begin{lem}
+	Let $T$ be a rooted tree with $n \ge 3$ leaves.
+	Let $N_S(T)$ be the neighbors of $T$ induced by small SPR moves. 
+	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
+	\label{lem:spr_size}
+\end{lem}
+\begin{proof}
+	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
+	Each subtree at a depth $ > 1$ from the root induces $2n - x - 5$ SPR neighbors, where $x$ is the number of nodes in the subtree, while the children of the root induce $2n - x - 3$ neighbors.
+	This implies that the number of small SPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
+	There are thus two extremes for the distribution of rSPR moves induced by small subtrees.
+	A ladder tree has the smallest proportion of small subtrees.
+	One 1-leaf subtree in a ladder tree is a child of the root, but this is counterbalanced by the fact that any other tree has at least one more subtree with 2 leaves.
+	A balanced tree has the greatest proportion of small subtrees.
+	Thus, a balanced tree has the greatest number of small SPR moves.
+
+	We first bound the number of small subtree moves that can be applied to a ladder tree.
+	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.
+	One of the 1-leaf subtrees is a child of the root.
+	A 2-leaf subtree contains three nodes.
+	Therefore, a ladder tree induces $n (2n - 1 - 5) + 2 + (2n - 3 - 5) = 2n^2 - 4n - 6$ small subtree moves.
+
+	Now, recall that the maximum degree of any tree is less than $4n^2 - 16n + 16$ by Lemma~\ref{lem:degree_extremes}.
+	We then have that $2(2n^2 - 4n - 6) = 4n^2 - 8n - 12 \ge 4n^2 - 8n - 12 + (24 - 8n) = 4n^2 -16n + 16$, for all $n \ge 3$.
+	Therefore more than half of the neighborhood of any tree is induced by small SPR moves.
+\end{proof}
+
+Now, we show that the average size of the moved subtrees is small.
+
+\begin{lem}
+	Let $T$ be a rooted tree with $n \ge 4$ leaves.
+	Let $N_i(T)$ be the neighbors of $T$ induced by SPR moves of an $i$-leaf subtree.
+	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 2}{4} .$$
+	\label{lem:spr_size_mean}
+\end{lem}
+\begin{proof}
+	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
+	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size.
+	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
+	We ignore the root subtree, which does not induce any SPR moves.
+	By Lemma~\ref{lem:spr_size}, at least half of the SPR moves involve a 1-leaf subtree.
+	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
+	This bounds the mean moved subtree size from above because each successively larger subtree contributes fewer moves.
+	The mean subtree size is $(2 + n - 2) / 2 = n / 2$.
+	Thus, the average size of a moved subtree is at most the weighted average of $1 / 2 + (n / 2) / 2$ = $n / 4 + 1 / 2$.
+
+%	The average moved subtree size for a ladder tree is then:
+%	\begin{align}
+%	(n (2n - 1 - 5) + 2) + (2 (2n - 3 - 5)) + \ldots \\
+%	n (2n - 1 - 5) + 2) + 2 (2n - 8) + 3 (2n - 10) + \ldots \\
+%	(2n^2 -6n + 2) + (4n - 16) + 6n - 30 + 8n - 48 + \ldots \\
+%	\le 3n^2 \\
+%\end{align}
+
+\end{proof}
+
 \section{Curvature}
@@
 
-This flatness-in-the-limit theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
-%EM If desired, could we push this statement a little bit by describing the expected size of subtrees between neighbors.
+As we already showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean}, most adjacent trees differ by a small subtree.
+This flatness-in-the-limit theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
 

### 2016-04-28 curvature.tex
Some revisions to the size distribution of subtrees.

@@
 
+\todo[inline]{Please define x in the text of the following lemma.}
 \begin{lem}
@@
 
-In our final two lemmas of this section, we consider the distribution of trees in a neighborhood with respect to the size of the subtrees moved by SPR moves.
-First, we bound the number of neighbors of a tree that are induced by SPR moves involving a subtree with three or more leaves.
-In particular, we show that the majority of neighbors are induced by small subtree moves (one or two leaves).
-Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR move depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
+In our final two lemmas of this section, we consider the distribution of sizes of subtrees moved by SPR operations from a given tree.
+First, we bound the number of neighbors of a tree that are induced by SPR operations involving a subtree with three or more leaves.
+In particular, we show that the majority of neighbors are induced by moving small subtrees (one or two leaves).
+Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR operation depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
 	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} SPR move is one that moves a small subtree.
@@
 	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
-	Each subtree at a depth $ > 1$ from the root induces $2n - x - 5$ SPR neighbors, where $x$ is the number of nodes in the subtree, while the children of the root induce $2n - x - 3$ neighbors.
+	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ SPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
 	This implies that the number of small SPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
-	There are thus two extremes for the distribution of rSPR moves induced by small subtrees.
-	A ladder tree has the smallest proportion of small subtrees.
-	One 1-leaf subtree in a ladder tree is a child of the root, but this is counterbalanced by the fact that any other tree has at least one more subtree with 2 leaves.
-	A balanced tree has the greatest proportion of small subtrees.
-	Thus, a balanced tree has the greatest number of small SPR moves.
-
-	We first bound the number of small subtree moves that can be applied to a ladder tree.
+
+	A ladder tree has the smallest proportion of small subtrees, so we first bound the number of small subtree moves that can be applied to a ladder tree.
 	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.

### 2016-04-28 curvature.tex
define x in lem:compute_degree

@@
 
-\todo[inline]{Please define x in the text of the following lemma.}
 \begin{lem}
@@
 	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
-	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
+	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, $x$ is the number of nodes in the subtree rooted at $u$, and:
 	$$\size{N(T,u)} = \begin{cases}

### 2016-04-28 curvature.tex
checked for proper usage of rSPR vs SPR

@@
 These methods are known to mix quickly when sampling from the simple uniform distribution of trees but may become stuck in the local optima of multi-modal posterior distributions for real data induced by non-uniform likelihoods.
-The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating study of the underlying \emph{rSPR graph} in greater~detail.
-
-In this paper, we investigate the rSPR graph in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an SPR neighbor of a tree uniformly at random to $O(n)$ time.
-We then develop a closed form solution to characterize how the number of SPR neighbors of a tree changes after an SPR operation is applied to that tree.
+The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating study of the underlying \emph{SPR graph} in greater~detail.
+
+In this paper, we investigate the SPR graph of rooted trees (rSPR graph) in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
+These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
+We then develop a closed form solution to characterize how the number of rSPR neighbors of a tree changes after an rSPR operation is applied to that tree.
 This gives bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
@@
 On the other hand, fast mixing has been identified with exceptionally well-behaved data~\citep{Stefankovic2011-hu} or with a uniform distribution~\citep{spade2014note}.
-% Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\citep{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\citep{Evans2006-xh,Athreya2014-de}.
+ Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\citep{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\citep{Evans2006-xh,Athreya2014-de}.
 Studies on real data~\citep{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
@@
 In previous work, we computed mean access times (MAT) between pairs of trees in  MCMC random walks: the mean number of iterations required to move from one tree to the other.
-We applied this work to demonstrate the influence of SPR graph structure on real MCMC posteriors sampled with MrBayes~\citep{Whidden2015-yi} using \texttt{sprspace}~\citep{sprspace}.
+We applied this work to demonstrate the influence of (unrooted) SPR graph structure on real MCMC posteriors sampled with MrBayes~\citep{Whidden2015-yi} using \texttt{sprspace}~\citep{sprspace}.
 
@@
 
-The maximum degree difference from Lemma~\ref{lem:degree_max_delta} occurs between trees that differ by many SPR operations.
+The maximum degree difference from Lemma~\ref{lem:degree_max_delta} occurs between trees that differ by many rSPR operations.
 In our next lemma, we characterize how the degree of a tree changes after a single rSPR operation.
@@
 
-In our final two lemmas of this section, we consider the distribution of sizes of subtrees moved by SPR operations from a given tree.
-First, we bound the number of neighbors of a tree that are induced by SPR operations involving a subtree with three or more leaves.
+In our final two lemmas of this section, we consider the distribution of sizes of subtrees moved by rSPR operations from a given tree.
+First, we bound the number of neighbors of a tree that are induced by rSPR operations involving a subtree with three or more leaves.
 In particular, we show that the majority of neighbors are induced by moving small subtrees (one or two leaves).
-Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR operation depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
-	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} SPR move is one that moves a small subtree.
+Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an rSPR operation depends on the subtree size and height change, this implies that most rSPR neighbors have a similar degree.
+	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} rSPR move is one that moves a small subtree.
 
@@
 	Let $T$ be a rooted tree with $n \ge 3$ leaves.
-	Let $N_S(T)$ be the neighbors of $T$ induced by small SPR moves.
+	Let $N_S(T)$ be the neighbors of $T$ induced by small rSPR moves.
 	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
@@
 	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
-	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ SPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
-	This implies that the number of small SPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
+	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ rSPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
+	This implies that the number of small rSPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
 
@@
 	We then have that $2(2n^2 - 4n - 6) = 4n^2 - 8n - 12 \ge 4n^2 - 8n - 12 + (24 - 8n) = 4n^2 -16n + 16$, for all $n \ge 3$.
-	Therefore more than half of the neighborhood of any tree is induced by small SPR moves.
+	Therefore more than half of the neighborhood of any tree is induced by small rSPR moves.
 \end{proof}
@@
 	Let $T$ be a rooted tree with $n \ge 4$ leaves.
-	Let $N_i(T)$ be the neighbors of $T$ induced by SPR moves of an $i$-leaf subtree.
+	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
 	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 2}{4} .$$
@@
 	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
-	We ignore the root subtree, which does not induce any SPR moves.
-	By Lemma~\ref{lem:spr_size}, at least half of the SPR moves involve a 1-leaf subtree.
+	We ignore the root subtree, which does not induce any rSPR moves.
+	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a 1-leaf subtree.
 	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.

### 2016-04-28 curvature.tex
revised subtree size proof, fixed off by one error

@@
 % regular margins
-\usepackage{thmtools}
-\usepackage{thm-restate}
+%\usepackage{thmtools}
+%\usepackage{thm-restate}
 
@@
 	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
-	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 2}{4} .$$
+	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
 	\label{lem:spr_size_mean}
@@
 	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
-	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size.
+	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
 	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
@@
 	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
-	This bounds the mean moved subtree size from above because each successively larger subtree contributes fewer moves.
-	The mean subtree size is $(2 + n - 2) / 2 = n / 2$.
-	Thus, the average size of a moved subtree is at most the weighted average of $1 / 2 + (n / 2) / 2$ = $n / 4 + 1 / 2$.
+	In particular, we assume that each subtree contributes $2n - 3 - 5$ moves.
+	This underestimates the number of small moved subtrees and overestimates the number of larger moved subtrees.
+	Thus, this estimate bounds the mean moved subtree size from above.
+	The mean subtree size, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
+	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
 

### 2016-04-28 curvature.tex
fixed usage of small
A proof EM can understand.

@@
 
-Now, we show that the average size of the moved subtrees is small.
+Now, we bound the average size of the moved subtrees.
 
@@
 	We ignore the root subtree, which does not induce any rSPR moves.
-	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a 1-leaf subtree.
-	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
-	In particular, we assume that each subtree contributes $2n - 3 - 5$ moves.
-	This underestimates the number of small moved subtrees and overestimates the number of larger moved subtrees.
-	Thus, this estimate bounds the mean moved subtree size from above.
-	The mean subtree size, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
-	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
+	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
+    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we increase the weight for $i \leq 2$ and decrease the weight on $i > 2$.
+	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
+	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
+	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
+	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
 
@@
 As we already showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean}, most adjacent trees differ by a small subtree.
-This flatness-in-the-limit theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
+We next show that curvature is small in such a case for pairs of trees with a sufficiently large number of leaves.
+Thus this ``flatness-in-the-limit'' theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
 

### 2016-04-28 curvature.tex
reference tweaks

--- 2016-04-28 curvature.tex
@@
 \label{sec:random_walks}
-The uniform random walk moves from one vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
+The uniform random walk moves from a vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
 In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
@@
 	We ignore the root subtree, which does not induce any rSPR moves.
-	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a 1-leaf subtree.
+	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
 	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.

### 2016-04-28 curvature.tex
Switched increase vs decrease.

--- 2016-04-28 curvature.tex
@@
 	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
-    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we increase the weight for $i \leq 2$ and decrease the weight on $i > 2$.
+    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
 	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.

### 2016-04-28 curvature.tex
Ah, this is a better way to phrase flatness connection.

@@
 
+We showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean} that most adjacent trees differ by a small subtree, raising the question of what the curvature is between such pairs.
 In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the size of the subtree moved to transition between the two trees remains the same size.
@@
 The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
-
-As we already showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean}, most adjacent trees differ by a small subtree.
-We next show that curvature is small in such a case for pairs of trees with a sufficiently large number of leaves.
 Thus this ``flatness-in-the-limit'' theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.

### 2016-04-29 curvature.tex
added funding to acknowledgements as well

@@
 \section{Acknowledgements}
+This work was funded by National Science Foundation award 1223057.
+Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.
+
 The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.

### 2016-05-10 curvature.tex
atkins expected distance ref

@@
 \end{equation}
+\citep{atkins2015extremal} showed that the expected distance between a pair of trees is $n - \Theta(n^{2/3})$.
 We are not aware of any further work investigating properties of the SPR graph, which may be due to its complexity.

### 2016-11-24 response_to_reviewers.txt
revised the first 6 pages, added response text file

Reviewer 2

Page 1

27: The spelling 'multimodal' seems to be more common than 'multi-modal' (and would be in line with other uses in your paper; e.g. page 3, line 12).
* corrected

30: The word 'the' seems to be missing before 'study'.
* corrected

32: The word 'the' seems to be missing before 'Ricci-Ollivier'.
* corrected

36: Here, I would suggest clarifying the phrase 'and n their number of leaves' a bit. In its current state, it might not be clear from the wording whether each tree has n leaves or whether n is the total number of leaves of all trees.
* revised to "... and n is the number of leaves in each tree, ..."

44-46: The phrasing 'We show ... showing ...' sounds a bit odd because 'show' is used twice.
* revised "showing" to "demonstrating"

Page 2
10: It looks like some kind of connecting phrase is missing between 'particular' and 'properties' (e.g.  'with respect to').
* we added the suggested connecting phrase

11: The word 'the' seems to be missing before 'investigation'.
* corrected

52-53: As you start this part of this sentence with 'both', an 'and' seems to be more fitting in 'leaves [12, 13] or for SPR'.
* corrected

54: Do you rather mean 'an infinite number of leaves' or 'infinitely many leaves' instead of 'infinite leaves'?
* corrected to "infinitely many leaves"

Page 3

15: A more fitting position of 'still' seems to be between 'is' and 'known'.
* corrected

18: A word like 'the' or 'its' seems to be missing before 'degree'.
* corrected to "its".

30: The phrase 'begun to allow' sounds a bit odd. Swapping the verbs ('allowed to begin') might be preferable.
* The phrase "allowed to begin" would change our intended meaning, so we instead changed this to "... efficient fixed-parameter algorithms have recently been developed [17, 24, 25] that begin to allow such investigations." .

31: It looks like the word 'investigation' is missing the indefinite article or should be in the plural form.
* corrected, as noted above

33: The word 'the' seems to be missing before 'Ricci'.
* corrected

47: The word 'the' seems to be missing before 'curvature'.
* corrected

56: The same problem as on page 1, line 36.
* corrected to "... each of which has n leaves".

Page 4
44: In this context, 'linear-time' seems to be more common without a hyphen.
* corrected

47-52: Both sentences use the phrase 'we found'. Varying the wording a bit would make it sound better.
* We revised the second sentence with "we found" to begin with "Our results demonstrated that ...".

48: The word 'the' seems to be missing before 'distance'.
* corrected

Page 5

27: The word 'the' seems to be missing before 'proba-'.
* corrected

29: The word 'the' seems to be missing before 'knowledge'.
* corrected

Page 6
30: Is it really necessary to define the LCA of two or more subtrees as the LCA of their parent nodes (and not as the LCA of the root nodes of the subtrees)? It seems a bit unconventional (also w.r.t. your LCA definition of one or more nodes beforehand).
* Corrected. We did indeed mean the root nodes of the subtrees by "parent".

40: Either the word 'the' seems to be missing before 'arbitrary' or it has to be 'arbitrarily' (in which case the 'of' before 'trees' has to be removed).
* corrected by adding "the".

Page 7
13: The wording 'we will specialize to that space being a graph' sounds odd. Maybe something
along the lines of 'we restrict ourselves to ...' is preferable.
20: It looks like the apostrophe in 'movers' is missing (it should be 'mover's').
Page 8
14: It looks like there is too much white space behind 'e.g.'.
24: In this context, 'previously' seems more appropriate than 'previous'.
28-33: Both points are possibly related.
1. In the final sentence of the first paragraph, m is assumed to be (2n-3)!! ('m is the rapidly
growing function (2n-3)!!'). You continue using m in the second paragraph. Is it an arbitrary
value not larger than (2n-3)!! (or still equal to the maximum)? I would suggest clarifying the
transition.
2. The phrase 'graphs from subsets of m high probability trees' is a bit ambiguous. Does it
mean that the graphs are constructed from subsets which themselves consist of m trees or are
the graphs constructed from subsets of the m-element set of high probability trees?47-48: What is the unit of the length of the Newick representation?
51: To which list does 'in this list' refer to?
Page 9
44-45: The statement starts with 'requires only O(n)-time for each tree' (of which there are m).
Therefore, it is not quite clear (even though not wrong), why the total time of this step is bound by
the unnecessarily unsharp term 'O(mn3)'.
Page 10
42: The wording 'within a subtree' seems a bit imprecise and could be clarified by changing it to
something along the lines of 'within the subtree to be moved / regrafted'.
55: The term 'every' seems to be more fitting than 'each' in this context.
Page 11
9-12: The wording seems a bit imprecise because the target of a move is not the node itself but the
node's edge.
23-25: The case distinction requires that the extra root node ρ has depth -1, the actual root depth 0
and so on, but this is stated nowhere in the paper.
33: Probably, the set-difference operator ('\') is supposed to denote the pruning (including the
merging of the incoming and outgoing edge of the resulting unary node). Unfortunately, this is not
explained. A short explanation would make this more obvious to the reader.
Page 12
15: It looks like that '-th' is missing behind the '|r|' in order to turn it into an ordinal number.
23: What is the unit of the 'O(n) space'?
52-55: The significance of this paragraph could be increased by stating the special features,
properties or outcomes of these runs instead of barely stating that you have performed them.
Page 13
32: Which graph is 'the rSPR graph ... as discussed in Section 3.1'? The complete one with (2n-3)!!
trees? This is not quite clear since section 3.1 also considers other values for m.
33: It is stated that the curvature is computed for 'given pairs of trees', but who / what gives these
pairs? The wording is a bit odd here.55: The word 'tempered' is slightly ambiguous, because it can have the meaning 'to moderate' as
well as 'to harden'.
Page 14
56-: The connection between a smaller κ and a right-shift of the access time distribution is difficult
to see as differences in the saturation are hardly noticable.
Page 15
Plots of Figure 5:
1. It is very hard to extract any information from the plots, because the plots (as soon as there
are a few more lines) quickly get congested due to overlaps and hardly (or not at all) visible
differences in saturation.
2. Adding axis labels (and not just explaining them in the figure caption) could help the
readability of the plots.
3. It might be a bit counterintuitive to associate smaller κ values with an increased saturation.
4. Furthermore, the print version of articles in this journal usually come with monochrome
images, (then) making it basically impossible to distinguish the lines by colour and
saturation.
Page 16
Plots of Figure 6:
1. Except for the axis labels, the plots have the similar problems as those of Figure 5.
2. Plots (c) and (d) are especially problematic w.r.t. overlaps.
3. Adding brightness (blue vs. light blue) as another distinctive feature (besides saturation) can
also hamper the readability (when the differences are as small as they are here).
45-56:
1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of
pairs of trees with a smaller number of taxa'?
Page 17
12-15: Adding hyphens (e.g. 'distance-1 pairs') might increase the readability.
15-23: The traceability / comprehensibility of the statements would strongly benefit from improving
the readability of the plots in Figure 5 and 6.
53: It looks like there is too much white space behind 'w.l.o.g.'.
Page 18
Proof of (i):The proof only works when the '+14' in the numerator is not reduced to '+12' (which is also not
necessary, because bounding the sum of the logarithms in the denominator is sufficient.
Proof of (ii):
There seems to be a small error in the format. The penultimate equality is not aligned with the other
equal and less-than signs.
42-43: The statement 'The maximum degree difference from Lemma 5.2. occurs between trees that
differ by many rSPR operations.' is (in this generality) not correct or at least misleading. Consider
the rSPR graph for X-trees with four leaves from Figure 3(a). The ladder tree and the balanced tree
have the maximum degree difference but are only separated by a single rSPR move while there are
other tree pairs separated by two moves (which, in this context / example, should be considered as
many).
Page 19
39:
1. There seems to be an inconsistency between the definition of internal edges and their
number stated in this line. On page 5, internal edges are defined as those that do not contact
a leaf or ρ. Then, a 'rooted tree with k leaves' cannot have 2(k-1) internal edges as stated here
(because this is the number of all edges (also those leading to leaves), disregarding the extra
root edge). So, either the definition of 'internal edge' has to be adopted or the word 'internal'
has to be dropped here.
2. The white space after 'edges' is missing.
Page 20
9-19: While Corollary 5.4 appears correct, the given justification seems to be a bit problematic.
Based on the definition of γ and the excluded (subtracted) moves, it appears that these γ moves still
include some which move R. For example, from the set of all possible moves for T (deg(T)), only
some moves into R (2kb, case (ii)) and some moves of R's parent subtree (2(j-1), case (iv)) are
excluded.
However, it is then stated that these moves 'can be applied to T and S with the same source and
target nodes'. The minor (but arguable) problem concerns the part 'same source nodes': When are
they the same? Is the parent node of R the same although its position differs in T and S? If not, then
not all γ moves can be applied with the same source nodes. The bigger problem is created by the
fact that moves moving R with the same target node eliminate the only difference between T and S
and, thus, the difference between the obtained neighbours is zero (not one).
Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be
obtained.
14: The phrase 'the pairings are disjoint' might need some more explanation as 'disjoint' might get
interpreted in several ways (e.g. no tree is allowed to occur in more than one pairing vs. the pairings
are pairwise distinct).
37: The word 'and' seems to be missing before 'then' .
Lemma 5.5:
1. Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is
necessary (Why is it necessary to increase j? Why are the subcases of (not) moving theadditional leaf necessary?) in order to clarify the line of argument.
2. Concerning statement (ii): Is there a particular reason why the quotient is bounded for n ≥ 4?
The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
Page 21
49:
1. The meaning of the function L(.) can be guessed from the context without too much effort,
but providing a short definition / explanation would help avoiding any misconceptions.
2. Does this equality really hold between the restricted versions of T and S , which differ only in
R and, thus, should still be different after restricting them to X \ L(Q)?
51-53: Using U as a subtree while it also denotes a whole tree in the neighbourhood of T and S just
a few lines before can be slightly confusing.
Page 22
30-32: The statement on the proportionality of small moves and small subtrees seems to be a bit
troubled.
1. For example, if both children of the root are not small, each small subtree with one leaf
contributes 2n-6 moves, while each small subtree with two leaves contributes 2n-8 moves.
So, what is the factor of proportionality?
2. In the phrase 'where one of the root's children is small', is the intended meaning exactly one
or at least one? Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the
majority of) the leaves), the statement does not hold in general (neither for exactly nor for at
least). Consider, e.g., a tree with a leaf as the only small child of the root and at least three
small subtrees with two leaves (for the case exactly) respectively the balanced tree with four
leaves (for the case at least).
36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more
consistent with the previous use of numerals).
38-42:
1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n2-4n-6) is not
correct for all intended n. The reason is that for the ladder tree with three leaves, the small
subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5)
moves. Hence, a case distinction seems to be necessary here and for the inequality in the
subsequent paragraph (lines 40-42)
2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads
to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case
distinction n = 3 vs. n ≥ 4).
45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45
possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the
typical notion of size in mind).
54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to
hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3
for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).
This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
56: Similar to line 36, 'one leaf' and 'one subtree' appear more readable / consistent.
Page 23
11-13: The vertical bars denoting the size of N i (T) are missing at all three occurrences.
11-20:
1. It is stated that |N i (T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N 1 (T)| = (n-1)(2n-1-5) +
(2n-1-3) > 2n-3-5.
2. Also, it is stated that |N i (T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n-
(2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-
3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
4. Is it actually obvious enough that setting all |N i (T)| to 2n-3-5 leads to an upper for the
quotient representing the average moved subtree size? A short explanation might make this
point clearer.
5. In the last sentence of the proof, the weighted average uses a proportion of 1⁄2 for both small
and large subtree (which is okay), but why is the average moved subtree size of small
substrees assumed to be one? Shouldn't the one small subtree with two leaves (which always
exists) make it slightly larger than one?
32: Is there are special reason for (re)defining the mass transport term W 1,n ?
37: The verb 'increase' misses the third-person singular s (as it refers to 'the size of the trees').
45: Only one of the two occurrences of 'thus' is needed.
Page 24
14: Adding a short explanation why the number of unmatched trees is in O(kn) might help the
reader understanding the argument at this point better.
19: The innermost pair of parentheses around 'O(n)' is not necessary.
30: Replacing the word 'a' before 'distance' by 'at' or 'of' conveys the meaning better.
31: The equality sign in the equation might be less fitting than a <, because it is previously (line 25)
stated that S n has a strictly larger neighbourhood.
46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to
bounds for W 1 (m T ,m S ) could make it easier for the reader to grasp the idea how the proof works.
Page 25
29: Possibly it should be 'each other' instead of 'each either'.40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree
neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
Page 26
10: A short remark on how the term '1 + μ' is actually derived would make this passage easier /
faster to comprehend
14: Following Corollary 5.4 it should be 'at least γ' and not 'at most γ' paired trees (and, otherwise,
the argumentation is not sensible at this point).
19: The rationale for bounding the degree by using the maximum possible absolute difference
(irrespective of the actual move) seems a bit unclear. Currently, it does not look like the line of
argument leads to the maximum μ: using the maximum degree of T (instead of the actual,
potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily
large denominator.
20:
1. The last inequality ('≤ 1⁄2n2-n') holds only if n is even. For odd n, 1⁄2 has to be added to the
upper bound (which is then also okay for the even case). In the end, this change also affects
the curvature bound.
2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would
not be wrong, it is a potential source of confusion.
39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the
proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to
it.
Page 27
35: The upper bound does not seem to follow as analogously to the lower bound as it is stated.
Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to
κ p (T , S )
2
⩽ ...⩽ κ (T , S)+
p
p∗max (∣ N (T )∣ , ∣ N (S )∣)
where the factor p in the denominator causes a problem, because simply dropping it (in order to
arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
Page 28
11-12: According to Lemma 5.5, the quotient of degrees for two adjacent trees is at least 5/6 only
for n ≥ 4, but there is no such assumption in Lemma 6.9 or its proof.
14-18: Using W 1 (T,S) (resp. W' 1 (T,S)) is slightly imprecise as T and S are the trees and not the
probability densities.
26: Within the min-term of the last case, w is the argument of all four probability densities.
Probably, two of them should have z as their argument.29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is
slightly imprecise.
50: The word 'the' seems to be missing before 'understanding'.
55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the
regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven
lemma) the changes depend linearly on the product (which in turn is quadratic in the number of
leaves).
Page 29
13-17: The same problem as on page 17, lines 15-23.
29: Changing 'achieve' to 'to be achieved' seems a bit more readable.

### 2016-11-24 curvature.tex
revised the first 6 pages, added response text file

@@
 Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
-These methods are known to mix quickly when sampling from the simple uniform distribution of trees but may become stuck in the local optima of multi-modal posterior distributions for real data induced by non-uniform likelihoods.
-The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating study of the underlying \emph{SPR graph} in greater~detail.
-
-In this paper, we investigate the SPR graph of rooted trees (rSPR graph) in a new way: by calculating Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ their number of leaves, and reducing the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
+These methods are known to mix quickly when sampling from the simple uniform distribution of trees but may become stuck in the local optima of multimodal posterior distributions for real data induced by non-uniform likelihoods.
+The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating the study of the underlying \emph{SPR graph} in greater~detail.
+
+In this paper, we investigate the SPR graph of rooted trees (rSPR graph) in a new way: by calculating the Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
+These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ is the number of leaves in each tree, and reducing the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
 We then develop a closed form solution to characterize how the number of rSPR neighbors of a tree changes after an rSPR operation is applied to that tree.
@@
 However, we find that large topology changes (i.e. moving a large subtree) give pairs of trees with negative curvature.
-We show using simulation that mean access time distributions depend on distance, degree, and curvature, showing the relevance of these curvature results to stochastic tree search.
+We show using simulation that mean access time distributions depend on distance, degree, and curvature, demonstrating the relevance of these curvature results to stochastic tree search.
 Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
-This work extends our knowledge of the rSPR graph, in particular properties that are relevant for investigation of sampling the rSPR graph.
+This work extends our knowledge of the rSPR graph, in particular with respect to properties that are relevant for the investigation of sampling the rSPR graph.
 \end{abstract}
@@
 On the other hand, fast mixing has been identified with exceptionally well-behaved data~\citep{Stefankovic2011-hu} or with a uniform distribution~\citep{spade2014note}.
- Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\citep{Aldous2000-vg,Diaconis2002-gy} or for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinite leaves~\citep{Evans2006-xh,Athreya2014-de}.
+ Probabilists have also approached the problem using related frameworks that are more amenable to proving theorems, both for other sets of moves on trees with a finite number of leaves~\citep{Aldous2000-vg,Diaconis2002-gy} and for SPR and related moves on a continuous tree-like object which formalizes the notion of a tree with infinitely many leaves~\citep{Evans2006-xh,Athreya2014-de}.
 Studies on real data~\citep{beiko2006searching, lakner2008efficiency}, however, have identified posteriors which are difficult to sample using MCMC.
@@
 
-Although the SPR graph is thus very important in determining the success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself.
-\citep{Song2003-gf}~developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
+Although the SPR graph is thus very important in determining the success of phylogenetic inference procedures, little is still known about the rooted or unrooted versions of the SPR graph itself.
+\citep{Song2003-gf}~developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on its degree.
 \citep{Ding2011-bj}~showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
@@
 Indeed, even computing the distance between topologies in terms of SPR operations (rooted and unrooted) is NP-hard~\citep{bordewich05,hickey2008sdc}.
-Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and unrooted case~\citep{bonet2010complexity} and efficient fixed-parameter algorithms have recently been developed~\citep{Whidden2015-yi,whidden2013hybridization,whidden2016calculating} and begun to allow such investigation.
-
-Ollivier and colleagues recently pioneered a new approach to calculating Ricci curvature on a general type of metric space, including graphs~\citep{Ollivier2009-bw,Joulin2010-jg}.
+Fortunately, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05} and unrooted case~\citep{bonet2010complexity} and efficient fixed-parameter algorithms have recently been developed~\citep{Whidden2015-yi,whidden2013hybridization,whidden2016calculating} that begin to allow such investigations.
+
+Ollivier and colleagues recently pioneered a new approach to calculating the Ricci curvature on a general type of metric space, including graphs~\citep{Ollivier2009-bw,Joulin2010-jg}.
 In this framework, local information about the metric space is given by a random walk (rather than a Riemann tensor) such that their notion of curvature formalizes the notion of to what extent random walking brings neighborhoods together.
 Applying the framework to Brownian motion on a manifold returns the classical definition of Ricci curvature.
-Curvature is determined by the ratio of the earth mover's distance~\citep{rubner2000earth} between neighborhoods of a pair of vertices given by a random walk and the distance between the vertices.
+The curvature is determined by the ratio of the earth mover's distance~\citep{rubner2000earth} between neighborhoods of a pair of vertices given by a random walk and the distance between the vertices.
 Here the term \emph{random walk} on a space $X$ simply denotes a family of probability measures parameterized by points of $X$ satisfying reasonable assumptions, which includes biased walks such as MCMC.
@@
 We required several new computational tools.
-We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
+We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees each of which has $n$ leaves.
 As the full rSPR graph on trees with $n$ leaves contains $(2n-3)!!
@@
 By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
-By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
+By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear time without explicitly generating the graph.
 This stands in contrast to the sampling methods used in current software such as MrBayes, which do not propose SPR moves uniformly.
@@
 
-Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
-Moreover, we found that rSPR graphs for trees with 7 or more leaves have tree pairs with negative curvature, corresponding to direct paths that are difficult to traverse stochastically.
+Using our methods to simulate these random walks, we found that the distribution of access times between pairs of trees can be described by the distance between the trees, the degrees of the trees, and the curvature.
+Our results demonstrated that rSPR graphs for trees with 7 or more leaves have tree pairs with negative curvature, corresponding to direct paths that are difficult to traverse stochastically.
 By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on the earth mover's distance in this context and thus curvatures under these random walks.
@@
 Pairs of trees separated by such moves correspond to the pairs with negative curvature identified in our simulation results.
-These pairs occur infrequently in these well-connected graphs, however, they may be more problematic in real posterior distributions where the majority of probability is spread over a relatively small number of trees~\citep{Whidden2015-yi}.
-In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
+These pairs occur infrequently in these well-connected graphs, however, they may be more problematic in real posterior distributions where the majority of the probability is spread over a relatively small number of trees~\citep{Whidden2015-yi}.
+In summary, we extend the knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.
 
@@
 The \emph{least common ancestor} (LCA) of a set $R$ of two or more nodes is the unique node that is an ancestor of each node $r \in R$ and at maximum depth.
-Similarly, the LCA of two or more subtrees is the LCA of their parent nodes.
+Similarly, the LCA of two or more subtrees is the LCA of their root nodes.
 
@@
 See Figure~\ref{fig:spr} for an example.
-The inclusion of $\rho$ allows for rSPR moves which move subtrees to the root of the tree, however, this definition does not allow arbitrary re-rooting of trees by moving $\rho$.
+The inclusion of $\rho$ allows for rSPR moves which move subtrees to the root of the tree, however, this definition does not allow the arbitrary re-rooting of trees by moving $\rho$.
 
@@
 
-Let $m_x$ and $m_y$ be probability densities of the position of a specified random walk after one step of the random walk, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
+Let $m_x$ and $m_y$ be the probability densities of the position of a specified random walk after one step of the random walk, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
 The transportation distance~\citep{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\citep{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
@@
 
-To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing rSPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
+To efficiently simulate the MH random walk, we developed a linear time algorithm for proposing rSPR moves that does not require the rSPR graph to be explicitly built and stored in memory.
 A na\"ive approach would require $\OhOf{n^3}$ time: $\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree so that one could be picked uniformly at random.

### 2016-11-29 response_to_reviewers.txt
revised up to page 10

@@
 Page 6
+
 30: Is it really necessary to define the LCA of two or more subtrees as the LCA of their parent nodes (and not as the LCA of the root nodes of the subtrees)?
@@
 Page 7
-13: The wording 'we will specialize to that space being a graph' sounds odd. Maybe something
-along the lines of 'we restrict ourselves to ...' is preferable.
+
+13: The wording 'we will specialize to that space being a graph' sounds odd. Maybe something along the lines of 'we restrict ourselves to ...' is preferable.
+* revised to "... we restrict ourselves to graphs equipped with a shortest-path distance."
+
 20: It looks like the apostrophe in 'movers' is missing (it should be 'mover's').
+* corrected
+
 Page 8
+
 14: It looks like there is too much white space behind 'e.g.'.
+* corrected
+
 24: In this context, 'previously' seems more appropriate than 'previous'.
+* corrected
+
 28-33: Both points are possibly related.
-1. In the final sentence of the first paragraph, m is assumed to be (2n-3)!! ('m is the rapidly
-growing function (2n-3)!!'). You continue using m in the second paragraph. Is it an arbitrary
-value not larger than (2n-3)!! (or still equal to the maximum)? I would suggest clarifying the
-transition.
-2. The phrase 'graphs from subsets of m high probability trees' is a bit ambiguous. Does it
-mean that the graphs are constructed from subsets which themselves consist of m trees or are
-the graphs constructed from subsets of the m-element set of high probability trees?47-48: What is the unit of the length of the Newick representation?
+
+1.
+In the final sentence of the first paragraph, m is assumed to be (2n-3)!!
+('m is the rapidly growing function (2n-3)!!').
+You continue using m in the second paragraph.
+Is it an arbitrary value not larger than (2n-3)!!
+(or still equal to the maximum)?
+I would suggest clarifying the transition.
+* As stated in the second sentence of the first paragraph, $m$ is the number of trees in the graph.
+This is (2n-3)!!
+for the full rSPR graph, as noted in that final sentence.
+In the second paragraph, $m$ still refers to the number of trees in the graphs in question, which is upper bounded by (2n-3)!!
+as you note.
+We have revised these sentences to limit confusion and address your next point.
+We now note that, for the full rSPR graph, the value of $m$ is given by $(2n-3)!!
+in the first paragraph.
+We removed the redundant reference to $m$ in the first sentence of the second paragraph.
+
+2.
+The phrase 'graphs from subsets of m high probability trees' is a bit ambiguous.
+Does it mean that the graphs are constructed from subsets which themselves consist of m trees or are the graphs constructed from subsets of the m-element set of high probability trees?
+* The subsets consist of m trees. As noted above, m is simply the number of trees in a given graph and not a fixed value. We removed the redundant reference to m here.
+
+47-48: What is the unit of the length of the Newick representation?
+* The unit of the length is the number of characters in the string (equivalently wrt big O, the size in bytes).
+We realized in a follow up paper that, although the length of a Newick string is typically assumed to be linear, this does require some (reasonable) assumptions.
+We now indicate so here and cite the preprint of that paper.
+
 51: To which list does 'in this list' refer to?
+* This statement refers to the order of the tree in the specified ordering. We revised "in this list" to "in this ordering" to prevent confusion.
+
 Page 9
 44-45: The statement starts with 'requires only O(n)-time for each tree' (of which there are m).
-Therefore, it is not quite clear (even though not wrong), why the total time of this step is bound by
-the unnecessarily unsharp term 'O(mn3)'.
+Therefore, it is not quite clear (even though not wrong), why the total time of this step is bound by the unnecessarily unsharp term 'O(mn^3)'.
+* This is O(n)-time for each neighbor tree, of which there are O(mn^2), for a total of O(mn^3)-time. We added the word neighbor to clarify.
+
 Page 10

### 2016-11-29 curvature.tex
revised up to page 10

@@
 Ricci-Ollivier curvature provides a rigorous yet intuitive formalization of the shape of a metric space with respect to a random walk.
-For the purposes of this paper, we will specialize to that space being a graph equipped with the shortest-path distance.
+For the purposes of this paper, we restrict ourselves to graphs equipped with a shortest-path distance.
 For a more rigorous presentation in the more general setting of a Polish metric space, see~\citep{Ollivier2009-bw} or the survey~\citep{Ollivier2010-ao}.
@@
 Let $m_x$ and $m_y$ be the probability densities of the position of a specified random walk after one step of the random walk, starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
-The transportation distance~\citep{Villani2003-wv} (equivalently Wasserstein distance, or ``earth movers distance''~\citep{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
+The transportation distance~\citep{Villani2003-wv} (equivalently Wasserstein distance, or ``earth mover's distance''~\citep{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ along edges of the graph, that is
 \vspace{-0.5em}
@@
 As above for $\kappa$, we use $\ric{x, y}$ as shorthand for $\ric{m; x, y}$ when $m$ is the uniform lazy random walk, and $\ric{\MH; x, y}$ when $m$ is the Metropolis-Hastings random walk sampling the uniform distribution (Section~\ref{sec:random_walks}).
-This definition of curvature is invariant of $p$ for small $p$~\citep{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.
-Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
+This definition of curvature is invariant of $p$ for small $p$~\citep{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g.~Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
 As we prove in Lemma~\ref{lem:asymptotic}, the notions of coarse and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
@@
 It is necessary to have an efficient method of constructing the full rSPR graph for a fixed number of leaves in order to study it.
-The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\citep{Whidden2015-yi}.
+The previously best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\citep{Whidden2015-yi}.
 Here we reduce that time to $\OhOf{mn^3}$.
-Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement in practice, as we demonstrate below.
-
-In previous work~\citep{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of $m$ high probability trees sampled from phylogenetic posteriors to compare mixing and identify local maxima.
+Note that for the full rSPR graph, the value of $m$ is given by the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement in practice, as we demonstrate below.
+
+In previous work~\citep{Whidden2015-yi}, we constructed (unrooted) SPR graphs from subsets of high probability trees sampled from phylogenetic posteriors to compare mixing and identify local maxima.
 Although the SPR distance (rooted and unrooted) is NP-hard to compute~\citep{bordewich05,hickey2008sdc}, it is fixed-parameter tractable with respect to the distance in the rooted case~\citep{bordewich05}.
 In particular, one can determine in $\OhOf{n}$-time whether two rooted phylogenetic trees are adjacent in the rSPR graph ($\OhOf{n^2}$-time for unrooted trees) using the algorithms of Whidden et al.~\citep{whidden2009unifying,whidden2010fast, whidden2013hybridization,Whidden2015-yi}.
-We previously applied this method comparing each of the $m$ trees pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
+We previously applied this method comparing each of the $m$ trees in a given graph pairwise to identify adjacencies, requiring a total of $\OhOf{m^2n}$-time ($\OhOf{m^2 n^2}$-time in the unrooted case).
 However, this method is impractical when applied to construct graphs with 7 or more leaves, due to the rapidly growing $\OhOf{m^2}$ factor.
@@
 The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
-The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string.
+The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string (assuming integer labels and bounded reasonable values of $n$, see~\citep{whidden2016efficiently}).
 These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
-We construct a mapping from each tree $T_i$ to its order index in this list $i$.
+We construct a mapping from each tree $T_i$ to its order index in this ordering i$i$.
 Begin with an empty graph $G$.
@@
 
-	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.
-red-black trees~\citep{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
+	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g.~red-black trees~\citep{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
 	As described above, the integer labels are simply the order of the input trees.
@@
 	We store the tree to index mappings for current vertices of $G$ in a trie~\citep{fredkin1960trie} using the Newick representation.
-	This requires only $\OhOf{n}$-time for each tree (i.e.
+	This requires only $\OhOf{n}$-time for each neighbor tree (i.e.
 a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).

### 2016-11-29 response_to_reviewers.txt
(most) revisions up to page 20

@@
 Page 10
-42: The wording 'within a subtree' seems a bit imprecise and could be clarified by changing it to
-something along the lines of 'within the subtree to be moved / regrafted'.
+
+42: The wording 'within a subtree' seems a bit imprecise and could be clarified by changing it to something along the lines of 'within the subtree to be moved / regrafted'.
+* revised to "within the subtree to be moved".
+
 55: The term 'every' seems to be more fitting than 'each' in this context.
+* corrected
+
 Page 11
-9-12: The wording seems a bit imprecise because the target of a move is not the node itself but the
-node's edge.
-23-25: The case distinction requires that the extra root node ρ has depth -1, the actual root depth 0
-and so on, but this is stated nowhere in the paper.
-33: Probably, the set-difference operator ('\') is supposed to denote the pruning (including the
-merging of the incoming and outgoing edge of the resulting unary node). Unfortunately, this is not
-explained. A short explanation would make this more obvious to the reader.
+
+9-12: The wording seems a bit imprecise because the target of a move is not the node itself but the node's edge.
+* The wording was indeed confusing. We are assigning moves for counting
+	purposes to one possible source node, not specifying a move to any
+	given node. We added the sentence "To do so, we assign each of the
+	neighbors to one node of the tree, such that any given neighbor can be
+	obtained by moving the subtree rooted at its assigned node." We then
+	changed each mention of assigning moves to assigning neighbors.
+
+23-25: The case distinction requires that the extra root node ρ has depth -1, the actual root depth 0 and so on, but this is stated nowhere in the paper.
+* We added a sentence defining the depth of the rho node to be -1 to the
+	preliminaries: "The depth of \rho is considered to be -1, the depth of
+	the original root is 0, and the depth of each other node is given by
+	its distance from the original root.
+
+33: Probably, the set-difference operator ('\') is supposed to denote the pruning (including the merging of the incoming and outgoing edge of the resulting unary node).
+Unfortunately, this is not explained.
+A short explanation would make this more obvious to the reader.
+* We have added an explanation here: "This implies that both T \U = S \U
+	and T \V = S \V by the definition of an rSPR operation, where \ here
+	indicates removing the subtree and its parent edge and then replacing
+	its former parent node and both adjacent edges with a single edge."
+
 Page 12
+
 15: It looks like that '-th' is missing behind the '|r|' in order to turn it into an ordinal number.
+* corrected
+
 23: What is the unit of the 'O(n) space'?
-52-55: The significance of this paragraph could be increased by stating the special features,
-properties or outcomes of these runs instead of barely stating that you have performed them.
+* bits of memory, as is standard for the Big-O notation with respect to space.
+
+52-55: The significance of this paragraph could be increased by stating the special features, properties or outcomes of these runs instead of barely stating that you have performed them.
+* The runs are analyzed in the next section. To avoid confusion we moved
+	the description of the runs to the Access time simulation section and
+	simply say "We used this procedure to sample random walks on varying
+	size rSPR graphs for our analysis in the next section" here.
+
 Page 13
-32: Which graph is 'the rSPR graph ... as discussed in Section 3.1'? The complete one with (2n-3)!!
-trees? This is not quite clear since section 3.1 also considers other values for m.
-33: It is stated that the curvature is computed for 'given pairs of trees', but who / what gives these
-pairs? The wording is a bit odd here.55: The word 'tempered' is slightly ambiguous, because it can have the meaning 'to moderate' as
-well as 'to harden'.
+32: Which graph is 'the rSPR graph ... as discussed in Section 3.1'? The complete one with (2n-3)!!  trees? This is not quite clear since section 3.1 also considers other values for m.
+* This refers to the complete graph. We revised our use of m as
+	described above, to limit this type of confusion. In addition we added
+	the word full ("full rSPR graph") to this sentence.
+
+33: It is stated that the curvature is computed for 'given pairs of trees', but who / what gives these pairs? The wording is a bit odd here.
+* This is explained in the next paragraph. We could not compute
+	curvatures for all pairs of trees directly, but instead compute
+	curvatures for each tanglegram. The procedure described here is to
+	used to compute the curvature of any given pair of trees.
+
+55: The word 'tempered' is slightly ambiguous, because it can have the meaning 'to moderate' as well as 'to harden'.
+* revised to "reduced"
+
+****TODO FIGURES START*****
+
 Page 14
-56-: The connection between a smaller κ and a right-shift of the access time distribution is difficult
-to see as differences in the saturation are hardly noticable.
+
+56-: The connection between a smaller κ and a right-shift of the access time distribution is difficult to see as differences in the saturation are hardly noticable.
+
 Page 15
+
 Plots of Figure 5:
-1. It is very hard to extract any information from the plots, because the plots (as soon as there
-are a few more lines) quickly get congested due to overlaps and hardly (or not at all) visible
-differences in saturation.
-2. Adding axis labels (and not just explaining them in the figure caption) could help the
-readability of the plots.
+1.
+It is very hard to extract any information from the plots, because the plots (as soon as there are a few more lines) quickly get congested due to overlaps and hardly (or not at all) visible differences in saturation.
+
+2. Adding axis labels (and not just explaining them in the figure caption) could help the readability of the plots.
+
 3. It might be a bit counterintuitive to associate smaller κ values with an increased saturation.
-4. Furthermore, the print version of articles in this journal usually come with monochrome
-images, (then) making it basically impossible to distinguish the lines by colour and
-saturation.
+
+4. Furthermore, the print version of articles in this journal usually come with monochrome images, (then) making it basically impossible to distinguish the lines by colour and saturation.
+
 Page 16
@@
 also hamper the readability (when the differences are as small as they are here).
+
+****TODO FIGURES END*****
+
 45-56:
+
 1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
-2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of
-pairs of trees with a smaller number of taxa'?
+* This referred to both of the cases mentioned in text (early and later
+	access times). We revised the phrase to "these plots" to avoid
+	confusion.
+
+2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
+* These plots demonstrate the effect of curvature on access times when variables such as degree are controlled.
+Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.
+	A pair of trees that both contain leaves 1 and 2 as siblings are an
+	example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves.
+
 Page 17
+
 12-15: Adding hyphens (e.g. 'distance-1 pairs') might increase the readability.
-15-23: The traceability / comprehensibility of the statements would strongly benefit from improving
-the readability of the plots in Figure 5 and 6.
+* We have made the suggested change
+
+15-23: The traceability / comprehensibility of the statements would strongly benefit from improving the readability of the plots in Figure 5 and 6.
+* TODO
+
 53: It looks like there is too much white space behind 'w.l.o.g.'.
+* corrected
+
 Page 18
-Proof of (i):The proof only works when the '+14' in the numerator is not reduced to '+12' (which is also not
-necessary, because bounding the sum of the logarithms in the denominator is sufficient.
+
+Proof of (i):The proof only works when the '+14' in the numerator is not reduced to '+12' (which is also not necessary, because bounding the sum of the logarithms in the denominator is sufficient.
+* TODO: Do you see a problem here Erick? Reducing the numerator from 14
+	to 12 is valid. Maybe he is confused by the last step, where I borrow
+	from the reduced -n in the denominator to reduce the 20 to 18?
+
 Proof of (ii):
-There seems to be a small error in the format. The penultimate equality is not aligned with the other
-equal and less-than signs.
-42-43: The statement 'The maximum degree difference from Lemma 5.2. occurs between trees that
-differ by many rSPR operations.' is (in this generality) not correct or at least misleading. Consider
-the rSPR graph for X-trees with four leaves from Figure 3(a). The ladder tree and the balanced tree
-have the maximum degree difference but are only separated by a single rSPR move while there are
-other tree pairs separated by two moves (which, in this context / example, should be considered as
-many).
+There seems to be a small error in the format. The penultimate equality is not aligned with the other equal and less-than signs.
+* corrected
+
+42-43: The statement 'The maximum degree difference from Lemma 5.2.
+occurs between trees that differ by many rSPR operations.' is (in this generality) not correct or at least misleading.
+Consider the rSPR graph for X-trees with four leaves from Figure 3(a).
+The ladder tree and the balanced tree have the maximum degree difference but are only separated by a single rSPR move while there are other tree pairs separated by two moves (which, in this context / example, should be considered as many).
+* revised to "may occur". Our intent here is simply to lead up to our
+	improved bounds for adjacent trees.
+
 Page 19
+
 39:
-1. There seems to be an inconsistency between the definition of internal edges and their
-number stated in this line. On page 5, internal edges are defined as those that do not contact
-a leaf or ρ. Then, a 'rooted tree with k leaves' cannot have 2(k-1) internal edges as stated here
-(because this is the number of all edges (also those leading to leaves), disregarding the extra
-root edge). So, either the definition of 'internal edge' has to be adopted or the word 'internal'
-has to be dropped here.
+1.
+There seems to be an inconsistency between the definition of internal edges and their number stated in this line.
+On page 5, internal edges are defined as those that do not contact a leaf or ρ.
+Then, a 'rooted tree with k leaves' cannot have 2(k-1) internal edges as stated here (because this is the number of all edges (also those leading to leaves), disregarding the extra root edge).
+So, either the definition of 'internal edge' has to be adopted or the word 'internal' has to be dropped here.
+* We have dropped the word "internal" here.
+
 2. The white space after 'edges' is missing.
+* corrected
+
 Page 20
+
 9-19: While Corollary 5.4 appears correct, the given justification seems to be a bit problematic.
-Based on the definition of γ and the excluded (subtracted) moves, it appears that these γ moves still
-include some which move R. For example, from the set of all possible moves for T (deg(T)), only
-some moves into R (2kb, case (ii)) and some moves of R's parent subtree (2(j-1), case (iv)) are
-excluded.
-However, it is then stated that these moves 'can be applied to T and S with the same source and
-target nodes'. The minor (but arguable) problem concerns the part 'same source nodes': When are
-they the same? Is the parent node of R the same although its position differs in T and S? If not, then
-not all γ moves can be applied with the same source nodes. The bigger problem is created by the
-fact that moves moving R with the same target node eliminate the only difference between T and S
-and, thus, the difference between the obtained neighbours is zero (not one).
-Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be
-obtained.
-14: The phrase 'the pairings are disjoint' might need some more explanation as 'disjoint' might get
-interpreted in several ways (e.g. no tree is allowed to occur in more than one pairing vs. the pairings
-are pairwise distinct).
+Based on the definition of γ and the excluded (subtracted) moves, it appears that these γ moves still include some which move R.
+For example, from the set of all possible moves for T (deg(T)), only some moves into R (2kb, case (ii)) and some moves of R's parent subtree (2(j-1), case (iv)) are excluded.
+However, it is then stated that these moves 'can be applied to T and S with the same source and target nodes'.
+The minor (but arguable) problem concerns the part 'same source nodes': When are they the same?
+Is the parent node of R the same although its position differs in T and S?
+If not, then not all γ moves can be applied with the same source nodes.
+The bigger problem is created by the fact that moves moving R with the same target node eliminate the only difference between T and S and, thus, the difference between the obtained neighbours is zero (not one).
+Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be obtained.
+* revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.". 
+* TODO: We can either change the distance bound here to <= 0 or subtract
+	out the distance 0 pairs. Check which is better after completing the
+	other revisions.
+
+14: The phrase 'the pairings are disjoint' might need some more explanation as 'disjoint' might get interpreted in several ways (e.g.
+no tree is allowed to occur in more than one pairing vs.
+the pairings are pairwise distinct).
+* revised to note "...  each neighbor of T is paired
+with a single neighbor of S, and vice versa. Note that a tree may be a
+neighbor of both T and S and thus involved in two different pairings."
+
 37: The word 'and' seems to be missing before 'then' .
+* corrected
+
 Lemma 5.5:
 1. Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is
-necessary (Why is it necessary to increase j? Why are the subcases of (not) moving theadditional leaf necessary?) in order to clarify the line of argument.
+necessary (Why is it necessary to increase j? Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
+* TODO
+
 2. Concerning statement (ii): Is there a particular reason why the quotient is bounded for n ≥ 4?
 The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
+*
+
 Page 21

### 2016-11-29 curvature.tex
(most) revisions up to page 20

@@
 As in~\citep{bordewich05,whidden2013hybridization,Whidden2015-yi}, the tree is augmented with a labelled root node $\rho$ and $\rho$ is considered a member of $X$ (Fig.~\ref{fig:x-tree}).
+The depth of $\rho$ is considered to be $-1$, the depth of the original root is 0, and the depth of each other node is given by its distance from the original root.
 We generally use $n$ to refer to the number of leaves in an $X$-tree.
@@
 The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
-The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string (assuming integer labels and bounded reasonable values of $n$, see~\citep{whidden2016efficiently}).
+The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string of characters (assuming integer labels and bounded reasonable values of $n$, see~\citep{whidden2016efficiently}).
 These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
@@
 
-We consider the distribution of rSPR moves in terms of the number of nodes contained within a subtree.
+We consider the distribution of rSPR moves in terms of the number of nodes contained within the subtree to be moved.
 Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
@@
 \end{enumerate}
-We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is, each other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
-We assign $(2n-1-x)-2$ moves to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ moves to each other non-root node.
+We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is, every other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
+To do so, we assign each of the neighbors to one node of the tree, such that any given neighbor can be obtained by moving the subtree rooted at its assigned node.
+We assign $(2n-1-x)-2$ neighbors to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ neighbors to each other non-root node.
 Let $N(T,u)$ denote the neighbors of $T$ assigned to node $u$, obtained by moving the subtree $R$ rooted at $u$.
@@
 	Call these $U$ and $V$, respectively.
-	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
+	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation, where $\setminus$ here indicates removing the subtree and its parent edge and then replacing its former parent node and both adjacent edges with a single edge.
 	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
@@
 		\item Decrease $r$ by $\size{N(T,u)}$.
-		\item If $r < 0$, let $S$ be the $\abs{r}$ member of $N(T,u)$ and terminate the for loop.
+		\item If $r < 0$, let $S$ be the $\abs{r}$-th member of $N(T,u)$ and terminate the for loop.
 	\end{enumerate}
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\citep{random_spr_walk}.
-We sampled a 200,000-iteration random walk on the 5-leaf rSPR graph, a 50,000-iteration random walk on the 6-leaf rSPR graph, and a 5,000-iteration random walk on the 7-leaf rSPR graph.
+We used this procedure to sample random walks on varying size rSPR graphs four our analysis in the next section.
 
@@
 \label{sec:results:curvature}
-To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
+To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the full rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
 We then computed curvatures for given pairs of trees directly, by using linear programming~\citep{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \citep{SAGE} front-end to the GLPK~\citep{glpk} solver; code can be found in \citep{gricci} which grew from the code described in \citep{Loisel2014-hu}.
@@
 Neighboring trees achieved minimum curvature values for a given number of leaves, and we found maximum curvature values between trees at maximum distance or one rSPR move closer than the maximum.
-This suggests that the increased difficulty of moving between trees with a random walk due to distance may be tempered somewhat by larger curvature in the highly connected rSPR~graph.
+This suggests that the increased difficulty of moving between trees with a random walk due to distance may be reduced somewhat by larger curvature in the highly connected rSPR~graph.
 
@@
 Here, to gain more insight, we used simulation to approximate the entire access time distribution.
+Using the procedure of Corollary~\ref{cor:enumerate_neighbors}, we sampled a 200,000-iteration random walk on the 5-leaf rSPR graph, a 50,000-iteration random walk on the 6-leaf rSPR graph, and a 5,000-iteration random walk on the 7-leaf rSPR graph.
 Again we use the idea that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
@@
         Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-taxon trees with degree 24 and (d) 6-taxon trees with degree 44.
-				In both cases we exclude pairs that are simple inclusions of pairs of trees with a smaller number of taxa.
+				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of taxa (e.g.
+pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
 We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
-For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 2 minus the count for time 3.
+For example, $\delta_1$ for distance-1 pairs (green lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance-3 pairs (blue lines in Fig.~\ref{fig:kappa-access-distribution}) is the count for time 2 minus the count for time 3.
 Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
@@
 		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
-		=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
+		&=\ 4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
 		&=\ n^2 - 5n + 6.
@@
 
-The maximum degree difference from Lemma~\ref{lem:degree_max_delta} occurs between trees that differ by many rSPR operations.
+The maximum degree difference from Lemma~\ref{lem:degree_max_delta} may occur between trees that differ by many rSPR operations.
 In our next lemma, we characterize how the degree of a tree changes after a single rSPR operation.
@@
 
-Recall that a rooted tree with $k$ leaves has $2(k-1)$ internal edges(recall that we are excluding any ``root edge'' in these calculations).
+Recall that a rooted tree with $k$ leaves has $2(k-1)$ edges (recall that we are excluding any ``root edge'' in these calculations).
 In the first case there are $a$ subtrees that can now be moved onto the $2k$ edges in $R$ (including its newly introduced parent edge and one of the newly subdivided root edges of $V$) for a total gain of $2ka$ distinct moves.
@@
 $$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
-trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
+trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that $\dspr{T',S'} = 1$ for each $(T',S')$ pair, each neighbor of $T$ is paired with a single neighbor of $S$, and vice versa.
+Note that a tree may be a neighbor of both $T$ and $S$ and thus involved in two different pairings.
 \end{cor}
@@
 For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
+We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
 \end{proof}
@@
 	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
-	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.
+	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) and then maximally balancing the terms in the product $ka$ as follows.
 

### 2016-11-30 response_to_reviewers.txt
finished first pass of revisions

@@
 The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
-*
+* 5/6 is closer to the limit of 6/7 than it is to 4/5 so this gives a
+	closer bound at the small cost of excluding the trivial 3-leaf tree graph.
 
 Page 21
+
 49:
-1. The meaning of the function L(.) can be guessed from the context without too much effort,
-but providing a short definition / explanation would help avoiding any misconceptions.
-2. Does this equality really hold between the restricted versions of T and S , which differ only in
-R and, thus, should still be different after restricting them to X \ L(Q)?
-51-53: Using U as a subtree while it also denotes a whole tree in the neighbourhood of T and S just
-a few lines before can be slightly confusing.
+
+1. The meaning of the function L(.) can be guessed from the context without too much effort, but providing a short definition / explanation would help avoiding any misconceptions.
+* We added "... , where L(Q) is the label set of Q, ..." to the sentence.
+
+2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
+* Yes, because of the triangle formed by T, S, and U, as we state in the
+	proof. T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position.
+
+51-53: Using U as a subtree while it also denotes a whole tree in the neighbourhood of T and S just a few lines before can be slightly confusing.
+* Indeed, thanks. We changed the second use of U to V and u' to v'.
+
 Page 22
-30-32: The statement on the proportionality of small moves and small subtrees seems to be a bit
-troubled.
-1. For example, if both children of the root are not small, each small subtree with one leaf
-contributes 2n-6 moves, while each small subtree with two leaves contributes 2n-8 moves.
+
+30-32: The statement on the proportionality of small moves and small subtrees seems to be a bit troubled.
+
+1. For example, if both children of the root are not small, each small subtree with one leaf contributes 2n-6 moves, while each small subtree with two leaves contributes 2n-8 moves.
 So, what is the factor of proportionality?
-2. In the phrase 'where one of the root's children is small', is the intended meaning exactly one
-or at least one? Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the
-majority of) the leaves), the statement does not hold in general (neither for exactly nor for at
-least). Consider, e.g., a tree with a leaf as the only small child of the root and at least three
-small subtrees with two leaves (for the case exactly) respectively the balanced tree with four
-leaves (for the case at least).
-36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more
-consistent with the previous use of numerals).
+* TODO
+
+2.
+In the phrase 'where one of the root's children is small', is the intended meaning exactly one or at least one?
+Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the majority of) the leaves), the statement does not hold in general (neither for exactly nor for at least).
+Consider, e.g., a tree with a leaf as the only small child of the root and at least three small subtrees with two leaves (for the case exactly) respectively the balanced tree with four leaves (for the case at least).
+* TODO
+
+36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more consistent with the previous use of numerals).
+* We have made the suggested change.
+
 38-42:
-1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n2-4n-6) is not
-correct for all intended n. The reason is that for the ladder tree with three leaves, the small
-subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5)
-moves. Hence, a case distinction seems to be necessary here and for the inequality in the
-subsequent paragraph (lines 40-42)
-2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads
-to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case
-distinction n = 3 vs. n ≥ 4).
-45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45
-possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the
-typical notion of size in mind).
-54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to
-hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3
-for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).
+
+1.
+Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n2-4n-6) is not correct for all intended n.
+The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves.
+Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
+* TODO
+
+2.
+The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads to the n-free term '+12' (not '+16').
+But this can be 'repaired' through the necessary case distinction n = 3 vs.
+n ≥ 4).
+* TODO
+
+45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
+* we added the statement to line 45.
+
+54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to hold for all n ≥ 4.
+For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3 for the balanced tree.
+Similarly, the balanced tree has a larger one for n = 8 (1.55 vs.
+1.39).
 This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
+* TODO
+
 56: Similar to line 36, 'one leaf' and 'one subtree' appear more readable / consistent.
+* We have made the suggested change.
+
 Page 23
+
 11-13: The vertical bars denoting the size of N i (T) are missing at all three occurrences.
+* corrected
+
 11-20:
-1. It is stated that |N i (T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N 1 (T)| = (n-1)(2n-1-5) +
-(2n-1-3) > 2n-3-5.
-2. Also, it is stated that |N i (T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n-
-(2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
-3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-
-3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
-4. Is it actually obvious enough that setting all |N i (T)| to 2n-3-5 leads to an upper for the
-quotient representing the average moved subtree size? A short explanation might make this
-point clearer.
-5. In the last sentence of the proof, the weighted average uses a proportion of 1⁄2 for both small
-and large subtree (which is okay), but why is the average moved subtree size of small
-substrees assumed to be one? Shouldn't the one small subtree with two leaves (which always
-exists) make it slightly larger than one?
-32: Is there are special reason for (re)defining the mass transport term W 1,n ?
+
+1. It is stated that |N_i(T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N_1(T)| = (n-1)(2n-1-5) + (2n-1-3) > 2n-3-5.
+* TODO
+
+2. Also, it is stated that |N_i(T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n- (2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
+* TODO
+
+3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
+* TODO
+
+4. Is it actually obvious enough that setting all |N_i (T)| to 2n-3-5 leads to an upper for the quotient representing the average moved subtree size? A short explanation might make this point clearer.
+* TODO
+
+5.
+In the last sentence of the proof, the weighted average uses a proportion of 1⁄2 for both small and large subtree (which is okay), but why is the average moved subtree size of small substrees assumed to be one?
+Shouldn't the one small subtree with two leaves (which always exists) make it slightly larger than one?
+* TODO
+
+32: Is there are special reason for (re)defining the mass transport term
+W_{1,n} ?
+* W_{1,n} is used in Theorem 6.1 to distinguish between the sequential
+	mass transport terms, but this this one should just be W_1. Corrected.
+
 37: The verb 'increase' misses the third-person singular s (as it refers to 'the size of the trees').
+* Corrected.
+
 45: Only one of the two occurrences of 'thus' is needed.
+* We removed the second occurence of "thus".
+
 Page 24
-14: Adding a short explanation why the number of unmatched trees is in O(kn) might help the
-reader understanding the argument at this point better.
+
+14: Adding a short explanation why the number of unmatched trees is in O(kn) might help the reader understanding the argument at this point better.
+* We added the short explanation "There are O(kn) trees unmatched by
+	this pairing, because b and a (in the nomenclature of Corollary 5.4)
+	are bounded by n."
+
 19: The innermost pair of parentheses around 'O(n)' is not necessary.
+* Corrected.
+
 30: Replacing the word 'a' before 'distance' by 'at' or 'of' conveys the meaning better.
-31: The equality sign in the equation might be less fitting than a <, because it is previously (line 25)
-stated that S n has a strictly larger neighbourhood.
-46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to
-bounds for W 1 (m T ,m S ) could make it easier for the reader to grasp the idea how the proof works.
+* Replaced with "of".
+
+31: The equality sign in the equation might be less fitting than a <, because it is previously (line 25) stated that S_n has a strictly larger neighbourhood.
+* It is counterintuitive at first glance, but the equal sign is correct
+	because T_n and S_n differ by an O(kn) factor.
+
+46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to bounds for W_1 (m_T ,m_S ) could make it easier for the reader to grasp the idea how the proof works.
+* TODO
+
 Page 25
-29: Possibly it should be 'each other' instead of 'each either'.40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree
-neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
+29: Possibly it should be 'each other' instead of 'each either'.
+* corrected
+
+40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
+* TODO
+
 Page 26
-10: A short remark on how the term '1 + μ' is actually derived would make this passage easier /
-faster to comprehend
-14: Following Corollary 5.4 it should be 'at least γ' and not 'at most γ' paired trees (and, otherwise,
-the argumentation is not sensible at this point).
-19: The rationale for bounding the degree by using the maximum possible absolute difference
-(irrespective of the actual move) seems a bit unclear. Currently, it does not look like the line of
-argument leads to the maximum μ: using the maximum degree of T (instead of the actual,
-potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily
-large denominator.
+
+10: A short remark on how the term '1 + μ' is actually derived would make this passage easier / faster to comprehend
+* TODO
+
+14: Following Corollary 5.4 it should be 'at least γ' and not 'at most γ' paired trees (and, otherwise, the argumentation is not sensible at this point).
+* Corrected.
+
+19: The rationale for bounding the degree by using the maximum possible absolute difference (irrespective of the actual move) seems a bit unclear.
+Currently, it does not look like the line of argument leads to the maximum μ: using the maximum degree of T (instead of the actual, potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily large denominator.
+* TODO
+
 20:
-1. The last inequality ('≤ 1⁄2n2-n') holds only if n is even. For odd n, 1⁄2 has to be added to the
-upper bound (which is then also okay for the even case). In the end, this change also affects
-the curvature bound.
-2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would
-not be wrong, it is a potential source of confusion.
-39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the
-proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to
-it.
+
+1.
+The last inequality ('≤ 1⁄2n2-n') holds only if n is even.
+For odd n, 1⁄2 has to be added to the upper bound (which is then also okay for the even case).
+In the end, this change also affects the curvature bound.
+* TODO
+
+2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would not be wrong, it is a potential source of confusion.
+* TODO
+
+39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to it.
+* TODO
+
 Page 27
+
 35: The upper bound does not seem to follow as analogously to the lower bound as it is stated.
-Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to
-κ p (T , S )
-2
-⩽ ...⩽ κ (T , S)+
-p
-p∗max (∣ N (T )∣ , ∣ N (S )∣)
-where the factor p in the denominator causes a problem, because simply dropping it (in order to
-arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
+Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to κ p (T , S ) 2 ⩽ ...⩽ κ (T , S)+ p p∗max (∣ N (T )∣ , ∣ N (S )∣) where the factor p in the denominator causes a problem, because simply dropping it (in order to arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
+* TODO (needs checking from the PDF)
+
 Page 28
-11-12: According to Lemma 5.5, the quotient of degrees for two adjacent trees is at least 5/6 only
-for n ≥ 4, but there is no such assumption in Lemma 6.9 or its proof.
-14-18: Using W 1 (T,S) (resp. W' 1 (T,S)) is slightly imprecise as T and S are the trees and not the
-probability densities.
-26: Within the min-term of the last case, w is the argument of all four probability densities.
-Probably, two of them should have z as their argument.29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is
-slightly imprecise.
+
+11-12: According to Lemma 5.5, the quotient of degrees for two adjacent trees is at least 5/6 only for n ≥ 4, but there is no such assumption in Lemma 6.9 or its proof.
+* TODO, add assumption
+
+14-18: Using W 1 (T,S) (resp. W' 1 (T,S)) is slightly imprecise as T and S are the trees and not the probability densities.
+* TODO
+
+26: Within the min-term of the last case, w is the argument of all four probability densities.  Probably, two of them should have z as their argument.
+* TODO
+
+29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is slightly imprecise.
+* TODO
+
 50: The word 'the' seems to be missing before 'understanding'.
-55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the
-regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven
-lemma) the changes depend linearly on the product (which in turn is quadratic in the number of
-leaves).
+* Corrected.
+
+55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven lemma) the changes depend linearly on the product (which in turn is quadratic in the number of leaves).
+* TODO
+
 Page 29
+
 13-17: The same problem as on page 17, lines 15-23.
+* TODO (address the plots)
+
 29: Changing 'achieve' to 'to be achieved' seems a bit more readable.
+* Corrected

### 2016-11-30 curvature.tex
finished first pass of revisions

@@
 	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
-	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
-	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
+	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
+	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
 	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
@@
 	A ladder tree has the smallest proportion of small subtrees, so we first bound the number of small subtree moves that can be applied to a ladder tree.
-	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.
+	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and one 2-leaf subtree.
 	One of the 1-leaf subtrees is a child of the root.
@@
 
-Now, we bound the average size of the moved subtrees.
+Now, we bound the average size of the moved subtrees (in terms of the number of leaves).
 
@@
 	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
-	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
+	There are $n$ subtrees with one leaf, and one subtree each with $2,3, \ldots n - 1$ leaves.
 	We ignore the root subtree, which does not induce any rSPR moves.
 	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
-    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
+    $\size{N_i(T)} \leq 2n - 3 - 5$ for the first set of terms, and $\size{N_i(T)} > 2n - 3 - 5$ for the second, so by replacing $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
 	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
@@
 Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
-Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
+Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_1$ is the mass transport term \eqref{eq:curvatureDef}.
 For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.
@@
 We showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean} that most adjacent trees differ by a small subtree, raising the question of what the curvature is between such pairs.
-In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the size of the subtree moved to transition between the two trees remains the same size.
+In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increases but the size of the subtree moved to transition between the two trees remains the same size.
 To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
 The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
-Thus this ``flatness-in-the-limit'' theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
+Thus this ``flatness-in-the-limit'' theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
 
@@
 We pair these trees in the mass transport.
-There are $\OhOf{kn}$ trees unmatched by this pairing, and we can pair each of them arbitrarily with another tree of distance at most 3.
+There are $\OhOf{kn}$ trees unmatched by this pairing, because $b$ and $a$ (in the nomenclature of Corollary~\ref{cor:paired_neighbors}) are bounded by $n$.
+We can pair each of the unmatched trees arbitrarily with another tree of distance at most 3.
 Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.
@@
 A lower bound is also available because we can't do better than distance 1 for all trees except for shared neighbors, of which there are $\OhOf{n}$ by Lemma~\ref{lem:shared_neighbors}.
-By ignoring these trees we get a lower bound of $(\degree{T_n} - (\OhOf{n}))/\degree{T_n}$ for $W_{1,n}$.
+By ignoring these trees we get a lower bound of $(\degree{T_n} - \OhOf{n})/\degree{T_n}$ for $W_{1,n}$.
 
@@
 Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
-We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
+We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees of distance at most 3.
 Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
@@
 Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
-Moreover, nearly all of the neighbors of these pairs approach each either.
+Moreover, nearly all of the neighbors of these pairs approach each other.
 We thus conjecture the following:
@@
 	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
-	By Corollary~\ref{cor:paired_neighbors}, there are at most $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
+	By Corollary~\ref{cor:paired_neighbors}, there are at least $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
 	We now construct a pair of trees that maximizes both effects.
@@
 \section{Conclusion and future work}
-In summary, we have gone beyond graph diameter and vertex degree to substantially advance understanding of the phylogenetic rSPR graph.
+In summary, we have gone beyond graph diameter and vertex degree to substantially advance the understanding of the phylogenetic rSPR graph.
 We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
@@
 In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
-Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
+Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to be achieved, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
 One avenue of exploration will be to follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.

### 2016-12-01 response_to_reviewers.txt
most comments addressed, some figure comments remaining, two other TODOs remaining

@@
 27: The spelling 'multimodal' seems to be more common than 'multi-modal' (and would be in line with other uses in your paper; e.g. page 3, line 12).
-* corrected
+* Corrected.
 
 30: The word 'the' seems to be missing before 'study'.
-* corrected
+* Corrected.
 
 32: The word 'the' seems to be missing before 'Ricci-Ollivier'.
-* corrected
+* Corrected.
 
@@
 In its current state, it might not be clear from the wording whether each tree has n leaves or whether n is the total number of leaves of all trees.
-* revised to "... and n is the number of leaves in each tree, ..."
+* Revised to "... and n is the number of leaves in each tree, ...".
 
 44-46: The phrasing 'We show ... showing ...' sounds a bit odd because 'show' is used twice.
-* revised "showing" to "demonstrating"
+* Revised "showing" to "demonstrating".
 
@@
 10: It looks like some kind of connecting phrase is missing between 'particular' and 'properties' (e.g.  'with respect to').
-* we added the suggested connecting phrase
+* We added the suggested connecting phrase.
 
 11: The word 'the' seems to be missing before 'investigation'.
-* corrected
+* Corrected.
 
 52-53: As you start this part of this sentence with 'both', an 'and' seems to be more fitting in 'leaves [12, 13] or for SPR'.
-* corrected
+* Corrected.
 
 54: Do you rather mean 'an infinite number of leaves' or 'infinitely many leaves' instead of 'infinite leaves'?
-* corrected to "infinitely many leaves"
+* Corrected to "infinitely many leaves"
 
@@
 15: A more fitting position of 'still' seems to be between 'is' and 'known'.
-* corrected
+* Corrected.
 
 18: A word like 'the' or 'its' seems to be missing before 'degree'.
-* corrected to "its".
+* Corrected to "its".
 
@@
 31: It looks like the word 'investigation' is missing the indefinite article or should be in the plural form.
-* corrected, as noted above
+* Corrected, as noted above.
 
 33: The word 'the' seems to be missing before 'Ricci'.
-* corrected
+* Corrected.
 
 47: The word 'the' seems to be missing before 'curvature'.
-* corrected
+* Corrected.
 
 56: The same problem as on page 1, line 36.
-* corrected to "... each of which has n leaves".
+* Corrected to "... each of which has n leaves".
 
@@
 44: In this context, 'linear-time' seems to be more common without a hyphen.
-* corrected
+* Corrected.
 
@@
 48: The word 'the' seems to be missing before 'distance'.
-* corrected
+* Corrected.
 
@@
 27: The word 'the' seems to be missing before 'proba-'.
-* corrected
+* Corrected.
 
 29: The word 'the' seems to be missing before 'knowledge'.
-* corrected
+* Corrected.
 
@@
 40: Either the word 'the' seems to be missing before 'arbitrary' or it has to be 'arbitrarily' (in which case the 'of' before 'trees' has to be removed).
-* corrected by adding "the".
+* Corrected by adding "the".
 
@@
 13: The wording 'we will specialize to that space being a graph' sounds odd. Maybe something along the lines of 'we restrict ourselves to ...' is preferable.
-* revised to "... we restrict ourselves to graphs equipped with a shortest-path distance."
+* Revised to "... we restrict ourselves to graphs equipped with a shortest-path distance."
 
 20: It looks like the apostrophe in 'movers' is missing (it should be 'mover's').
-* corrected
+* Corrected.
 
@@
 14: It looks like there is too much white space behind 'e.g.'.
-* corrected
+* Corrected.
 
 24: In this context, 'previously' seems more appropriate than 'previous'.
-* corrected
+* Corrected.
 
@@
 42: The wording 'within a subtree' seems a bit imprecise and could be clarified by changing it to something along the lines of 'within the subtree to be moved / regrafted'.
-* revised to "within the subtree to be moved".
+* Revised to "within the subtree to be moved".
 
 55: The term 'every' seems to be more fitting than 'each' in this context.
-* corrected
+* Corrected.
 
@@
 15: It looks like that '-th' is missing behind the '|r|' in order to turn it into an ordinal number.
-* corrected
+* Corrected.
 
 23: What is the unit of the 'O(n) space'?
-* bits of memory, as is standard for the Big-O notation with respect to space.
+* Bits of memory, as is standard for the Big-O notation with respect to space.
 
@@
 55: The word 'tempered' is slightly ambiguous, because it can have the meaning 'to moderate' as well as 'to harden'.
-* revised to "reduced"
+* Revised to "reduced".
 
@@
 53: It looks like there is too much white space behind 'w.l.o.g.'.
-* corrected
+* Corrected.
 
@@
 There seems to be a small error in the format. The penultimate equality is not aligned with the other equal and less-than signs.
-* corrected
+* Corrected.
 
@@
 The ladder tree and the balanced tree have the maximum degree difference but are only separated by a single rSPR move while there are other tree pairs separated by two moves (which, in this context / example, should be considered as many).
-* revised to "may occur". Our intent here is simply to lead up to our
+* Revised to "may occur". Our intent here is simply to lead up to our
 	improved bounds for adjacent trees.
@@
 2. The white space after 'edges' is missing.
-* corrected
+* Corrected.
 
@@
 Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be obtained.
-* revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.". 
+* Revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.". 
 * TODO: We can either change the distance bound here to <= 0 or subtract
 	out the distance 0 pairs. Check which is better after completing the
-	other revisions.
+	other revisions. The argument I added after Corollary 6.6 that the Lemma 6.5 pairs of trees have negative curvature needs to be checked. Lemma 6.5 might need to take a change here in to account.
 
@@
 the pairings are pairwise distinct).
-* revised to note "...  each neighbor of T is paired
+* Revised to note "...  each neighbor of T is paired
 with a single neighbor of S, and vice versa. Note that a tree may be a
@@
 37: The word 'and' seems to be missing before 'then' .
-* corrected
+* Corrected.
 
@@
 necessary (Why is it necessary to increase j? Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
-* TODO
+* We have revised the sentence introducing the cases to state "There are two cases, depending on whether the subtree of k leaves is moved to the root (thereby introducing a new level to the tree) or not.
 
@@
 So, what is the factor of proportionality?
-* TODO
+* Proportional was too strong a word here, so we changed "directly proportional to" to "determined by".
+We also added the sentence "Moreover, every tree has n 1-leaf subtrees." to make the line of reasoning more clear.
 
@@
 Consider, e.g., a tree with a leaf as the only small child of the root and at least three small subtrees with two leaves (for the case exactly) respectively the balanced tree with four leaves (for the case at least).
-* TODO
+* See the previous response.
 
@@
 1.
-Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n2-4n-6) is not correct for all intended n.
+Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n^2-4n-6) is not correct for all intended n.
 The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves.
 Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
-* TODO
+* To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.
 
@@
 n ≥ 4).
-* TODO
+* See our response above.
 
 45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
-* we added the statement to line 45.
+* We added the statement to line 45.
 
@@
 This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
-* TODO
+* There appears to be some confusion on what constitutes a moved subtree here.
+From our distribution of rSPR moves, the balanced tree with n = 4 leaves does not move any subtrees of size 2.
+There are moves of such trees that are equivalent to moves of leaves which appears to be the source of the confusion.
+As such, the average moved subtree size for n = 8 is 1.4 for the ladder tree and approximately 1.379 (160/116) for the balanced tree.
+We have added the following explanation to make this more clear to the reader:
+
+"Now, we bound the average size of the moved subtrees (in terms of the number of leaves) from rSPR moves according to our distribution.
+Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees.
+For example, from the balanced tree with 4 leaves, we do not consider 22moves of the root’s 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution."
 
@@
 11-13: The vertical bars denoting the size of N i (T) are missing at all three occurrences.
-* corrected
+* Corrected.
 
@@
 1. It is stated that |N_i(T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N_1(T)| = (n-1)(2n-1-5) + (2n-1-3) > 2n-3-5.
-* TODO
-
 2. Also, it is stated that |N_i(T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n- (2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
-* TODO
-
 3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
-* TODO
+* Corrected.
 
 4. Is it actually obvious enough that setting all |N_i (T)| to 2n-3-5 leads to an upper for the quotient representing the average moved subtree size? A short explanation might make this point clearer.
-* TODO
+* We have moved the reference to Lemma 5.7 earler (upon which this argument relies) and revised the proof for clarity.
 
@@
 Shouldn't the one small subtree with two leaves (which always exists) make it slightly larger than one?
-* TODO
+* Yes, we are reducing the contribution of the small subtrees to achieve an upper bound.
 
@@
 
-46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to bounds for W_1 (m_T ,m_S ) could make it easier for the reader to grasp the idea how the proof works.
-* TODO
+46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to bounds for W_1 (m_T, m_S ) could make it easier for the reader to grasp the idea how the proof works.
+* We have expanded the proof with such an explanation.
 
@@
 29: Possibly it should be 'each other' instead of 'each either'.
-* corrected
+* Corrected.
 
 40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
-* TODO
+* Revised to "However, we suspect that this will require a greater understanding of the distribution of pairwise distances between tree neighborhoods than is currently known.".
 
@@
 10: A short remark on how the term '1 + μ' is actually derived would make this passage easier / faster to comprehend
-* TODO
+* Revised to "Pairs of neighbors of adjacent trees are at most distance 3 apart, so μ of the mass is moved a distance at most 3.
+The remaining 1−μ of the mass is moved between adjacent trees.
+Thus, we can bound the mass transport cost from above by 3μ + 1(1 − μ) = 1 + 2μ."
 
@@
 Currently, it does not look like the line of argument leads to the maximum μ: using the maximum degree of T (instead of the actual, potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily large denominator.
-* TODO
+* This pair of trees has the maximum possible degree change and S has the smallest possible degree.
+Any pair of trees with a smaller denominator will necessarily have a numerator that is reduced by at least the same amount, resulting in a smaller value of mu.
+We've added a sentence to this effect to make this more clear.
 
@@
 In the end, this change also affects the curvature bound.
-* TODO
+* Corrected.
 
 2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would not be wrong, it is a potential source of confusion.
-* TODO
+* Corrected
 
 39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to it.
-* TODO
+* We added such a statement
 
@@
 Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to κ p (T , S ) 2 ⩽ ...⩽ κ (T , S)+ p p∗max (∣ N (T )∣ , ∣ N (S )∣) where the factor p in the denominator causes a problem, because simply dropping it (in order to arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
-* TODO (needs checking from the PDF)
+* Indeed, the stated observation was incorrect by a p factor in the numerator, because a fraction of p (the amount of mass that is moved) divided by the number of neighbors is the amount of mass that is moved from e.g.
+T to S and can therefore be matched with the lazy portion that remains at s.
+We have corrected the error.
 
@@
 11-12: According to Lemma 5.5, the quotient of degrees for two adjacent trees is at least 5/6 only for n ≥ 4, but there is no such assumption in Lemma 6.9 or its proof.
-* TODO, add assumption
+* We added that assumption that n is at least 4.
 
 14-18: Using W 1 (T,S) (resp. W' 1 (T,S)) is slightly imprecise as T and S are the trees and not the probability densities.
-* TODO
+* Corrected.
 
 26: Within the min-term of the last case, w is the argument of all four probability densities.  Probably, two of them should have z as their argument.
-* TODO
+* Corrected.
 
 29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is slightly imprecise.
-* TODO
+* Corrected.
 
@@
 55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven lemma) the changes depend linearly on the product (which in turn is quadratic in the number of leaves).
-* TODO
+* Revised to "Moreover, we proved that rSPR graph degree changes depend on the product of the size of the regrafted subtree with its change in depth.
+This product is quadratic in the number of leaves in the trees, the same order as the degree."
 

### 2016-12-01 curvature.tex
most comments addressed, some figure comments remaining, two other TODOs remaining

@@
 
-	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root or not.
+	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root (thereby introducing a new level to the tree) or not.
     If not, then we set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
@@
 \begin{lem}
-	Let $T$ be a rooted tree with $n \ge 3$ leaves.
+	Let $T$ be a rooted tree with $n \ge 4$ leaves.
 	Let $N_S(T)$ be the neighbors of $T$ induced by small rSPR moves.
@@
 	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ rSPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
-	This implies that the number of small rSPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
+	Moreover, every tree has $n$ 1-leaf subtrees.
+	This implies that the number of small rSPR moves is determined by the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
 
@@
 
-Now, we bound the average size of the moved subtrees (in terms of the number of leaves).
+Now, we bound the average size of the moved subtrees (in terms of the number of leaves) from rSPR moves according to our distribution.
+Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees.
+For example, from the balanced tree with 4 leaves, we do not consider moves of the root's 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution.
 
@@
 	We ignore the root subtree, which does not induce any rSPR moves.
-	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
-    $\size{N_i(T)} \leq 2n - 3 - 5$ for the first set of terms, and $\size{N_i(T)} > 2n - 3 - 5$ for the second, so by replacing $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
-	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
-	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
+	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms for each subtree contributing to the values for $i \leq 2$ and those with $i > 2$.
+    The contribution to $\size{N_i(T)}$ is $\geq 2n - 3 - 5$ for the first set of terms, and the contribution to $\size{N_i(T)} < 2n - 3 - 5$ for the second.
+		By replacing the contribution to $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
+		Note that this replacement does not modify the $i = 2$ term (which was already exactly $2n - 3 - 5$ and decreases the weight of only the smallest terms (which must already be below the average size of the moved subtrees), so this provides an upper bound on the mean moved subtree size.
+	The bound for large subtrees only is $(3 + (n-2)) / 2 = (n + 1) / 2$.
 	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
@@
 	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
+	Then, any mass transported between a probability distribution $m_T$ on neighbors of $T$ and $m_S$ on neighbors of $S$ is transported a distance at least $\dspr{T,S} - 2$.
+	This implies that the mass transport cost $W_1(m_T,m_S)$ is at least $\dspr{T,S} - 2$, for any $m_T$ and $m_S$.
 	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
-	The lower bound follows similarly.
+	The lower bound follows similarly by using the maximum distance of $\dspr{T,S} + 2$ .
 \end{proof}
@@
 Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
-However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.
+However, we suspect that this will require a greater understanding of the distribution of pairwise distances between tree neighborhoods than is currently known.
 
@@
 	The curvature between adjacent trees with $n$ leaves is at least
-	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
+	$$\frac{-n^2 + 2n + 1}{3.5n^2 - 15n + 16}.$$
 \end{lem}
@@
     To bound these effects, let $\mu$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
-    We can bound the mass transport cost from above by $1 + 2\mu$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
+		Pairs of neighbors of adjacent trees are at most distance 3 apart, so $\mu$ of the mass is moved a distance at most 3.
+		The remaining $1-\mu$ of the mass is moved between adjacent trees.
+		Thus, we can bound the mass transport cost from above by $3\mu + 1(1-\mu) = 1 + 2\mu$.
 	This gives a lower bound of $1 - (1 + 2\mu) / 1) = -2\mu$ on the curvature.
@@
 	$T$ has degree at most $3.5n^2 -15n + 16$.
-	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
+	There are thus $2ka + 2(i-1) = 2\left(\floor{\frac{n}{2}}\ceil{\frac{n-2}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n + \frac{1}{2}$ unpaired neighbors, the maximum possible.
 	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
 	Thus, the maximum $\mu$ is:
-	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
+	$$\frac{\frac{1}{2}n^2 - n + \frac{1}{2}}{3.5n^2 - 15n + 16}.$$
+	Note that any pair of trees with a smaller degree change will necessarily reduce the numerator by at least the same amount as the denominator, allowing us to use the degree of $T$ here rather than assuming the minimum degree of $S$.
 	The claim follows from multiplying this value by $-2$.
@@
 However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
+This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.
 
@@
 	\begin{align*}
-		p &W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},
+		p &W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2p}{\max(\degree{T}, \degree{S})},
 	\end{align*}
-	as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
+	as at most $p / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
 	The upper bound then follows analogously to the lower bound.
@@
 	\label{lem:mh_curvature}
-	Let $T$ and $S$ be trees with $n$ leaves. Then:
+	Let $T$ and $S$ be trees with $n \ge 4$ leaves. Then:
 	\begin{align*}
@@
 	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
-	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
-	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
-	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
-	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
+	Let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
+	We construct an upper bound on $W_1(m'_T,m'_S)$ by moving mass according to $W_1(m_T,m_S)$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
+	That is, for each $W_1(m_T,m_S)$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
 	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions
@@
 	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
-	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
-	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
+	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(z)}{m_T(z)}\right)$.
+	The maximum possible mass that is not moved according to $W_1(m_T,m_S)$ is $\frac{1}{6}$.
 	Moreover, the affected mass must be moved through at most two additional trees.
-	Then, $W_1' \le W_1 + \frac{2}{6}$.
+	Then, $W_1(m'_T,m'_S) \le W_1(m_T,m_S) + \frac{2}{6}$.
 	We now have:
 	\begin{align*}
-		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
+		\curvature{\MH; T,S} & \ge 1 - \frac{W_1(m_T,m_S) + \frac{1}{3}}{\dspr{T,S}} \\
 		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}.
@@
 We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
-Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth.
+Moreover, we proved that rSPR graph degree changes depend on the product of the size of the regrafted subtree with its change in depth.
+This product is quadratic in the number of leaves in the trees, the same order as the degree.
 As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.

### 2016-12-01 response_to_reviewers.txt
addressed 2 out of 3 of reviewer 1's comments

@@
+Reviewer 1
+
+I think that the abstract is too long and should be three-time shorter.
+* TODO, any comments Erick?
+
+There are several occurrence of the word “taxon” or “taxa” while there is no definition: I guessed it was the leaves, and maybe the authors want to replace every occurrence.
+* replaced each occurence of taxon or taxa with leaf or leaves.
+
+Section 6 and Th 6.1 – in the paragraphs before Th 6.1 and most importantly in Th 6.1, it would be good to say that n is the number of leaves (even if n is always the number of leaves in the article).
+In particular, the fact that S_n and T_n both have n leaves should appear in the Th.
+* We have made the suggested change.
+
 Reviewer 2

### 2016-12-01 curvature.tex
addressed 2 out of 3 of reviewer 1's comments

@@
 We implemented this procedure in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\citep{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
-The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
-Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
-Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
+The construction procedure reduced the time required to compute the 10,395-vertex 7-leaf rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
+Moreover, although we do not study the 135,135-vertex 8-leaf rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
+Constructing the 8-leaf rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
 Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets induced by MCMC credible sets similar to~\citep{Whidden2015-yi}, as the algorithm can quickly construct rSPR graphs for any given subset of trees.
@@
 	\hspace*{\stretch{1}}
-	\subfigure[6 taxa]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter6}}
+	\subfigure[6 leaves]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter6}}
 	\hspace*{\stretch{2}}
-	\subfigure[7 taxa]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter7}}
+	\subfigure[7 leaves]{\includegraphics[width=2.3in]{figs/rspr-upmh-scatter7}}
 	\hspace*{\stretch{1}}
@@
 \caption{p-values for ordinary least squares linear multiple regression of rSPR mean access time against degree and distance (two-tailed $t$-test of regression coefficient).
-The p-values for 7 taxa are smaller than the machine precision used to calculate them.}
+The p-values for 7 leaves are smaller than the machine precision used to calculate them.}
 \csvautotabular{tables/mean-access.csv}
@@
 	\caption{\
-        Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-taxon trees, grouped by the degree of the trees.
+        Distribution of rSPR $\MH$ access times (x-axis) for all pairs of 5-leaf trees, grouped by the degree of the trees.
 				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
     \caption{\
-        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-taxon trees with degree 44 and (b) 7-taxon trees with degree 70.
-        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-taxon trees with degree 24 and (d) 6-taxon trees with degree 44.
-				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of taxa (e.g.
+        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
+        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
+				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g.
 pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
@@
 To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
-The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
+The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with the number of leaves $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
 Thus this ``flatness-in-the-limit'' theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
@@
 Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
-Let $\{T_n \mid n > k\}$ be a sequence of trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
+Let $\{T_n \mid n > k\}$ be a sequence of trees with $n$ leaves all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
 Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$ for the uniform random walk on the rSPR graph.

### 2016-12-01 response_to_reviewers.txt
A few fixes.

@@
+We thank the reviewers for their suggestions.
+
+We are especially grateful to Reviewer 2, who has offered the most detailed and constructive review we may have ever received.
+
 Reviewer 1
@@
 There are several occurrence of the word “taxon” or “taxa” while there is no definition: I guessed it was the leaves, and maybe the authors want to replace every occurrence.
-* replaced each occurence of taxon or taxa with leaf or leaves.
+* replaced each occurrence of taxon or taxa with leaf or leaves.
 
@@
 33: It is stated that the curvature is computed for 'given pairs of trees', but who / what gives these pairs? The wording is a bit odd here.
-* This is explained in the next paragraph. We could not compute
+* As we now clarify with revised wording, this is explained in the next paragraph. We could not compute
 	curvatures for all pairs of trees directly, but instead compute
@@
 45: Only one of the two occurrences of 'thus' is needed.
-* We removed the second occurence of "thus".
+* We removed the second occurrence of "thus".
 

### 2016-12-01 curvature.tex
A few fixes.

@@
 To compute curvature values, we first used \texttt{dense\_spr\_graph} to compute the full rSPR graph for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
-We then computed curvatures for given pairs of trees directly, by using linear programming~\citep{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \citep{SAGE} front-end to the GLPK~\citep{glpk} solver; code can be found in \citep{gricci} which grew from the code described in \citep{Loisel2014-hu}.
-
-This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves, even for the small values of $n$ we consider here.
+We then computed curvatures for pairs of trees directly, by using linear programming~\citep{Loisel2014-hu} to compute the minimal mass transport $W_1$ using the SAGE \citep{SAGE} front-end to the GLPK~\citep{glpk} solver; code can be found in \citep{gricci} which grew from the code described in \citep{Loisel2014-hu}.
+Specifically, we did this calculation for all representative pairs of trees as described next.
+
+To directly compute curvatures for all $((2n-3)!!)^2$ pairs of trees with $n$ leaves would have required an enormous amount of computation, even for the small values of $n$ we consider here.
 We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
@@
 are the same after relabeling, so their curvatures are the same.
-We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in \citep{Matsen2015-yv} with a closed formula in \citep{billey2016trees}, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
+We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} \citep{Venkatachalam2010-zh}; the group-theoretic enumeration methods are described in \citep{Matsen2016-cv} with a closed formula in \citep{Billey2017-ci}, and the SAGE \citep{SAGE} and GAP4 \citep{GAP4} code is at \citep{tangle}.
 

### 2016-12-01 response_to_reviewers.txt
A few notes in response_to_reviewers

@@
 
+**** EM We need to revise the text including your response rather than just respond.
+
 Page 17
@@
 * TODO
+*** EM Perhaps we can respond by just saying that all of the statements regarding the influence of curvature are confirmed by the regression analysis.
 
@@
 	from the reduced -n in the denominator to reduce the 20 to 18?
+**** EM: I'm a little confused by the very last step. Don't we want the 18 to be a 16?
 
@@
 	proof. T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position.
+**** EM: perhaps worth throwing an extra hint in?
 
@@
 Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
-* To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.
+* Thank you for pointing this out. To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.
 
@@
 n ≥ 4).
-* See our response above.
+* As described above we restrict to n>3.
 
@@
 Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees.
-For example, from the balanced tree with 4 leaves, we do not consider 22moves of the root’s 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution."
+For example, from the balanced tree with 4 leaves, we do not consider moves of the root’s 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution."
 

### 2016-12-01 response_to_reviewers.txt
fixed Lemma 5.2 proof and response
fixed line breaks in response document. Added a couple review comments to the text

@@
 33: It is stated that the curvature is computed for 'given pairs of trees', but who / what gives these pairs? The wording is a bit odd here.
-* As we now clarify with revised wording, this is explained in the next paragraph. We could not compute
-	curvatures for all pairs of trees directly, but instead compute
-	curvatures for each tanglegram. The procedure described here is to
-	used to compute the curvature of any given pair of trees.
+* As we now clarify with revised wording, this is explained in the next paragraph.
+We could not compute curvatures for all pairs of trees directly, but instead compute curvatures for each tanglegram.
+The procedure described here is to used to compute the curvature of any given pair of trees.
 
@@
 Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.
-	A pair of trees that both contain leaves 1 and 2 as siblings are an
-	example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves.
-
-**** EM We need to revise the text including your response rather than just respond.
+A pair of trees that both contain leaves 1 and 2 as siblings are an example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves.
+We rephrased this sentence to state "To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g.
+pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."
 
@@
 Proof of (i):The proof only works when the '+14' in the numerator is not reduced to '+12' (which is also not necessary, because bounding the sum of the logarithms in the denominator is sufficient.
-* TODO: Do you see a problem here Erick? Reducing the numerator from 14
-	to 12 is valid. Maybe he is confused by the last step, where I borrow
-	from the reduced -n in the denominator to reduce the 20 to 18?
-**** EM: I'm a little confused by the very last step. Don't we want the 18 to be a 16?
+* We corrected the mistake.
 
@@
 the pairings are pairwise distinct).
-* Revised to note "...  each neighbor of T is paired
-with a single neighbor of S, and vice versa. Note that a tree may be a
-neighbor of both T and S and thus involved in two different pairings."
+* Revised to note "...  each neighbor of T is paired with a single neighbor of S, and vice versa. Note that a tree may be a neighbor of both T and S and thus involved in two different pairings."
 
@@
 Lemma 5.5:
-1. Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is
-necessary (Why is it necessary to increase j? Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
+1.
+Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is necessary (Why is it necessary to increase j?
+Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
 * We have revised the sentence introducing the cases to state "There are two cases, depending on whether the subtree of k leaves is moved to the root (thereby introducing a new level to the tree) or not.
 
-2. Concerning statement (ii): Is there a particular reason why the quotient is bounded for n ≥ 4?
-The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
-* 5/6 is closer to the limit of 6/7 than it is to 4/5 so this gives a
-	closer bound at the small cost of excluding the trivial 3-leaf tree graph.
+2. Concerning statement (ii): Is there a particular reason why the quotient is bounded for n ≥ 4? The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
+* 5/6 is closer to the limit of 6/7 than it is to 4/5 so this gives a closer bound at the small cost of excluding the trivial 3-leaf tree graph.
 
@@
 2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
-* Yes, because of the triangle formed by T, S, and U, as we state in the
-	proof. T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position.
-**** EM: perhaps worth throwing an extra hint in?
+* Yes, because of the triangle formed by T, S, and U, as we state in the proof.
+T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position.
+We now state: In the latter case, S is reached from U by moving Q in a manner that displaces T from its original position, because T and S differ only in the location of R and d_SPR(T, U ) = d_SPR(S, U ) = 1.
+Thus, T |(X \ L(Q)) = S|(X \ L(Q)), where L(Q) is the label set of Q."
 
@@
 W_{1,n} ?
-* W_{1,n} is used in Theorem 6.1 to distinguish between the sequential
-	mass transport terms, but this this one should just be W_1. Corrected.
+* W_{1,n} is used in Theorem 6.1 to distinguish between the sequential mass transport terms, but this this one should just be W_1. Corrected.
 
@@
 14: Adding a short explanation why the number of unmatched trees is in O(kn) might help the reader understanding the argument at this point better.
-* We added the short explanation "There are O(kn) trees unmatched by
-	this pairing, because b and a (in the nomenclature of Corollary 5.4)
-	are bounded by n."
+* We added the short explanation "There are O(kn) trees unmatched by this pairing, because b and a (in the nomenclature of Corollary 5.4) are bounded by n."
 

### 2016-12-01 curvature.tex
fixed Lemma 5.2 proof and response
fixed line breaks in response document. Added a couple review comments to the text

@@
 		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
-		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
-		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
-		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
-		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
+		\ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2(n-2)} \\
+		= \ &\frac{3n^2 - 13n + 14}{4n^2 - 16n + 16 - 2(n-2)} \\
+		= \ &\frac{3n^2 - 13n + 14}{4n^2 - 18n + 20} \\
+		\ge\ &\frac{3n^2 - 13n + 14}{4n^2 - 17\frac{1}{3}n + 18\frac{2}{3}} &\forall n \ge 3, \\
 	\end{align*}
-	which is greater than 3/4 when $n \ge 3$.
+	which is at least 3/4 when $n \ge 3$.
     Similarly for (ii):
@@
 	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
-	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
+	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
+	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$. 
 	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.

### 2016-12-01 curvature.tex
cut down abstract

@@
 In this paper, we investigate the SPR graph of rooted trees (rSPR graph) in a new way: by calculating the Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-These calculations require fast new algorithms for constructing and sampling these graphs, reducing the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ is the number of leaves in each tree, and reducing the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
+These calculations require fast new algorithms.
+Wereduce the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ is the number of leaves in each tree, and we reduce the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
 We then develop a closed form solution to characterize how the number of rSPR neighbors of a tree changes after an rSPR operation is applied to that tree.
@@
 However, we find that large topology changes (i.e. moving a large subtree) give pairs of trees with negative curvature.
-We show using simulation that mean access time distributions depend on distance, degree, and curvature, demonstrating the relevance of these curvature results to stochastic tree search.
-Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
-This work extends our knowledge of the rSPR graph, in particular with respect to properties that are relevant for the investigation of sampling the rSPR graph.
+We show using simulation that mean access time distributions depend on distance, degree, and curvature, demonstrating the relevance of these results to stochastic tree search.
+%Although these pairs of trees with negative curvature do not impede mixing in this simple well-connected space, they may manifest as bottlenecks in the much smaller credible sets induced by phylogenetic posteriors with a likelihood function.
+%This work extends our knowledge of the rSPR graph, in particular with respect to properties that are relevant for the investigation of sampling the rSPR graph.
 \end{abstract}

### 2016-12-02 curvature.tex
Abstract edits.

@@
 Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
-These methods are known to mix quickly when sampling from the simple uniform distribution of trees but may become stuck in the local optima of multimodal posterior distributions for real data induced by non-uniform likelihoods.
+These methods may become stuck in the local optima of multimodal posterior distributions for real data induced by non-uniform likelihoods.
 The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating the study of the underlying \emph{SPR graph} in greater~detail.
@@
 In this paper, we investigate the SPR graph of rooted trees (rSPR graph) in a new way: by calculating the Ricci-Ollivier curvature with respect to uniform and Metropolis-Hastings random walks.
-These calculations require fast new algorithms.
-Wereduce the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ is the number of leaves in each tree, and we reduce the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
-We then develop a closed form solution to characterize how the number of rSPR neighbors of a tree changes after an rSPR operation is applied to that tree.
-This gives bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
+This value quantifies the degree to which a pair of random walkers from specified points move towards each other; negative curvature means that they move away from one another on average, while positive curvature means that they move towards each other.
+In order to calculate this curvature, we develop fast new algorithms for rSPR graph computation.
+% We reduce the time required to compute an rSPR graph from $O(m^2n)$-time to $O(mn^3)$, where $m$ is the (often large) number of trees in the graph and $n$ is the number of leaves in each tree, and we reduce the time required to select an rSPR neighbor of a tree uniformly at random to $O(n)$ time.
+We then develop formulas characterizing how the number of rSPR neighbors of a tree changes after an rSPR operation is applied to that tree.
+These give bounds on the curvature, as well as a flatness-in-the-limit theorem indicating that paths of small topology changes are easy to traverse.
 However, we find that large topology changes (i.e. moving a large subtree) give pairs of trees with negative curvature.

### 2016-12-02 curvature.tex
Adding my fancygrant.

@@
 
-	\tnotetext[t1]{This work was funded by National Science Foundation award 1223057.}
+	\tnotetext[t1]{This work was funded by National Science Foundation award 1223057.
+Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.
+The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.
+    }
 	\tnotetext[t2]{A preliminary version of this work appeared in the 2016 Proceedings of the Thirteenth Workshop on Analytic Algorithmics and Combinatorics (ANALCO).}
@@
 \cortext[cor1]{Corresponding author}
-\fntext[fn1]{Chris Whidden is a Simons Foundation Fellow of the Life Sciences Research Foundation.}
 

### 2016-12-02 response_to_reviewers.txt
Addressing figure comments.

@@
 I think that the abstract is too long and should be three-time shorter.
-* TODO, any comments Erick?
+* The abstract is now substantially shorter. We have cut it down as much as we felt possible without losing clarity.
 
@@
 
-****TODO FIGURES START*****
-
 Page 14
@@
 
+* General response to the following figure-related comments:
+* Thank you for the comments and we agree that it is difficult to parse out the various dimensions shown here.
+However, we are faced with a rather difficult visualization problem that has many variables: access time, count, degrees, and curvature.
+After quite a bit of effort, this was the best that we could do.
+To make it possible for people such as this reviewer to dig a little more deeply into the data, we have now linked to an archived version of the exact data points used to make the plots, in their five-dimensional overlapping glory.
+We also note that the statements regarding the influence of curvature are confirmed by the regression analysis.
+
 Plots of Figure 5:
@@
 2. Adding axis labels (and not just explaining them in the figure caption) could help the readability of the plots.
+
+**** TODO: Chris, could you add these axis labels with the same fonts?
 
@@
 also hamper the readability (when the differences are as small as they are here).
-
-****TODO FIGURES END*****
 
@@
 15-23: The traceability / comprehensibility of the statements would strongly benefit from improving the readability of the plots in Figure 5 and 6.
-* TODO
-*** EM Perhaps we can respond by just saying that all of the statements regarding the influence of curvature are confirmed by the regression analysis.
+* See above, also noting that Tables 1 and 2 confirm several of the statements made in a very readable form.
 
@@
 13-17: The same problem as on page 17, lines 15-23.
-* TODO (address the plots)
+* Addressed above.
 

### 2016-12-02 response_to_reviewers.txt
fix proof of cor:paired_neighbors

@@
 1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
-* This referred to both of the cases mentioned in text (early and later
-	access times). We revised the phrase to "these plots" to avoid
-	confusion.
+* This referred to both of the cases mentioned in text (early and later access times). We revised the phrase to "these plots" to avoid confusion.
 
@@
 Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be obtained.
-* Revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.".
-* TODO: We can either change the distance bound here to <= 0 or subtract
-	out the distance 0 pairs. Check which is better after completing the
-	other revisions. The argument I added after Corollary 6.6 that the Lemma 6.5 pairs of trees have negative curvature needs to be checked. Lemma 6.5 might need to take a change here in to account.
+* We revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.".
+* For the second part, we revised the proof of Corollary 5.4 to pair the set of shared neighbors with each other such that each pair of trees is at distance 1 rather than distance 0.
+As no optimal mass transport pairing would do this, we also added the following note before Lemma 5.6: "Note that these shared neighbors are exactly the trees from the second case of Corollary 5.4 and an optimal mass transport pairing will pair mass at these trees".
 

### 2016-12-02 curvature.tex
Addressing figure comments.
fix proof of cor:paired_neighbors

@@
 				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+    See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
 	}
@@
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
+        See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
         }
@@
 By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
-For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
 We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
+For each such $(T',S')$ pair, there are two cases, depending on whether $T'$ differs from $T$ by $R$ or another subtree.
+If $T'$ and $T$ differ by a different subtree, then we can move $R$ in $T'$ or $S'$ to obtain the other member of the pair.
+Thus, $\dspr{T',S'} = 1$.
+If $T'$ and $T$ differ by $R$, however, then $T' = S'$.
+Consider the set of such pairs $(T'_1, S'_1), (T'_2, S'_2), \ldots, (T'_p, S'_p)$.
+Each tree of the pairs can be obtained from the others by moving $R$, so we simply pair $T'_1$ with $S'_2$, $T'_2$ with $S'_3$ and so on, finally pairing $T'_p$ with $S'_1$.
 \end{proof}
@@
 The number of such ``triangles'' involving two adjacent trees has a key role in determining whether their curvature is positive, as mass assigned to a shared neighbor may not require a mass transport cost.
+Note that these shared neighbors are exactly the trees from the second case of Corollary~\ref{cor:paired_neighbors} and an optimal mass transport pairing will pair mass at these trees.
 We show that the maximum number of shared neighbors is linear in the number of leaves.

### 2016-12-03 response_to_reviewers.txt
Some small tweaks.

@@
 
-We are especially grateful to Reviewer 2, who has offered the most detailed and constructive review we may have ever received.
+We are especially grateful to Reviewer 2, who may have offered the most detailed and constructive review we have ever received.
 
@@
 
+**** TODO: EM modified the below, though you can probably do better. Note that the text hasn't been modified accordingly yet.
+
 2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
@@
 Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.
-A pair of trees that both contain leaves 1 and 2 as siblings are an example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves.
-We rephrased this sentence to state "To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g.
-pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."
+For example, a pair of trees that both contain leaves 1 and 2 as siblings have very similar characteristics as the pair obtained by merging each sibling pair into a single leaf and relabeling the leaves.
+We rephrased this sentence to state "To limit clutter, these plots exclude pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g.\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."
 
@@
 2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
+***** TODO this needs to get updated if the text is.
 * Yes, because of the triangle formed by T, S, and U, as we state in the proof.
@@
 This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
-* There appears to be some confusion on what constitutes a moved subtree here.
+* We should have been more clear on what constitutes a moved subtree here.
 From our distribution of rSPR moves, the balanced tree with n = 4 leaves does not move any subtrees of size 2.
@@
 * Corrected
+
+* Again, thank you very much!

### 2016-12-03 curvature.tex
Some small tweaks.

@@
 These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
-We construct a mapping from each tree $T_i$ to its order index in this ordering i$i$.
+We construct a mapping from each tree $T_i$ to its order index in this ordering $i$.
 Begin with an empty graph $G$.
@@
 
+%EM "distribution" here and below seems strange to me because we aren't always sampling. Wouldn't "enumeration" work?
 We consider the distribution of rSPR moves in terms of the number of nodes contained within the subtree to be moved.
@@
 \begin{proof}
+%EM Unlike you and the reviewer, I can't figure out what source and target mean here-- are they defined?
 By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
@@
 	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
+    %EM Don't you mean displaces $R$ from its original position?
 	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
@@
 We pair these trees in the mass transport.
+%EM Sorry, I don't see b and a in cor:paired_neighbors.
 There are $\OhOf{kn}$ trees unmatched by this pairing, because $b$ and $a$ (in the nomenclature of Corollary~\ref{cor:paired_neighbors}) are bounded by $n$.
@@
 However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
-This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.
+This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is, the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.
 

### 2016-12-05 curvature.tex
protect floating urls, try to add axis labels to access time plots

@@
 				Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
-    See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
+		See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
 	}
@@
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
-        See online version for color figures; data tables available at \url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
+				See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
         }

### 2016-12-05 scripts/plot-access.py

@@ -96,6 +96,8 @@ for (t1_deg, t2_deg, dist), group in catted.groupby(['t1_deg','t2_deg', 'dist'])
         sub_d = access[access['tangle'] == idx]
         if args.maxtime:
             sub_d = sub_d[sub_d['time'] < args.maxtime]
+        d_axs.set_xlabel('access time')
+        d_axs.set_ylabel('count')
         d_axs[i][j].plot(sub_d['time'], sub_d['count'],
                          color=m.cmap(kappa), label=kappa)
         # d_axs[i][j].legend()

### 2016-12-05 response_to_reviewers.txt
protect floating urls, try to add axis labels to access time plots
added axis labels to access5 plots, revised text with R fix that was already in the response document

@@
 There are several occurrence of the word “taxon” or “taxa” while there is no definition: I guessed it was the leaves, and maybe the authors want to replace every occurrence.
-* replaced each occurrence of taxon or taxa with leaf or leaves.
+* We replaced each occurrence of taxon or taxa with leaf or leaves.
 
@@
 2. Adding axis labels (and not just explaining them in the figure caption) could help the readability of the plots.
-
-**** TODO: Chris, could you add these axis labels with the same fonts?
+* We added axis labels to the plots.
 
@@
 * This referred to both of the cases mentioned in text (early and later access times). We revised the phrase to "these plots" to avoid confusion.
-
-**** TODO: EM modified the below, though you can probably do better. Note that the text hasn't been modified accordingly yet.
 
@@
 2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
-***** TODO this needs to get updated if the text is.
 * Yes, because of the triangle formed by T, S, and U, as we state in the proof.

### 2016-12-05 curvature.tex
added axis labels to access5 plots, revised text with R fix that was already in the response document

@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{figs/access5-distribution.png}
+	\includegraphics[width=\textwidth]{figs/access5-distribution-alabels.png}
 
@@
         Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
-				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g.
-pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
+				To limit clutter, these plots exclude pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings) as their access time distributions do not change substantially.
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
     %EM Don't you mean displaces $R$ from its original position?
-	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
+	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $R$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
 	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$.

### 2017-01-19 response_to_reviewers_2.txt
minor revisions suggested by reviewer 2

We again thank Reviewer 2 for their detailed suggestions that have greatly improved the paper.

Anonymous comments from Referee #2:

Page 12, line 23 [page 12, line 25]:
I would still suggest adding the word bits to make it clearer, because the Big-O notation is also used with the unit 'words'.
* We revised the sentence to say "... using O(n) space (in bits).".

Page 12, line 52-55 [page 12, line 57]:
The correction is fine, but in the new last sentence on page 12, 'four' should probably be 'for'.
* Corrected.

Page 16, lines 45-46 [page 16, lines 43-46]:
The added explanation is helpful. However, I am not sure whether it is obvious enough that the "access times distributions do not change substantially" through the exclusions. Maybe this also requires some proof and a more precise statement of the effect (compared to just saying "substantial").
* We apologize for the misunderstanding. This figure is simply meant to visually demonstrate the effect of curvature on access times when all other variables are held constant. The excluded pairs have much smaller access times than those shown but which are not substantially different (visually) from those of the trees from which they can be obtained via replacement. We were not claiming that the excluded trees have the same access times as the included trees, and, as such there is nothing to prove here. As the replaced trees have a smaller number of leaves (and therefore much smaller access times) the effect is a messy plot with multiple curves that essentially includes the results of trees with 3 to say 7 leaves (for (b)). Each individual curve shows a similar influence from curvature.
* We revised these sentences to say: "To limit clutter, these plots exclude the relatively small number of pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g we exclude pairs of trees that both contain leaves 1 and 2 as siblings). Such pairs of trees have much lower access times that are visually identical to the pairs from which they can be obtained via replacement. The net effect is that a distinct curve would appear for each smaller number of leaves. Each individual curve would show a similar influence from curvature"

Page 20, Lemma 5.5, point 1 [page 20, Lemma 5.5]:
Unfortunately, the revised sentence introducing the cases is not much more helpful than its predecessor (for me). The main source of confusion is this: "we do not need to exclude the target branch from k and a, gaining an additional leaf to balance the product ka at the cost of increasing j" (from the case where the subtree is moved to the root). Probably I am missing a point somewhere, but I cannot make much sense of this sentence. Exclude the target branch from k and a? Gaining an additional leaf - where / how?
* At this point in the proof, we are balancing the number of leaves between the values k and a to maximize the degree change. If we do not move the subtree to the root, then the j leaves of the new neighbor subtree V cannot be part of k or a. If we do move the subtree to the root then the new neighbor subtree V is actually the entire tree barring the moved subtree R, so we can include those leaves when balancing k and a at the cost of increasing j.
* We have revised these two paragraphs of the proof to state:

"There are two cases, depending on whether the subtree of k leaves is moved to the root (thereby introducing a new level to the tree) or not. If not, then the j leaves of the new neighbor subtree V cannot be part of k or a. We thus set j = 1 and split the remaining n − b − i − j = n − 2 leaves between k and a in as balanced a way as possible, giving (i). Note that this corresponds to moving the bottom subtree of floor((n−2)/2) or ceiling((n-2)/2) leaves in a ladder tree to the root-most leaf of the tree.

If the subtree of k leaves is moved to the root, then the new neighbor subtree V is the entire tree excluding R itself. We can thus include the root-most leaf of the tree in a. In this case, we gain an additional leaf over the previous case to balance the product ka at the cost of increasing j. This corresponds to moving the bottom subtree of floor(n/2)) or ceiling(n/2) leaves in a ladder tree to the root. ..."

Page 22, lines 30-32, point 2 [page 22, lines 49-50]:
I think that the last part of the sentence "with a possible difference of two moves for a tree where one of the root's children is small" does still not fit perfectly. T can still have two small children at the root node (resulting in the balanced tree with four leaves) and then there is a difference of four moves from the 'general' case (= no small root children). Maybe a wording along the lines "with a possible difference of two moves per small child of the root" covers (all) the different cases better.
* We adopted the suggested wording.

Page 22, lines 38-42, point 2 [page23, lines 10-11]:
The restriction to n > 3, however, does not remove the arithmetical error. '4n^2-8n-12 + (24-8n)' is still not '4n^2-16n+16' but '4n^2-16n+12'. So, you would probably need to add the term '(28-8n)' and say 'n>=4' at the end of the sentence to fix this.
* Corrected.

Page 22, lines 54-55 [page 23, lines 14-19]:
The added explanations makes it much clearer. However, is it 'fair' to change the 'ranking' (w.r.t. the average moved subtree size) of the trees by ignoring equivalent moves? And is it obvious that the ladder tree will be the tree with the highest average moved subtree size after ignoring equivalent moves? Or does this maybe need a proof?
* We again apologize for the misunderstanding. Our intention was not to change the ranking in such a manner, but in fact the opposite: we merely analyze the distribution of the subtree sizes according to our specific distribution. We show in the previous proof that the the ladder tree has the smallest proportion of small subtrees *given our distribution* which implies that it has the highest average moved subtree size when ignoring equivalent moves in this way. We have made the following changes to make this more clear:
* In the paragraph before the lemma we now say (in reference to ignoring equivalent moves) "This occurs because multiple NNI moves may lead to the same tree, as described previously, and each equivalent move may involve a subtree with a different number of leaves. Our distribution includes only one of these equivalent moves."
* We revised the lemma to state: "Let T be a rooted tree with >= 4 leaves. Let N_i(T) be the neighbors of T induced by rSPR moves of an i-leaf subtree, where the moved subtree is defined according to our distribution of rSPR moves."
* In the proof of the lemma we now say "Recall from the proof of Lemma 5.7 that a ladder tree has the smallest proportion of small subtrees. This implies that a ladder tree has the greatest average moved subtree size (in terms of the number of leaves) with respect to our distribution."
* In the conclusion of the proof we now say "Thus, the average size of a moved subtree over all rSPR operations from our distribution on a given tree is at most the weighted average of ..."
* Following the proof we finally say "Further, note that our distribution does not necessarily minimize the moved subtree size among equivalent moves. It may be interesting future work to improve our bound by explicitly minimizing the moved subtree size in such a manner."

Page 23, line 32 [page 24, line 9]:
While this occurrence is fixed, I would strongly suggest stating (in the proof of Theorem 6.1) that 'W_{1,n}' is the short form for the mass transport term when considering T_n and S_n (if I figure it now correctly, which took far too long).
* We now begin the proof with the sentence "Let W_{1,n} be the mass transport term with respect to T_n and S_n, that is W_1(m_{T_n}, m_{S_n}).".

### 2017-01-19 curvature.tex
potentially controversial change to get the keywords back on the title page
keep an algorithm with its name/declaration
disable todonotes
minor revisions suggested by reviewer 2

@@
 \usepackage{graphicx}
-\usepackage{todonotes}
 \usepackage{url}
@@
 Statistical phylogenetic inference methods use tree rearrangement operations such as subtree-prune-regraft (SPR) to perform Markov chain Monte Carlo (MCMC) across tree topologies.
-These methods may become stuck in the local optima of multimodal posterior distributions for real data induced by non-uniform likelihoods.
+%These methods may become stuck in the local optima of multimodal posterior distributions for real data induced by non-uniform likelihoods.
 The structure of the graph induced by tree rearrangement operations is an important determinant of the mixing properties of MCMC, motivating the study of the underlying \emph{SPR graph} in greater~detail.
@@
 \vspace{1em}
+\begin{samepage}
 \textsc{Select-rSPR-Neighbor($T$)}
@@
 \end{enumerate}
+\end{samepage}
 
@@
 	\label{lem:select_random_neighbor}
-	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
+	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space (in bits).
 \end{lem}
@@
 We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\citep{random_spr_walk}.
-We used this procedure to sample random walks on varying size rSPR graphs four our analysis in the next section.
+We used this procedure to sample random walks on varying size rSPR graphs for our analysis in the next section.
 
@@
         Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
-				To limit clutter, these plots exclude pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings) as their access time distributions do not change substantially.
+				To limit clutter, these plots exclude the relatively small number of pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
+Such pairs of trees have much lower access times that are visually identical to the pairs from which they can be obtained via replacement.
+The net effect is that a distinct curve would appear for each smaller number of leaves.
+Each individual curve would show a similar influence from curvature.
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
 	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root (thereby introducing a new level to the tree) or not.
-    If not, then we set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
+    If not, then the $j$ leaves of the new neighbor subtree $V$ cannot be part of $k$ or $a$.
+We thus set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
 	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.
 
-	If the subtree of $k$ leaves is moved to the root, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf to balance the product $ka$ at the cost of increasing $j$.
+	If the subtree of $k$ leaves is moved to the root, then the new neighbor subtree $V$ is the entire tree excluding $R$ itself.
+We can thus include the root-most leaf of the tree in $a$.
+In this case, we gain an additional leaf over the previous case to balance the product $ka$ at the cost of increasing $j$.
 	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
@@
 	Moreover, every tree has $n$ 1-leaf subtrees.
-	This implies that the number of small rSPR moves is determined by the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
+	This implies that the number of small rSPR moves is determined by the number of small subtrees, with a possible difference of two moves per small child of the root.
 
@@
 	Now, recall that the maximum degree of any tree is less than $4n^2 - 16n + 16$ by Lemma~\ref{lem:degree_extremes}.
-	We then have that $2(2n^2 - 4n - 6) = 4n^2 - 8n - 12 \ge 4n^2 - 8n - 12 + (24 - 8n) = 4n^2 -16n + 16$, for all $n \ge 3$.
+	We then have that $2(2n^2 - 4n - 6) = 4n^2 - 8n - 12 \ge 4n^2 - 8n - 12 + (28 - 8n) = 4n^2 -16n + 16$, for all $n \ge 4$.
 	Therefore more than half of the neighborhood of any tree is induced by small rSPR moves.
@@
 Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees.
+This occurs because multiple NNI moves may lead to the same tree, as described previously, and each equivalent move may involve a subtree with a different number of leaves.
+Our distribution includes only one of these equivalent moves.
 For example, from the balanced tree with 4 leaves, we do not consider moves of the root's 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution.
@@
 	Let $T$ be a rooted tree with $n \ge 4$ leaves.
-	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
+	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree, where the moved subtree is defined according to our distribution of rSPR moves.
 	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
@@
 	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
-	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
+	Recall from the proof of Lemma~\ref{lem:spr_size} that a ladder tree has the smallest proportion of small subtrees.
+This implies that a ladder tree has the greatest average moved subtree size (in terms of the number of leaves) with respect to our distribution.
 	There are $n$ subtrees with one leaf, and one subtree each with $2,3, \ldots n - 1$ leaves.
@@
 	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
-	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
+	Thus, the average size of a moved subtree over all SPR operations from our distribution on a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
 
@@
 
+Further, note that our distribution does not necessarily minimize the moved subtree size among equivalent moves.
+It may be interesting future work to improve our bound by explicitly minimizing the moved subtree size in such a manner.
+
 \section{Curvature}
@@
 \begin{proof}
+	Let $W_{1,n}$ be the mass transport term with respect to $T_n$ and $S_n$, that is $W_1(m_{T_n}, m_{S_n})$.
 Because $d(T_n, S_n) = 1$, we will prove the theorem by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.

### 2017-01-19 response_to_reviewers_2.txt
A few tweaks and a suggestion.

@@
-We again thank Reviewer 2 for their detailed suggestions that have greatly improved the paper.
+We again thank Reviewer 2 for their remarkably detailed suggestions; these have greatly improved the paper.
 
@@
 Each individual curve shows a similar influence from curvature.
-* We revised these sentences to say: "To limit clutter, these plots exclude the relatively small number of pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
-Such pairs of trees have much lower access times that are visually identical to the pairs from which they can be obtained via replacement.
-The net effect is that a distinct curve would appear for each smaller number of leaves.
-Each individual curve would show a similar influence from curvature"
+* We revised these sentences to say: "To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
+These additional curves show a similar influence from curvature."
 
@@
 Gaining an additional leaf - where / how?
+EM-- I think this makes sense, but the wording "then the j leaves of the new neighbor subtree V cannot be part of k or a" could be made a little more clear IMHO.
+k and a aren't sets-- you are saying that they are not part of the sets enumerated by k and a.
+Alternatively you could have an explanatory sentence at the beginning describing what you mean by membership in k and a, and leave things how they are.
 * At this point in the proof, we are balancing the number of leaves between the values k and a to maximize the degree change.
@@
 Maybe a wording along the lines "with a possible difference of two moves per small child of the root" covers (all) the different cases better.
-* We adopted the suggested wording.
+* We adopted the suggested wording, thanks.
 
@@
 * Following the proof we finally say "Further, note that our distribution does not necessarily minimize the moved subtree size among equivalent moves.
-It may be interesting future work to improve our bound by explicitly minimizing the moved subtree size in such a manner."
+It may be interesting in future work to improve our bound by explicitly minimizing the moved subtree size in such a manner."
 

### 2017-01-19 curvature.tex
A few tweaks and a suggestion.

@@
         Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
-				To limit clutter, these plots exclude the relatively small number of pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
-Such pairs of trees have much lower access times that are visually identical to the pairs from which they can be obtained via replacement.
-The net effect is that a distinct curve would appear for each smaller number of leaves.
-Each individual curve would show a similar influence from curvature.
+				To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
+These additional curves show a similar influence from curvature.
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
 Further, note that our distribution does not necessarily minimize the moved subtree size among equivalent moves.
-It may be interesting future work to improve our bound by explicitly minimizing the moved subtree size in such a manner.
+It may be interesting in future work to improve our bound by explicitly minimizing the moved subtree size in such a manner.
 

### 2017-01-20 response_to_reviewers_2.txt
addressed suggestion

@@
 * We revised these sentences to say: "To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
-These additional curves show a similar influence from curvature."
+The additional curves induced by the excluded pairs show a similar influence from curvature."
 
@@
 Gaining an additional leaf - where / how?
-EM-- I think this makes sense, but the wording "then the j leaves of the new neighbor subtree V cannot be part of k or a" could be made a little more clear IMHO.
-k and a aren't sets-- you are saying that they are not part of the sets enumerated by k and a.
-Alternatively you could have an explanatory sentence at the beginning describing what you mean by membership in k and a, and leave things how they are.
-* At this point in the proof, we are balancing the number of leaves between the values k and a to maximize the degree change.
-If we do not move the subtree to the root, then the j leaves of the new neighbor subtree V cannot be part of k or a.
-If we do move the subtree to the root then the new neighbor subtree V is actually the entire tree barring the moved subtree R, so we can include those leaves when balancing k and a at the cost of increasing j.
-* We have revised these two paragraphs of the proof to state:
+* We apologize for the confusion, which was likely caused by our incorrect definition of a and b in Lemma 5.3.
+In the special case that R is moved to an ancestor, the newly introduced node becomes L in the new tree S.
+Thus, the node "L" of T should also be part of the path that defines a, because moves of the subtree rooted at "L" can now be attached to R.
+In the case described in Lemma 5.5 we thus increase the length of a by 1 when we move R to the root.
+We originally defined (a - b) in terms of the height change experienced by R but felt the path-based definition was more intuitive and thereby made this error.
+Thus, this does not change any of our results.
+* We have revised the statement and proof of Lemma 5.3, introducing separate least common ancestors L_1 in T and L_2 in S, defining a using the path from U to L_2, and defining b using the path from V to L_1.
+* We have revised the proof of Lemma 5.5 to state:
 
-"There are two cases, depending on whether the subtree of k leaves is moved to the root (thereby introducing a new level to the tree) or not.
-If not, then the j leaves of the new neighbor subtree V cannot be part of k or a.
-We thus set j = 1 and split the remaining n − b − i − j = n − 2 leaves between k and a in as balanced a way as possible, giving (i).
+"Recall that k is the number of leaves in the moved subtree R, i is the number of leaves in R’s neighbor U and j is the number of leaves in R’s new neighbor V after the SPR move.
+The length of the path from U ’s root to the least common ancestor L 2 of U and V in S is a, while b is the length of the path from V ’s root to the least common ancestor L 1 of U and V in T .
+Call the former path A.
+
+The degree change is maximized by making L_1 the root of T and thereby minimizing b, namely setting b = 0.
+The resulting equation 2(ka + i − j) is similarly maximized by including only one leaf in V, setting i = 1.
+We then maximally balance the terms in the product ka by adding as many leaves as possible to either R or as children of the path A.
+Each new leaf of R increases k by 1 and each new child of A lengthens the path and increases a by 1.
+
+There are two cases, depending on whether R is moved to the root (thereby introducing a new level to the tree) or not.
+If not, then the j leaves of the new neighbor subtree V cannot contribute to increasing k or a.
+We thus set j = 1 and split the remaining n − b − i − j = n − 2 leaves between R and A in as balanced a way as possible, maximizing the product ka and giving (i).
 Note that this corresponds to moving the bottom subtree of floor((n−2)/2) or ceiling((n-2)/2) leaves in a ladder tree to the root-most leaf of the tree.
 
-If the subtree of k leaves is moved to the root, then the new neighbor subtree V is the entire tree excluding R itself.
-We can thus include the root-most leaf of the tree in a.
-In this case, we gain an additional leaf over the previous case to balance the product ka at the cost of increasing j.
+If R is moved to the root, then the new neighbor subtree V is the entire tree excluding R itself.
+The root-most leaf of the tree is on the path from U to L in S and thus a child of A, despite also being a child of L in T .
+In this case, we have an additional leaf over the previous case that can be added to R or lengthen A, increasing the product ka at the cost of increasing j.
 This corresponds to moving the bottom subtree of floor(n/2)) or ceiling(n/2) leaves in a ladder tree to the root.

### 2017-01-20 curvature.tex
addressed suggestion

@@
 				To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings).
-These additional curves show a similar influence from curvature.
+The additional curves induced by the excluded pairs show a similar influence from curvature.
         Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
@@
 Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
-Let $L$ be the $\lca{U,V}$ in $T$.
-Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
-Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
+Let $L_1$ be the $\lca{U,V}$ in $T$.
+	Let $L_2$ be the $\lca{U,V}$ in $S$ (typically $L_1 = L_2$).
+Let $a$ be the number of intermediate nodes on the path from $U$ to $L_2$ in $S$, excluding endpoints.
+Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L_1$ in $T$, excluding endpoints.
 Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
@@
 The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
-(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
-(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
+(i) subtrees that include nodes on the path from $U$ to $L_2$ may now be moved into $R$ and its newly introduced parent node,
+(ii) subtrees that include nodes on the path from $V$ to $L_1$ may no longer be moved into $R$ and its parent node,
 (iii) $R$'s parent subtree may now be moved into $U$, and
@@
 By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
-	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
-	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) and then maximally balancing the terms in the product $ka$ as follows.
-
-	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root (thereby introducing a new level to the tree) or not.
-    If not, then the $j$ leaves of the new neighbor subtree $V$ cannot be part of $k$ or $a$.
-We thus set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
+Recall that $k$ is the number of leaves in the moved subtree $R$, $i$ is the number of leaves in $R$'s neighbor $U$ and $j$ is the number of leaves in $R$'s new neighbor $V$ after the SPR move.
+The length of the path from $U$'s root to the least common ancestor $L_2$ of $U$ and $V$ in $S$ is $a$, while $b$ is the length of the path from $V$'s root to the least common ancestor $L_1$ of $U$ and $V$ in $T$.
+Call the former path $A$.
+
+The degree change is maximized by making $L_1$ the root of T and thereby minimizing $b$, namely setting $b=0$.
+The resulting equation $2(ka + i - j)$ is similarly maximized by including only one leaf in $V$, setting $i=1$.
+We then maximally balance the terms in the product $ka$ by adding as many leaves as possible to either $R$ or as children of the path $A$.
+Each new leaf of $R$ increases $k$ by 1 and each new child of $A$ lengthens the path and increases $a$ by 1.
+
+There are two cases, depending on whether $R$ is moved to the root (thereby introducing a new level to the tree) or not.
+If not, then the $j$ leaves of the new neighbor subtree $V$ cannot contribute to increasing $k$ or $a$.
+We thus set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $R$ and $A$ in as balanced a way as possible, maximizing the product $ka$ and giving (i).
 Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.
 
-	If the subtree of $k$ leaves is moved to the root, then the new neighbor subtree $V$ is the entire tree excluding $R$ itself.
-We can thus include the root-most leaf of the tree in $a$.
-In this case, we gain an additional leaf over the previous case to balance the product $ka$ at the cost of increasing $j$.
+If $R$ is moved to the root, then the new neighbor subtree $V$ is the entire tree excluding $R$ itself.
+The root-most leaf of the tree is on the path from $U$ to $L$ in $S$ and thus a child of $A$, despite also being a child of $L$ in $T$.
+In this case, we have an additional leaf over the previous case that can be added to $R$ or lengthen $A$, increasing the product $ka$ at the cost of increasing $j$.
 This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.

### 2017-01-20 response_to_reviewers_2.txt
minor edits to response copied text formatting, one more use each of L_1 and L_2

@@
 
-"Recall that k is the number of leaves in the moved subtree R, i is the number of leaves in R’s neighbor U and j is the number of leaves in R’s new neighbor V after the SPR move.
-The length of the path from U ’s root to the least common ancestor L 2 of U and V in S is a, while b is the length of the path from V ’s root to the least common ancestor L 1 of U and V in T .
+"Recall that k is the number of leaves in the moved subtree R, i is the number of leaves in R's neighbor U and j is the number of leaves in R's new neighbor V after the SPR move.
+The length of the path from U's root to the least common ancestor L_2 of U and V in S is a, while b is the length of the path from V's root to the least common ancestor L_1 of U and V in T.
 Call the former path A.
@@
 If R is moved to the root, then the new neighbor subtree V is the entire tree excluding R itself.
-The root-most leaf of the tree is on the path from U to L in S and thus a child of A, despite also being a child of L in T .
+The root-most leaf of the tree is on the path from U to L_2 in S and thus a child of A, despite also being a child of L_1 in T .
 In this case, we have an additional leaf over the previous case that can be added to R or lengthen A, increasing the product ka at the cost of increasing j.
@@
 Our intention was not to change the ranking in such a manner, but in fact the opposite: we merely analyze the distribution of the subtree sizes according to our specific distribution.
-We show in the previous proof that the the ladder tree has the smallest proportion of small subtrees *given our distribution* which implies that it has the highest average moved subtree size when ignoring equivalent moves in this way.
+We show in the previous proof that the ladder tree has the smallest proportion of small subtrees *given our distribution* which implies that it has the highest average moved subtree size when ignoring equivalent moves in this way.
 We have made the following changes to make this more clear:
